### 2017-12-22 tex/MVMetaLearnMRP2017OctoberMRPMM.tex
Minor text updates

\documentclass{nature}
\usepackage[T1]{fontenc}
\usepackage[dvips]{graphicx}
\usepackage[usenames]{color}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amscd}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{eucal}
% Please add the following required packages to your document preamble:
 \usepackage{booktabs}
\usepackage{rotating}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{bbm}
\usepackage[font=small]{caption}
\usepackage[font=small]{subcaption}
\usepackage{color}
\usepackage{lineno}
\usepackage{url}
\usepackage{lscape}
\usepackage{breqn}
\usepackage{enumitem}
\usepackage{listings}
\def\bs{\boldsymbol}
\def\til{\widetilde}
\def\MM{\mathcal{M}}
\def\Vy{\widehat{\textrm{V}}_Y}

\bibliographystyle{naturemag}
\title{Bayesian model comparison for rare variant association studies of multiple phenotypes}
\author{Christopher DeBoever$^{1}$, Matthew Aguirre$^{1}$, Yosuke Tanigawa$^{1}$, Adam Lavertu$^{1}$, Greg McInnes$^{1}$, Timothy Poterba$^{2}$, Mark J. Daly$^{2,3}$,  Matti Pirinen$^{4}$\thanks{matti.pirinen@helsinki.fi}, and Manuel A. Rivas$^{1}$\thanks{mrivas@stanford.edu}}
\usepackage{setspace}   
\linenumbers
\begin{document}
\maketitle
\begin{abstract}
\singlespacing
Whole genome sequencing studies applied to large populations or biobanks with extensive phenotyping raise new analytic challenges. The need to consider many variants at a locus or group of genes simultaneously and the potential to study many correlated phenotypes with shared genetic architecture provide opportunities for discovery and inference that are not addressed by the traditional one variant-one phenotype association study. Here we introduce a model comparison approach we refer to as MRP for rare variant association studies that considers correlation, scale, and location of genetic effects across a group of genetic variants, phenotypes, and studies. In so doing we consider the use of summary statistic data to 1) apply standard univariate and multivariate gene-based meta-analysis models, 2) models for assessing heterogeneity of genetic effects, which may be used in practice for downstream quality control, and 3) models for identifying protective protein-truncating variants, which can expedite drug discovery.  Through simulation studies, we demonstrate that the proposed model comparison approach may improve ability to detect rare variant association signals. Finally, we demonstrate its use with rare variant data combined with asthma diagnosis and haematological and spirometry measures for individuals in the UK Biobank. We show that we are able to retain useful features from widely-used meta-analysis approaches and prioritize protective modifiers of disease risk.
\end{abstract}

\begin{affiliations}
\item Department of Biomedical Data Science, Stanford University, Stanford, CA, USA
\item Broad Institute of MIT and Harvard, Cambridge, MA, USA
\item Wellcome Trust Centre for Human Genetics, University of Oxford, Oxford, UK
\item Analytic and Translational Genetics Unit, Massachusetts General Hospital, Boston, MA, USA
\item Oxford Centre for Diabetes, Endocrinology and Metabolism, Oxford, UK
\item Institute for Molecular Medicine Finland, University of Helsinki, Finland
\end{affiliations}
\newpage

\section{Introduction} 
%% Sequencing technologies
Sequencing technologies are quickly transforming human genetic studies of complex traits: it is increasingly possible to obtain whole genome sequence data on thousands of samples at manageable costs. As a result, the genome-wide study of rare variants (minor allele frequency [MAF] $< 1\%$) and their contribution to disease susceptibility and phenotype variation is now feasible\cite{ifih1,altshuler2010map,rivas2011deep,10002012integrated}. 
%% Rare variant methods
In genetic studies of diseases or continuous phenotypes, rare variants are hard to assess individually due to the limited number of copies. Hence, to boost the ability to detect a signal, evidence is usually `aggregated' across variants. When designing an `aggregation' method, there are three questions that are usually considered. First, across which biological units should variants be combined; second, which variants mapping within those units should be included\cite{majithia2014rare}; and third, which statistical model should be used\cite{lee2014rare}. Given the widespread observations of shared genetic risk factors across distinct diseases, there is also considerable motivation to proactively modify gene discovery approaches accordingly. In other words, rather than only aggregating variants that may have effects on a single phenotype, we can also bring together sets of phenotypes for which a single variant or sets of variants might have effects.

%% The model comparison method
In this paper, we present a Bayesian multiple rare variants and phenotypes (MRP) model comparison approach, as an alternative to current widely-used statistical tests. The MRP framework, as implemented, can exploit correlation, scale, or location (direction) of genetic effects in a broad range of rare variant association study designs including: case-control; multiple diseases and shared controls; single continuous phenotype; multiple continuous phenotypes; or a mixture of case-control and multiple continuous phenotypes (Fig.~\ref{Fig1}).

In Bayesian model comparison, we compute a Bayes Factor (BF) defined as the ratio of the marginal likelihoods of the observed data under two models: 1) a pre-specified null where all genetic effects are zero; and 2) an alternative model where factors like correlation, scale, or location of genetic effects are considered.

%% Use of summary statistic data
In practice, sharing of individual genotype and phenotype data across groups in large genetic consortium studies is difficult to achieve.  We present a derivation where summary statistics, such as estimates of effect size and the corresponding standard error from single variant and single phenotype analysis, are the input data. Furthermore, we use insights from Liu et al.\cite{liu2014meta} and Cichonska et al.\cite{cichonska2016metacca} who suggest the use of additional summary statistics, like covariance estimates across variants and studies, respectively, that would enable lossless ability to detect gene-based association signals using summary statistics alone.

%% Genome variant annotation and protein truncating variants
Genome variant annotation is critical for the interpretation of genetic findings. In the MRP model comparison we present approaches for including priors on spread of effect sizes. An example of an important class of variants where functional prediction is much more straightforward is the class of variants predicted to truncate the protein product, commonly referred to as protein truncating variants or PTV\cite{rivas2013assessing,rivas2015effect}, and in this paper we place special emphasis on the interpretation of PTVs.

%% Protective modifiers of disease risk
Given the translational potential of identifying protective modifiers of human disease risk we characterize how model comparison may be used to improve discovery of such protective signals. We present an extension allowing for modeling of location (direction) of genetic effects, which may be useful for prioritizing variants or genes that are consistent with a protective profile of disease\cite{pcsk9,cohen2006sequence,sullivan2012effect}.

%%Comparing to other methods 
To evaluate the performance of MRP and to study its behavior we use simulations and compare it to other commonly used approaches. Some simple alternatives to MRP include univariate approaches for rare variant association studies including the sequence kernel association test (SKAT)\cite{skat}, and the burden test, which we show are special cases of the MRP model comparison when we assign the prior correlation of genetic effects across different variants to be zero or one. Furthermore, we show that MRP is able to detect the presence of heterogeneity of effects, which in some circumstances, such as those where effects are observed to be heterogeneous across studies, may indicate the presence of technical sources of error.

%%Apply methods to the GLGC data and exome sequencing data sets. 
We apply MRP to four phenotypes: asthma (HC382: the corresponding phenotype label in Global Biobank Engine [\url{https://biobankengine.stanford.edu}]), eosinophil count (INI30150), Forced Expiratory Volume in 1-second (FEV1, INI3063), and Forced Vital Capacity (FVC, INI3062) in the UK Biobank dataset. We show that we are able to retain useful features from single variant and single phenotype meta-analysis, rare variant analysis, and prioritize protective modifiers of disease risk.

%% FIGURE 1 
\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{../figures/FIG1FINAL.pdf}
  \caption{Schematic overview of MRP. A. MRP is suitable for a broad range of rare variant association study designs including (from left to right): i) case-control, ii) multiple diseases with shared controls, iii) single quantitative phenotype, and iv) mixture of case-control and quantitative phenotypes. B. Diagram of factors considered in rare variant association analysis including the correlation matrices: $\mathbf{R_{\textrm{study}}}$  (expected correlation of genetic effects among a group of studies), $\mathbf{R_{\textrm{var}}}$ (expected correlation of genetic effects among a group of variants), and $\mathbf{R_{\textrm{phen}}}$ (expected correlation of genetic effects among a group of phenotypes); the scale parameter for genetic variant annotation; and the location of genetic effects, which may be used to prioritize or identify protective modifiers of disease risk.}
  \label{Fig1}
  \end{figure}

\section{Results}

\subsection{Methods overview}
A complete description of MRP, along with an analytical derivation, is given in the Online Methods and Appendix. Briefly, we describe approach.

\subsubsection{Model comparison}
In Bayesian model comparison, we compare the null and alternative models via the Bayes Factor (BF) defined as the ratio of the marginal likelihoods of the observed data under two models. Here, the null model is that the effect sizes obtained across all studies for a group of variants and a group of phenotypes is zero. To complete the specification of the alternative model we consider the correlation structure, the scale, and the location of the effect sizes (which we find to be informative for alternative models where we seek to identify variants with protective effects).

Let $N$ be the number of individuals and $K$ the number of quantitative phenotype measurements on each individual, for example, here in the applications we focus on asthma, eosinophil count, Forced Expiratory Volume in 1-second, and Forced Vital Capacity, in which case $K=4$. Let $M$ be the number of variants in a testing unit ${\bs G}$, where ${\bs G}$ can be, for example, a gene, pathway, or a network. Let $S$ be the number of studies where data is obtained from - this data may be in the form of raw genotypes and phenotypes or summary statistics including linkage-disequilibrium, effect sizes (or odds ratio), and standard error of the effect size.

In the multivariate setting we define the prior correlation structure of the effect sizes, denoted by a SMK-by-SMK matrix $\mathbf{U}$. In practice, we define $\mathbf{U}$ as a Kronecker product, an operation of matrices of arbitrary size, of three sub-matrices: an S-by-S matrix $\mathbf{R_{\textrm{study}}}$ being the correlations of genetic effects among studies where different values can be used to compare different models of association, such as for identifying heterogeneity of effect sizes between populations~\cite{band2013imputation}; an M-by-M matrix $\mathbf{R_{\textrm{var}}}$ being the correlations of genetic effects among genetic variants, which may reflect the assumption that all the PTVs in a gene may have the same biological consequence\cite{macarthur,rivas2013assessing,rivas2015effect} or prior information obtained through integration of additional data sources, such as functional assay data\cite{majithia2014rare,findlay2014saturation}, otherwise zero correlation of genetic effects may be assumed, which is used in dispersion tests like C-alpha\cite{calpha,clarke2013flexible} and SKAT\cite{skat}; and a K-by-K matrix $\mathbf{R_{\textrm{phen}}}$ being the correlations of genetic effects among phenotypes, which may be obtained from common variant data\cite{cotsapas2011pervasive,solovieff2013pleiotropy,gencorr2015}. The variance-covariance matrix of the effect sizes may be obtained from readily available summary statistic data such as in-study LD matrices, effect size estimate (or log odds ratio), and the standard error of the effect size estimate (Online Methods).

For gene discovery, the scale may be used to denote the prior spread of effect sizes. For instance, emerging empirical genetic studies have shown that within a gene protein truncating variants may have stronger effects than missense variants\cite{do2015exome}.

Thus far we have assumed that the prior mean, or location, of genetic effects is zero. This makes it feasible to analyze a large number of phenotypes without enumerating the prior mean across all phenotypes. However, in practice, we may want to proactively identify genetic variants that have effects that are consistent with a protective profile for a disease, which may exploit information from mendelian randomization studies of common variants, such as recent rare variant findings in {\it PCSK9} where truncating loss-of-function variants are found to decrease LDL and triglyceride levels, and decrease CAD risk\cite{cohen2005low,pcsk9,do2013common,cohorts2014loss}. This is straightforward to consider by including a non-zero vector as a prior mean of genetic effect (Online Methods).

Although we see advantages in adopting a Bayesian perspective, our approach could be used in a frequentist context by calculating a Bayes factor and using it as a test statistic to compute p-values (Methods, Fig.~\ref{Fig2}).

%\subsubsection{Mixture model for rare variant effects across multiple phenotypes}
%In Bayesian model comparison, the alternative model must be pre-specified. In practice, for gene discovery efforts, this has the attractive property that when we contrast it against a standard pre-specified null, i.e. genetic effects are zero across all studies in a group of variants and phenotypes, an analytical derivation is obtained making it computationally flexible and efficient. However, a limitation of this approach is that interpretation may not be straightforward. To make statistical inferences about the properties of the underlying mixture of genetic effects, we present a mixture model for estimating rare variant effects. Our goal is to cluster the variants into groups based on their effects on the multivariate phenotype. In particular, we allow variants to share clusters both within a gene and across the genes using a mixture model formulation of a clustering model where we also estimate the number of clusters in the data using Bayesian Information Criterion (BIC). More specifically, we can estimate (1) the proportion of non-null variants, (2) whether variants with the same annotation in one gene behave similarly, (3) whether variants across genes share effect profiles, or (4) whether different annotations differ in the magnitude of their effects.

\newpage 
%% FIGURE 2
\begin{figure}
\centering
   \includegraphics[width=\textwidth]{../figures/Figure2Nov2017.pdf}
  \caption{Simulation studies. A. Comparison of $-\log_{10}$(p-values) from frequentist $\textrm{BF}_{\textrm{MRP}}$ approximation for an independent effects and a similar effects model to commonly used gene-based statistical tests (skat and burden). B. Comparison of log10(Bayes Factors) obtained when raw genotype and phenotype data is available to a scenario where summary statistics only was available and similar effects across studies is assumed. C. From single variant and single phenotype to multiple variants and multiple phenotypes gene discovery: ROC curves for detecting gene association to any of the phenotypes using single variant/single phenotype association (blue) to multiple variants and multiple phenotypes association (purple). D. ROC curves for detecting gene association when incorporating prior mean of genetic effects (orange) to identify protective alleles. }
  \label{Fig2}
\end{figure}

\subsection{Simulation studies}
We first verified the analytical derivations and examined the properties of the approach under a simulation framework.

%% Comparison to other widely used methods
\subsubsection{Comparison to frequentist gene tests}
For the analysis of multiple rare variants and a single phenotype we compared it to the burden test and the SKAT test, commonly used statistical tests in rare variant association studies of a single phenotype. We observe concordance between the frequentist methods and the Bayesian models. To compare the Bayesian models we compute p-values by using the Bayes Factor as the test statistic and approximating it using distribution properties of quadratic forms (Appendix). As expected, an independent effects model has high correlation with the gene-based test SKAT ($r^2 = 0.99$), whereas the similar effects model has high correlation with the burden test ($r^2 = 0.93$, Fig.~\ref{Fig2}A).

\subsubsection{Summary statistic data}
To study the behavior of MRP using summary statistics we simulate two scenarios: first, the scenario where analysts have access to all the raw genotype and phenotype data; and second, the scenario where analysts only have access to summary statistics data\cite{liu2014meta}. We conducted $1000$ simulation experiments where we let $K$ (the number of phenotypes) $=3$, $M$ (the number of variants) $=10$, $S$ (the number of studies) $=2$, $N_0$ (number of individuals in study with access to all the data) $= 10000$, $N_1$ (meta-analysis study 1) $= 5000$, $N_2$ (meta-analysis study 2) $= 5000$. We find that, under the scenario where similar effects are assumed across studies, the Bayes Factors obtained using summary statistics alone are strongly correlated ($r^2 = 1$) to Bayes Factors obtained by the full genotype and phenotype data (Fig.~\ref{Fig2}B).

\subsubsection{From single variant and single phenotype analysis to multiple variants and multiple phenotypes}
To validate the flexibility of the approach we conducted a simulation experiment where we assumed an allelic architecture consistent to that discovered for {\it APOC3} in relation to coronary artery disease (CAD), triglycerides (TG), low-density lipoprotein cholesterol (LDL-C), and high-density lipoprotein cholesterol (HDL-C)\cite{apoc3,apoc32,jorgensen2014loss,cohorts2014loss}. We simulated three studies and applied the model comparison unit jointly to summary statistic data obtained for each study (Supplementary Note). Overall, we observed that considering the joint effects across multiple studies in a group of variants and phenotypes may improve ability to detect gene-based signals (Fig.~\ref{Fig2}C), and that considering prior mean of genetic effects should aid in efforts to identify protective modifiers of disease risk (Fig.~\ref{Fig2}D).

\subsection{Applications}
We demonstrate its use in rare variant analysis in the UK Biobank.

\subsubsection{Model comparison applied to asthma, eosinophil count, forced expiratory volume in 1-second, and forced vital capacity summary statistic data}
 First, we applied the model comparison component to summary statistic data from the exome content of the UK Biobank array. The summary statistic data was generated from single variant logistic regression and linear regression analysis applied to asthma, eosinophil count, forced expiratory volume in 1-second, and forced vital capacity phenotype files.

%% FIGURE 3
\begin{figure}
\centering
\includegraphics[width=\textwidth]{../figures/manhattan.pdf}
  \caption{MRP application to summary statistics from (A) 
  MRP application to summary statistic data from the GLGC and EOMI exome-chip studies. (A-E). Here, we show manhattan plots of meta-analysis results obtained by applying gene-based MRP across PTVs for (A) triglycerides (TG), (B) low-density lipoprotein-cholesterol (LDL-C), (C) high-density lipoprotein cholesterol (HDL-C), and (D) EOMI separately. (E-F) Gene-based MRP results when multiple phenotypes are analyzed and prior mean of genetic effect is included to prioritize protective and risk signals. For each gene the the $\log_{10}\left(\textrm{Bayes Factor}\right)$ is plotted.}
  \label{Fig3}
\end{figure}

\section{Discussion}
We developed a Bayesian model comparison approach for rare variant association studies by .

In this study, we applied the model comparison unit of the MRP framework to summary statistic data from

\newpage
\section{Methods}

\subsection{MRP model comparison for association testing}
We consider the multivariate linear regression model 
$$\underset{\left(N\times K\right)}{\mathbf{\textrm{Y}}} 
= \underset{\left(N\times K\right)}{\mathbf{\Psi}}  + \underset{\left(N\times M\right)}{\mathbf{\textrm{X}}}\underset{\left(M \times K\right)}{\mathbf{\textrm{B}}} 
+ \underset{\left(N\times K\right)}{\mathbf{\textrm{E}}},$$
where the matrices $\mathbf{\textrm{Y}} = \left[{y}_{i k}\right]$, $\mathbf{\textrm{X}} = \left[ x_{i m}\right]$, 
$\mathbf{\textrm{B}} = \left[ \beta_{m k} \right]$ and $\mathbf{\textrm{E}} = \left[e_{i k}\right]$  describe the phenotype values ($y_{i k}$),
copies of minor allele ($x_{i m}$), variant-phenotype effects ($\beta_{m k}$), and residual errors ($e_{i k}$), 
for individual $i$, phenotype $k$, and variant $m$. 
We assume that each phenotype has been transformed to a standard normal
distribution and that the columns of ${\mathbf{{\textrm{X}}}}$ 
have been centered, which means that the estimate for the intercept term ${\mathbf{\Psi}}$ is 0 and independent of the estimate of ${\mathbf{\textrm{B}}}$.
We use vectorized notation where the rows of ${\mathbf{\textrm{B}}}$ form vector 
$\bs {{ \beta}} = \left(\bs {{ \beta}}_{1},\dots,\bs {{\beta}}_{M}\right)^{\intercal}$
of length $MK$.

\noindent We define the MRP model comparison as a Bayes factor (BF) between the alternative model, where at least one variant
affects at least one phenotype, and the null model, where all variant-phenotype effects are zero.  
\noindent BF is the ratio of the marginal likelihoods for these two models: 
$$ \textrm{BF} = \frac{\int_{{\bs {\beta}}}p\left(\mathbf{\textrm{Data}}|{\bs {\beta}}\right)p\left({\bs {\beta}}|\textrm{ALT}\right)d{\bs {\beta}}}{\int_{{\bs {\beta}}}p\left(\mathbf{\textrm{Data}}|{\bs {\beta}}\right)p\left({\bs {\beta}}|\textrm{NULL}\right)d{\bs {\beta}}},$$ 
where Data can correspond either to the effect size estimates $\bs {\widehat{ \beta}}$ and the estimated variance-covariance matrix of  $\bs {\widehat{ \beta}}$, $\widehat{\mathbf{\textrm{V}}}_{\bs{\beta}}$, or to the original phenotypes and genotypes, 
$\underset{\left(N\times K\right)}{\mathbf{\textrm{\textrm{Y}}}}$ and $ \underset{\left(N\times M\right)}{\mathbf{\textrm{\textrm{X}}}}$, and any other covariates that we
want to regress out from the phenotypes.

The prior distribution for the null model, $p\left({\bs {\beta}}|\textrm{NULL}\right)$, is simply the point mass at $\bs \beta = 0$.
In subsection~\ref{approxlikelihood} we show how we approximate the likelihood function for $\bs \beta$, $p\left(\mathbf{\textrm{Data}} | {\bs {\beta}}\right)$,
in subsection~\ref{prioralt} we define the prior distribution $p\left({\bs {\beta}}|\textrm{ALT}\right)$ for the alternative model, 
and finally, in subsection~\ref{bfmodel}, we compute the BF.

\subsubsection{Likelihood function} 
\label{approxlikelihood}
\noindent A maximum likelihood estimator of $\mathbf{\textrm{B}}$ is given by the ordinary least-squares method
$$\widehat{\mathbf{\textrm{B}}} = \left(\mathbf{\textrm{X}}^\intercal\mathbf{\textrm{X}}\right)^{-1}\mathbf{\textrm{X}}^{\intercal}\mathbf{\textrm{Y}},$$
that in vectorized form is denoted by   
$\bs {\widehat{ \beta}} = \left(\bs {\widehat{ \beta}}_{1},\dots,\bs {\widehat{\beta}}_{M}\right)^{\intercal}.$
An estimator of the variance-covariance of $\widehat{\bs \beta}$  is given by 
$$\widehat{\textrm{V}}_{\mathbf{\bs \beta}} =  \left(\mathbf{\textrm{X}}^\intercal\mathbf{\textrm{X}}\right)^{-1}\otimes \widehat{\textrm{V}}_{\mathbf{\textrm{Y}}},$$
where $\widehat{\textrm{V}}_{\mathbf{\textrm{Y}}}$ is the estimated residual variance-covariance matrix of $\mathbf{\textrm{Y}}$ given $\mathbf{\textrm{X}}$.

\noindent 
%Let $\bs {\widehat{ \beta}} = \left(\bs {\widehat{ \beta}}^{\intercal}_{1},\dots,\bs {\widehat{\beta}}^{\intercal}_{m}\right)^{\intercal}$ be a vector of point estimates of all $MK$ effects. 
Following Band et al.\cite{band2013imputation}, we approximate the likelihood function of $\bs {\beta}$ by a multivariate normal distribution with mean $\bs {\widehat{\beta}}$ and variance-covariance matrix $\widehat{\textrm{V}}_{\bs {\beta}}$. Note that by approximating $\Vy$ by the trait correlation matrix, this likelihood approximation does not require access to the individual level data $\mathbf{\textrm{X}}$ and $\mathbf{\textrm{Y}}$ but only to the 
summary data of effect sizes $\bs {\widehat{\beta}}$, LD-matrix $\mathbf{\textrm{X}}^\intercal\mathbf{\textrm{X}}$ and a trait correlation estimate.

\subsubsection{Prior of $\bs{\beta}$ in the alternative model} 
\label{prioralt}
\noindent We construct the prior distribution $p\left({\bs {\beta}} | \textrm{ALT} \right)$ for the alternative model in three steps allowing 
user to specify correlations between effects of different variants on different traits across different studies.

\noindent In a single study, the prior density for ${\bs {\beta}}$ incorporates the expected correlation of genetic effects among 
a group of variants ($\mathbf{R}_{\textrm{var}}$) and among a group of phenotypes ($\mathbf{R}_{\textrm{phen}}$). 
In addition, we  incorporate an expected spread of the effect size of each variant by scaling $\mathbf{R}_{\textrm{var}}$ as 
$$\mathbf{S}_{\textrm{var}} = \Delta\left(\sigma_m\right) \mathbf{R}_{\textrm{var}} \Delta\left(\sigma_m\right),$$
\noindent where $\Delta\left(\sigma_m\right)$ is a diagonal matrix with entries $\sigma_m$ determining the spread of the effect size distribution for each variant $m \leq M$. Thus, we can model settings where, e.g., protein-truncating variants have larger effect sizes ($\sigma = 1$) than missense variants ($\sigma = 0.1$). 
Note that when $\sigma_m = 1$ for all $m$ then $\mathbf{S}_{\textrm{var}} = \mathbf{R}_{\textrm{var}}$. 
\noindent All in all, our prior density for ${\bs \beta}$ under alternative model is 
$${\bs {\beta}}|\textrm{ALT} \sim \mathcal{N}\left(\mathbf{0}, \mathbf{U}\right), \textrm{ where } \mathbf{U} = \mathbf{S}_{\textrm{var}} \otimes \mathbf{R}_{\textrm{phen}}.$$

\noindent When we have data from multiple studies 
we allow for possible differences in genetic effects across ethnicities or populations 
extending the Approximate Bayes Factors of Band et al.~\cite{band2013imputation}
and the summary statistics approach of RAREMETAL ~\cite{liu2014meta} from univariate to multivariate phenotypes.
Let $\bs {\widehat{ \beta}} = \left(\widehat{ \beta}_{s,m,k}\right)=\left(\widehat{\beta}_{1,1,1}, \widehat{\beta}_{1,1,2}, \dots, \widehat{\beta}_{1,1,K}, \widehat{\beta}_{1,2,1}, \dots, \widehat{\beta}_{1,2,K}, \dots, \widehat{\beta}_{1,M,K}, \widehat{\beta}_{2,1,1}, \dots, \widehat{\beta}_{S,M,K}\right),$ where $S$ is the number of studies, $M$ is the number of variants, and $K$ is the number of phenotypes. 
As with a single study, we incorporate the expected correlation of genetic effects between a pair of variants and a single phenotype using the matrix $\mathbf{S_{\textrm{var}}}$, between a variant and a pair of phenotypes using the matrix $\mathbf{R_{\textrm{phen}}}$, and we introduce the matrix $\mathbf{R_{\textrm{study}}}$ to specify prior on the similarity in effect sizes across the studies.
Thus, the prior is
$${\bs {\beta}} \sim \mathcal{N}\left(\mathbf{0}, \mathbf{U}\right), \textrm{ where } 
\mathbf{U} = \mathbf{R}_{\textrm{study}}\otimes\left(\mathbf{S}_{\textrm{var}} \otimes \mathbf{R}_{\textrm{phen}}\right).$$

\noindent It is straightforward to include a non-zero vector ${\bs\mu}$ as a prior mean of genetic effects, in which case the prior is
$${\bs {\beta}} \sim \mathcal{N}\left({\bs \mu},\mathbf{U}\right).$$
We use this, for example, when screening for protective rare variants that have a pre-specified beneficial profile on a set of risk factors.

\subsubsection{$\textrm{BF}_{\textrm{MRP}}$} 
\label{bfmodel}
\noindent The Bayes Factor is the ratio of the marginal likelihoods between the alternative and the null model. 
\noindent The marginal likelihood for the alternative model is 
$$ \int_{{\bs {\beta}}} p\left(\mathbf{\textrm{Data}}|{\bs {\beta}}\right) p\left({\bs {\beta}}|\textrm{ALT}\right)d{\bs {\beta}} = c \times \mathcal{N}\left({ {\widehat{\bs \beta}}}; {\bs \mu},  \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U}\right)$$
\noindent and the marginal likelihood for the null model is 
$$\int_{{\bs {\beta}}} p\left(\mathbf{\textrm{Data}}|{\bs {\beta}}\right) p\left({\bs {\beta}}|\textrm{NULL}\right)d{\bs {\beta}} = c \times \mathcal{N}\left({ {\widehat{\bs \beta}}}; 0,  \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}\right).$$
\noindent In Appendix (\ref{app1}) we show how we compute the Bayes Factor
$$\textrm{BF}_{\textrm{MRP}} = \frac{\det\left(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U} \right)^{-\frac{1}{2}}
\exp\left[-\frac{1}{2}\left({\bs {\widehat{\beta}} - {\bs \mu}}\right)^{\intercal}\left(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U}\right)^{-1}\left({\bs {\widehat{\beta}}} - {\bs \mu}\right)\right]}
{\det \left( \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}\right)^{-\frac{1}{2}}\exp \left[-\frac{1}{2}{\bs {\widehat{\beta}}}^{\intercal}\, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} 
\,{\bs {\widehat{\beta}}}\right]}.$$

When $\bs \mu=0$, $\textrm{BF}_{\textrm{MRP}}$ is an increasing function of the following quadratic form
\begin{equation}\label{quadratic_form}
Q(\bs{\widehat{\beta}};\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}},\mathbf{U})=
\bs{\widehat{\beta}^\intercal} \left(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}^{-1} - (\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}+\mathbf{U})^{-1}\right) \bs{\widehat{\beta}}.
\end{equation}
Furthermore, this quadratic form is the only part of the $\textrm{BF}_{\textrm{MRP}}$ that depends on $\bs {\widehat{\beta}}$.
Thus, by deriving a distribution of $Q(\bs{\widehat{\beta}};\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}},\mathbf{U})$
under the null model we can compute a p-value when $\textrm{BF}_{\textrm{MRP}}$ is used as a test statistic.
According to basic properties of quadratic forms of Gaussian variables,
$Q(\bs{\widehat{\beta}};\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}},\mathbf{U}) \sim \sum_{i=1}^n d_i \chi^2_i$, where $\chi_i^2$
are an independent sample from $\chi_1^2$ distribution (chi-square with one degree of freedom),
and $d_i$ are the eigenvalues of matrix
$I-(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}+\mathbf{U})^{-1}\, \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}$.
The distribution function for a mixture of chi-squares can be numerically evaluated by
the R-package `CompQuadForm'~\cite{Duchesne2010}.

%%%
%
%\subsection{MRP mixture modeling for effect size estimation} 
%\label{mrpmm}
%Index genes by $j = 1,\dots, J$ and denote variant $m$ in gene $j$ by $v_{jm}$. Our goal is to cluster the variants into groups based on their effects on the multivariate phenotype. To compare models with different numbers of clusters we use the Bayesian Information Criterion (BIC).  
%
%We represent each cluster $c$ by a mean effect size 
%$$ \bs b_c \sim \mathcal{N}\left(0,\bs \Theta_0\right),$$
%where $\bs \Theta_0$ is our prior estimate of genetic correlation across the traits; 
%and a covariance structure
%$\bs \Sigma_c = \widehat{\mathbf{\textrm{V}}}_Y $ (when phenotype data available) or $\bs \Sigma_c = \widehat{\mathbf{\textrm{V}}}_{jm}$ (when only summary statistics available),  
%where $\widehat{\mathbf{\textrm{V}}}_Y$ is an estimate of the phenotype correlation and remains fixed across the clusters, and $\widehat{\mathbf{\textrm{V}}}_{jm}$ will be defined below. 
%(Extensions to cluster-specific $\bs \Sigma_c$ are possible but not considered in this work.)
%
%Consider the model with $C>1$ clusters. To model the sharing of clusters across the genes we first draw a $C$ dimensional probability vector 
%$$\bs \pi_{0} \sim \mathrm{Dirichlet}\left(1,1,1,\ldots ,1\right).$$ Next, for each gene $j$, we draw a probability vector 
%$$\bs \pi_{j} | \bs \pi_{0} \sim \mathrm{Dirichlet}\left(\alpha \bs \pi_{0}\right)$$
%to determine the mixture proportions $\pi_{jc}$ that tell how the variants in gene $j$ are distributed across the clusters $c$. The parameter $\alpha$ governs how much sharing of clusters exists across genes with prior 
%$$\alpha \sim \textrm{Inv-Gamma}\left(1,1\right).$$ 
%To relate the clusters to the observed data, the phenotype of
%individual $i$ with a rare allele of variant $m$ with annotation $a$ in gene $j$ is
%
%$$\bs y_{i} | \left( \bs \pi_j, {\bs b}, \sigma_a \right) \sim \sum_{c=1}^{C}\pi_{jc}\mathcal{N}\left(\sigma_{a} \bs b_c, \bs \Sigma_c\right) $$
%where $\sigma_a$ corresponds to a scale parameter for annotation $a$ across all the clusters, where annotations may be, e.g., 
%protein truncating variant (PTV), missense variant, or other functional annotation category, with an inverse-gamma prior
%$$\sigma^2_{a} \sim \textrm{Inv-Gamma}\left(sh_a, sc_a\right),$$
%with hyperparameter values $sh_a$ (shape) and $sc_a$ (scale).
%At the above formulation, we assume that each individual carriers at most one of the rare variants considered.
%
%If we have access only to summary statistics of estimated effect sizes 
%$\widehat{\bs \beta}_{jm}$ and estimated variance-covariance 
%matrices $\widehat{\mathbf{\textrm{V}}}_{jm}$ for each variant $m$ in gene $j$, our sampling model for the data is
%$$\widehat{\bs \beta}_{jm} | \left( \bs \pi_j, {\bs b}, \sigma_a \right) \sim \sum_{c=1}^{C}\pi_{jc}\mathcal{N}\left(\sigma_{a} \bs b_c, \widehat{\mathbf{\textrm{V}}}_{jm} \right).$$
%Similar to the individual level data model above, this formulation assumes independence between the variants, which is 
%approximately true when each individual carries at most one of the rare variants considered.
%A connection between the two data types is
%that $\widehat{\bs \beta}_{jm} \approx \frac{1}{1-2f_{jm}} \overline{\bs y}_{jm}$ and
%$\widehat{\mathbf{\textrm{V}}}_{jm} \approx \frac{1}{2 N f_{v_{jm}}(1-2f_{jm})} \widehat{\mathbf{\textrm{V}}}_{Y}$ where $f_{jm}$ is the frequency of the 
%rare allele at variant $m$ of gene $j$ among the $2N$ haplotypes 
%and $\overline{\bs y}_{jm}$ is the mean phenotype of the carrier individuals of that allele.
%
%In the Appendix (\ref{app2}) we give an algorithm to analyze this mixture model using summary statistic data.
%
%To compare the models with different numbers of clusters we use BIC\cite{schwarz1978estimating}, defined for the model $\mathcal{M}_C$ with $C$ clusters, as
%$$BIC_C = -2\log\left(p\left(D | \widehat{\theta_C}, \mathcal{M}_C\right)\right) + \nu_C \log\left(n\right)$$ where $D$ denotes the observed data, $\widehat{\theta_C}$ is the vector of maximum likelihood estimates of the parameters of $\mathcal{M}_C$, $\nu_c$ is the number of independent parameters of $\mathcal{M}_c$ and $n$ is the number of data points. The difference in BIC values between two models approximates twice the logarithm of the Bayes factor between the models.
%
%In the Appendix (\ref{app2}) we give an algorithm to analyze this mixture model using summary statistic data.

\section{Appendix}
\subsection{MRP Bayes Factor computation} \label{app1}
To compute the Bayes Factor 
$$\textrm{BF}_{\textrm{MRP}} = \frac{\det\left(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U} \right)^{-\frac{1}{2}}
\exp\left[-\frac{1}{2}\left({\bs {\widehat{\beta}} - {\bs \mu}}\right)^{\intercal}\left(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U}\right)^{-1}\left({\bs {\widehat{\beta}}} - {\bs \mu}\right)\right]}
{\det \left( \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}\right)^{-\frac{1}{2}}\exp \left[-\frac{1}{2}{\bs {\widehat{\beta}}}^{\intercal}\, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} 
\,{\bs {\widehat{\beta}}}\right]},$$
we first consider the term inside the exponential function:
$$ \mathcal{E} \left(\widehat{\bs \beta}, \bs \mu, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}, \mathbf{U}\right) = 
\frac{1}{2}\, {\bs {\widehat{\beta}}}^{\intercal}\, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} \,{\bs {\widehat{\beta}}}
-\frac{1}{2}\left({\bs {\widehat{\beta}} - {\bs \mu}}\right)^{\intercal}\left(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U}\right)^{-1}\left({\bs {\widehat{\beta}}} - {\bs \mu}\right).
$$
Since $\widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}$ and $\mathbf{U}$ are typically defined through Kronecker products of smaller matrices, their inverses
are easier to compute than the inverse of their sum. Hence we use Woodbury matrix identity to write
$$ 
\mathcal{E} \left(\widehat{\bs \beta}, \bs \mu, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}, \mathbf{U}\right) = 
\frac{1}{2}\, {\bs {\widehat{\beta}}}^{\intercal}\, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} \,{\bs {\widehat{\beta}}}
-\frac{1}{2}\left({\bs {\widehat{\beta}} - {\bs \mu}}\right)^{\intercal} \left( \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} - \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} 
\left(\mathbf{U}^{-1} + \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} \right)^{-1} \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} \right ) 
\left({\bs {\widehat{\beta}} - {\bs \mu}}\right).
$$

To simplify the determinant calculation we write 
$$\det\left( \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U} \right) = \det\left( \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}\right) \det\left( \mathbf{I} + \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}^{-1} \mathbf{U}\right).$$
The logarithm of the Bayes Factor is then
$$\log \left( \textrm{BF}_{\textrm{MRP}} \right) = 
-\frac{1}{2} \log \left( \det\left( \mathbf{I} + \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}^{-1} \mathbf{U} \right) \right)  + \mathcal{E} \left(\widehat{\bs \beta}, \bs \mu, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}, \mathbf{U}\right).$$

If studies do not share individuals, $\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}$ is a block-diagonal matrix 
$$\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}=
 \left[\begin{array}{c|c|c|c}
 \widehat{\mathbf{\textrm{V}}}^{1}_{{\bs \beta}} & 0 & \cdots &0\\ \hline
0 & \widehat{\mathbf{\textrm{V}}}^{2}_{{\bs \beta}}&\cdots &0 \\ \hline
\vdots & &\ddots &\vdots \\ \hline
0 & 0 &\cdots & \widehat{\mathbf{\textrm{V}}}^{S}_{{\bs {\beta}}}\\
\end{array}\right].$$
If studies share individuals, e.g., controls, 
we take the approach of Cichonska et al.~\cite{cichonska2016metacca} to use summary level data
to estimate the correlation structure of the non-diagonal blocks caused by overlapping individuals.

%\subsection{MRP Mixture model computation} \label{app2}
%For multiparameter models, the joint posterior distribution may be difficult to sample from directly. 
%Often, it is easier to sample sequentially from the full conditional distribution of each parameter~\cite{hoff2009first} using a 
%Gibbs sampler, a Markov Chain Monte Carlo (MCMC) algorithm, that constructs a dependent sequence of parameter 
%values whose distribution approximates the joint posterior distribution~\cite{geman1984stochastic,casella1992explaining,hoff2009first}. 
%We propose the following algorithm that is augmented by a Metropolis-Hastings step to update the annotation scaling parameters $\bs \sigma_a^2$ and the sharing intensity parameter $\alpha$. 
%
%%% 03.01.2016 
%The implementation of the Gibbs sampler requires the following data structure: 
%Within gene $j$, the variants $1,\ldots,M_j$ are assigned to clusters $c=1,...,C$,
%each of which is represented by a parameter $(\bs b_c, \bs \Sigma_c)$ as described in Section~\ref{mrpmm} with
%$\bs \Sigma_c = \Vy$ being fixed to an estimate of the trait covariance across the clusters or $\widehat{\mathbf{\textrm{V}}}_{jm}$ for each allele when using summary statistics.
%
%The Gibbs sampler is run for $n_{\text{burn}} + n_{\text{iter}}$ iterations, of which the first $n_{\text{burn}}$
%are discarded as an initial "burn-in" part. 
%With the prior distributions given in Section~\ref{mrpmm}, and superscripts in parentheses denoting iteration, the algorithm is as follows:
%
%\begin{enumerate}
%\item Initialize 
%	\begin{enumerate}
%	\item Draw $\alpha^{\left(0\right)} \sim \textrm{Inv-Gamma}\left(1,1\right);$
%	\item Draw $\bs \pi_0^{\left(0\right)} \sim \textrm{Dirichlet}\left(1,1,1,\ldots, 1\right);$ %OK
%	\item Draw for each gene $j$: $\bs \pi_j^{\left(0\right)} \sim \textrm{Dirichlet}\left(\alpha^{(0)} \bs \pi_0^{\left(0\right)} \right);$
%	\item ${\bs b}^{\left(0\right)}_{1} = \bs 0;$
%	\item For each $c$ in $2,\dots,C$ draw $\bs b_c^{\left(0\right)} \sim \mathcal{N}\left(\bs 0, \bs \Theta_0\right);$
%	\item $\left(\sigma^2_{a}\right)^{\left(0\right)} = 0.2^2$ for all $a$.  
%	\end{enumerate}
%
%	\item Repeat for iteration $t = 1, 2, \dots, n_{\text{burn}} + n_{\text{iter}}$
%	\begin{enumerate}
%	\item Update $\bs \pi_0$. We use a Metropolis-Hastings sub-step~\cite{hastings1970monte} using a proposal centred around the current value. 
%	We sample a proposal value for 
%	$$\bs \pi_0^{'} \sim \textrm{Dirichlet}\left(\gamma \bs \pi_0^{(t-1)} \right).$$ 
%	To calculate the acceptance probability, we define the normalizing constant for the C-dimensional Dirichlet distribution with parameters $\bs z$ as
%	$$ D(\bs z) = \frac{\Gamma \left( \sum_{c=1}^C z_c \right) }{\prod_{c=1}^C \Gamma\left ( z_c \right)}$$ and the density at point $\bs x$ as
%	$$p_{\textit Dir}(\bs x | \bs z ) = D(\bs z) \prod_{c=1}^C x_c^{ z_c -1}.$$
%	With this notation the Metropolis-Hastings transition probability from $\bs \pi_0^{(t-1)}$ to $\bs \pi_0^{'}$ is $p_{\textit Dir}\left (\bs \pi_0^{'} | \gamma \bs \pi_0^{(t-1)} \right)$
%	and the density of the observed data depends on $\bs \pi_0^{(t-1)}$ only through the product $\prod_{j=1}^J p_{\textit Dir}\left (\bs \pi_j | \alpha \bs \pi_0^{(t-1)} \right ).$
%	Hence, the acceptance probability of the proposal is
%	$$
%	\lambda = \min\left( 1, \frac{  p_{\textit Dir}\left (\bs \pi_0^{(t-1)} | \gamma \bs \pi_0^{'} \right ) \prod_{j=1}^J p_{\textit Dir}\left (\bs \pi_j | \alpha \bs \pi_0^{'} \right )  }
%	{p_{\textit Dir}\left (\bs \pi_0^{'} | \gamma \bs \pi_0^{(t-1)} \right ) \prod_{j=1}^J p_{\textit Dir}\left (\bs \pi_j | \alpha \bs \pi_0^{(t-1)} \right ) } \right)$$
% Thus, with probability $\lambda$, we set $\bs \pi^{\left(t \right)}_0 = \bs \pi^{'}_0$ and with probability $1-\lambda$ we set $\bs \pi^{\left(t \right)}_0 = \bs \pi^{(t-1)}_0$.
%	
%	\item For each gene $j=1,\ldots , J$ update $\bs \pi_j$: $$\bs \pi_j \sim \textrm{Dirichlet}\left(\alpha \bs \pi_0 + \left(\sum_{m=1}^{M_j}\delta_{jm} = 1, 
%	\sum_{m=1}^{M_j}\delta_{jm} = 2, \dots, \sum_{m=1}^{M_j} \delta_{jm} = C\right)\right),$$
%	where $\delta_{jm}$ is the index of the cluster to which the variant $m$ of gene $j$ belongs to.
%	
%	%% update cluster membership 
%	\item Update $\delta_{jm}$ for all $j=1,\ldots, J$ and $m=1,\ldots,M_j$. 
%	For all $c$ compute $$p^'_{jmc} = p\left(\hat{\bs \beta}_{jm}; \bs b_c, \bs \Sigma_c\right)\pi_{jc},$$
%	and renormalize in such a way that $p_{jmc} \propto p^'_{jmc}$ sums to 1 over $c$.
%	Then we sample $$\delta_{jm} \sim \textrm{Discrete} (p_{jmc}).$$
%		
%	%% update cluster effect profile
%	\item Update $\bs b_{c}$ using a Gibbs update (see derivation below) from a Gaussian distribution with
%	\begin{eqnarray*}	
%	\textrm{mean}&=&\left(\bs \Theta_0^{-1} + \sum_{v_{jm} \in c}\sigma^2_{a_{jm}} \bs \Sigma^{-1}_{c} \right)^{-1} \left(\sum_{v_{jm} \in c}\sigma_{a_{jm}}\bs \Sigma_{c}^{-1}\hat{\bs \beta}_{jm}\right)\\
%	\textrm{var }&=& \left(\bs \Theta_0^{-1} + \sum_{v_{jm} \in c}\sigma^2_{a_{jm}}\bs \Sigma_{c}^{-1}\right)^{-1}.
%	\end{eqnarray*}
%	
%	%% Update sigma^2_a
%	\item Update $\sigma^2_{a}$. We use a Metropolis-Hastings sub-step~\cite{hastings1970monte} using a random-walk proposal. That is, sequentially for 
%	each annotation $a$, we sample a proposal value
%	$$
%	\sigma^{'}_{a} = |\eta|, \textrm{ where } \eta \sim \mathcal{N}\left(\sigma^{\left(t - 1\right)}_{a}, \xi_0\right),
%	$$
%	where $\xi_0$ is a hyperparameter controlling the spread of the proposals. In the examples we have used 
%	$\xi_0=?$. Then we calculate the acceptance probability  
%% Summary statistic data
%%% INSERT EQUATION FOR SUMMARY STATISTICS HERE. 
%	$$\lambda = \min\left(1,  
%	\frac{p\left(\sigma^{'}_a | sh_{a}, sc_{a} \right)
%	\prod_{\textrm{anno}(v_{jm})=a}\mathcal{N}\left(\widehat{\bs \beta}_{jm} | \sigma^{'}_{a}\bs b_{c_{jm}},\widehat{\bs V}_{jm} \right)}
%	{p\left(\sigma^{(t-1)}_a | sh_{a}, sc_{a} \right)
%	\prod_{\textrm{anno}(v_{jm})=a}\mathcal{N}\left(\widehat{\bs \beta}_{jm} | \sigma^{(t-1)}_{a}\bs b_{c_{jm}},\widehat{\bs V}_{jm} \right)}\right),$$
%	where the products are over those variants $v_{jm}$ whose annotation is $a$ and $c_{jm}$ is the cluster of variant $v_{jm}.$
%	With probability $\lambda$, we set $\sigma^{(t)}_{a} = \sigma^{'}_{a}$ and with probability
%	$1-\lambda$ we set $\sigma^{(t)}_{a} =\sigma^{(t-1)}_{a}.$
%
%	\item Update $\alpha$. We use a Metropolis-Hastings sub-step using a random-walk proposal. That is, we sample 
%	a proposal value $\alpha^' = |\eta|$, where $\eta \sim \mathcal{N}(\alpha^{(t-1)}, \xi_\alpha )$ where 
%	$\xi_\alpha$ is a fixed value controlling the variance of the proposal distribution. 
%	Then we compute the acceptance probability 
%	$$\lambda = \min\left(1,  
%	\frac{p\left(\alpha^{'} \right) \prod_{j=1}^J p_{\textit Dir}\left(\bs \pi_j^{(t)}  | \alpha^{'} \bs\pi_0^{(t)}\right) }
%	{p\left(\alpha^{(t-1)} \right)  \prod_{j=1}^J p_{\textit Dir}\left(\bs \pi_j^{(t)}  | \alpha^{(t-1)} \bs\pi_0^{(t)} \right) } \right),$$
%	where the prior for $\alpha$ is Inv-Gamma(1,1), i.e., $p(\alpha) = \alpha^{-2}\exp\left(-\alpha^{-1}\right)$, and
%	$p_{\textit Dir}$ is the density of the Dirichlet distribution as defined above in part (a).
%	With probability $\lambda$, we set $\alpha^{(t)} = \alpha^{'}$ and with probability
%	$1-\lambda$ we set $\alpha^{(t)} = \alpha^{(t-1)}.$
%
%	\end{enumerate}
%
%
%\end{enumerate}

\subsection{Implementation details: HDF5 tables}
Although summary statistics are quicker to read and process than raw data, the number of studies meta-analyzed in this work is expected to be sufficiently large to require optimizations in data representation and processing. Our solution was the use of the HDF5 (Hierarchical Data Format 5) data representation to enable rapid processing of effect size, uncertainty, and cross-trait estimate data. HDF5 is a fast and lightweight file format designed for scientific data. It has bindings for R, Python, C/C++, Java, and nearly every other population programming language. Reading data from a table within a HDF5 file can be an order of magnitude faster than reading text files from a Unix file, and it makes it easier to organize data within an internal structure.

%% FIGURE 3
\begin{figure}
\centering
\includegraphics[width=\textwidth]{/Users/mrivas/Dropbox/MultiplePhenotypesManuscript/hdf5MAMBAMRP.pdf}
  \caption{Our HDF5 implementation contained the following components: first, a group with one table per annotation file. All effect size (beta) values and study-specific annotations were contained here, and the number of tables is limited by $S\ \textrm{(the number of studies)} \times K\ \textrm{(the number of traits)}$. Second, a group with site-site covariance data. While these covariance matrices may have dimension $M\ \textrm{(the number of variants)} \times M$, we store the data as tables, each row specifiying the covariance between two variants. The number of tables should be the same as the previous set, capped by $S\ \textrm{(the number of studies)} \times K\ \textrm{(the number of traits)}$. Third, we store one table with sigma values for each study/phenotype combination. In the event that the traits were rank-normal transformation was performed these sigma values are equal to 1. These are used to compute correlation between two datasets. Finally, we store a matrix/table pair for Vy null and its index. The Vy null matrix has dimensions $ (S \times K) \times (S \times K)$ each entry specifying the estimated correlation of effect sizes between two datasets. The index table encodes row/column position of each dataset.}
  \label{hdf5}
\end{figure}

\newpage

\bibliography{references}

\end{document}

### 2017-12-22 tex/MVMetaLearnMRP2017OctoberMRPMM.tex

@@
 \title{Bayesian model comparison for rare variant association studies of multiple phenotypes}
-\author{Christopher DeBoever$^{1}$, Matthew Aguirre$^{1}$, Yosuke Tanigawa$^{1}$, Adam Lavertu$^{1}$, Greg McInnes$^{1}$, Timothy Poterba$^{2}$, Mark J.
+\author{Chris Mark DeBoever$^{1}$, Matthew Aguirre$^{1}$, Yosuke Tanigawa$^{1}$, Adam Lavertu$^{1}$, Greg McInnes$^{1}$, Timothy Poterba$^{2}$, Mark J.
 Daly$^{2,3}$,  Matti Pirinen$^{4}$\thanks{matti.pirinen@helsinki.fi}, and Manuel A.
@@
 Through simulation studies, we demonstrate that the proposed model comparison approach may improve ability to detect rare variant association signals.
-Finally, we demonstrate its use with rare variant data combined with asthma diagnosis and haematological and spirometry measures for individuals in the UK Biobank.
+Finally, we demonstrate its use with rare variant data and haematological measures in individuals from multiple population groups in the UK Biobank.
 We show that we are able to retain useful features from widely-used meta-analysis approaches and prioritize protective modifiers of disease risk.
@@
 %% Sequencing technologies
-Sequencing technologies are quickly transforming human genetic studies of complex traits: it is increasingly possible to obtain whole genome sequence data on thousands of samples at manageable costs.
+Sequencing technologies are quickly transforming human genetic studies of complex traits: it is increasingly possible to obtain whole genome sequences on thousands of samples at manageable costs.
 As a result, the genome-wide study of rare variants (minor allele frequency [MAF] $< 1\%$) and their contribution to disease susceptibility and phenotype variation is now feasible\cite{ifih1,altshuler2010map,rivas2011deep,10002012integrated}.
@@
 Given the widespread observations of shared genetic risk factors across distinct diseases, there is also considerable motivation to proactively modify gene discovery approaches accordingly.
-In other words, rather than only aggregating variants that may have effects on a single phenotype, we can also bring together sets of phenotypes for which a single variant or sets of variants might have effects.
+In other words, rather than only bringing together variants that may have effects on a single phenotype also bring together sets of phenotypes for which a single variant or sets of variants might have effects.
 
@@
 In practice, sharing of individual genotype and phenotype data across groups in large genetic consortium studies is difficult to achieve.
-We present a derivation where summary statistics, such as estimates of effect size and the corresponding standard error from single variant and single phenotype analysis, are the input data.
+We present a derivation where summary statistics, such as estimates of effect size and standard error of a single variant and single phenotype analysis, are the input data.
 Furthermore, we use insights from Liu et al.\cite{liu2014meta} and Cichonska et al.\cite{cichonska2016metacca} who suggest the use of additional summary statistics, like covariance estimates across variants and studies, respectively, that would enable lossless ability to detect gene-based association signals using summary statistics alone.
@@
 Genome variant annotation is critical for the interpretation of genetic findings.
-In the MRP model comparison we present approaches for including priors on spread of effect sizes.
-An example of an important class of variants where functional prediction is much more straightforward is the class of variants predicted to truncate the protein product, commonly referred to as protein truncating variants or PTV\cite{rivas2013assessing,rivas2015effect}, and in this paper we place special emphasis on the interpretation of PTVs.
+In MRP model comparison we present approaches for including priors on spread of effect sizes.
+An example of an important class of variants where functional prediction is much more straightforward is that which is predicted to truncate the protein product, commonly referred to as protein truncating variants or PTV\cite{rivas2013assessing,rivas2015effect}, and in this paper we place special emphasis on the interpretation of PTVs.
 
@@
 %%Apply methods to the GLGC data and exome sequencing data sets. 
-We apply MRP to four phenotypes: asthma (HC382: the corresponding phenotype label in Global Biobank Engine [\url{https://biobankengine.stanford.edu}]), eosinophil count (INI30150), Forced Expiratory Volume in 1-second (FEV1, INI3063), and Forced Vital Capacity (FVC, INI3062) in the UK Biobank dataset.
+We apply MRP to four phenotypes: asthma (HC382: the corresponding phenotype label in Global Biobank Engine [\url{https://biobankengine.stanford.edu}]), Eosinophil count (INI30150), Forced Expiratory Volume in 1-second (FEV1, INI3063), and Forced Vital Capacity (FVC, INI3062) in the UK Biobank dataset.
 We show that we are able to retain useful features from single variant and single phenotype meta-analysis, rare variant analysis, and prioritize protective modifiers of disease risk.
@@
   \centering
-  \includegraphics[width=\textwidth]{../figures/FIG1FINAL.pdf}
+  \includegraphics[width=\textwidth]{/Users/mrivas/Dropbox/MultiplePhenotypesManuscript/figure/FIG1FINAL.pdf}
   \caption{Schematic overview of MRP.
@@
 
-Let $N$ be the number of individuals and $K$ the number of quantitative phenotype measurements on each individual, for example, here in the applications we focus on asthma, eosinophil count, Forced Expiratory Volume in 1-second, and Forced Vital Capacity, in which case $K=4$.
+Let $N$ be the number of individuals and $K$ the number of quantitative phenotype measurements on each individual, for example, here in the applications we focus on asthma, Eosinophil count, Forced Expiratory Volume in 1-second, and Forced Vital Capacity, in which case $K=4$.
 Let $M$ be the number of variants in a testing unit ${\bs G}$, where ${\bs G}$ can be, for example, a gene, pathway, or a network.
@@
 
-For gene discovery, the scale may be used to denote the prior spread of effect sizes.
-For instance, emerging empirical genetic studies have shown that within a gene protein truncating variants may have stronger effects than missense variants\cite{do2015exome}.
+For gene discovery, the scale, may be used to denote the prior spread of effect sizes.
+For instance, emerging empirical genetic studies have shown that with a gene protein truncating variants may have stronger effects than missense variants\cite{do2015exome}.
 
@@
 \centering
-   \includegraphics[width=\textwidth]{../figures/Figure2Nov2017.pdf}
+   \includegraphics[width=\textwidth]{/Users/mrivas/Dropbox/MultiplePhenotypesManuscript/Figure2Nov2017.pdf}
   \caption{Simulation studies.
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/manhattan.pdf}
-  \caption{MRP application to summary statistics from (A) 
-  MRP application to summary statistic data from the GLGC and EOMI exome-chip studies.
+\includegraphics[width=\textwidth]{/Users/mrivas/git/mrp/figures/manhattan.pdf}
+  \caption{MRP application to summary statistic data from the GLGC and EOMI exome-chip studies.
 (A-E).

### 2017-12-23 tex/MVMetaLearnMRP2017OctoberMRPMM.tex

--- 2017-12-22 tex/MVMetaLearnMRP2017OctoberMRPMM.tex
@@
 \includegraphics[width=\textwidth]{../figures/manhattan.pdf}
-  \caption{MRP application to summary statistics from (A) 
-  MRP application to summary statistic data from the GLGC and EOMI exome-chip studies.
-(A-E).
-Here, we show manhattan plots of meta-analysis results obtained by applying gene-based MRP across PTVs for (A) triglycerides (TG), (B) low-density lipoprotein-cholesterol (LDL-C), (C) high-density lipoprotein cholesterol (HDL-C), and (D) EOMI separately.
-(E-F) Gene-based MRP results when multiple phenotypes are analyzed and prior mean of genetic effect is included to prioritize protective and risk signals.
-For each gene the the $\log_{10}\left(\textrm{Bayes Factor}\right)$ is plotted.}
+  \caption{$\log_{10}$ Bayes Factors from applying MRP to summary statistics for missense and protein-truncating variants from (A) eosinophil counts (INI30150), (B) forced vital capacity (FVC, INI3062), (C) forced expiratory volume in 1-second (FEV1, INI3063), (D) asthma (HC382), and (E) all four traits jointly.
+The four genes outside of chromosome 6 with the largest Bayes Factors are labeled in each plot.
+Only $\log_{10}$ Bayes Factors greater than -5 are plotted.}
   \label{Fig3}
@@
 the R-package `CompQuadForm'~\cite{Duchesne2010}.
+
+\subsection{UK Biobank Data}
+\subsubsection{GWAS Summary Statistics}
+We performed genome-wide association analysis using PLINK v2.00a(17 July 2017) as described in ~\cite{DeBoever179762}.
+For asthma, we used the Firth-fallback in PLINK, a hybrid algorithm which normally uses the logistic regression code described in (Hill 2017), but switches to a port of logistf() (https://cran.r-project.org/web/packages/logistf/index.html) in two cases: (1) one of the cells in the 2x2 allele count by case/control status contingency table is empty (2) logistic regression was attempted since all the contingency table cells were nonzero, but it failed to converge within the usual number of steps.
+We used the following covariates in our analysis: age, sex, array type, and the first four principal components, where array type is a binary variable that represents whether an individual was genotyped with UK Biobank Axiom Array or UK BiLEVE Axiom Array.
+For variants that were specific to one array, we did not use array as a covariate.
+Asthma cases were defined using both Hospital Episode Statistics and verbal questionnaire responses.
+We used the provided values from the UK Biobank for eosinophil counts, forced vital capacity (FVC), and forced expiratory volume in 1-second (FEV1).
+The phenotype codes used throughout (asthma=HC382, eosinophil count=INI30150, Forced Expiratory Volume in 1-second (FEV1)=INI3063, and Forced Vital Capacity (FVC)=INI3062) correspond to the phenotype codes used in on the Global Biobank Engine [\url{https://biobankengine.stanford.edu}].
+
+\subsubsection{Genetic Correlations}
+We calculated the genetic between asthma, eosinophil counts, forced vital capacity (FVC), and forced expiratory volume in 1-second (FEV1) using the MultiVariate Polygenic Mixture Model (MVPMM) described in ~\cite{DeBoever2017}.
+Briefly, MVPMM estimates genetic correlation given GWAS summary statistics (effect size and standard error of effect size estimate) by modeling GWAS summary statistics as generated from one of two mixture components.
+Summary statistics from variants in the null component are modeled as being drawn from a multivariate normal distribution with zero mean and covariance matrix that captures correlation in the summary statistics due to the use of shared subjects or other sources of correlation.
+Summary statistics from variants in the non-null component are modeled as being drawn from a multivariate normal distribution with zero mean as well.
+The covariance matrix for the non-null component combines the covariance matrix from the null component with another covariance matrix that captures the genetic correlation between the phenotypes being considered.
 

### 2017-12-23 tex/MVMetaLearnMRP2017OctoberMRPMM.tex
Text updates

@@
 \subsubsection{Model comparison applied to asthma, eosinophil count, forced expiratory volume in 1-second, and forced vital capacity summary statistic data}
- First, we applied the model comparison component to summary statistic data from the exome content of the UK Biobank array.
-The summary statistic data was generated from single variant logistic regression and linear regression analysis applied to asthma, eosinophil count, forced expiratory volume in 1-second, and forced vital capacity phenotype files.
+We applied the model comparison component to summary statistic data from the exome content of the UK Biobank array.
+The summary statistic data was generated from single variant logistic regression and linear regression analysis applied to asthma, eosinophil count, forced expiratory volume in 1-second, and forced vital capacity phenotypes.
+We applied the model to missense variants and PTVs for each phenotype separately (\ref{Fig3}A-D) as well as to all phenotypes jointly (\ref{Fig3}E) and obtained $\log_{10}$ BF for each gene.
+We observed evidence that missense variants and PTVs in \textit{IL33} affect both eosinophil counts and asthma from the single-phenotype analyses, though the evidence was strongest for the joint analysis (\ref{Tab1}).
 
@@
 \end{figure}
+
+%% TABLE 3
+\begin{table}
+\begin{tabular}{llllll}
+\toprule
+{} & Joint & Eosinophil count &   FVC &  FEV1 & Asthma \\
+\midrule
+\textit{IL33}   &  36.6 &             31.5 &  -2.3 &  -1.9 &    9.4 \\
+\textit{GPR20}  &  10.2 &             10.4 &  -0.1 &  -0.2 &        \\
+\textit{ZFAT}   &  10.1 &             -3.1 &   7.8 &   5.9 &   -0.5 \\
+\textit{CSF2RB} &   9.0 &              9.1 &   0.2 &  -0.1 &   -0.2 \\
+\textit{SCMH1}  &   7.1 &             -1.5 &   6.8 &   2.6 &   -0.9 \\
+\textit{PACS2}  &   6.2 &              9.6 &  -1.3 &  -1.6 &   -0.5 \\
+\textit{CCR3}   &   6.0 &             14.0 &  -3.3 &  -3.7 &   -1.0 \\
+\textit{ELN}    &   5.0 &             -1.5 &   4.9 &   2.3 &   -0.6 \\
+\bottomrule
+\end{tabular}
+\caption{$\log_{10}$ Bayes Factors for genes outside of chromosome 6 with $\log_{10}$ Bayes Factors greater than 5 in the joint analysis.}
+\label{Tab1}
+\end{table}
 

### 2017-12-23 tex/MVMetaLearnMRP2017OctoberMRPMM.tex
Text updates

@@
 %% Rare variant methods
-In genetic studies of diseases or continuous phenotypes, rare variants are hard to assess individually due to the limited number of copies.
+In genetic studies of diseases or continuous phenotypes, rare variants are hard to assess individually due to the limited number of copies of each rare variant.
 Hence, to boost the ability to detect a signal, evidence is usually `aggregated' across variants.
 When designing an `aggregation' method, there are three questions that are usually considered.
-First, across which biological units should variants be combined; second, which variants mapping within those units should be included\cite{majithia2014rare}; and third, which statistical model should be used\cite{lee2014rare}.
+First, across which biological units should variants be combined; second, which variants mapping within those units should be included\cite{majithia2014rare}; and third, which statistical model should be used\cite{lee2014rare}?
 Given the widespread observations of shared genetic risk factors across distinct diseases, there is also considerable motivation to proactively modify gene discovery approaches accordingly.
@@
 %% The model comparison method
-In this paper, we present a Bayesian multiple rare variants and phenotypes (MRP) model comparison approach, as an alternative to current widely-used statistical tests.
+In this paper, we present a Bayesian multiple rare variants and phenotypes (MRP) model comparison approach for identifying rare variant associations as an alternative to current widely-used statistical tests.
 The MRP framework, as implemented, can exploit correlation, scale, or location (direction) of genetic effects in a broad range of rare variant association study designs including: case-control; multiple diseases and shared controls; single continuous phenotype; multiple continuous phenotypes; or a mixture of case-control and multiple continuous phenotypes (Fig.~\ref{Fig1}).
@@
 \subsection{Methods overview}
-A complete description of MRP, along with an analytical derivation, is given in the Online Methods and Appendix. Briefly, we describe approach.
+A complete description of MRP, along with an analytical derivation, is given in the Online Methods and Appendix. Briefly, we describe the approach.
 
@@
 \subsubsection{Model comparison applied to asthma, eosinophil count, forced expiratory volume in 1-second, and forced vital capacity summary statistic data}
-We applied the model comparison component to summary statistic data from the exome content of the UK Biobank array.
+We applied the MRP model comparison approach to summary statistic data for coding variants on the UK Biobank array.
 The summary statistic data was generated from single variant logistic regression and linear regression analysis applied to asthma, eosinophil count, forced expiratory volume in 1-second, and forced vital capacity phenotypes.
 We applied the model to missense variants and PTVs for each phenotype separately (\ref{Fig3}A-D) as well as to all phenotypes jointly (\ref{Fig3}E) and obtained $\log_{10}$ BF for each gene.
-We observed evidence that missense variants and PTVs in \textit{IL33} affect both eosinophil counts and asthma from the single-phenotype analyses, though the evidence was strongest for the joint analysis (\ref{Tab1}).
+We observed evidence that missense variants and PTVs in \textit{IL33} affect eosinophil counts and offer protection from asthma from the single-phenotype analyses\cite{DeBoever179762, 10.1371/journal.pgen.1006659}, though the evidence of association was strongest for the joint analysis (\ref{Tab1}).
+We also found evidence for associations between missense variants and PTVs in several other genes and the four phenotypes in the joint model that were not identified when considering asthma alone.
+For instance, \textit{GPR20}, \textit{CSF2RB}, \textit{PACS2}, and \textit{CCR3} all had $\log_{10}$ BF greater than 9 in the eosinophil count-only analysis but $\log_{10}$ BF less than 0 for the asthma-only analysis.
+In the joint model, the $\log_{10}$ BFs for these genes were all greater than six indicating possible associations with these traits.
+Besides \textit{IL33}, none of the genes with $\log_{10}$ BF greater than six in the joint model were reported in a large GWAS for allergic disease including asthma\cite{Ferreira:2017ba} though \textit{CCR3} is near a locus associated with atopy in a previous meta-analysis\cite{Ober:2011jk}.
+These results illustrate the ability of applying model comparison to multiple traits simultaneously to identify new genes involved in disease.
 
@@
 \section{Discussion}
-We developed a Bayesian model comparison approach for rare variant association studies by .
+In this study, we developed a Bayesian model comparison approach MRP that shares information across both variants and phenotypes to identify rare variant associations.
+We used simulations to compare MRP to the widely used burden and SKAT tests for identifying rare variant associations and found that jointly considering both variants and phenotypes can improve the ability to detect associations.
+We also applied the MRP model comparison framework to summary statistic data from four traits from the UK Biobank: asthma, eosinophil counts, forced expiratory volume in 1-second (FEV1), and forced vital capacity (FVC).
+We identified strong evidence for the previously described association between \textit{IL33} and asthma\cite{DeBoever179762, 10.1371/journal.pgen.1006659} as well as evidence of association for several additional genes that have not been previously reported \ref{Tab1}.
 
-In this study, we applied the model comparison unit of the MRP framework to summary statistic data from
+As genetic data linked to high-dimensional phenotype data continues to be made available through biobanks, health systems, and research programs, there is a large need for statistical approaches that can leverage information across different genetic variants, phenotypes, and studies to make strong inferences about disease-associated genes.
+The approach presented here relies only on summary statistics from marginal association analyses which can be shared with less privacy concerns compared to raw genotype and phenotype data.
+Combining joint analysis of variants and phenotypes with meta-analysis across studies offers new opportunities to identify gene-disease associations.
 
@@
 \subsubsection{GWAS Summary Statistics}
-We performed genome-wide association analysis using PLINK v2.00a(17 July 2017) as described in ~\cite{DeBoever179762}.
+We performed genome-wide association analysis using PLINK v2.00a(17 July 2017) as previously described~\cite{DeBoever179762}.
 For asthma, we used the Firth-fallback in PLINK, a hybrid algorithm which normally uses the logistic regression code described in (Hill 2017), but switches to a port of logistf() (https://cran.r-project.org/web/packages/logistf/index.html) in two cases: (1) one of the cells in the 2x2 allele count by case/control status contingency table is empty (2) logistic regression was attempted since all the contingency table cells were nonzero, but it failed to converge within the usual number of steps.
@@
 \subsubsection{Genetic Correlations}
-We calculated the genetic between asthma, eosinophil counts, forced vital capacity (FVC), and forced expiratory volume in 1-second (FEV1) using the MultiVariate Polygenic Mixture Model (MVPMM) described in ~\cite{DeBoever2017}.
+We calculated the genetic between asthma, eosinophil counts, forced vital capacity (FVC), and forced expiratory volume in 1-second (FEV1) using the MultiVariate Polygenic Mixture Model (MVPMM)~\cite{DeBoever2017}.
 Briefly, MVPMM estimates genetic correlation given GWAS summary statistics (effect size and standard error of effect size estimate) by modeling GWAS summary statistics as generated from one of two mixture components.
 Summary statistics from variants in the null component are modeled as being drawn from a multivariate normal distribution with zero mean and covariance matrix that captures correlation in the summary statistics due to the use of shared subjects or other sources of correlation.
-Summary statistics from variants in the non-null component are modeled as being drawn from a multivariate normal distribution with zero mean as well.
-The covariance matrix for the non-null component combines the covariance matrix from the null component with another covariance matrix that captures the genetic correlation between the phenotypes being considered.
+Summary statistics from variants in the non-null component are modeled as being drawn from a multivariate normal distribution with zero mean, but the covariance matrix for the non-null component combines the covariance matrix from the null component with another covariance matrix that captures the genetic correlation between the phenotypes being considered.
 

### 2017-12-23 tex/MVMetaLearnMRP2017OctoberMRPMM.tex
Minor text update

@@
 
+\subsubsection{UK Biobank Asthma Application}
+For the Manhattan plots \ref{Fig3} and table \ref{Tab1}, we removed any genes with non-unique gene symbols.
+We also removed the sense overlapping gene \textit{CTD-3064M3.3} and and the antisense gene \textit{ZFAT-AS1} since these genes overlap \textit{GPR20} and \textit{ZFAT} respectively and therefore have the same $\log_{10}$ Bayes Factors as those genes.
+
 %%%
@@
 \centering
-\includegraphics[width=\textwidth]{/Users/mrivas/Dropbox/MultiplePhenotypesManuscript/hdf5MAMBAMRP.pdf}
+\includegraphics[width=\textwidth]{../figures/hdf5MAMBAMRP.pdf}
   \caption{Our HDF5 implementation contained the following components: first, a group with one table per annotation file.

### 2018-01-15 tex/MVMetaLearnMRP2017OctoberMRPMM.tex
adding editing nature class that permits figure inclusion
Text updates

@@
 \usepackage[T1]{fontenc}
-\usepackage[dvips]{graphicx}
 \usepackage[usenames]{color}
@@
 \usepackage{graphicx}
+\graphicspath{{../figures/}}
 \usepackage{multirow}
@@
 \def\Vy{\widehat{\textrm{V}}_Y}
-
 \bibliographystyle{naturemag}
 \title{Bayesian model comparison for rare variant association studies of multiple phenotypes}
-\author{Christopher DeBoever$^{1}$, Matthew Aguirre$^{1}$, Yosuke Tanigawa$^{1}$, Adam Lavertu$^{1}$, Greg McInnes$^{1}$, Timothy Poterba$^{2}$, Mark J.
+\author{Christopher DeBoever$^{1}$, Matthew Aguirre$^{1}$, Yosuke Tanigawa$^{1}$, Timothy Poterba$^{2}$, Mark J.
 Daly$^{2,3}$,  Matti Pirinen$^{4}$\thanks{matti.pirinen@helsinki.fi}, and Manuel A.
@@
 \linenumbers
+
 \begin{document}
@@
 %%Apply methods to the GLGC data and exome sequencing data sets. 
-We apply MRP to four phenotypes: asthma (HC382: the corresponding phenotype label in Global Biobank Engine [\url{https://biobankengine.stanford.edu}]), eosinophil count (INI30150), Forced Expiratory Volume in 1-second (FEV1, INI3063), and Forced Vital Capacity (FVC, INI3062) in the UK Biobank dataset.
+We also apply MRP to two groups of related phenotypes.
+First, we apply MRP to asthma (HC382: the corresponding phenotype label in Global Biobank Engine [\url{https://biobankengine.stanford.edu}]), eosinophil count (INI30150), Forced Expiratory Volume in 1-second (FEV1, INI3063), and Forced Vital Capacity (FVC, INI3062) in the UK Biobank dataset.
+We next apply MRP to glaucoma (HC276), intra-ocular pressure (INI5263), and corneal resistance factor (INI5265).
 We show that we are able to retain useful features from single variant and single phenotype meta-analysis, rare variant analysis, and prioritize protective modifiers of disease risk.
@@
   \centering
-  \includegraphics[width=\textwidth]{../figures/FIG1FINAL.pdf}
+  \includegraphics[width=\textwidth]{FIG1FINAL.pdf}
   \caption{Schematic overview of MRP.
@@
 \centering
-   \includegraphics[width=\textwidth]{../figures/Figure2Nov2017.pdf}
+   \includegraphics[width=\textwidth]{Figure2Nov2017.pdf}
   \caption{Simulation studies.
@@
 
-\subsubsection{Model comparison applied to asthma, eosinophil count, forced expiratory volume in 1-second, and forced vital capacity summary statistic data}
-We applied the MRP model comparison approach to summary statistic data for coding variants on the UK Biobank array.
-The summary statistic data was generated from single variant logistic regression and linear regression analysis applied to asthma, eosinophil count, forced expiratory volume in 1-second, and forced vital capacity phenotypes.
-We applied the model to missense variants and PTVs for each phenotype separately (\ref{Fig3}A-D) as well as to all phenotypes jointly (\ref{Fig3}E) and obtained $\log_{10}$ BF for each gene.
-We observed evidence that missense variants and PTVs in \textit{IL33} affect eosinophil counts and offer protection from asthma from the single-phenotype analyses\cite{DeBoever179762, 10.1371/journal.pgen.1006659}, though the evidence of association was strongest for the joint analysis (\ref{Tab1}).
+\subsubsection{Model comparison applied to summary statistic from the UK Biobank}
+We applied the MRP model comparison approach to summary statistic data generated from single variant logistic regression and linear regression analysis for coding variants on the UK Biobank array (Methods).
+We first applied MRP to the asthma, eosinophil count, forced expiratory volume in 1-second, and forced vital capacity phenotypes.
+We applied the model to missense variants and PTVs for each phenotype separately (Fig.~\ref{asthma}A-D) as well as to all phenotypes jointly (Fig.~\ref{asthma}E) and obtained $\log_{10}$ BF for each gene.
+As expected from recent studies \cite{DeBoever179762, 10.1371/journal.pgen.1006659}, we observed evidence that missense variants and PTVs in \textit{IL33} affect eosinophil counts and offer protection from asthma from the single-phenotype analyses, though the evidence of association was strongest for the joint analysis (Table~\ref{asthma_table}).
 We also found evidence for associations between missense variants and PTVs in several other genes and the four phenotypes in the joint model that were not identified when considering asthma alone.
@@
 Besides \textit{IL33}, none of the genes with $\log_{10}$ BF greater than six in the joint model were reported in a large GWAS for allergic disease including asthma\cite{Ferreira:2017ba} though \textit{CCR3} is near a locus associated with atopy in a previous meta-analysis\cite{Ober:2011jk}.
-These results illustrate the ability of applying model comparison to multiple traits simultaneously to identify new genes involved in disease.
+
+We also applied MRP to missense variants and PTVs for glaucoma, intra-ocular pressure, and corneal resistance factor as well as performing a joint analysis and a joint analysis focused on protective associations.
+While the individual glaucoma analysis did not yield any associations with $\log_{10}$ BF greater than 5, the joint analyses identified missense and PTVs in both \textit{ANGPTL7} and \textit{WNT10A} as associated with protection against glaucoma.
+These results illustrate the ability of applying model comparison to multiple traits simultaneously to identify protective modifiers of disease.
 
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/manhattan.pdf}
-  \caption{$\log_{10}$ Bayes Factors from applying MRP to summary statistics for missense and protein-truncating variants from (A) eosinophil counts (INI30150), (B) forced vital capacity (FVC, INI3062), (C) forced expiratory volume in 1-second (FEV1, INI3063), (D) asthma (HC382), and (E) all four traits jointly.
-The four genes outside of chromosome 6 with the largest Bayes Factors are labeled in each plot.
-Only $\log_{10}$ Bayes Factors greater than -5 are plotted.}
-  \label{Fig3}
+\includegraphics[width=\textwidth]{glaucoma_manhattan.pdf}
+  \caption{$\log_{10}$ Bayes Factors from applying MRP to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
+(E) shows the results of a joint analysis focused on finding rare variants that protect against glaucoma.
+The genes outside of chromosome 6 with with Bayes Factor greater than three are indicated by arrows.
+Only $\log_{10}$ Bayes Factors greater than zero are plotted.}
+  \label{glaucoma}
 \end{figure}
-
-%% TABLE 3
-\begin{table}
-\begin{tabular}{llllll}
-\toprule
-{} & Joint & Eosinophil count &   FVC &  FEV1 & Asthma \\
-\midrule
-\textit{IL33}   &  36.6 &             31.5 &  -2.3 &  -1.9 &    9.4 \\
-\textit{GPR20}  &  10.2 &             10.4 &  -0.1 &  -0.2 &        \\
-\textit{ZFAT}   &  10.1 &             -3.1 &   7.8 &   5.9 &   -0.5 \\
-\textit{CSF2RB} &   9.0 &              9.1 &   0.2 &  -0.1 &   -0.2 \\
-\textit{SCMH1}  &   7.1 &             -1.5 &   6.8 &   2.6 &   -0.9 \\
-\textit{PACS2}  &   6.2 &              9.6 &  -1.3 &  -1.6 &   -0.5 \\
-\textit{CCR3}   &   6.0 &             14.0 &  -3.3 &  -3.7 &   -1.0 \\
-\textit{ELN}    &   5.0 &             -1.5 &   4.9 &   2.3 &   -0.6 \\
-\bottomrule
-\end{tabular}
-\caption{$\log_{10}$ Bayes Factors for genes outside of chromosome 6 with $\log_{10}$ Bayes Factors greater than 5 in the joint analysis.}
-\label{Tab1}
-\end{table}
 
@@
 We also applied the MRP model comparison framework to summary statistic data from four traits from the UK Biobank: asthma, eosinophil counts, forced expiratory volume in 1-second (FEV1), and forced vital capacity (FVC).
-We identified strong evidence for the previously described association between \textit{IL33} and asthma\cite{DeBoever179762, 10.1371/journal.pgen.1006659} as well as evidence of association for several additional genes that have not been previously reported \ref{Tab1}.
+We identified strong evidence for the previously described association between \textit{IL33} and asthma\cite{DeBoever179762, 10.1371/journal.pgen.1006659} as well as evidence of association for several additional genes that have not been previously reported (Table~\ref{asthma_table}).
 
@@
 \subsubsection{UK Biobank Asthma Application}
-For the Manhattan plots \ref{Fig3} and table \ref{Tab1}, we removed any genes with non-unique gene symbols.
+For the Manhattan plots and Table \ref{asthma_table}, we removed any genes with non-unique gene symbols.
 We also removed the sense overlapping gene \textit{CTD-3064M3.3} and and the antisense gene \textit{ZFAT-AS1} since these genes overlap \textit{GPR20} and \textit{ZFAT} respectively and therefore have the same $\log_{10}$ Bayes Factors as those genes.
@@
 %In the Appendix (\ref{app2}) we give an algorithm to analyze this mixture model using summary statistic data.
+
+\subsection{Author Contributions}
+M.A.R.
+and M.P.
+designed the method and derived all analytical calculations.
+M.A.R., M.P., and C.D.
+wrote the manuscript.
+Y.T., M.A., and C.D.
+provided analysis and designed figures.
+T.P.
+designed hDF5 tables and implementation of loaders.
+M.J.D.
+provided critical feedback on methodology.
+
+\subsection{Acknowledgments}
+This research has been conducted using the UK Biobank Resource under Application Number 24983.
+We thank all the participants in the UK Biobank study.
+M.A.R.
+and C.D.
+are supported by Stanford University and a National Institute of Health center for Multi- and Trans-ethnic Mapping of Mendelian and Complex Diseases grant (5U01 HG009080).
+C.D.
+is supported by a postdoctoral fellowship from the Stanford Center for Computational, Evolutionary, and Human Genomics and the Stanford ChEM-H Institute.
+Y .T.
+is supported by Funai Overseas Scholarship from Funai Foundation for Information Technology and the Stanford University Biomedical Informatics Training Program.
+The primary and processed data used to generate the analyses presented here are available in the UK Biobank access management system (https://amsportal.ukbiobank.ac.uk/) for application 24983, ?Generating effective therapeutic hypotheses from genomic and hospital linkage data?
+(http://www.ukbiobank.ac.uk/wp-content/uploads/2017/06/24983-Dr-Manuel-Rivas.pdf), and the results are displayed in the Global Biobank Engine (https://biobankengine.stanford.edu).
+We would like to thank the Customer Solutions Team from Paradigm4 who helped us implement efficient databases for queries and application of inference methods to the data, and also implemented optimized versions of truncated singular value decomposition.
+M.A.R.
+is a paid consultant in Genomics PLC.
 
@@
 \subsection{Implementation details: HDF5 tables}
-Although summary statistics are quicker to read and process than raw data, the number of studies meta-analyzed in this work is expected to be sufficiently large to require optimizations in data representation and processing.
+Although summary statistics are quicker to read and process than raw data, the number of studies meta-analyzed in this work is expected to be sufficiently large to require optimizations in data representation and processing (Figure \ref{hdf5}).
 Our solution was the use of the HDF5 (Hierarchical Data Format 5) data representation to enable rapid processing of effect size, uncertainty, and cross-trait estimate data.
@@
 
+ %% FIGURE: ASTHMA MANHATTAN
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{asthma_manhattan.pdf}
+  \caption{$\log_{10}$ Bayes Factors from applying MRP to summary statistics for missense and protein-truncating variants from (A) eosinophil counts (INI30150), (B) forced vital capacity (FVC, INI3062), (C) forced expiratory volume in 1-second (FEV1, INI3063), (D) asthma (HC382), and (E) all four traits jointly.
+The four genes outside of chromosome 6 with the largest Bayes Factors are labeled in each plot.
+Only $\log_{10}$ Bayes Factors greater than -5 are plotted.}
+  \label{asthma}
+\end{figure}
+
+%% TABLE 3
+\begin{table}
+\begin{tabular}{llllll}
+\toprule
+{} & Joint & Eosinophil count &   FVC &  FEV1 & Asthma \\
+\midrule
+\textit{IL33}   &  36.6 &             31.5 &  -2.3 &  -1.9 &    9.4 \\
+\textit{GPR20}  &  10.2 &             10.4 &  -0.1 &  -0.2 &        \\
+\textit{ZFAT}   &  10.1 &             -3.1 &   7.8 &   5.9 &   -0.5 \\
+\textit{CSF2RB} &   9.0 &              9.1 &   0.2 &  -0.1 &   -0.2 \\
+\textit{SCMH1}  &   7.1 &             -1.5 &   6.8 &   2.6 &   -0.9 \\
+\textit{PACS2}  &   6.2 &              9.6 &  -1.3 &  -1.6 &   -0.5 \\
+\textit{CCR3}   &   6.0 &             14.0 &  -3.3 &  -3.7 &   -1.0 \\
+\textit{ELN}    &   5.0 &             -1.5 &   4.9 &   2.3 &   -0.6 \\
+\bottomrule
+\end{tabular}
+\caption{$\log_{10}$ Bayes Factors for genes outside of chromosome 6 with $\log_{10}$ Bayes Factors greater than 5 in the joint analysis.}
+\label{asthma_table}
+\end{table}
+
 %% FIGURE 3
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/hdf5MAMBAMRP.pdf}
+\includegraphics[width=\textwidth]{hdf5MAMBAMRP.pdf}
   \caption{Our HDF5 implementation contained the following components: first, a group with one table per annotation file.

### 2018-01-15 tex/MVMetaLearnMRP2017OctoberMRPMM.tex

--- 2017-12-23 tex/MVMetaLearnMRP2017OctoberMRPMM.tex
@@
 The summary statistic data was generated from single variant logistic regression and linear regression analysis applied to asthma, eosinophil count, forced expiratory volume in 1-second, and forced vital capacity phenotypes.
-We applied the model to missense variants and PTVs for each phenotype separately (\ref{Fig3}A-D) as well as to all phenotypes jointly (\ref{Fig3}E) and obtained $\log_{10}$ BF for each gene.
-We observed evidence that missense variants and PTVs in \textit{IL33} affect eosinophil counts and offer protection from asthma from the single-phenotype analyses\cite{DeBoever179762, 10.1371/journal.pgen.1006659}, though the evidence of association was strongest for the joint analysis (\ref{Tab1}).
+We applied the model to missense variants and PTVs for each phenotype separately (Fig.~\ref{Fig3}A-D) as well as to all phenotypes jointly (Fig.~\ref{Fig3}E) and obtained $\log_{10}$ BF for each gene.
+We observed evidence that missense variants and PTVs in \textit{IL33} affect eosinophil counts and offer protection from asthma from the single-phenotype analyses\cite{DeBoever179762, 10.1371/journal.pgen.1006659}, though the evidence of association was strongest for the joint analysis (Table~\ref{Tab1}).
 We also found evidence for associations between missense variants and PTVs in several other genes and the four phenotypes in the joint model that were not identified when considering asthma alone.
@@
 \centering
-\includegraphics[width=\textwidth]{../figures/manhattan.pdf}
+\includegraphics[width=\textwidth]{asthma_manhattan.pdf}
   \caption{$\log_{10}$ Bayes Factors from applying MRP to summary statistics for missense and protein-truncating variants from (A) eosinophil counts (INI30150), (B) forced vital capacity (FVC, INI3062), (C) forced expiratory volume in 1-second (FEV1, INI3063), (D) asthma (HC382), and (E) all four traits jointly.
@@
 We also applied the MRP model comparison framework to summary statistic data from four traits from the UK Biobank: asthma, eosinophil counts, forced expiratory volume in 1-second (FEV1), and forced vital capacity (FVC).
-We identified strong evidence for the previously described association between \textit{IL33} and asthma\cite{DeBoever179762, 10.1371/journal.pgen.1006659} as well as evidence of association for several additional genes that have not been previously reported \ref{Tab1}.
+We identified strong evidence for the previously described association between \textit{IL33} and asthma\cite{DeBoever179762, 10.1371/journal.pgen.1006659} as well as evidence of association for several additional genes that have not been previously reported (Table~\ref{Tab1}).
 
@@
 
+\subsection{Author Contributions}
+M.A.R.
+and M.P.
+designed the method and derived all analytical calculations.
+M.A.R., M.P., and C.D.
+wrote the manuscript.
+Y.T., M.A., and C.D.
+provided analysis and designed figures.
+T.P.
+designed hDF5 tables and implementation of loaders.
+M.J.D.
+provided critical feedback on methodology.
+
+\subsection{Acknowledgments}
+This research has been conducted using the UK Biobank Resource under Application Number 24983.
+We thank all the participants in the UK Biobank study.
+M.A.R.
+and C.D.
+are supported by Stanford University and a National Institute of Health center for Multi- and Trans-ethnic Mapping of Mendelian and Complex Diseases grant (5U01 HG009080).
+C.D.
+is supported by a postdoctoral fellowship from the Stanford Center for Computational, Evolutionary, and Human Genomics and the Stanford ChEM-H Institute.
+Y .T.
+is supported by Funai Overseas Scholarship from Funai Foundation for Information Technology and the Stanford University Biomedical Informatics Training Program.
+The primary and processed data used to generate the analyses presented here are available in the UK Biobank access management system (https://amsportal.ukbiobank.ac.uk/) for application 24983, ?Generating effective therapeutic hypotheses from genomic and hospital linkage data?
+(http://www.ukbiobank.ac.uk/wp-content/uploads/2017/06/24983-Dr-Manuel-Rivas.pdf), and the results are displayed in the Global Biobank Engine (https://biobankengine.stanford.edu).
+We would like to thank the Customer Solutions Team from Paradigm4 who helped us implement efficient databases for queries and application of inference methods to the data, and also implemented optimized versions of truncated singular value decomposition.
+M.A.R.
+is a paid consultant in Genomics PLC.
+
 \section{Appendix}

### 2018-01-15 tex/MVMetaLearnMRP2017OctoberMRPMM.tex
Text updates

--- 2018-01-15 tex/MVMetaLearnMRP2017OctoberMRPMM.tex
@@
 Here we introduce a model comparison approach we refer to as MRP for rare variant association studies that considers correlation, scale, and location of genetic effects across a group of genetic variants, phenotypes, and studies.
-In so doing we consider the use of summary statistic data to 1) apply standard univariate and multivariate gene-based meta-analysis models, 2) models for assessing heterogeneity of genetic effects, which may be used in practice for downstream quality control, and 3) models for identifying protective protein-truncating variants, which can expedite drug discovery.
+In so doing we consider the use of summary statistic data to apply 1) standard univariate and multivariate gene-based meta-analysis models, 2) models for assessing heterogeneity of genetic effects, which may be used in practice for downstream quality control, and 3) models for identifying protective protein-truncating variants, which can expedite drug discovery.
 Through simulation studies, we demonstrate that the proposed model comparison approach may improve ability to detect rare variant association signals.
@@
 As a result, the genome-wide study of rare variants (minor allele frequency [MAF] $< 1\%$) and their contribution to disease susceptibility and phenotype variation is now feasible\cite{ifih1,altshuler2010map,rivas2011deep,10002012integrated}.
+
 %% Rare variant methods
@@
 First, across which biological units should variants be combined; second, which variants mapping within those units should be included\cite{majithia2014rare}; and third, which statistical model should be used\cite{lee2014rare}?
-Given the widespread observations of shared genetic risk factors across distinct diseases, there is also considerable motivation to proactively modify gene discovery approaches accordingly.
+Given the widespread observations of shared genetic risk factors across distinct diseases, there is also considerable motivation to use gene discovery approaches that leverage the information from multiple phenotypes jointly.
 In other words, rather than only aggregating variants that may have effects on a single phenotype, we can also bring together sets of phenotypes for which a single variant or sets of variants might have effects.
@@
 In this paper, we present a Bayesian multiple rare variants and phenotypes (MRP) model comparison approach for identifying rare variant associations as an alternative to current widely-used statistical tests.
-The MRP framework, as implemented, can exploit correlation, scale, or location (direction) of genetic effects in a broad range of rare variant association study designs including: case-control; multiple diseases and shared controls; single continuous phenotype; multiple continuous phenotypes; or a mixture of case-control and multiple continuous phenotypes (Fig.~\ref{Fig1}).
-
-In Bayesian model comparison, we compute a Bayes Factor (BF) defined as the ratio of the marginal likelihoods of the observed data under two models: 1) a pre-specified null where all genetic effects are zero; and 2) an alternative model where factors like correlation, scale, or location of genetic effects are considered.
+The MRP framework exploits correlation, scale, or location (direction) of genetic effects in a broad range of rare variant association study designs including: case-control; multiple diseases and shared controls; single continuous phenotype; multiple continuous phenotypes; or a mixture of case-control and multiple continuous phenotypes (Fig.~\ref{Fig1}).
+MRP makes use of Bayesian model comparison, whereby we compute a Bayes Factor (BF) defined as the ratio of the marginal likelihoods of the observed data under two models: 1) a pre-specified null where all genetic effects are zero; and 2) an alternative model where factors like correlation, scale, or location of genetic effects are considered.
+The BF is an alternative to $p$-values from traditional hypothesis testing.
+For MRP, the BF represents the statistical evidence for a non-zero effect for a particular group of rare variants on the phenotype(s) of interest.
 
 %% Use of summary statistic data
-In practice, sharing of individual genotype and phenotype data across groups in large genetic consortium studies is difficult to achieve.
-We present a derivation where summary statistics, such as estimates of effect size and the corresponding standard error from single variant and single phenotype analysis, are the input data.
+While many large genetic consortia generally collect both raw genotype and phenotype data, in practice, sharing of individual genotype and phenotype data across groups is difficult to achieve.
+To address this, MRP can take summary statistics, such as estimates of effect size and the corresponding standard error from typical single variant-single phenotype linear or logistic regressions, as input data.
 Furthermore, we use insights from Liu et al.\cite{liu2014meta} and Cichonska et al.\cite{cichonska2016metacca} who suggest the use of additional summary statistics, like covariance estimates across variants and studies, respectively, that would enable lossless ability to detect gene-based association signals using summary statistics alone.
 
-%% Genome variant annotation and protein truncating variants
-Genome variant annotation is critical for the interpretation of genetic findings.
-In the MRP model comparison we present approaches for including priors on spread of effect sizes.
-An example of an important class of variants where functional prediction is much more straightforward is the class of variants predicted to truncate the protein product, commonly referred to as protein truncating variants or PTV\cite{rivas2013assessing,rivas2015effect}, and in this paper we place special emphasis on the interpretation of PTVs.
-
-%% Protective modifiers of disease risk
-Given the translational potential of identifying protective modifiers of human disease risk we characterize how model comparison may be used to improve discovery of such protective signals.
-We present an extension allowing for modeling of location (direction) of genetic effects, which may be useful for prioritizing variants or genes that are consistent with a protective profile of disease\cite{pcsk9,cohen2006sequence,sullivan2012effect}.
+%% Genome variant annotation and protein truncating variants and  protective modifiers of disease risk
+Aggregation techniques rely on variant annotations to assign variants to groups for analysis.
+MRP allows for the inclusion of priors on the spread of effect sizes that can be adjusted depending on what type of variants are included in the analysis.
+For instance, protein truncating variants (PTVs)\cite{rivas2013assessing,rivas2015effect} are an important class of variants that are more likely to be functional because they often disrupt the normal function of a gene.
+This biological knowledge can be reflected in the choices of priors for PTVs in MRP.
+Since PTVs typically abolish gene function, there is particular interest in identifying protective PTV modifiers of human disease risk that may serve as targets for therapeutics\cite{pcsk9,cohen2006sequence,sullivan2012effect}.
+We therefore demonstrate how the MRP model comparison approach can improve discovery of such protective signals by modeling the location (direction) of genetic effects which prioritizes variants or genes that are consistent with protecting against disease.
 
@@
 %%Apply methods to the GLGC data and exome sequencing data sets. 
-We also apply MRP to two groups of related phenotypes.
-First, we apply MRP to asthma (HC382: the corresponding phenotype label in Global Biobank Engine [\url{https://biobankengine.stanford.edu}]), eosinophil count (INI30150), Forced Expiratory Volume in 1-second (FEV1, INI3063), and Forced Vital Capacity (FVC, INI3062) in the UK Biobank dataset.
-We next apply MRP to glaucoma (HC276), intra-ocular pressure (INI5263), and corneal resistance factor (INI5265).
-We show that we are able to retain useful features from single variant and single phenotype meta-analysis, rare variant analysis, and prioritize protective modifiers of disease risk.
+We also apply MRP to summary statistics for two groups of related phenotypes from the UK Biobank.
+First, we apply MRP to asthma (HC382: the corresponding phenotype label in Global Biobank Engine [\url{https://biobankengine.stanford.edu}]), eosinophil count (INI30150), Forced Expiratory Volume in 1-second (FEV1, INI3063), and Forced Vital Capacity (FVC, INI3062).
+We find that XXX.
+We next apply MRP to glaucoma (HC276), intra-ocular pressure (INI5263), and corneal resistance factor (INI5265) and find strong evidence that rare PTVs in \textit{ANGPTL7} protect against glaucoma.
+These analyses show that MRP recovers results from typical single variant-single phenotype analyses while identifying new rare variant associations that include protective modifiers of disease risk.
 
@@
 extending the Approximate Bayes Factors of Band et al.~\cite{band2013imputation}
-and the summary statistics approach of RAREMETAL ~\cite{liu2014meta} from univariate to multivariate phenotypes.
+and the summary statistics approach of RAREMETAL~\cite{liu2014meta} from univariate to multivariate phenotypes.
 Let $\bs {\widehat{ \beta}} = \left(\widehat{ \beta}_{s,m,k}\right)=\left(\widehat{\beta}_{1,1,1}, \widehat{\beta}_{1,1,2}, \dots, \widehat{\beta}_{1,1,K}, \widehat{\beta}_{1,2,1}, \dots, \widehat{\beta}_{1,2,K}, \dots, \widehat{\beta}_{1,M,K}, \widehat{\beta}_{2,1,1}, \dots, \widehat{\beta}_{S,M,K}\right),$ where $S$ is the number of studies, $M$ is the number of variants, and $K$ is the number of phenotypes.

### 2018-01-18 tex/MVMetaLearnMRP2017OctoberMRPMM.tex
Updates to figures and tables

@@
 We first applied MRP to the asthma, eosinophil count, forced expiratory volume in 1-second, and forced vital capacity phenotypes.
-We applied the model to missense variants and PTVs for each phenotype separately (Fig.~\ref{asthma}A-D) as well as to all phenotypes jointly (Fig.~\ref{asthma}E) and obtained $\log_{10}$ BF for each gene.
+We applied the model to missense variants and PTVs for each phenotype separately (Fig.~\ref{asthma_manhattan}A-D) as well as to all phenotypes jointly (Fig.~\ref{asthma_manhattan}E) and obtained $\log_{10}$ BF for each gene.
 As expected from recent studies \cite{DeBoever179762, 10.1371/journal.pgen.1006659}, we observed evidence that missense variants and PTVs in \textit{IL33} affect eosinophil counts and offer protection from asthma from the single-phenotype analyses, though the evidence of association was strongest for the joint analysis (Table~\ref{asthma_table}).
@@
 Only $\log_{10}$ Bayes Factors greater than zero are plotted.}
+  \label{glaucoma_manhattan}
+\end{figure}
+
+%% FIGURE 4
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{glaucoma_forest.pdf}
+  \caption{$-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{ANGPTL7} for glaucoma, intra-ocular pressure, and corneal resistance factor.}
   \label{glaucoma}
 \end{figure}
+
+%% TABLE: GLAUCOMA GENES OF INTEREST
+\begin{table}
+\begin{tabular}{l p{3cm} p{1cm} p{1.5cm} p{3.8cm} p{3.5cm}}
+\toprule
+{} & Joint, protective & Joint & Glaucoma & Intra-ocular pressure, Goldmann-correlated & Corneal resistance factor \\
+\midrule
+\textit{ANGPTL7} &              10.1 &   9.4 &      1.7 &                                       10.1 &                       6.2 \\
+\textit{KLHL22}  &               3.1 &   3.6 &     -0.2 &                                        3.9 &                       2.2 \\
+\textit{GLI3}    &               2.9 &   2.9 &     -0.3 &                                       -0.7 &                       3.2 \\
+\textit{WNT10A}  &               2.2 &   3.0 &     -0.3 &                                       -0.5 &                       3.5 \\
+\textit{RNASE2}  &              -1.3 &  -0.3 &      0.4 &                                       -1.0 &                      -1.1 \\
+\textit{FUK}     &              -6.5 &  -2.3 &      2.5 &                                       -2.1 &                      -2.1 \\
+\bottomrule
+\end{tabular}
+\caption{$\log_{10}$ Bayes Factors for genes highlighted in Figure \ref{glaucoma_manhattan}.}
+\label{glaucoma_table}
+\end{table}
 
@@
 
- %% FIGURE: ASTHMA MANHATTAN
-\begin{figure}
-\centering
-\includegraphics[width=\textwidth]{asthma_manhattan.pdf}
-  \caption{$\log_{10}$ Bayes Factors from applying MRP to summary statistics for missense and protein-truncating variants from (A) eosinophil counts (INI30150), (B) forced vital capacity (FVC, INI3062), (C) forced expiratory volume in 1-second (FEV1, INI3063), (D) asthma (HC382), and (E) all four traits jointly.
-The four genes outside of chromosome 6 with the largest Bayes Factors are labeled in each plot.
-Only $\log_{10}$ Bayes Factors greater than -5 are plotted.}
-  \label{asthma}
-\end{figure}
-
-%% TABLE 3
-\begin{table}
-\begin{tabular}{llllll}
-\toprule
-{} & Joint & Eosinophil count &   FVC &  FEV1 & Asthma \\
-\midrule
-\textit{IL33}   &  36.6 &             31.5 &  -2.3 &  -1.9 &    9.4 \\
-\textit{GPR20}  &  10.2 &             10.4 &  -0.1 &  -0.2 &        \\
-\textit{ZFAT}   &  10.1 &             -3.1 &   7.8 &   5.9 &   -0.5 \\
-\textit{CSF2RB} &   9.0 &              9.1 &   0.2 &  -0.1 &   -0.2 \\
-\textit{SCMH1}  &   7.1 &             -1.5 &   6.8 &   2.6 &   -0.9 \\
-\textit{PACS2}  &   6.2 &              9.6 &  -1.3 &  -1.6 &   -0.5 \\
-\textit{CCR3}   &   6.0 &             14.0 &  -3.3 &  -3.7 &   -1.0 \\
-\textit{ELN}    &   5.0 &             -1.5 &   4.9 &   2.3 &   -0.6 \\
-\bottomrule
-\end{tabular}
-\caption{$\log_{10}$ Bayes Factors for genes outside of chromosome 6 with $\log_{10}$ Bayes Factors greater than 5 in the joint analysis.}
-\label{asthma_table}
-\end{table}
-
-%% FIGURE 3
+%% FIGURE: HDF5 TABLE
 \begin{figure}
@@
 
+ %% FIGURE: ASTHMA MANHATTAN
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{asthma_manhattan.pdf}
+  \caption{$\log_{10}$ Bayes Factors from applying MRP to summary statistics for missense and protein-truncating variants from (A) eosinophil counts (INI30150), (B) forced vital capacity (FVC, INI3062), (C) forced expiratory volume in 1-second (FEV1, INI3063), (D) asthma (HC382), and (E) all four traits jointly.
+The four genes outside of chromosome 6 with the largest Bayes Factors are labeled in each plot.
+Only $\log_{10}$ Bayes Factors greater than -5 are plotted.}
+  \label{asthma_manhattan}
+\end{figure}
+
+%% FIGURE: ASTHMA FOREST PLOT
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{asthma_forest.pdf}
+  \caption{$-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{IL33} for eosinophil counts, forced vital capacity (FVC), forced expiratory volume in 1-second (FEV1), and asthma.}
+  \label{asthma_forest}
+\end{figure}
+
+%% TABLE: ASTHMA GENES OF INTEREST
+\begin{table}
+\begin{tabular}{lllllll}
+\toprule
+{} & Joint, protective & Joint & Eosinophil count &   FVC &  FEV1 & Asthma \\
+\midrule
+\textit{IL33}         &              30.1 &  29.8 &             30.9 &  -3.1 &  -2.7 &    9.5 \\
+\textit{CCR3}         &               5.8 &   6.3 &              9.1 &  -1.7 &  -2.3 &   -0.5 \\
+\textit{IL17RA}       &               4.6 &   6.1 &             14.1 &  -4.1 &  -3.9 &   -1.4 \\
+\textit{AL161450.1}   &               3.2 &   3.8 &              4.9 &  -0.1 &  -0.4 &    0.3 \\
+\textit{RP11-39K24.9} &               3.2 &   3.8 &              4.9 &  -0.1 &  -0.4 &    0.3 \\
+\textit{ELN}          &               2.9 &   3.2 &             -1.0 &   5.6 &   2.9 &   -0.6 \\
+\textit{SCMH1}        &               2.7 &   3.9 &             -2.1 &   7.5 &   3.3 &   -0.7 \\
+\textit{TGFB2}        &               2.6 &   4.1 &             -2.8 &   1.7 &  -2.4 &    0.1 \\
+\textit{ZFAT}         &               1.3 &   1.9 &             -2.8 &   6.7 &   4.6 &   -0.4 \\
+\textit{STC2}         &               0.2 &   1.3 &             -0.8 &   3.9 &   2.3 &   -0.7 \\
+\textit{P4HTM}        &              -1.3 &  -0.7 &             -1.6 &   2.9 &   4.3 &   -1.0 \\
+\textit{IL18RAP}      &              -2.2 &  -1.0 &             -0.8 &  -2.0 &  -2.2 &    3.6 \\
+\textit{WDR6}         &              -4.1 &  -2.9 &             -2.3 &   1.9 &   3.4 &   -0.9 \\
+\textit{ATP2A3}       &              -4.2 &  -1.5 &             -1.8 &  -3.1 &  -3.1 &    5.9 \\
+\textit{CHI3L1}       &              -4.4 &  -2.0 &             -2.0 &  -1.9 &  -2.0 &    3.0 \\
+\bottomrule
+\end{tabular}
+\caption{$\log_{10}$ Bayes Factors for genes highlighted in Figure \ref{asthma_manhattan}.}
+\label{asthma_table}
+\end{table}
+
 \newpage

### 2018-01-23 tex/manuscript.tex
Working on moving to PLOS format
Not done yet.

--- 2018-01-18 tex/MVMetaLearnMRP2017OctoberMRPMM.tex
@@
-\documentclass{nature}
-\usepackage[T1]{fontenc}
-\usepackage[usenames]{color}
-\usepackage{amsthm}
-\usepackage{amsmath}
-\usepackage{amscd}
-\usepackage{amssymb}
-\usepackage{amsfonts}
-\usepackage{eucal}
-% Please add the following required packages to your document preamble:
- \usepackage{booktabs}
-\usepackage{rotating}
-\usepackage{graphicx}
-\graphicspath{{../figures/}}
-\usepackage{multirow}
-\usepackage{bbm}
-\usepackage[font=small]{caption}
-\usepackage[font=small]{subcaption}
-\usepackage{color}
-\usepackage{lineno}
-\usepackage{url}
-\usepackage{lscape}
-\usepackage{breqn}
-\usepackage{enumitem}
-\usepackage{listings}
+% Template for PLoS
+% Version 3.4 January 2017
+%
+% % % % % % % % % % % % % % % % % % % % % %
+%
+% -- IMPORTANT NOTE
+%
+% This template contains comments intended 
+% to minimize problems and delays during our production 
+% process. Please follow the template instructions
+% whenever possible.
+%
+% % % % % % % % % % % % % % % % % % % % % % % 
+%
+% Once your paper is accepted for publication, 
+% PLEASE REMOVE ALL TRACKED CHANGES in this file 
+% and leave only the final text of your manuscript. 
+% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
+% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
+%
+%
+% There are no restrictions on package use within the LaTeX files except that 
+% no packages listed in the template may be deleted.
+%
+% Please do not include colors or graphics in the text.
+%
+% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
+%
+% % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- FIGURES AND TABLES
+%
+% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
+%
+% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
+% - Figures should be uploaded separately from your manuscript file. 
+% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
+% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
+% For figure citations, please use "Fig" instead of "Figure".
+% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
+%
+% Tables should be cell-based and may not contain:
+% - spacing/line breaks within cells to alter layout or alignment
+% - do not nest tabular environments (no tabular environments within tabular environments)
+% - no graphics or colored text (cell background color/shading OK)
+% See http://journals.plos.org/plosone/s/tables for table guidelines.
+%
+% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
+%
+% IMPORTANT
+% Below are a few tips to help format your equations and other special characters according to our specifications.
+For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
+%
+% For inline equations, please be sure to include all portions of an equation in the math environment.
+For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
+%
+% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
+%
+% Please add line breaks to long display equations when possible in order to fit size of the column. 
+%
+% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
+%
+% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
+%
+% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
+%
+% % % % % % % % % % % % % % % % % % % % % % % % 
+%
+% Please contact latex@plos.org with any questions.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+
+\documentclass[10pt,letterpaper]{article}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+
+% amsmath and amssymb packages, useful for mathematical formulas and symbols
+\usepackage{amsmath,amssymb}
+
+% Use adjustwidth environment to exceed column width (see example table in text)
+\usepackage{changepage}
+
+% Use Unicode characters when possible
+\usepackage[utf8x]{inputenc}
+
+% textcomp package and marvosym package for additional characters
+\usepackage{textcomp,marvosym}
+
+% cite package, to clean up citations in the main text. Do not remove.
+\usepackage{cite}
+
+% Use nameref to cite supporting information files (see Supporting Information section for more info)
+\usepackage{nameref,hyperref}
+
+% line numbers
+\usepackage[right]{lineno}
+
+% ligatures disabled
+\usepackage{microtype}
+\DisableLigatures[f]{encoding = *, family = * }
+
+% color can be used to apply background shading to table cells only
+\usepackage[table]{xcolor}
+
+% array package and thick rules for tables
+\usepackage{array}
+
+% create "+" rule type for thick vertical lines
+\newcolumntype{+}{!{\vrule width 2pt}}
+
+% create \thickcline for thick horizontal lines of variable length
+\newlength\savedwidth
+\newcommand\thickcline[1]{%
+  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+  \cline{#1}%
+  \noalign{\vskip\arrayrulewidth}%
+  \noalign{\global\arrayrulewidth\savedwidth}%
+}
+
+% \thickhline command for thick horizontal lines that span the table
+\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+\hline
+\noalign{\global\arrayrulewidth\savedwidth}}
+
+% Remove comment for double spacing
+%\usepackage{setspace} 
+%\doublespacing
+
+% Text layout
+\raggedright
+\setlength{\parindent}{0.5cm}
+\textwidth 5.25in 
+\textheight 8.75in
+
+% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
+% Captions will be left justified
+\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
+\renewcommand{\figurename}{Fig}
+
+% Use the PLoS provided BiBTeX style
+\bibliographystyle{plos2015}
+
+% Remove brackets from numbering in List of References
+\makeatletter
+\renewcommand{\@biblabel}[1]{\quad#1.}
+\makeatother
+
+% Leave date blank
+\date{}
+
+% Header and Footer with logo
+\usepackage{lastpage,fancyhdr,graphicx}
+\usepackage{epstopdf}
+\pagestyle{myheadings}
+\pagestyle{fancy}
+\fancyhf{}
+\setlength{\headheight}{27.023pt}
+\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
+\rfoot{\thepage/\pageref{LastPage}}
+\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
+\fancyheadoffset[L]{2.25in}
+\fancyfootoffset[L]{2.25in}
+\lfoot{\sf PLOS}
+
+%% Include all macros below
+
+\newcommand{\lorem}{{\bf LOREM}}
+\newcommand{\ipsum}{{\bf IPSUM}}
+
 \def\bs{\boldsymbol}
-\def\til{\widetilde}
-\def\MM{\mathcal{M}}
-\def\Vy{\widehat{\textrm{V}}_Y}
-\bibliographystyle{naturemag}
-\title{Bayesian model comparison for rare variant association studies of multiple phenotypes}
-\author{Christopher DeBoever$^{1}$, Matthew Aguirre$^{1}$, Yosuke Tanigawa$^{1}$, Timothy Poterba$^{2}$, Mark J.
-Daly$^{2,3}$,  Matti Pirinen$^{4}$\thanks{matti.pirinen@helsinki.fi}, and Manuel A.
-Rivas$^{1}$\thanks{mrivas@stanford.edu}}
-\usepackage{setspace}   
-\linenumbers
+
+%% END MACROS SECTION
 
 \begin{document}
-\maketitle
-\begin{abstract}
-\singlespacing
+\vspace*{0.2in}
+
+% Title must be 250 characters or less.
+\begin{flushleft}
+{\Large
+\textbf\newline{Bayesian model comparison for rare variant association studies of multiple phenotypes} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+}
+\newline
+% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
+\\
+\author{Christopher DeBoever$^{1}$, Matthew Aguirre$^{1}$, Yosuke Tanigawa$^{1}$, Chris C.
+A.
+Spencer$^{2}$, Timothy Poterba$^{3}$, Mark J.
+Daly$^{3,4}$,  Matti Pirinen$^{5}$\thanks{matti.pirinen@helsinki.fi}, and Manuel A.
+Rivas$^{1}$\thanks{mrivas@stanford.edu}}
+Christopher DeBoever\textsuperscript{1},
+Matthew Aguirre\textsuperscript{1},
+Yosuke Tanigawa\textsuperscript{1,2},
+Chris C. A. Spencer\textsuperscript{3},
+Timothy Poterba\textsuperscript{4},
+Mark J. Daly\textsuperscript{4,5},
+Matti Pirinen\textsuperscript{6},
+Manuel A. Rivas\textsuperscript{1*},
+\\
+\bigskip
+\textbf{1} Department of Biomedical Data Science, Stanford University, Stanford, CA, USA
+\\
+\textbf{2} Biomedical Informatics Training Program, Stanford University, Stanford, CA, USA
+\\
+\textbf{3} Genomics plc, Oxford, UK
+\\
+\textbf{4} Broad Institute of MIT and Harvard, Cambridge, MA, USA
+\\
+\textbf{5} Wellcome Trust Centre for Human Genetics, University of Oxford, Oxford, UK
+\\
+\textbf{6} Analytic and Translational Genetics Unit, Massachusetts General Hospital, Boston, MA, USA
+\\
+\textbf{7} Oxford Centre for Diabetes, Endocrinology and Metabolism, Oxford, UK
+\\
+\textbf{8} Institute for Molecular Medicine Finland, University of Helsinki, Finland
+\\
+% \textbf{1} Affiliation Dept/Program/Center, Institution Name, City, State, Country
+% \\
+\bigskip
+
+% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
+% 
+% Remove or comment out the author notes below if they aren't used.
+%
+% Primary Equal Contribution Note
+% \Yinyang These authors contributed equally to this work.
+
+% Additional Equal Contribution Note
+% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
+% \ddag These authors also contributed equally to this work.
+
+% Current address notes
+% \textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
+% \textcurrency b Insert second current address 
+% \textcurrency c Insert third current address
+
+% Deceased author note
+% \dag Deceased
+
+% Group/Consortium Author Note
+% \textpilcrow Membership list can be found in the Acknowledgments section.
+
+% Use the asterisk to denote corresponding authorship and provide email address in note below.
+* mrivas@stanford.edu
+
+\end{flushleft}
+% Please keep the abstract below 300 words
+\section*{Abstract}
 Whole genome sequencing studies applied to large populations or biobanks with extensive phenotyping raise new analytic challenges.
@@
 We show that we are able to retain useful features from widely-used meta-analysis approaches and prioritize protective modifiers of disease risk.
-\end{abstract}
 
-\begin{affiliations}
-\item Department of Biomedical Data Science, Stanford University, Stanford, CA, USA
-\item Broad Institute of MIT and Harvard, Cambridge, MA, USA
-\item Wellcome Trust Centre for Human Genetics, University of Oxford, Oxford, UK
-\item Analytic and Translational Genetics Unit, Massachusetts General Hospital, Boston, MA, USA
-\item Oxford Centre for Diabetes, Endocrinology and Metabolism, Oxford, UK
-\item Institute for Molecular Medicine Finland, University of Helsinki, Finland
-\end{affiliations}
-\newpage
+% Please keep the Author Summary between 150 and 200 words
+% Use first person. PLOS ONE authors please skip this step. 
+% Author Summary not valid for PLOS ONE submissions.   
+\section*{Author summary}
+TODO
 
-\section{Introduction} 
+\linenumbers
+
+% Use "Eq" instead of "Equation" for equation citations.
+\section*{Introduction}
 %% Sequencing technologies
@@
 In this paper, we present a Bayesian multiple rare variants and phenotypes (MRP) model comparison approach for identifying rare variant associations as an alternative to current widely-used statistical tests.
-The MRP framework exploits correlation, scale, or location (direction) of genetic effects in a broad range of rare variant association study designs including: case-control; multiple diseases and shared controls; single continuous phenotype; multiple continuous phenotypes; or a mixture of case-control and multiple continuous phenotypes (Fig.~\ref{Fig1}).
+The MRP framework exploits correlation, scale, or location (direction) of genetic effects in a broad range of rare variant association study designs including: case-control; multiple diseases and shared controls; single continuous phenotype; multiple continuous phenotypes; or a mixture of case-control and multiple continuous phenotypes (Fig \ref{fig1}).
 MRP makes use of Bayesian model comparison, whereby we compute a Bayes Factor (BF) defined as the ratio of the marginal likelihoods of the observed data under two models: 1) a pre-specified null where all genetic effects are zero; and 2) an alternative model where factors like correlation, scale, or location of genetic effects are considered.
@@
 For MRP, the BF represents the statistical evidence for a non-zero effect for a particular group of rare variants on the phenotype(s) of interest.
+
+% Place figure captions after the first paragraph in which they are cited.
+\begin{figure}[!h]
+% \includegraphics[width=\textwidth]{FIG1FINAL.pdf}
+\caption{{\bf Schematic overview of MRP.}
+A: MRP is suitable for a broad range of rare variant association study designs including (from left to right): i) case-control, ii) multiple diseases with shared controls, iii) single quantitative phenotype, and iv) mixture of case-control and quantitative phenotypes.
+B: Diagram of factors considered in rare variant association analysis including the correlation matrices: $\mathbf{R_{\textrm{study}}}$  (expected correlation of genetic effects among a group of studies), $\mathbf{R_{\textrm{var}}}$ (expected correlation of genetic effects among a group of variants), and $\mathbf{R_{\textrm{phen}}}$ (expected correlation of genetic effects among a group of phenotypes); the scale parameter for genetic variant annotation; and the location of genetic effects, which may be used to prioritize or identify protective modifiers of disease risk.}
+\label{fig1}
+\end{figure}
 
@@
 
-%% FIGURE 1 
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{FIG1FINAL.pdf}
-  \caption{Schematic overview of MRP.
-A.
-MRP is suitable for a broad range of rare variant association study designs including (from left to right): i) case-control, ii) multiple diseases with shared controls, iii) single quantitative phenotype, and iv) mixture of case-control and quantitative phenotypes.
-B.
-Diagram of factors considered in rare variant association analysis including the correlation matrices: $\mathbf{R_{\textrm{study}}}$  (expected correlation of genetic effects among a group of studies), $\mathbf{R_{\textrm{var}}}$ (expected correlation of genetic effects among a group of variants), and $\mathbf{R_{\textrm{phen}}}$ (expected correlation of genetic effects among a group of phenotypes); the scale parameter for genetic variant annotation; and the location of genetic effects, which may be used to prioritize or identify protective modifiers of disease risk.}
-  \label{Fig1}
-  \end{figure}
+% Equation example
+% \begin{eqnarray}
+% \label{eq:schemeP}
+% 	\mathrm{P_Y} = \underbrace{H(Y_n) - H(Y_n|\mathbf{V}^{Y}_{n})}_{S_Y} + \underbrace{H(Y_n|\mathbf{V}^{Y}_{n})- H(Y_n|\mathbf{V}^{X,Y}_{n})}_{T_{X\rightarrow Y}},
+% \end{eqnarray}
 
-\section{Results}
-
-\subsection{Methods overview}
+\section*{Materials and methods}
 A complete description of MRP, along with an analytical derivation, is given in the Online Methods and Appendix. Briefly, we describe the approach.
 
-\subsubsection{Model comparison}
 In Bayesian model comparison, we compare the null and alternative models via the Bayes Factor (BF) defined as the ratio of the marginal likelihoods of the observed data under two models.
@@
 
-Although we see advantages in adopting a Bayesian perspective, our approach could be used in a frequentist context by calculating a Bayes factor and using it as a test statistic to compute p-values (Methods, Fig.~\ref{Fig2}).
+Although we see advantages in adopting a Bayesian perspective, our approach could be used in a frequentist context by calculating a Bayes factor and using it as a test statistic to compute p-values (Methods, Fig \ref{fig2}).
 
-%\subsubsection{Mixture model for rare variant effects across multiple phenotypes}
-%In Bayesian model comparison, the alternative model must be pre-specified.
-In practice, for gene discovery efforts, this has the attractive property that when we contrast it against a standard pre-specified null, i.e.
-genetic effects are zero across all studies in a group of variants and phenotypes, an analytical derivation is obtained making it computationally flexible and efficient.
-However, a limitation of this approach is that interpretation may not be straightforward.
-To make statistical inferences about the properties of the underlying mixture of genetic effects, we present a mixture model for estimating rare variant effects.
-Our goal is to cluster the variants into groups based on their effects on the multivariate phenotype.
-In particular, we allow variants to share clusters both within a gene and across the genes using a mixture model formulation of a clustering model where we also estimate the number of clusters in the data using Bayesian Information Criterion (BIC).
-More specifically, we can estimate (1) the proportion of non-null variants, (2) whether variants with the same annotation in one gene behave similarly, (3) whether variants across genes share effect profiles, or (4) whether different annotations differ in the magnitude of their effects.
+% For figure citations, please use "Fig" instead of "Figure".
 
-\newpage 
-%% FIGURE 2
-\begin{figure}
-\centering
-   \includegraphics[width=\textwidth]{Figure2Nov2017.pdf}
-  \caption{Simulation studies.
-A.
-Comparison of $-\log_{10}$(p-values) from frequentist $\textrm{BF}_{\textrm{MRP}}$ approximation for an independent effects and a similar effects model to commonly used gene-based statistical tests (skat and burden).
-B.
-Comparison of log10(Bayes Factors) obtained when raw genotype and phenotype data is available to a scenario where summary statistics only was available and similar effects across studies is assumed.
-C.
-From single variant and single phenotype to multiple variants and multiple phenotypes gene discovery: ROC curves for detecting gene association to any of the phenotypes using single variant/single phenotype association (blue) to multiple variants and multiple phenotypes association (purple).
-D.
-ROC curves for detecting gene association when incorporating prior mean of genetic effects (orange) to identify protective alleles.
-}
-  \label{Fig2}
+% % Place figure captions after the first paragraph in which they are cited.
+% \begin{figure}[!h]
+% \caption{{\bf Bold the figure title.}
+% Figure caption text here, please use this space for the figure panel descriptions instead of using subfigure commands. A: Lorem ipsum dolor sit amet. B: Consectetur adipiscing elit.}
+% \label{fig1}
+% \end{figure}
+
+% Results and Discussion can be combined.
+\section*{Results}
+
+\begin{figure}[!h]
+\caption{{\bf Simulation studies.}
+A: Comparison of $-\log_{10}$(p-values) from frequentist $\textrm{BF}_{\textrm{MRP}}$ approximation for an independent effects and a similar effects model to commonly used gene-based statistical tests (skat and burden).
+B: Comparison of log10(Bayes Factors) obtained when raw genotype and phenotype data is available to a scenario where summary statistics only was available and similar effects across studies is assumed.
+C: From single variant and single phenotype to multiple variants and multiple phenotypes gene discovery: ROC curves for detecting gene association to any of the phenotypes using single variant/single phenotype association (blue) to multiple variants and multiple phenotypes association (purple).
+D: ROC curves for detecting gene association when incorporating prior mean of genetic effects (orange) to identify protective alleles.}
+\label{fig2}
 \end{figure}
 
-\subsection{Simulation studies}
+\subsection*{Simulation studies}
 We first verified the analytical derivations and examined the properties of the approach under a simulation framework.
@@
 %% Comparison to other widely used methods
-\subsubsection{Comparison to frequentist gene tests}
+\subsubsection*{Comparison to frequentist gene tests}
 For the analysis of multiple rare variants and a single phenotype we compared it to the burden test and the SKAT test, commonly used statistical tests in rare variant association studies of a single phenotype.
@@
 To compare the Bayesian models we compute p-values by using the Bayes Factor as the test statistic and approximating it using distribution properties of quadratic forms (Appendix).
-As expected, an independent effects model has high correlation with the gene-based test SKAT ($r^2 = 0.99$), whereas the similar effects model has high correlation with the burden test ($r^2 = 0.93$, Fig.~\ref{Fig2}A).
+As expected, an independent effects model has high correlation with the gene-based test SKAT ($r^2 = 0.99$), whereas the similar effects model has high correlation with the burden test ($r^2 = 0.93$, Fig \ref{fig2}A).
 
-\subsubsection{Summary statistic data}
+\subsubsection*{Summary statistic data}
 To study the behavior of MRP using summary statistics we simulate two scenarios: first, the scenario where analysts have access to all the raw genotype and phenotype data; and second, the scenario where analysts only have access to summary statistics data\cite{liu2014meta}.
 We conducted $1000$ simulation experiments where we let $K$ (the number of phenotypes) $=3$, $M$ (the number of variants) $=10$, $S$ (the number of studies) $=2$, $N_0$ (number of individuals in study with access to all the data) $= 10000$, $N_1$ (meta-analysis study 1) $= 5000$, $N_2$ (meta-analysis study 2) $= 5000$.
-We find that, under the scenario where similar effects are assumed across studies, the Bayes Factors obtained using summary statistics alone are strongly correlated ($r^2 = 1$) to Bayes Factors obtained by the full genotype and phenotype data (Fig.~\ref{Fig2}B).
+We find that, under the scenario where similar effects are assumed across studies, the Bayes Factors obtained using summary statistics alone are strongly correlated ($r^2 = 1$) to Bayes Factors obtained by the full genotype and phenotype data (Fig \ref{fig2}B).
 
-\subsubsection{From single variant and single phenotype analysis to multiple variants and multiple phenotypes}
+\subsubsection*{From single variant and single phenotype analysis to multiple variants and multiple phenotypes}
 To validate the flexibility of the approach we conducted a simulation experiment where we assumed an allelic architecture consistent to that discovered for {\it APOC3} in relation to coronary artery disease (CAD), triglycerides (TG), low-density lipoprotein cholesterol (LDL-C), and high-density lipoprotein cholesterol (HDL-C)\cite{apoc3,apoc32,jorgensen2014loss,cohorts2014loss}.
 We simulated three studies and applied the model comparison unit jointly to summary statistic data obtained for each study (Supplementary Note).
-Overall, we observed that considering the joint effects across multiple studies in a group of variants and phenotypes may improve ability to detect gene-based signals (Fig.~\ref{Fig2}C), and that considering prior mean of genetic effects should aid in efforts to identify protective modifiers of disease risk (Fig.~\ref{Fig2}D).
+Overall, we observed that considering the joint effects across multiple studies in a group of variants and phenotypes may improve ability to detect gene-based signals (Fig \ref{fig2}C), and that considering prior mean of genetic effects should aid in efforts to identify protective modifiers of disease risk (Fig \ref{fig2}D).
 
-\subsection{Applications}
+\subsection*{Applications}
 We demonstrate its use in rare variant analysis in the UK Biobank.
 
-\subsubsection{Model comparison applied to summary statistic from the UK Biobank}
+\subsubsection*{Model comparison applied to summary statistic from the UK Biobank}
 We applied the MRP model comparison approach to summary statistic data generated from single variant logistic regression and linear regression analysis for coding variants on the UK Biobank array (Methods).
 We first applied MRP to the asthma, eosinophil count, forced expiratory volume in 1-second, and forced vital capacity phenotypes.
-We applied the model to missense variants and PTVs for each phenotype separately (Fig.~\ref{asthma_manhattan}A-D) as well as to all phenotypes jointly (Fig.~\ref{asthma_manhattan}E) and obtained $\log_{10}$ BF for each gene.
+We applied the model to missense variants and PTVs for each phenotype separately (Fig \ref{asthma_combined}A-D) as well as to all phenotypes jointly (Fig \ref{asthma_combined}E) and obtained $\log_{10}$ BF for each gene.
 As expected from recent studies \cite{DeBoever179762, 10.1371/journal.pgen.1006659}, we observed evidence that missense variants and PTVs in \textit{IL33} affect eosinophil counts and offer protection from asthma from the single-phenotype analyses, though the evidence of association was strongest for the joint analysis (Table~\ref{asthma_table}).
@@
 
-%% FIGURE 3
-\begin{figure}
-\centering
-\includegraphics[width=\textwidth]{glaucoma_manhattan.pdf}
-  \caption{$\log_{10}$ Bayes Factors from applying MRP to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
+% \begin{figure}
+% \centering
+% \includegraphics[width=\textwidth]{glaucoma_combined.pdf}
+%   \caption{$\log_{10}$ Bayes Factors from applying MRP to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
 (E) shows the results of a joint analysis focused on finding rare variants that protect against glaucoma.
 The genes outside of chromosome 6 with with Bayes Factor greater than three are indicated by arrows.
-Only $\log_{10}$ Bayes Factors greater than zero are plotted.}
-  \label{glaucoma_manhattan}
+Only $\log_{10}$ Bayes Factors greater than zero are plotted.
+(F) $-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{ANGPTL7} for all three phenotypes.}
+%   \label{glaucoma_combined}
+% \end{figure}
+
+\begin{figure}[!h]
+% \includegraphics[width=\textwidth]{FIG1FINAL.pdf}
+\caption{{\bf Results for glaucoma application.}
+$\log_{10}$ Bayes Factors from applying MRP to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
+(E) shows the results of a joint analysis focused on finding rare variants that protect against glaucoma.
+The genes outside of chromosome 6 with with Bayes Factor greater than three are indicated by arrows.
+Only $\log_{10}$ Bayes Factors greater than zero are plotted.
+F: $-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{ANGPTL7} for all three phenotypes.}
+\label{fig3}
 \end{figure}
 
-%% FIGURE 4
-\begin{figure}
+% %% TABLE: GLAUCOMA GENES OF INTEREST
+\begin{table}[!ht]
+\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
 \centering
-\includegraphics[width=\textwidth]{glaucoma_forest.pdf}
-  \caption{$-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{ANGPTL7} for glaucoma, intra-ocular pressure, and corneal resistance factor.}
-  \label{glaucoma}
-\end{figure}
-
-%% TABLE: GLAUCOMA GENES OF INTEREST
-\begin{table}
-\begin{tabular}{l p{3cm} p{1cm} p{1.5cm} p{3.8cm} p{3.5cm}}
-\toprule
-{} & Joint, protective & Joint & Glaucoma & Intra-ocular pressure, Goldmann-correlated & Corneal resistance factor \\
-\midrule
+\caption{
+{\bf $\log_{10}$ Bayes Factors for genes highlighted in Figure \ref{fig3}.}}
+\begin{tabular}{|l+p{3cm}|p{1cm}|p{1.5cm}|p{3.8cm}|p{3.5cm}|}
+\hline
+{\bf Gene} & {\bf Joint, protective} & {\bf Joint} & {\bf Glaucoma} & {\bf Intra-ocular pressure, Goldmann-correlated} & {\bf Corneal resistance factor} \\
+\thickhline
 \textit{ANGPTL7} &              10.1 &   9.4 &      1.7 &                                       10.1 &                       6.2 \\
+\hline
 \textit{KLHL22}  &               3.1 &   3.6 &     -0.2 &                                        3.9 &                       2.2 \\
+\hline
 \textit{GLI3}    &               2.9 &   2.9 &     -0.3 &                                       -0.7 &                       3.2 \\
+\hline
 \textit{WNT10A}  &               2.2 &   3.0 &     -0.3 &                                       -0.5 &                       3.5 \\
+\hline
 \textit{RNASE2}  &              -1.3 &  -0.3 &      0.4 &                                       -1.0 &                      -1.1 \\
+\hline
 \textit{FUK}     &              -6.5 &  -2.3 &      2.5 &                                       -2.1 &                      -2.1 \\
-\bottomrule
+\hline
 \end{tabular}
-\caption{$\log_{10}$ Bayes Factors for genes highlighted in Figure \ref{glaucoma_manhattan}.}
-\label{glaucoma_table}
+\label{table1}
+\end{adjustwidth}
 \end{table}
 
-\section{Discussion}
+\begin{table}[!ht]
+\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+\centering
+\caption{
+{\bf $\log_{10}$ Bayes Factors for genes highlighted in Figure \ref{fig4}.}}
+\begin{tabular}{|l+l|l|l|l|l|l|}
+\hline
+{\bf Gene} & {\bf Joint, protective} & {\bf Joint} & {\bf Eosinophil count} & {\bf FVC} & {\bf FEV1} & {\bf Asthma} \\
+\thickhline
+\textit{IL33}         &              30.1 &  29.8 &             30.9 &  -3.1 &  -2.7 &    9.5 \\
+\hline
+\textit{CCR3}         &               5.8 &   6.3 &              9.1 &  -1.7 &  -2.3 &   -0.5 \\
+\hline
+\textit{IL17RA}       &               4.6 &   6.1 &             14.1 &  -4.1 &  -3.9 &   -1.4 \\
+\hline
+\textit{AL161450.1}   &               3.2 &   3.8 &              4.9 &  -0.1 &  -0.4 &    0.3 \\
+\hline
+\textit{RP11-39K24.9} &               3.2 &   3.8 &              4.9 &  -0.1 &  -0.4 &    0.3 \\
+\hline
+\textit{ELN}          &               2.9 &   3.2 &             -1.0 &   5.6 &   2.9 &   -0.6 \\
+\hline
+\textit{SCMH1}        &               2.7 &   3.9 &             -2.1 &   7.5 &   3.3 &   -0.7 \\
+\hline
+\textit{TGFB2}        &               2.6 &   4.1 &             -2.8 &   1.7 &  -2.4 &    0.1 \\
+\hline
+\textit{ZFAT}         &               1.3 &   1.9 &             -2.8 &   6.7 &   4.6 &   -0.4 \\
+\hline
+\textit{STC2}         &               0.2 &   1.3 &             -0.8 &   3.9 &   2.3 &   -0.7 \\
+\hline
+\textit{P4HTM}        &              -1.3 &  -0.7 &             -1.6 &   2.9 &   4.3 &   -1.0 \\
+\hline
+\textit{IL18RAP}      &              -2.2 &  -1.0 &             -0.8 &  -2.0 &  -2.2 &    3.6 \\
+\hline
+\textit{WDR6}         &              -4.1 &  -2.9 &             -2.3 &   1.9 &   3.4 &   -0.9 \\
+\hline
+\textit{ATP2A3}       &              -4.2 &  -1.5 &             -1.8 &  -3.1 &  -3.1 &    5.9 \\
+\hline
+\textit{CHI3L1}       &              -4.4 &  -2.0 &             -2.0 &  -1.9 &  -2.0 &    3.0 \\
+\hline
+\end{tabular}
+\label{table2}
+\end{adjustwidth}
+\end{table}
+
+% % Place tables after the first paragraph in which they are cited.
+% \begin{table}[!ht]
+% \begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+% \centering
+% \caption{
+% {\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
+% \begin{tabular}{|l+l|l|l|l|l|l|l|}
+% \hline
+% \multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \thickhline
+% $cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
+% $cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
+% $cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
+% \end{tabular}
+% \begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl.
+Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
+% \end{flushleft}
+% \label{table1}
+% \end{adjustwidth}
+% \end{table}
+
+%PLOS does not support heading levels beyond the 3rd (no 4th level headings).
+% \subsection*{\lorem\ and \ipsum\ nunc blandit a tortor}
+% \subsubsection*{3rd level heading}
+
+% \begin{enumerate}
+% 	\item{react}
+% 	\item{diffuse free particles}
+% 	\item{increment time by dt and go to 1}
+% \end{enumerate}
+
+% \begin{itemize}
+% 	\item First bulleted item.
+% 	\item Second bulleted item.
+% 	\item Third bulleted item.
+% \end{itemize}
+
+\section*{Discussion}
 In this study, we developed a Bayesian model comparison approach MRP that shares information across both variants and phenotypes to identify rare variant associations.
@@
 
-\newpage
-\section{Methods}
+% Conclusion is optional
+% \section*{Conclusion}
 
-\subsection{MRP model comparison for association testing}
-We consider the multivariate linear regression model 
-$$\underset{\left(N\times K\right)}{\mathbf{\textrm{Y}}} 
-= \underset{\left(N\times K\right)}{\mathbf{\Psi}}  + \underset{\left(N\times M\right)}{\mathbf{\textrm{X}}}\underset{\left(M \times K\right)}{\mathbf{\textrm{B}}} 
-+ \underset{\left(N\times K\right)}{\mathbf{\textrm{E}}},$$
-where the matrices $\mathbf{\textrm{Y}} = \left[{y}_{i k}\right]$, $\mathbf{\textrm{X}} = \left[ x_{i m}\right]$, 
-$\mathbf{\textrm{B}} = \left[ \beta_{m k} \right]$ and $\mathbf{\textrm{E}} = \left[e_{i k}\right]$  describe the phenotype values ($y_{i k}$),
-copies of minor allele ($x_{i m}$), variant-phenotype effects ($\beta_{m k}$), and residual errors ($e_{i k}$), 
-for individual $i$, phenotype $k$, and variant $m$. 
-We assume that each phenotype has been transformed to a standard normal
-distribution and that the columns of ${\mathbf{{\textrm{X}}}}$ 
-have been centered, which means that the estimate for the intercept term ${\mathbf{\Psi}}$ is 0 and independent of the estimate of ${\mathbf{\textrm{B}}}$.
-We use vectorized notation where the rows of ${\mathbf{\textrm{B}}}$ form vector 
-$\bs {{ \beta}} = \left(\bs {{ \beta}}_{1},\dots,\bs {{\beta}}_{M}\right)^{\intercal}$
-of length $MK$.
+% CO\textsubscript{2}
 
-\noindent We define the MRP model comparison as a Bayes factor (BF) between the alternative model, where at least one variant
-affects at least one phenotype, and the null model, where all variant-phenotype effects are zero.  
-\noindent BF is the ratio of the marginal likelihoods for these two models: 
-$$ \textrm{BF} = \frac{\int_{{\bs {\beta}}}p\left(\mathbf{\textrm{Data}}|{\bs {\beta}}\right)p\left({\bs {\beta}}|\textrm{ALT}\right)d{\bs {\beta}}}{\int_{{\bs {\beta}}}p\left(\mathbf{\textrm{Data}}|{\bs {\beta}}\right)p\left({\bs {\beta}}|\textrm{NULL}\right)d{\bs {\beta}}},$$
-where Data can correspond either to the effect size estimates $\bs {\widehat{ \beta}}$ and the estimated variance-covariance matrix of  $\bs {\widehat{ \beta}}$, $\widehat{\mathbf{\textrm{V}}}_{\bs{\beta}}$, or to the original phenotypes and genotypes,
-$\underset{\left(N\times K\right)}{\mathbf{\textrm{\textrm{Y}}}}$ and $ \underset{\left(N\times M\right)}{\mathbf{\textrm{\textrm{X}}}}$, and any other covariates that we
-want to regress out from the phenotypes.
+\section*{Supporting information}
 
-The prior distribution for the null model, $p\left({\bs {\beta}}|\textrm{NULL}\right)$, is simply the point mass at $\bs \beta = 0$.
-In subsection~\ref{approxlikelihood} we show how we approximate the likelihood function for $\bs \beta$, $p\left(\mathbf{\textrm{Data}} | {\bs {\beta}}\right)$,
-in subsection~\ref{prioralt} we define the prior distribution $p\left({\bs {\beta}}|\textrm{ALT}\right)$ for the alternative model, 
-and finally, in subsection~\ref{bfmodel}, we compute the BF.
+% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
+\paragraph*{S1 Fig.}
+\label{S1_Fig}
+{\bf Bold the title sentence.} Add descriptive text after the title of the item (optional).
 
-\subsubsection{Likelihood function} 
-\label{approxlikelihood}
-\noindent A maximum likelihood estimator of $\mathbf{\textrm{B}}$ is given by the ordinary least-squares method
-$$\widehat{\mathbf{\textrm{B}}} = \left(\mathbf{\textrm{X}}^\intercal\mathbf{\textrm{X}}\right)^{-1}\mathbf{\textrm{X}}^{\intercal}\mathbf{\textrm{Y}},$$
-that in vectorized form is denoted by   
-$\bs {\widehat{ \beta}} = \left(\bs {\widehat{ \beta}}_{1},\dots,\bs {\widehat{\beta}}_{M}\right)^{\intercal}.$
-An estimator of the variance-covariance of $\widehat{\bs \beta}$  is given by 
-$$\widehat{\textrm{V}}_{\mathbf{\bs \beta}} =  \left(\mathbf{\textrm{X}}^\intercal\mathbf{\textrm{X}}\right)^{-1}\otimes \widehat{\textrm{V}}_{\mathbf{\textrm{Y}}},$$
-where $\widehat{\textrm{V}}_{\mathbf{\textrm{Y}}}$ is the estimated residual variance-covariance matrix of $\mathbf{\textrm{Y}}$ given $\mathbf{\textrm{X}}$.
+\paragraph*{S2 Fig.}
+\label{S2_Fig}
+{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
 
-\noindent 
-%Let $\bs {\widehat{ \beta}} = \left(\bs {\widehat{ \beta}}^{\intercal}_{1},\dots,\bs {\widehat{\beta}}^{\intercal}_{m}\right)^{\intercal}$ be a vector of point estimates of all $MK$ effects. 
-Following Band et al.\cite{band2013imputation}, we approximate the likelihood function of $\bs {\beta}$ by a multivariate normal distribution with mean $\bs {\widehat{\beta}}$ and variance-covariance matrix $\widehat{\textrm{V}}_{\bs {\beta}}$.
-Note that by approximating $\Vy$ by the trait correlation matrix, this likelihood approximation does not require access to the individual level data $\mathbf{\textrm{X}}$ and $\mathbf{\textrm{Y}}$ but only to the
-summary data of effect sizes $\bs {\widehat{\beta}}$, LD-matrix $\mathbf{\textrm{X}}^\intercal\mathbf{\textrm{X}}$ and a trait correlation estimate.
+\paragraph*{S1 File.}
+\label{S1_File}
+{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
 
-\subsubsection{Prior of $\bs{\beta}$ in the alternative model} 
-\label{prioralt}
-\noindent We construct the prior distribution $p\left({\bs {\beta}} | \textrm{ALT} \right)$ for the alternative model in three steps allowing 
-user to specify correlations between effects of different variants on different traits across different studies.
+\paragraph*{S1 Video.}
+\label{S1_Video}
+{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
 
-\noindent In a single study, the prior density for ${\bs {\beta}}$ incorporates the expected correlation of genetic effects among 
-a group of variants ($\mathbf{R}_{\textrm{var}}$) and among a group of phenotypes ($\mathbf{R}_{\textrm{phen}}$). 
-In addition, we  incorporate an expected spread of the effect size of each variant by scaling $\mathbf{R}_{\textrm{var}}$ as 
-$$\mathbf{S}_{\textrm{var}} = \Delta\left(\sigma_m\right) \mathbf{R}_{\textrm{var}} \Delta\left(\sigma_m\right),$$
-\noindent where $\Delta\left(\sigma_m\right)$ is a diagonal matrix with entries $\sigma_m$ determining the spread of the effect size distribution for each variant $m \leq M$.
-Thus, we can model settings where, e.g., protein-truncating variants have larger effect sizes ($\sigma = 1$) than missense variants ($\sigma = 0.1$).
-Note that when $\sigma_m = 1$ for all $m$ then $\mathbf{S}_{\textrm{var}} = \mathbf{R}_{\textrm{var}}$. 
-\noindent All in all, our prior density for ${\bs \beta}$ under alternative model is 
-$${\bs {\beta}}|\textrm{ALT} \sim \mathcal{N}\left(\mathbf{0}, \mathbf{U}\right), \textrm{ where } \mathbf{U} = \mathbf{S}_{\textrm{var}} \otimes \mathbf{R}_{\textrm{phen}}.$$
+\paragraph*{S1 Appendix.}
+\label{S1_Appendix}
+{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
 
-\noindent When we have data from multiple studies 
-we allow for possible differences in genetic effects across ethnicities or populations 
-extending the Approximate Bayes Factors of Band et al.~\cite{band2013imputation}
-and the summary statistics approach of RAREMETAL~\cite{liu2014meta} from univariate to multivariate phenotypes.
-Let $\bs {\widehat{ \beta}} = \left(\widehat{ \beta}_{s,m,k}\right)=\left(\widehat{\beta}_{1,1,1}, \widehat{\beta}_{1,1,2}, \dots, \widehat{\beta}_{1,1,K}, \widehat{\beta}_{1,2,1}, \dots, \widehat{\beta}_{1,2,K}, \dots, \widehat{\beta}_{1,M,K}, \widehat{\beta}_{2,1,1}, \dots, \widehat{\beta}_{S,M,K}\right),$ where $S$ is the number of studies, $M$ is the number of variants, and $K$ is the number of phenotypes.
-As with a single study, we incorporate the expected correlation of genetic effects between a pair of variants and a single phenotype using the matrix $\mathbf{S_{\textrm{var}}}$, between a variant and a pair of phenotypes using the matrix $\mathbf{R_{\textrm{phen}}}$, and we introduce the matrix $\mathbf{R_{\textrm{study}}}$ to specify prior on the similarity in effect sizes across the studies.
-Thus, the prior is
-$${\bs {\beta}} \sim \mathcal{N}\left(\mathbf{0}, \mathbf{U}\right), \textrm{ where } 
-\mathbf{U} = \mathbf{R}_{\textrm{study}}\otimes\left(\mathbf{S}_{\textrm{var}} \otimes \mathbf{R}_{\textrm{phen}}\right).$$
+\paragraph*{S1 Table.}
+\label{S1_Table}
+{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
 
-\noindent It is straightforward to include a non-zero vector ${\bs\mu}$ as a prior mean of genetic effects, in which case the prior is
-$${\bs {\beta}} \sim \mathcal{N}\left({\bs \mu},\mathbf{U}\right).$$
-We use this, for example, when screening for protective rare variants that have a pre-specified beneficial profile on a set of risk factors.
-
-\subsubsection{$\textrm{BF}_{\textrm{MRP}}$} 
-\label{bfmodel}
-\noindent The Bayes Factor is the ratio of the marginal likelihoods between the alternative and the null model. 
-\noindent The marginal likelihood for the alternative model is 
-$$ \int_{{\bs {\beta}}} p\left(\mathbf{\textrm{Data}}|{\bs {\beta}}\right) p\left({\bs {\beta}}|\textrm{ALT}\right)d{\bs {\beta}} = c \times \mathcal{N}\left({ {\widehat{\bs \beta}}}; {\bs \mu},  \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U}\right)$$
-\noindent and the marginal likelihood for the null model is 
-$$\int_{{\bs {\beta}}} p\left(\mathbf{\textrm{Data}}|{\bs {\beta}}\right) p\left({\bs {\beta}}|\textrm{NULL}\right)d{\bs {\beta}} = c \times \mathcal{N}\left({ {\widehat{\bs \beta}}}; 0,  \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}\right).$$
-\noindent In Appendix (\ref{app1}) we show how we compute the Bayes Factor
-$$\textrm{BF}_{\textrm{MRP}} = \frac{\det\left(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U} \right)^{-\frac{1}{2}}
-\exp\left[-\frac{1}{2}\left({\bs {\widehat{\beta}} - {\bs \mu}}\right)^{\intercal}\left(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U}\right)^{-1}\left({\bs {\widehat{\beta}}} - {\bs \mu}\right)\right]}
-{\det \left( \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}\right)^{-\frac{1}{2}}\exp \left[-\frac{1}{2}{\bs {\widehat{\beta}}}^{\intercal}\, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} 
-\,{\bs {\widehat{\beta}}}\right]}.$$
-
-When $\bs \mu=0$, $\textrm{BF}_{\textrm{MRP}}$ is an increasing function of the following quadratic form
-\begin{equation}\label{quadratic_form}
-Q(\bs{\widehat{\beta}};\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}},\mathbf{U})=
-\bs{\widehat{\beta}^\intercal} \left(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}^{-1} - (\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}+\mathbf{U})^{-1}\right) \bs{\widehat{\beta}}.
-\end{equation}
-Furthermore, this quadratic form is the only part of the $\textrm{BF}_{\textrm{MRP}}$ that depends on $\bs {\widehat{\beta}}$.
-Thus, by deriving a distribution of $Q(\bs{\widehat{\beta}};\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}},\mathbf{U})$
-under the null model we can compute a p-value when $\textrm{BF}_{\textrm{MRP}}$ is used as a test statistic.
-According to basic properties of quadratic forms of Gaussian variables,
-$Q(\bs{\widehat{\beta}};\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}},\mathbf{U}) \sim \sum_{i=1}^n d_i \chi^2_i$, where $\chi_i^2$
-are an independent sample from $\chi_1^2$ distribution (chi-square with one degree of freedom),
-and $d_i$ are the eigenvalues of matrix
-$I-(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}+\mathbf{U})^{-1}\, \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}$.
-The distribution function for a mixture of chi-squares can be numerically evaluated by
-the R-package `CompQuadForm'~\cite{Duchesne2010}.
-
-\subsection{UK Biobank Data}
-\subsubsection{GWAS Summary Statistics}
-We performed genome-wide association analysis using PLINK v2.00a(17 July 2017) as previously described~\cite{DeBoever179762}.
-For asthma, we used the Firth-fallback in PLINK, a hybrid algorithm which normally uses the logistic regression code described in (Hill 2017), but switches to a port of logistf() (https://cran.r-project.org/web/packages/logistf/index.html) in two cases: (1) one of the cells in the 2x2 allele count by case/control status contingency table is empty (2) logistic regression was attempted since all the contingency table cells were nonzero, but it failed to converge within the usual number of steps.
-We used the following covariates in our analysis: age, sex, array type, and the first four principal components, where array type is a binary variable that represents whether an individual was genotyped with UK Biobank Axiom Array or UK BiLEVE Axiom Array.
-For variants that were specific to one array, we did not use array as a covariate.
-Asthma cases were defined using both Hospital Episode Statistics and verbal questionnaire responses.
-We used the provided values from the UK Biobank for eosinophil counts, forced vital capacity (FVC), and forced expiratory volume in 1-second (FEV1).
-The phenotype codes used throughout (asthma=HC382, eosinophil count=INI30150, Forced Expiratory Volume in 1-second (FEV1)=INI3063, and Forced Vital Capacity (FVC)=INI3062) correspond to the phenotype codes used in on the Global Biobank Engine [\url{https://biobankengine.stanford.edu}].
-
-\subsubsection{Genetic Correlations}
-We calculated the genetic between asthma, eosinophil counts, forced vital capacity (FVC), and forced expiratory volume in 1-second (FEV1) using the MultiVariate Polygenic Mixture Model (MVPMM)~\cite{DeBoever2017}.
-Briefly, MVPMM estimates genetic correlation given GWAS summary statistics (effect size and standard error of effect size estimate) by modeling GWAS summary statistics as generated from one of two mixture components.
-Summary statistics from variants in the null component are modeled as being drawn from a multivariate normal distribution with zero mean and covariance matrix that captures correlation in the summary statistics due to the use of shared subjects or other sources of correlation.
-Summary statistics from variants in the non-null component are modeled as being drawn from a multivariate normal distribution with zero mean, but the covariance matrix for the non-null component combines the covariance matrix from the null component with another covariance matrix that captures the genetic correlation between the phenotypes being considered.
-
-\subsubsection{UK Biobank Asthma Application}
-For the Manhattan plots and Table \ref{asthma_table}, we removed any genes with non-unique gene symbols.
-We also removed the sense overlapping gene \textit{CTD-3064M3.3} and and the antisense gene \textit{ZFAT-AS1} since these genes overlap \textit{GPR20} and \textit{ZFAT} respectively and therefore have the same $\log_{10}$ Bayes Factors as those genes.
-
-%%%
-%
-%\subsection{MRP mixture modeling for effect size estimation} 
-%\label{mrpmm}
-%Index genes by $j = 1,\dots, J$ and denote variant $m$ in gene $j$ by $v_{jm}$.
-Our goal is to cluster the variants into groups based on their effects on the multivariate phenotype.
-To compare models with different numbers of clusters we use the Bayesian Information Criterion (BIC).
-%
-%We represent each cluster $c$ by a mean effect size 
-%$$ \bs b_c \sim \mathcal{N}\left(0,\bs \Theta_0\right),$$
-%where $\bs \Theta_0$ is our prior estimate of genetic correlation across the traits; 
-%and a covariance structure
-%$\bs \Sigma_c = \widehat{\mathbf{\textrm{V}}}_Y $ (when phenotype data available) or $\bs \Sigma_c = \widehat{\mathbf{\textrm{V}}}_{jm}$ (when only summary statistics available),  
-%where $\widehat{\mathbf{\textrm{V}}}_Y$ is an estimate of the phenotype correlation and remains fixed across the clusters, and $\widehat{\mathbf{\textrm{V}}}_{jm}$ will be defined below. 
-%(Extensions to cluster-specific $\bs \Sigma_c$ are possible but not considered in this work.)
-%
-%Consider the model with $C>1$ clusters. To model the sharing of clusters across the genes we first draw a $C$ dimensional probability vector 
-%$$\bs \pi_{0} \sim \mathrm{Dirichlet}\left(1,1,1,\ldots ,1\right).$$ Next, for each gene $j$, we draw a probability vector 
-%$$\bs \pi_{j} | \bs \pi_{0} \sim \mathrm{Dirichlet}\left(\alpha \bs \pi_{0}\right)$$
-%to determine the mixture proportions $\pi_{jc}$ that tell how the variants in gene $j$ are distributed across the clusters $c$.
-The parameter $\alpha$ governs how much sharing of clusters exists across genes with prior
-%$$\alpha \sim \textrm{Inv-Gamma}\left(1,1\right).$$ 
-%To relate the clusters to the observed data, the phenotype of
-%individual $i$ with a rare allele of variant $m$ with annotation $a$ in gene $j$ is
-%
-%$$\bs y_{i} | \left( \bs \pi_j, {\bs b}, \sigma_a \right) \sim \sum_{c=1}^{C}\pi_{jc}\mathcal{N}\left(\sigma_{a} \bs b_c, \bs \Sigma_c\right) $$
-%where $\sigma_a$ corresponds to a scale parameter for annotation $a$ across all the clusters, where annotations may be, e.g., 
-%protein truncating variant (PTV), missense variant, or other functional annotation category, with an inverse-gamma prior
-%$$\sigma^2_{a} \sim \textrm{Inv-Gamma}\left(sh_a, sc_a\right),$$
-%with hyperparameter values $sh_a$ (shape) and $sc_a$ (scale).
-%At the above formulation, we assume that each individual carriers at most one of the rare variants considered.
-%
-%If we have access only to summary statistics of estimated effect sizes 
-%$\widehat{\bs \beta}_{jm}$ and estimated variance-covariance 
-%matrices $\widehat{\mathbf{\textrm{V}}}_{jm}$ for each variant $m$ in gene $j$, our sampling model for the data is
-%$$\widehat{\bs \beta}_{jm} | \left( \bs \pi_j, {\bs b}, \sigma_a \right) \sim \sum_{c=1}^{C}\pi_{jc}\mathcal{N}\left(\sigma_{a} \bs b_c, \widehat{\mathbf{\textrm{V}}}_{jm} \right).$$
-%Similar to the individual level data model above, this formulation assumes independence between the variants, which is 
-%approximately true when each individual carries at most one of the rare variants considered.
-%A connection between the two data types is
-%that $\widehat{\bs \beta}_{jm} \approx \frac{1}{1-2f_{jm}} \overline{\bs y}_{jm}$ and
-%$\widehat{\mathbf{\textrm{V}}}_{jm} \approx \frac{1}{2 N f_{v_{jm}}(1-2f_{jm})} \widehat{\mathbf{\textrm{V}}}_{Y}$ where $f_{jm}$ is the frequency of the 
-%rare allele at variant $m$ of gene $j$ among the $2N$ haplotypes 
-%and $\overline{\bs y}_{jm}$ is the mean phenotype of the carrier individuals of that allele.
-%
-%In the Appendix (\ref{app2}) we give an algorithm to analyze this mixture model using summary statistic data.
-%
-%To compare the models with different numbers of clusters we use BIC\cite{schwarz1978estimating}, defined for the model $\mathcal{M}_C$ with $C$ clusters, as
-%$$BIC_C = -2\log\left(p\left(D | \widehat{\theta_C}, \mathcal{M}_C\right)\right) + \nu_C \log\left(n\right)$$ where $D$ denotes the observed data, $\widehat{\theta_C}$ is the vector of maximum likelihood estimates of the parameters of $\mathcal{M}_C$, $\nu_c$ is the number of independent parameters of $\mathcal{M}_c$ and $n$ is the number of data points.
-The difference in BIC values between two models approximates twice the logarithm of the Bayes factor between the models.
-%
-%In the Appendix (\ref{app2}) we give an algorithm to analyze this mixture model using summary statistic data.
-
-\subsection{Author Contributions}
+% I added in the Author Contributions section
+\section*{Author Contributions}
 M.A.R.
@@
 
-\subsection{Acknowledgments}
+\section*{Acknowledgments}
 This research has been conducted using the UK Biobank Resource under Application Number 24983.
@@
 is supported by Funai Overseas Scholarship from Funai Foundation for Information Technology and the Stanford University Biomedical Informatics Training Program.
-The primary and processed data used to generate the analyses presented here are available in the UK Biobank access management system (https://amsportal.ukbiobank.ac.uk/) for application 24983, ?Generating effective therapeutic hypotheses from genomic and hospital linkage data?
-(http://www.ukbiobank.ac.uk/wp-content/uploads/2017/06/24983-Dr-Manuel-Rivas.pdf), and the results are displayed in the Global Biobank Engine (https://biobankengine.stanford.edu).
+The primary and processed data used to generate the analyses presented here are available in the UK Biobank access management system (https://amsportal.ukbiobank.ac.uk/) for application 24983, ``Generating effective therapeutic hypotheses from genomic and hospital linkage data'' (http://www.ukbiobank.ac.uk/wp-content/uploads/2017/06/24983-Dr-Manuel-Rivas.pdf), and the results are displayed in the Global Biobank Engine (https://biobankengine.stanford.edu).
 We would like to thank the Customer Solutions Team from Paradigm4 who helped us implement efficient databases for queries and application of inference methods to the data, and also implemented optimized versions of truncated singular value decomposition.
@@
 
-\section{Appendix}
-\subsection{MRP Bayes Factor computation} \label{app1}
-To compute the Bayes Factor 
-$$\textrm{BF}_{\textrm{MRP}} = \frac{\det\left(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U} \right)^{-\frac{1}{2}}
-\exp\left[-\frac{1}{2}\left({\bs {\widehat{\beta}} - {\bs \mu}}\right)^{\intercal}\left(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U}\right)^{-1}\left({\bs {\widehat{\beta}}} - {\bs \mu}\right)\right]}
-{\det \left( \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}\right)^{-\frac{1}{2}}\exp \left[-\frac{1}{2}{\bs {\widehat{\beta}}}^{\intercal}\, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} 
-\,{\bs {\widehat{\beta}}}\right]},$$
-we first consider the term inside the exponential function:
-$$ \mathcal{E} \left(\widehat{\bs \beta}, \bs \mu, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}, \mathbf{U}\right) = 
-\frac{1}{2}\, {\bs {\widehat{\beta}}}^{\intercal}\, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} \,{\bs {\widehat{\beta}}}
--\frac{1}{2}\left({\bs {\widehat{\beta}} - {\bs \mu}}\right)^{\intercal}\left(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U}\right)^{-1}\left({\bs {\widehat{\beta}}} - {\bs \mu}\right).
-$$
-Since $\widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}$ and $\mathbf{U}$ are typically defined through Kronecker products of smaller matrices, their inverses
-are easier to compute than the inverse of their sum. Hence we use Woodbury matrix identity to write
-$$ 
-\mathcal{E} \left(\widehat{\bs \beta}, \bs \mu, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}, \mathbf{U}\right) = 
-\frac{1}{2}\, {\bs {\widehat{\beta}}}^{\intercal}\, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} \,{\bs {\widehat{\beta}}}
--\frac{1}{2}\left({\bs {\widehat{\beta}} - {\bs \mu}}\right)^{\intercal} \left( \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} - \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} 
-\left(\mathbf{U}^{-1} + \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} \right)^{-1} \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} \right ) 
-\left({\bs {\widehat{\beta}} - {\bs \mu}}\right).
-$$
+\nolinenumbers
 
-To simplify the determinant calculation we write 
-$$\det\left( \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U} \right) = \det\left( \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}\right) \det\left( \mathbf{I} + \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}^{-1} \mathbf{U}\right).$$
-The logarithm of the Bayes Factor is then
-$$\log \left( \textrm{BF}_{\textrm{MRP}} \right) = 
--\frac{1}{2} \log \left( \det\left( \mathbf{I} + \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}^{-1} \mathbf{U} \right) \right)  + \mathcal{E} \left(\widehat{\bs \beta}, \bs \mu, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}, \mathbf{U}\right).$$
+% From the PLOS website (http://journals.plos.org/plosone/s/latex#loc-plos-template):
+% As stated in the PLOS template, your reference information should be included in your .tex file (not submitted separately as .bib or .bbl).
+Here is a step-by-step way to include your reference list directly within your .tex file:
+% 
+% Compose your LaTeX manuscript as normal, using \cite{bibkey} for reference citations.
+% Compile your manuscript using your .bib file (\bibliography{references}) and the plos2015.bst style file.  This process should output a .bbl file into the same folder as your manuscript.
+% Open the .bbl file and copy/paste its contents into the appropriate position within the manuscript .tex.  Comment out or delete the \bibliography command.
+% Compile your manuscript again. The PDF output should be the same as before.
 
-If studies do not share individuals, $\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}$ is a block-diagonal matrix 
-$$\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}=
- \left[\begin{array}{c|c|c|c}
- \widehat{\mathbf{\textrm{V}}}^{1}_{{\bs \beta}} & 0 & \cdots &0\\ \hline
-0 & \widehat{\mathbf{\textrm{V}}}^{2}_{{\bs \beta}}&\cdots &0 \\ \hline
-\vdots & &\ddots &\vdots \\ \hline
-0 & 0 &\cdots & \widehat{\mathbf{\textrm{V}}}^{S}_{{\bs {\beta}}}\\
-\end{array}\right].$$
-If studies share individuals, e.g., controls, 
-we take the approach of Cichonska et al.~\cite{cichonska2016metacca} to use summary level data
-to estimate the correlation structure of the non-diagonal blocks caused by overlapping individuals.
+% Either type in your references using
+% \begin{thebibliography}{}
+% \bibitem{}
+% Text
+% \end{thebibliography}
+%
+% or
+%
+% Compile your BiBTeX database using our plos2015.bst
+% style file and paste the contents of your .bbl file
+% here. See http://journals.plos.org/plosone/s/latex for 
+% step-by-step instructions.
+%
 
-%\subsection{MRP Mixture model computation} \label{app2}
-%For multiparameter models, the joint posterior distribution may be difficult to sample from directly. 
-%Often, it is easier to sample sequentially from the full conditional distribution of each parameter~\cite{hoff2009first} using a 
-%Gibbs sampler, a Markov Chain Monte Carlo (MCMC) algorithm, that constructs a dependent sequence of parameter 
-%values whose distribution approximates the joint posterior distribution~\cite{geman1984stochastic,casella1992explaining,hoff2009first}. 
-%We propose the following algorithm that is augmented by a Metropolis-Hastings step to update the annotation scaling parameters $\bs \sigma_a^2$ and the sharing intensity parameter $\alpha$. 
+% \begin{thebibliography}{10}
 % 
-%%% 03.01.2016 
-%The implementation of the Gibbs sampler requires the following data structure: 
-%Within gene $j$, the variants $1,\ldots,M_j$ are assigned to clusters $c=1,...,C$,
-%each of which is represented by a parameter $(\bs b_c, \bs \Sigma_c)$ as described in Section~\ref{mrpmm} with
-%$\bs \Sigma_c = \Vy$ being fixed to an estimate of the trait covariance across the clusters or $\widehat{\mathbf{\textrm{V}}}_{jm}$ for each allele when using summary statistics.
+% \bibitem{bib1}
+% Conant GC, Wolfe KH.
+% \newblock {{T}urning a hobby into a job: how duplicated genes find new
+%   functions}.
+% \newblock Nat Rev Genet. 2008 Dec;9(12):938--950.
 % 
-%The Gibbs sampler is run for $n_{\text{burn}} + n_{\text{iter}}$ iterations, of which the first $n_{\text{burn}}$
-%are discarded as an initial "burn-in" part. 
-%With the prior distributions given in Section~\ref{mrpmm}, and superscripts in parentheses denoting iteration, the algorithm is as follows:
+% \bibitem{bib2}
+% Ohno S.
+% \newblock Evolution by gene duplication.
+% \newblock London: George Alien \& Unwin Ltd. Berlin, Heidelberg and New York:
+%   Springer-Verlag.; 1970.
 % 
-%\begin{enumerate}
-%\item Initialize 
-%	\begin{enumerate}
-%	\item Draw $\alpha^{\left(0\right)} \sim \textrm{Inv-Gamma}\left(1,1\right);$
-%	\item Draw $\bs \pi_0^{\left(0\right)} \sim \textrm{Dirichlet}\left(1,1,1,\ldots, 1\right);$ %OK
-%	\item Draw for each gene $j$: $\bs \pi_j^{\left(0\right)} \sim \textrm{Dirichlet}\left(\alpha^{(0)} \bs \pi_0^{\left(0\right)} \right);$
-%	\item ${\bs b}^{\left(0\right)}_{1} = \bs 0;$
-%	\item For each $c$ in $2,\dots,C$ draw $\bs b_c^{\left(0\right)} \sim \mathcal{N}\left(\bs 0, \bs \Theta_0\right);$
-%	\item $\left(\sigma^2_{a}\right)^{\left(0\right)} = 0.2^2$ for all $a$.  
-%	\end{enumerate}
+% \bibitem{bib3}
+% Magwire MM, Bayer F, Webster CL, Cao C, Jiggins FM.
+% \newblock {{S}uccessive increases in the resistance of {D}rosophila to viral
+%   infection through a transposon insertion followed by a {D}uplication}.
+% \newblock PLoS Genet. 2011 Oct;7(10):e1002337.
 % 
-%	\item Repeat for iteration $t = 1, 2, \dots, n_{\text{burn}} + n_{\text{iter}}$
-%	\begin{enumerate}
-%	\item Update $\bs \pi_0$. We use a Metropolis-Hastings sub-step~\cite{hastings1970monte} using a proposal centred around the current value. 
-%	We sample a proposal value for 
-%	$$\bs \pi_0^{'} \sim \textrm{Dirichlet}\left(\gamma \bs \pi_0^{(t-1)} \right).$$ 
-%	To calculate the acceptance probability, we define the normalizing constant for the C-dimensional Dirichlet distribution with parameters $\bs z$ as
-%	$$ D(\bs z) = \frac{\Gamma \left( \sum_{c=1}^C z_c \right) }{\prod_{c=1}^C \Gamma\left ( z_c \right)}$$ and the density at point $\bs x$ as
-%	$$p_{\textit Dir}(\bs x | \bs z ) = D(\bs z) \prod_{c=1}^C x_c^{ z_c -1}.$$
-%	With this notation the Metropolis-Hastings transition probability from $\bs \pi_0^{(t-1)}$ to $\bs \pi_0^{'}$ is $p_{\textit Dir}\left (\bs \pi_0^{'} | \gamma \bs \pi_0^{(t-1)} \right)$
-%	and the density of the observed data depends on $\bs \pi_0^{(t-1)}$ only through the product $\prod_{j=1}^J p_{\textit Dir}\left (\bs \pi_j | \alpha \bs \pi_0^{(t-1)} \right ).$
-%	Hence, the acceptance probability of the proposal is
-%	$$
-%	\lambda = \min\left( 1, \frac{  p_{\textit Dir}\left (\bs \pi_0^{(t-1)} | \gamma \bs \pi_0^{'} \right ) \prod_{j=1}^J p_{\textit Dir}\left (\bs \pi_j | \alpha \bs \pi_0^{'} \right )  }
-%	{p_{\textit Dir}\left (\bs \pi_0^{'} | \gamma \bs \pi_0^{(t-1)} \right ) \prod_{j=1}^J p_{\textit Dir}\left (\bs \pi_j | \alpha \bs \pi_0^{(t-1)} \right ) } \right)$$
-% Thus, with probability $\lambda$, we set $\bs \pi^{\left(t \right)}_0 = \bs \pi^{'}_0$ and with probability $1-\lambda$ we set $\bs \pi^{\left(t \right)}_0 = \bs \pi^{(t-1)}_0$.
-%	
-%	\item For each gene $j=1,\ldots , J$ update $\bs \pi_j$: $$\bs \pi_j \sim \textrm{Dirichlet}\left(\alpha \bs \pi_0 + \left(\sum_{m=1}^{M_j}\delta_{jm} = 1, 
-%	\sum_{m=1}^{M_j}\delta_{jm} = 2, \dots, \sum_{m=1}^{M_j} \delta_{jm} = C\right)\right),$$
-%	where $\delta_{jm}$ is the index of the cluster to which the variant $m$ of gene $j$ belongs to.
-%	
-%	%% update cluster membership 
-%	\item Update $\delta_{jm}$ for all $j=1,\ldots, J$ and $m=1,\ldots,M_j$. 
-%	For all $c$ compute $$p^'_{jmc} = p\left(\hat{\bs \beta}_{jm}; \bs b_c, \bs \Sigma_c\right)\pi_{jc},$$
-%	and renormalize in such a way that $p_{jmc} \propto p^'_{jmc}$ sums to 1 over $c$.
-%	Then we sample $$\delta_{jm} \sim \textrm{Discrete} (p_{jmc}).$$
-%		
-%	%% update cluster effect profile
-%	\item Update $\bs b_{c}$ using a Gibbs update (see derivation below) from a Gaussian distribution with
-%	\begin{eqnarray*}	
-%	\textrm{mean}&=&\left(\bs \Theta_0^{-1} + \sum_{v_{jm} \in c}\sigma^2_{a_{jm}} \bs \Sigma^{-1}_{c} \right)^{-1} \left(\sum_{v_{jm} \in c}\sigma_{a_{jm}}\bs \Sigma_{c}^{-1}\hat{\bs \beta}_{jm}\right)\\
-%	\textrm{var }&=& \left(\bs \Theta_0^{-1} + \sum_{v_{jm} \in c}\sigma^2_{a_{jm}}\bs \Sigma_{c}^{-1}\right)^{-1}.
-%	\end{eqnarray*}
-%	
-%	%% Update sigma^2_a
-%	\item Update $\sigma^2_{a}$. We use a Metropolis-Hastings sub-step~\cite{hastings1970monte} using a random-walk proposal. That is, sequentially for 
-%	each annotation $a$, we sample a proposal value
-%	$$
-%	\sigma^{'}_{a} = |\eta|, \textrm{ where } \eta \sim \mathcal{N}\left(\sigma^{\left(t - 1\right)}_{a}, \xi_0\right),
-%	$$
-%	where $\xi_0$ is a hyperparameter controlling the spread of the proposals. In the examples we have used 
-%	$\xi_0=?$. Then we calculate the acceptance probability  
-%% Summary statistic data
-%%% INSERT EQUATION FOR SUMMARY STATISTICS HERE. 
-%	$$\lambda = \min\left(1,  
-%	\frac{p\left(\sigma^{'}_a | sh_{a}, sc_{a} \right)
-%	\prod_{\textrm{anno}(v_{jm})=a}\mathcal{N}\left(\widehat{\bs \beta}_{jm} | \sigma^{'}_{a}\bs b_{c_{jm}},\widehat{\bs V}_{jm} \right)}
-%	{p\left(\sigma^{(t-1)}_a | sh_{a}, sc_{a} \right)
-%	\prod_{\textrm{anno}(v_{jm})=a}\mathcal{N}\left(\widehat{\bs \beta}_{jm} | \sigma^{(t-1)}_{a}\bs b_{c_{jm}},\widehat{\bs V}_{jm} \right)}\right),$$
-%	where the products are over those variants $v_{jm}$ whose annotation is $a$ and $c_{jm}$ is the cluster of variant $v_{jm}.$
-%	With probability $\lambda$, we set $\sigma^{(t)}_{a} = \sigma^{'}_{a}$ and with probability
-%	$1-\lambda$ we set $\sigma^{(t)}_{a} =\sigma^{(t-1)}_{a}.$
-%
-%	\item Update $\alpha$. We use a Metropolis-Hastings sub-step using a random-walk proposal. That is, we sample 
-%	a proposal value $\alpha^' = |\eta|$, where $\eta \sim \mathcal{N}(\alpha^{(t-1)}, \xi_\alpha )$ where 
-%	$\xi_\alpha$ is a fixed value controlling the variance of the proposal distribution. 
-%	Then we compute the acceptance probability 
-%	$$\lambda = \min\left(1,  
-%	\frac{p\left(\alpha^{'} \right) \prod_{j=1}^J p_{\textit Dir}\left(\bs \pi_j^{(t)}  | \alpha^{'} \bs\pi_0^{(t)}\right) }
-%	{p\left(\alpha^{(t-1)} \right)  \prod_{j=1}^J p_{\textit Dir}\left(\bs \pi_j^{(t)}  | \alpha^{(t-1)} \bs\pi_0^{(t)} \right) } \right),$$
-%	where the prior for $\alpha$ is Inv-Gamma(1,1), i.e., $p(\alpha) = \alpha^{-2}\exp\left(-\alpha^{-1}\right)$, and
-%	$p_{\textit Dir}$ is the density of the Dirichlet distribution as defined above in part (a).
-%	With probability $\lambda$, we set $\alpha^{(t)} = \alpha^{'}$ and with probability
-%	$1-\lambda$ we set $\alpha^{(t)} = \alpha^{(t-1)}.$
-%
-%	\end{enumerate}
-%
-%
-%\end{enumerate}
-
-\subsection{Implementation details: HDF5 tables}
-Although summary statistics are quicker to read and process than raw data, the number of studies meta-analyzed in this work is expected to be sufficiently large to require optimizations in data representation and processing (Figure \ref{hdf5}).
-Our solution was the use of the HDF5 (Hierarchical Data Format 5) data representation to enable rapid processing of effect size, uncertainty, and cross-trait estimate data.
-HDF5 is a fast and lightweight file format designed for scientific data.
-It has bindings for R, Python, C/C++, Java, and nearly every other population programming language.
-Reading data from a table within a HDF5 file can be an order of magnitude faster than reading text files from a Unix file, and it makes it easier to organize data within an internal structure.
-
-%% FIGURE: HDF5 TABLE
-\begin{figure}
-\centering
-\includegraphics[width=\textwidth]{hdf5MAMBAMRP.pdf}
-  \caption{Our HDF5 implementation contained the following components: first, a group with one table per annotation file.
-All effect size (beta) values and study-specific annotations were contained here, and the number of tables is limited by $S\ \textrm{(the number of studies)} \times K\ \textrm{(the number of traits)}$.
-Second, a group with site-site covariance data.
-While these covariance matrices may have dimension $M\ \textrm{(the number of variants)} \times M$, we store the data as tables, each row specifiying the covariance between two variants.
-The number of tables should be the same as the previous set, capped by $S\ \textrm{(the number of studies)} \times K\ \textrm{(the number of traits)}$.
-Third, we store one table with sigma values for each study/phenotype combination.
-In the event that the traits were rank-normal transformation was performed these sigma values are equal to 1.
-These are used to compute correlation between two datasets.
-Finally, we store a matrix/table pair for Vy null and its index.
-The Vy null matrix has dimensions $ (S \times K) \times (S \times K)$ each entry specifying the estimated correlation of effect sizes between two datasets.
-The index table encodes row/column position of each dataset.}
-  \label{hdf5}
-\end{figure}
-
- %% FIGURE: ASTHMA MANHATTAN
-\begin{figure}
-\centering
-\includegraphics[width=\textwidth]{asthma_manhattan.pdf}
-  \caption{$\log_{10}$ Bayes Factors from applying MRP to summary statistics for missense and protein-truncating variants from (A) eosinophil counts (INI30150), (B) forced vital capacity (FVC, INI3062), (C) forced expiratory volume in 1-second (FEV1, INI3063), (D) asthma (HC382), and (E) all four traits jointly.
-The four genes outside of chromosome 6 with the largest Bayes Factors are labeled in each plot.
-Only $\log_{10}$ Bayes Factors greater than -5 are plotted.}
-  \label{asthma_manhattan}
-\end{figure}
-
-%% FIGURE: ASTHMA FOREST PLOT
-\begin{figure}
-\centering
-\includegraphics[width=\textwidth]{asthma_forest.pdf}
-  \caption{$-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{IL33} for eosinophil counts, forced vital capacity (FVC), forced expiratory volume in 1-second (FEV1), and asthma.}
-  \label{asthma_forest}
-\end{figure}
-
-%% TABLE: ASTHMA GENES OF INTEREST
-\begin{table}
-\begin{tabular}{lllllll}
-\toprule
-{} & Joint, protective & Joint & Eosinophil count &   FVC &  FEV1 & Asthma \\
-\midrule
-\textit{IL33}         &              30.1 &  29.8 &             30.9 &  -3.1 &  -2.7 &    9.5 \\
-\textit{CCR3}         &               5.8 &   6.3 &              9.1 &  -1.7 &  -2.3 &   -0.5 \\
-\textit{IL17RA}       &               4.6 &   6.1 &             14.1 &  -4.1 &  -3.9 &   -1.4 \\
-\textit{AL161450.1}   &               3.2 &   3.8 &              4.9 &  -0.1 &  -0.4 &    0.3 \\
-\textit{RP11-39K24.9} &               3.2 &   3.8 &              4.9 &  -0.1 &  -0.4 &    0.3 \\
-\textit{ELN}          &               2.9 &   3.2 &             -1.0 &   5.6 &   2.9 &   -0.6 \\
-\textit{SCMH1}        &               2.7 &   3.9 &             -2.1 &   7.5 &   3.3 &   -0.7 \\
-\textit{TGFB2}        &               2.6 &   4.1 &             -2.8 &   1.7 &  -2.4 &    0.1 \\
-\textit{ZFAT}         &               1.3 &   1.9 &             -2.8 &   6.7 &   4.6 &   -0.4 \\
-\textit{STC2}         &               0.2 &   1.3 &             -0.8 &   3.9 &   2.3 &   -0.7 \\
-\textit{P4HTM}        &              -1.3 &  -0.7 &             -1.6 &   2.9 &   4.3 &   -1.0 \\
-\textit{IL18RAP}      &              -2.2 &  -1.0 &             -0.8 &  -2.0 &  -2.2 &    3.6 \\
-\textit{WDR6}         &              -4.1 &  -2.9 &             -2.3 &   1.9 &   3.4 &   -0.9 \\
-\textit{ATP2A3}       &              -4.2 &  -1.5 &             -1.8 &  -3.1 &  -3.1 &    5.9 \\
-\textit{CHI3L1}       &              -4.4 &  -2.0 &             -2.0 &  -1.9 &  -2.0 &    3.0 \\
-\bottomrule
-\end{tabular}
-\caption{$\log_{10}$ Bayes Factors for genes highlighted in Figure \ref{asthma_manhattan}.}
-\label{asthma_table}
-\end{table}
-
-\newpage
+% \end{thebibliography}
 

### 2018-01-23 tex/manuscript.tex
Updates to PLOS version of manuscript
I believe I have moved over all relevant text to the PLOS version of the
manuscript. There are a few in-text references to subsections (and maybe
to other things like equations that may not be working). I may also need
to reformat the equations into PLOS style. I am going to wait to fix
this stuff until I start working on the text again since some stuff
might move out of the Methods and into an appendix.

I've compiled the references and they are in the document, though before
submission we have to actually paste the contents of the .bbl file into
the .tex

@@
 % Please contact latex@plos.org with any questions.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+
+% % % % % % % % % % % % % % % % % % % % % % % %
+% Notes from Chris
+%
+% Example of a PLOS methods paper: http://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1005896
+%
+% Here are some examples I removed from the body of the document:
+%
+% Equation example
+% \begin{eqnarray}
+% \label{eq:schemeP}
+% 	\mathrm{P_Y} = \underbrace{H(Y_n) - H(Y_n|\mathbf{V}^{Y}_{n})}_{S_Y} + \underbrace{H(Y_n|\mathbf{V}^{Y}_{n})- H(Y_n|\mathbf{V}^{X,Y}_{n})}_{T_{X\rightarrow Y}},
+% \end{eqnarray}
+
+% For figure citations, please use "Fig" instead of "Figure".
+
+% % Place figure captions after the first paragraph in which they are cited.
+% \begin{figure}[!h]
+% \caption{{\bf Bold the figure title.}
+% Figure caption text here, please use this space for the figure panel descriptions instead of using subfigure commands. A: Lorem ipsum dolor sit amet. B: Consectetur adipiscing elit.}
+% \label{fig1}
+% \end{figure}
 %
@@
 \def\bs{\boldsymbol}
+\def\Vy{\widehat{\textrm{V}}_Y}
 
@@
 
-% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
-% 
-% Remove or comment out the author notes below if they aren't used.
-%
-% Primary Equal Contribution Note
-% \Yinyang These authors contributed equally to this work.
-
-% Additional Equal Contribution Note
-% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
-% \ddag These authors also contributed equally to this work.
-
-% Current address notes
-% \textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
-% \textcurrency b Insert second current address 
-% \textcurrency c Insert third current address
-
-% Deceased author note
-% \dag Deceased
-
-% Group/Consortium Author Note
-% \textpilcrow Membership list can be found in the Acknowledgments section.
-
 % Use the asterisk to denote corresponding authorship and provide email address in note below.
@@
 \begin{figure}[!h]
-% \includegraphics[width=\textwidth]{FIG1FINAL.pdf}
 \caption{{\bf Schematic overview of MRP.}
@@
 
-%%Apply methods to the GLGC data and exome sequencing data sets. 
+%%Apply methods to the UK Biobank
 We also apply MRP to summary statistics for two groups of related phenotypes from the UK Biobank.
@@
 
-% Equation example
-% \begin{eqnarray}
-% \label{eq:schemeP}
-% 	\mathrm{P_Y} = \underbrace{H(Y_n) - H(Y_n|\mathbf{V}^{Y}_{n})}_{S_Y} + \underbrace{H(Y_n|\mathbf{V}^{Y}_{n})- H(Y_n|\mathbf{V}^{X,Y}_{n})}_{T_{X\rightarrow Y}},
-% \end{eqnarray}
+\section*{Materials and methods}
+\subsection*{MRP model comparison for association testing}
+We consider the multivariate linear regression model 
+$$\underset{\left(N\times K\right)}{\mathbf{\textrm{Y}}} 
+= \underset{\left(N\times K\right)}{\mathbf{\Psi}}  + \underset{\left(N\times M\right)}{\mathbf{\textrm{X}}}\underset{\left(M \times K\right)}{\mathbf{\textrm{B}}} 
++ \underset{\left(N\times K\right)}{\mathbf{\textrm{E}}},$$
+where the matrices $\mathbf{\textrm{Y}} = \left[{y}_{i k}\right]$, $\mathbf{\textrm{X}} = \left[ x_{i m}\right]$, 
+$\mathbf{\textrm{B}} = \left[ \beta_{m k} \right]$ and $\mathbf{\textrm{E}} = \left[e_{i k}\right]$  describe the phenotype values ($y_{i k}$),
+copies of minor allele ($x_{i m}$), variant-phenotype effects ($\beta_{m k}$), and residual errors ($e_{i k}$), 
+for individual $i$, phenotype $k$, and variant $m$. 
+We assume that each phenotype has been transformed to a standard normal
+distribution and that the columns of ${\mathbf{{\textrm{X}}}}$ 
+have been centered, which means that the estimate for the intercept term ${\mathbf{\Psi}}$ is 0 and independent of the estimate of ${\mathbf{\textrm{B}}}$.
+We use vectorized notation where the rows of ${\mathbf{\textrm{B}}}$ form vector 
+$\bs {{ \beta}} = \left(\bs {{ \beta}}_{1},\dots,\bs {{\beta}}_{M}\right)^{\intercal}$
+of length $MK$.
 
-\section*{Materials and methods}
-A complete description of MRP, along with an analytical derivation, is given in the Online Methods and Appendix. Briefly, we describe the approach.
+We define the MRP model comparison as a Bayes factor (BF) between the alternative model, where at least one variant
+affects at least one phenotype, and the null model, where all variant-phenotype effects are zero.  
+BF is the ratio of the marginal likelihoods for these two models: 
+$$ \textrm{BF} = \frac{\int_{{\bs {\beta}}}p\left(\mathbf{\textrm{Data}}|{\bs {\beta}}\right)p\left({\bs {\beta}}|\textrm{ALT}\right)d{\bs {\beta}}}{\int_{{\bs {\beta}}}p\left(\mathbf{\textrm{Data}}|{\bs {\beta}}\right)p\left({\bs {\beta}}|\textrm{NULL}\right)d{\bs {\beta}}},$$
+where Data can correspond either to the effect size estimates $\bs {\widehat{ \beta}}$ and the estimated variance-covariance matrix of  $\bs {\widehat{ \beta}}$, $\widehat{\mathbf{\textrm{V}}}_{\bs{\beta}}$, or to the original phenotypes and genotypes,
+$\underset{\left(N\times K\right)}{\mathbf{\textrm{\textrm{Y}}}}$ and $ \underset{\left(N\times M\right)}{\mathbf{\textrm{\textrm{X}}}}$, and any other covariates that we
+want to regress out from the phenotypes.
+
+The prior distribution for the null model, $p\left({\bs {\beta}}|\textrm{NULL}\right)$, is simply the point mass at $\bs \beta = 0$.
+In subsection~\ref{approxlikelihood} we show how we approximate the likelihood function for $\bs \beta$, $p\left(\mathbf{\textrm{Data}} | {\bs {\beta}}\right)$,
+in subsection~\ref{prioralt} we define the prior distribution $p\left({\bs {\beta}}|\textrm{ALT}\right)$ for the alternative model, 
+and finally, in subsection~\ref{bfmodel}, we compute the BF.
+
+\subsubsection*{Likelihood function} 
+\label{approxlikelihood}
+A maximum likelihood estimator of $\mathbf{\textrm{B}}$ is given by the ordinary least-squares method
+$$\widehat{\mathbf{\textrm{B}}} = \left(\mathbf{\textrm{X}}^\intercal\mathbf{\textrm{X}}\right)^{-1}\mathbf{\textrm{X}}^{\intercal}\mathbf{\textrm{Y}},$$
+that in vectorized form is denoted by   
+$\bs {\widehat{ \beta}} = \left(\bs {\widehat{ \beta}}_{1},\dots,\bs {\widehat{\beta}}_{M}\right)^{\intercal}.$
+An estimator of the variance-covariance of $\widehat{\bs \beta}$  is given by 
+$$\widehat{\textrm{V}}_{\mathbf{\bs \beta}} =  \left(\mathbf{\textrm{X}}^\intercal\mathbf{\textrm{X}}\right)^{-1}\otimes \widehat{\textrm{V}}_{\mathbf{\textrm{Y}}},$$
+where $\widehat{\textrm{V}}_{\mathbf{\textrm{Y}}}$ is the estimated residual variance-covariance matrix of $\mathbf{\textrm{Y}}$ given $\mathbf{\textrm{X}}$.
+
+%Let $\bs {\widehat{ \beta}} = \left(\bs {\widehat{ \beta}}^{\intercal}_{1},\dots,\bs {\widehat{\beta}}^{\intercal}_{m}\right)^{\intercal}$ be a vector of point estimates of all $MK$ effects. 
+Following Band et al.\cite{band2013imputation}, we approximate the likelihood function of $\bs {\beta}$ by a multivariate normal distribution with mean $\bs {\widehat{\beta}}$ and variance-covariance matrix $\widehat{\textrm{V}}_{\bs {\beta}}$.
+Note that by approximating $\Vy$ by the trait correlation matrix, this likelihood approximation does not require access to the individual level data $\mathbf{\textrm{X}}$ and $\mathbf{\textrm{Y}}$ but only to the
+summary data of effect sizes $\bs {\widehat{\beta}}$, LD-matrix $\mathbf{\textrm{X}}^\intercal\mathbf{\textrm{X}}$ and a trait correlation estimate.
+
+\subsubsection*{Prior of $\bs{\beta}$ in the alternative model} 
+\label{prioralt}
+We construct the prior distribution $p\left({\bs {\beta}} | \textrm{ALT} \right)$ for the alternative model in three steps allowing 
+user to specify correlations between effects of different variants on different traits across different studies.
+
+In a single study, the prior density for ${\bs {\beta}}$ incorporates the expected correlation of genetic effects among 
+a group of variants ($\mathbf{R}_{\textrm{var}}$) and among a group of phenotypes ($\mathbf{R}_{\textrm{phen}}$). 
+In addition, we  incorporate an expected spread of the effect size of each variant by scaling $\mathbf{R}_{\textrm{var}}$ as 
+$$\mathbf{S}_{\textrm{var}} = \Delta\left(\sigma_m\right) \mathbf{R}_{\textrm{var}} \Delta\left(\sigma_m\right),$$
+where $\Delta\left(\sigma_m\right)$ is a diagonal matrix with entries $\sigma_m$ determining the spread of the effect size distribution for each variant $m \leq M$.
+Thus, we can model settings where, e.g., protein-truncating variants have larger effect sizes ($\sigma = 1$) than missense variants ($\sigma = 0.1$).
+Note that when $\sigma_m = 1$ for all $m$ then $\mathbf{S}_{\textrm{var}} = \mathbf{R}_{\textrm{var}}$.
+
+All in all, our prior density for ${\bs \beta}$ under alternative model is 
+$${\bs {\beta}}|\textrm{ALT} \sim \mathcal{N}\left(\mathbf{0}, \mathbf{U}\right), \textrm{ where } \mathbf{U} = \mathbf{S}_{\textrm{var}} \otimes \mathbf{R}_{\textrm{phen}}.$$
+
+When we have data from multiple studies 
+we allow for possible differences in genetic effects across ethnicities or populations 
+extending the Approximate Bayes Factors of Band et al.~\cite{band2013imputation}
+and the summary statistics approach of RAREMETAL~\cite{liu2014meta} from univariate to multivariate phenotypes.
+Let $\bs {\widehat{ \beta}} = \left(\widehat{ \beta}_{s,m,k}\right)=\left(\widehat{\beta}_{1,1,1}, \widehat{\beta}_{1,1,2}, \dots, \widehat{\beta}_{1,1,K}, \widehat{\beta}_{1,2,1}, \dots, \widehat{\beta}_{1,2,K}, \dots, \widehat{\beta}_{1,M,K}, \widehat{\beta}_{2,1,1}, \dots, \widehat{\beta}_{S,M,K}\right),$ where $S$ is the number of studies, $M$ is the number of variants, and $K$ is the number of phenotypes.
+As with a single study, we incorporate the expected correlation of genetic effects between a pair of variants and a single phenotype using the matrix $\mathbf{S_{\textrm{var}}}$, between a variant and a pair of phenotypes using the matrix $\mathbf{R_{\textrm{phen}}}$, and we introduce the matrix $\mathbf{R_{\textrm{study}}}$ to specify prior on the similarity in effect sizes across the studies.
+Thus, the prior is
+$${\bs {\beta}} \sim \mathcal{N}\left(\mathbf{0}, \mathbf{U}\right), \textrm{ where } 
+\mathbf{U} = \mathbf{R}_{\textrm{study}}\otimes\left(\mathbf{S}_{\textrm{var}} \otimes \mathbf{R}_{\textrm{phen}}\right).$$
+
+It is straightforward to include a non-zero vector ${\bs\mu}$ as a prior mean of genetic effects, in which case the prior is
+$${\bs {\beta}} \sim \mathcal{N}\left({\bs \mu},\mathbf{U}\right).$$
+We use this, for example, when screening for protective rare variants that have a pre-specified beneficial profile on a set of risk factors.
+
+\subsubsection*{$\textrm{BF}_{\textrm{MRP}}$} 
+\label{bfmodel}
+The Bayes Factor is the ratio of the marginal likelihoods between the alternative and the null model. 
+The marginal likelihood for the alternative model is 
+$$ \int_{{\bs {\beta}}} p\left(\mathbf{\textrm{Data}}|{\bs {\beta}}\right) p\left({\bs {\beta}}|\textrm{ALT}\right)d{\bs {\beta}} = c \times \mathcal{N}\left({ {\widehat{\bs \beta}}}; {\bs \mu},  \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U}\right)$$
+and the marginal likelihood for the null model is 
+$$\int_{{\bs {\beta}}} p\left(\mathbf{\textrm{Data}}|{\bs {\beta}}\right) p\left({\bs {\beta}}|\textrm{NULL}\right)d{\bs {\beta}} = c \times \mathcal{N}\left({ {\widehat{\bs \beta}}}; 0,  \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}\right).$$
+
+The Bayes Factor (derived is below) is given by
+$$\textrm{BF}_{\textrm{MRP}} = \frac{\det\left(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U} \right)^{-\frac{1}{2}}
+\exp\left[-\frac{1}{2}\left({\bs {\widehat{\beta}} - {\bs \mu}}\right)^{\intercal}\left(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U}\right)^{-1}\left({\bs {\widehat{\beta}}} - {\bs \mu}\right)\right]}
+{\det \left( \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}\right)^{-\frac{1}{2}}\exp \left[-\frac{1}{2}{\bs {\widehat{\beta}}}^{\intercal}\, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} 
+\,{\bs {\widehat{\beta}}}\right]}.$$
+
+When $\bs \mu=0$, $\textrm{BF}_{\textrm{MRP}}$ is an increasing function of the following quadratic form
+\begin{equation}\label{quadratic_form}
+Q(\bs{\widehat{\beta}};\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}},\mathbf{U})=
+\bs{\widehat{\beta}^\intercal} \left(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}^{-1} - (\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}+\mathbf{U})^{-1}\right) \bs{\widehat{\beta}}.
+\end{equation}
+Furthermore, this quadratic form is the only part of the $\textrm{BF}_{\textrm{MRP}}$ that depends on $\bs {\widehat{\beta}}$.
+Thus, by deriving a distribution of $Q(\bs{\widehat{\beta}};\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}},\mathbf{U})$
+under the null model we can compute a p-value when $\textrm{BF}_{\textrm{MRP}}$ is used as a test statistic.
+According to basic properties of quadratic forms of Gaussian variables,
+$Q(\bs{\widehat{\beta}};\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}},\mathbf{U}) \sim \sum_{i=1}^n d_i \chi^2_i$, where $\chi_i^2$
+are an independent sample from $\chi_1^2$ distribution (chi-square with one degree of freedom),
+and $d_i$ are the eigenvalues of matrix
+$I-(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}+\mathbf{U})^{-1}\, \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}$.
+The distribution function for a mixture of chi-squares can be numerically evaluated by
+the R-package `CompQuadForm'~\cite{Duchesne2010}.
+
+\subsubsection*{MRP Bayes Factor derivation} \label{app1}
+To compute the Bayes Factor 
+$$\textrm{BF}_{\textrm{MRP}} = \frac{\det\left(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U} \right)^{-\frac{1}{2}}
+\exp\left[-\frac{1}{2}\left({\bs {\widehat{\beta}} - {\bs \mu}}\right)^{\intercal}\left(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U}\right)^{-1}\left({\bs {\widehat{\beta}}} - {\bs \mu}\right)\right]}
+{\det \left( \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}\right)^{-\frac{1}{2}}\exp \left[-\frac{1}{2}{\bs {\widehat{\beta}}}^{\intercal}\, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} 
+\,{\bs {\widehat{\beta}}}\right]},$$
+we first consider the term inside the exponential function:
+$$ \mathcal{E} \left(\widehat{\bs \beta}, \bs \mu, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}, \mathbf{U}\right) = 
+\frac{1}{2}\, {\bs {\widehat{\beta}}}^{\intercal}\, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} \,{\bs {\widehat{\beta}}}
+-\frac{1}{2}\left({\bs {\widehat{\beta}} - {\bs \mu}}\right)^{\intercal}\left(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U}\right)^{-1}\left({\bs {\widehat{\beta}}} - {\bs \mu}\right).
+$$
+Since $\widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}$ and $\mathbf{U}$ are typically defined through Kronecker products of smaller matrices, their inverses
+are easier to compute than the inverse of their sum. Hence we use Woodbury matrix identity to write
+$$ 
+\mathcal{E} \left(\widehat{\bs \beta}, \bs \mu, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}, \mathbf{U}\right) = 
+\frac{1}{2}\, {\bs {\widehat{\beta}}}^{\intercal}\, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} \,{\bs {\widehat{\beta}}}
+-\frac{1}{2}\left({\bs {\widehat{\beta}} - {\bs \mu}}\right)^{\intercal} \left( \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} - \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} 
+\left(\mathbf{U}^{-1} + \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} \right)^{-1} \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} \right ) 
+\left({\bs {\widehat{\beta}} - {\bs \mu}}\right).
+$$
+
+To simplify the determinant calculation we write 
+$$\det\left( \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U} \right) = \det\left( \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}\right) \det\left( \mathbf{I} + \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}^{-1} \mathbf{U}\right).$$
+The logarithm of the Bayes Factor is then
+$$\log \left( \textrm{BF}_{\textrm{MRP}} \right) = 
+-\frac{1}{2} \log \left( \det\left( \mathbf{I} + \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}^{-1} \mathbf{U} \right) \right)  + \mathcal{E} \left(\widehat{\bs \beta}, \bs \mu, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}, \mathbf{U}\right).$$
+
+If studies do not share individuals, $\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}$ is a block-diagonal matrix 
+$$\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}=
+ \left[\begin{array}{c|c|c|c}
+ \widehat{\mathbf{\textrm{V}}}^{1}_{{\bs \beta}} & 0 & \cdots &0\\ \hline
+0 & \widehat{\mathbf{\textrm{V}}}^{2}_{{\bs \beta}}&\cdots &0 \\ \hline
+\vdots & &\ddots &\vdots \\ \hline
+0 & 0 &\cdots & \widehat{\mathbf{\textrm{V}}}^{S}_{{\bs {\beta}}}\\
+\end{array}\right].$$
+If studies share individuals, e.g., controls, 
+we take the approach of Cichonska et al.~\cite{cichonska2016metacca} to use summary level data
+to estimate the correlation structure of the non-diagonal blocks caused by overlapping individuals.
+
+\subsection*{HDF5 Tables}
+Although summary statistics are quicker to read and process than raw data, the number of studies meta-analyzed in this work is expected to be sufficiently large to require optimizations in data representation and processing (Figure \ref{S1_Fig}).
+Our solution was the use of the HDF5 (Hierarchical Data Format 5) data representation to enable rapid processing of effect size, uncertainty, and cross-trait estimate data.
+HDF5 is a fast and lightweight file format designed for scientific data.
+It has bindings for R, Python, C/C++, Java, and nearly every other population programming language.
+Reading data from a table within a HDF5 file can be an order of magnitude faster than reading text files from a Unix file, and it makes it easier to organize data within an internal structure.
+
+\subsection*{UK Biobank Data}
+\subsubsection*{GWAS Summary Statistics}
+We performed genome-wide association analysis using PLINK v2.00a(17 July 2017) as previously described~\cite{DeBoever179762}.
+For asthma, we used the Firth-fallback in PLINK, a hybrid algorithm which normally uses the logistic regression code described in (Hill 2017), but switches to a port of logistf() (https://cran.r-project.org/web/packages/logistf/index.html) in two cases: (1) one of the cells in the 2x2 allele count by case/control status contingency table is empty (2) logistic regression was attempted since all the contingency table cells were nonzero, but it failed to converge within the usual number of steps.
+We used the following covariates in our analysis: age, sex, array type, and the first four principal components, where array type is a binary variable that represents whether an individual was genotyped with UK Biobank Axiom Array or UK BiLEVE Axiom Array.
+For variants that were specific to one array, we did not use array as a covariate.
+Asthma cases were defined using both Hospital Episode Statistics and verbal questionnaire responses.
+We used the provided values from the UK Biobank for eosinophil counts, forced vital capacity (FVC), and forced expiratory volume in 1-second (FEV1).
+The phenotype codes used throughout (asthma=HC382, eosinophil count=INI30150, Forced Expiratory Volume in 1-second (FEV1)=INI3063, and Forced Vital Capacity (FVC)=INI3062) correspond to the phenotype codes used in on the Global Biobank Engine [\url{https://biobankengine.stanford.edu}].
+
+\subsubsection*{Genetic Correlations}
+We calculated the genetic between asthma, eosinophil counts, forced vital capacity (FVC), and forced expiratory volume in 1-second (FEV1) using the MultiVariate Polygenic Mixture Model (MVPMM)~\cite{DeBoever2017}.
+Briefly, MVPMM estimates genetic correlation given GWAS summary statistics (effect size and standard error of effect size estimate) by modeling GWAS summary statistics as generated from one of two mixture components.
+Summary statistics from variants in the null component are modeled as being drawn from a multivariate normal distribution with zero mean and covariance matrix that captures correlation in the summary statistics due to the use of shared subjects or other sources of correlation.
+Summary statistics from variants in the non-null component are modeled as being drawn from a multivariate normal distribution with zero mean, but the covariance matrix for the non-null component combines the covariance matrix from the null component with another covariance matrix that captures the genetic correlation between the phenotypes being considered.
+
+\subsubsection*{UK Biobank Asthma Application}
+For the Manhattan plots and Table \ref{asthma_table}, we removed any genes with non-unique gene symbols.
+We also removed the sense overlapping gene \textit{CTD-3064M3.3} and and the antisense gene \textit{ZFAT-AS1} since these genes overlap \textit{GPR20} and \textit{ZFAT} respectively and therefore have the same $\log_{10}$ Bayes Factors as those genes.
+
+% Results and Discussion can be combined.
+\section*{Results}
+
+\subsection*{Method overview}
+A complete description of MRP, along with an analytical derivation, is given in the Materials and Methods. Briefly, we describe the approach.
 
@@
 
-% For figure citations, please use "Fig" instead of "Figure".
-
-% % Place figure captions after the first paragraph in which they are cited.
-% \begin{figure}[!h]
-% \caption{{\bf Bold the figure title.}
-% Figure caption text here, please use this space for the figure panel descriptions instead of using subfigure commands. A: Lorem ipsum dolor sit amet. B: Consectetur adipiscing elit.}
-% \label{fig1}
-% \end{figure}
-
-% Results and Discussion can be combined.
-\section*{Results}
-
 \begin{figure}[!h]
@@
 We observe concordance between the frequentist methods and the Bayesian models.
-To compare the Bayesian models we compute p-values by using the Bayes Factor as the test statistic and approximating it using distribution properties of quadratic forms (Appendix).
+To compare the Bayesian models we compute p-values by using the Bayes Factor as the test statistic and approximating it using distribution properties of quadratic forms (Materials and Methods).
 As expected, an independent effects model has high correlation with the gene-based test SKAT ($r^2 = 0.99$), whereas the similar effects model has high correlation with the burden test ($r^2 = 0.93$, Fig \ref{fig2}A).
@@
 \label{S1_Fig}
-{\bf Bold the title sentence.} Add descriptive text after the title of the item (optional).
+{\bf HDF5 Implementation.} Our HDF5 implementation contained the following components: first, a group with one table per annotation file.
+All effect size (beta) values and study-specific annotations were contained here, and the number of tables is limited by $S\ \textrm{(the number of studies)} \times K\ \textrm{(the number of traits)}$.
+Second, a group with site-site covariance data.
+While these covariance matrices may have dimension $M\ \textrm{(the number of variants)} \times M$, we store the data as tables, each row specifiying the covariance between two variants.
+The number of tables should be the same as the previous set, capped by $S\ \textrm{(the number of studies)} \times K\ \textrm{(the number of traits)}$.
+Third, we store one table with sigma values for each study/phenotype combination.
+In the event that the traits were rank-normal transformation was performed these sigma values are equal to 1.
+These are used to compute correlation between two datasets.
+Finally, we store a matrix/table pair for Vy null and its index.
+The Vy null matrix has dimensions $ (S \times K) \times (S \times K)$ each entry specifying the estimated correlation of effect sizes between two datasets.
+The index table encodes row/column position of each dataset.
 
-\paragraph*{S2 Fig.}
-\label{S2_Fig}
-{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+% \paragraph*{S2 Fig.}
+% \label{S2_Fig}
+% {\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
 Etiam eget sapien nibh.
@@
 Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
-\paragraph*{S1 File.}
-\label{S1_File}
-{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
+% 
+% \paragraph*{S1 File.}
+% \label{S1_File}
+% {\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
 Etiam eget sapien nibh.
@@
 Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
-\paragraph*{S1 Video.}
-\label{S1_Video}
-{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
+% 
+% \paragraph*{S1 Video.}
+% \label{S1_Video}
+% {\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
 Etiam eget sapien nibh.
@@
 Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
-\paragraph*{S1 Appendix.}
-\label{S1_Appendix}
-{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+% 
+% \paragraph*{S1 Appendix.}
+% \label{S1_Appendix}
+% {\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
 Etiam eget sapien nibh.
@@
 Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
-\paragraph*{S1 Table.}
-\label{S1_Table}
-{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+% 
+% \paragraph*{S1 Table.}
+% \label{S1_Table}
+% {\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
 Etiam eget sapien nibh.
@@
 provided critical feedback on methodology.
+TODO: Add C.
+Spencer contributions.
 

### 2018-01-24 tex/old/MVMetaLearnMRP2017OctoberMRPMM.tex

--- 2018-01-18 tex/MVMetaLearnMRP2017OctoberMRPMM.tex
@@
 \title{Bayesian model comparison for rare variant association studies of multiple phenotypes}
-\author{Christopher DeBoever$^{1}$, Matthew Aguirre$^{1}$, Yosuke Tanigawa$^{1}$, Timothy Poterba$^{2}$, Mark J.
-Daly$^{2,3}$,  Matti Pirinen$^{4}$\thanks{matti.pirinen@helsinki.fi}, and Manuel A.
+\author{Christopher DeBoever$^{1}$, Matthew Aguirre$^{1}$, Yosuke Tanigawa$^{1}$, Chris C.
+A.
+Spencer$^{2}$, Timothy Poterba$^{3}$, Mark J.
+Daly$^{3,4}$,  Matti Pirinen$^{5}$\thanks{matti.pirinen@helsinki.fi}, and Manuel A.
 Rivas$^{1}$\thanks{mrivas@stanford.edu}}
@@
 \item Department of Biomedical Data Science, Stanford University, Stanford, CA, USA
+\item Genomics plc, Oxford, UK.
 \item Broad Institute of MIT and Harvard, Cambridge, MA, USA
@@
 We first applied MRP to the asthma, eosinophil count, forced expiratory volume in 1-second, and forced vital capacity phenotypes.
-We applied the model to missense variants and PTVs for each phenotype separately (Fig.~\ref{asthma_manhattan}A-D) as well as to all phenotypes jointly (Fig.~\ref{asthma_manhattan}E) and obtained $\log_{10}$ BF for each gene.
+We applied the model to missense variants and PTVs for each phenotype separately (Fig.~\ref{asthma_combined}A-D) as well as to all phenotypes jointly (Fig.~\ref{asthma_combined}E) and obtained $\log_{10}$ BF for each gene.
 As expected from recent studies \cite{DeBoever179762, 10.1371/journal.pgen.1006659}, we observed evidence that missense variants and PTVs in \textit{IL33} affect eosinophil counts and offer protection from asthma from the single-phenotype analyses, though the evidence of association was strongest for the joint analysis (Table~\ref{asthma_table}).
@@
 \centering
-\includegraphics[width=\textwidth]{glaucoma_manhattan.pdf}
+\includegraphics[width=\textwidth]{glaucoma_combined.pdf}
   \caption{$\log_{10}$ Bayes Factors from applying MRP to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
@@
 The genes outside of chromosome 6 with with Bayes Factor greater than three are indicated by arrows.
-Only $\log_{10}$ Bayes Factors greater than zero are plotted.}
-  \label{glaucoma_manhattan}
-\end{figure}
-
-%% FIGURE 4
-\begin{figure}
-\centering
-\includegraphics[width=\textwidth]{glaucoma_forest.pdf}
-  \caption{$-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{ANGPTL7} for glaucoma, intra-ocular pressure, and corneal resistance factor.}
-  \label{glaucoma}
+Only $\log_{10}$ Bayes Factors greater than zero are plotted.
+(F) $-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{ANGPTL7} for all three phenotypes.}
+  \label{glaucoma_combined}
 \end{figure}
@@
 \end{tabular}
-\caption{$\log_{10}$ Bayes Factors for genes highlighted in Figure \ref{glaucoma_manhattan}.}
+\caption{$\log_{10}$ Bayes Factors for genes highlighted in Figure \ref{glaucoma_combined}.}
 \label{glaucoma_table}
@@
 
- %% FIGURE: ASTHMA MANHATTAN
+ %% FIGURE: ASTHMA MANHATTAN AND FOREST
 \begin{figure}
 \centering
-\includegraphics[width=\textwidth]{asthma_manhattan.pdf}
-  \caption{$\log_{10}$ Bayes Factors from applying MRP to summary statistics for missense and protein-truncating variants from (A) eosinophil counts (INI30150), (B) forced vital capacity (FVC, INI3062), (C) forced expiratory volume in 1-second (FEV1, INI3063), (D) asthma (HC382), and (E) all four traits jointly.
+\includegraphics[width=\textwidth]{asthma_combined.pdf}
+  \caption{$\log_{10}$ Bayes Factors from applying MRP to summary statistics for missense and protein-truncating variants from (A) asthma (HC382), (B) eosinophil counts (INI30150), (C) forced vital capacity (FVC, INI3062), (D) forced expiratory volume in 1-second (FEV1, INI3063), (E) all four traits jointly, and (F) all four traits jointly with focus on protective effects.
 The four genes outside of chromosome 6 with the largest Bayes Factors are labeled in each plot.
-Only $\log_{10}$ Bayes Factors greater than -5 are plotted.}
-  \label{asthma_manhattan}
-\end{figure}
-
-%% FIGURE: ASTHMA FOREST PLOT
-\begin{figure}
-\centering
-\includegraphics[width=\textwidth]{asthma_forest.pdf}
-  \caption{$-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{IL33} for eosinophil counts, forced vital capacity (FVC), forced expiratory volume in 1-second (FEV1), and asthma.}
-  \label{asthma_forest}
+Only $\log_{10}$ Bayes Factors greater than -5 are plotted.
+(F) $-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{IL33} for each phenotype}
+  \label{asthma_combined}
 \end{figure}
@@
 \end{tabular}
-\caption{$\log_{10}$ Bayes Factors for genes highlighted in Figure \ref{asthma_manhattan}.}
+\caption{$\log_{10}$ Bayes Factors for genes highlighted in Figure \ref{asthma_combined}.}
 \label{asthma_table}

### 2018-01-26 tex/manuscript.tex
Text updates
I need to add the new stuff on Bayesian model averaging.

@@
 In this paper, we present a Bayesian multiple rare variants and phenotypes (MRP) model comparison approach for identifying rare variant associations as an alternative to current widely-used statistical tests.
-The MRP framework exploits correlation, scale, or location (direction) of genetic effects in a broad range of rare variant association study designs including: case-control; multiple diseases and shared controls; single continuous phenotype; multiple continuous phenotypes; or a mixture of case-control and multiple continuous phenotypes (Fig \ref{fig1}).
+The MRP framework exploits correlation, scale, or location (direction) of genetic effects in a broad range of rare variant association study designs including: case-control; multiple diseases and shared controls; single continuous phenotype; multiple continuous phenotypes; or a mixture of case-control and multiple continuous phenotypes (Fig~\ref{fig1}).
 MRP makes use of Bayesian model comparison, whereby we compute a Bayes Factor (BF) defined as the ratio of the marginal likelihoods of the observed data under two models: 1) a pre-specified null where all genetic effects are zero; and 2) an alternative model where factors like correlation, scale, or location of genetic effects are considered.
@@
 \subsection*{HDF5 Tables}
-Although summary statistics are quicker to read and process than raw data, the number of studies meta-analyzed in this work is expected to be sufficiently large to require optimizations in data representation and processing (Figure \ref{S1_Fig}).
+Although summary statistics are quicker to read and process than raw data, the number of studies meta-analyzed in this work is expected to be sufficiently large to require optimizations in data representation and processing (Fig~\ref{S1_Fig}).
 Our solution was the use of the HDF5 (Hierarchical Data Format 5) data representation to enable rapid processing of effect size, uncertainty, and cross-trait estimate data.
@@
 \subsubsection*{UK Biobank Asthma Application}
-For the Manhattan plots and Table \ref{asthma_table}, we removed any genes with non-unique gene symbols.
+For the Manhattan plots and Table~\ref{asthma_table}, we removed any genes with non-unique gene symbols.
 We also removed the sense overlapping gene \textit{CTD-3064M3.3} and and the antisense gene \textit{ZFAT-AS1} since these genes overlap \textit{GPR20} and \textit{ZFAT} respectively and therefore have the same $\log_{10}$ Bayes Factors as those genes.
@@
 
-Although we see advantages in adopting a Bayesian perspective, our approach could be used in a frequentist context by calculating a Bayes factor and using it as a test statistic to compute p-values (Methods, Fig \ref{fig2}).
+Although we see advantages in adopting a Bayesian perspective, our approach could be used in a frequentist context by calculating a Bayes factor and using it as a test statistic to compute p-values (Methods, Fig~\ref{fig2}).
 
@@
 To compare the Bayesian models we compute p-values by using the Bayes Factor as the test statistic and approximating it using distribution properties of quadratic forms (Materials and Methods).
-As expected, an independent effects model has high correlation with the gene-based test SKAT ($r^2 = 0.99$), whereas the similar effects model has high correlation with the burden test ($r^2 = 0.93$, Fig \ref{fig2}A).
+As expected, an independent effects model has high correlation with the gene-based test SKAT ($r^2 = 0.99$), whereas the similar effects model has high correlation with the burden test ($r^2 = 0.93$, Fig~\ref{fig2}A).
 
@@
 We conducted $1000$ simulation experiments where we let $K$ (the number of phenotypes) $=3$, $M$ (the number of variants) $=10$, $S$ (the number of studies) $=2$, $N_0$ (number of individuals in study with access to all the data) $= 10000$, $N_1$ (meta-analysis study 1) $= 5000$, $N_2$ (meta-analysis study 2) $= 5000$.
-We find that, under the scenario where similar effects are assumed across studies, the Bayes Factors obtained using summary statistics alone are strongly correlated ($r^2 = 1$) to Bayes Factors obtained by the full genotype and phenotype data (Fig \ref{fig2}B).
+We find that, under the scenario where similar effects are assumed across studies, the Bayes Factors obtained using summary statistics alone are strongly correlated ($r^2 = 1$) to Bayes Factors obtained by the full genotype and phenotype data (Fig~\ref{fig2}B).
 
@@
 We simulated three studies and applied the model comparison unit jointly to summary statistic data obtained for each study (Supplementary Note).
-Overall, we observed that considering the joint effects across multiple studies in a group of variants and phenotypes may improve ability to detect gene-based signals (Fig \ref{fig2}C), and that considering prior mean of genetic effects should aid in efforts to identify protective modifiers of disease risk (Fig \ref{fig2}D).
+Overall, we observed that considering the joint effects across multiple studies in a group of variants and phenotypes may improve ability to detect gene-based signals (Fig~\ref{fig2}C), and that considering prior mean of genetic effects should aid in efforts to identify protective modifiers of disease risk (Fig~\ref{fig2}D).
 
 \subsection*{Applications}
-We demonstrate its use in rare variant analysis in the UK Biobank.
+We also applied the MRP model comparison approach to summary statistic data generated from single variant logistic regression and linear regression analysis for coding variants on the UK Biobank array (Methods).
+We applied the MRP model comparison approach to asthma and three related traits and glaucoma and two related traits.
 
-\subsubsection*{Model comparison applied to summary statistic from the UK Biobank}
-We applied the MRP model comparison approach to summary statistic data generated from single variant logistic regression and linear regression analysis for coding variants on the UK Biobank array (Methods).
-We first applied MRP to the asthma, eosinophil count, forced expiratory volume in 1-second, and forced vital capacity phenotypes.
-We applied the model to missense variants and PTVs for each phenotype separately (Fig \ref{asthma_combined}A-D) as well as to all phenotypes jointly (Fig \ref{asthma_combined}E) and obtained $\log_{10}$ BF for each gene.
-As expected from recent studies \cite{DeBoever179762, 10.1371/journal.pgen.1006659}, we observed evidence that missense variants and PTVs in \textit{IL33} affect eosinophil counts and offer protection from asthma from the single-phenotype analyses, though the evidence of association was strongest for the joint analysis (Table~\ref{asthma_table}).
+\subsubsection*{Model comparison applied to asthma and related phenotypes}
+We first applied MRP to GWAS summary statistics for asthma, eosinophil count, forced expiratory volume in 1-second (FEV$_1$), and forced vital capacity (FVC) phenotypes.
+Recent work has identified associations between the PTV rs146597587 in \textit{IL33} and asthma and eosinophil counts \cite{DeBoever179762, 10.1371/journal.pgen.1006659}.
+FEV$_1$ and FVC are measures of pulmonary function that are used to diagnosis and classify pulmonary disease \cite{Swanney1046}.
+To demonstrate the advantage of considering the phenotypes jointly, we applied the model to missense variants and PTVs for each phenotype separately (Fig~\ref{asthma_combined}A-D) as well as to all phenotypes jointly (Fig~\ref{fig3}E) and obtained $\log_{10}$ BF for each gene.
+In agreement with previous studies, we observed evidence that missense variants and PTVs in \textit{IL33} affect eosinophil counts and offer protection from asthma from the single-phenotype analyses, though the evidence of association was strongest for the joint analysis (Table~\ref{asthma_table}) \cite{DeBoever179762, 10.1371/journal.pgen.1006659}.
+We also performed an analysis focused on identifying protective variants which also identified the \textit{IL33} association (Fig~\ref{fig3}).
+
 We also found evidence for associations between missense variants and PTVs in several other genes and the four phenotypes in the joint model that were not identified when considering asthma alone.
@@
 
-We also applied MRP to missense variants and PTVs for glaucoma, intra-ocular pressure, and corneal resistance factor as well as performing a joint analysis and a joint analysis focused on protective associations.
-While the individual glaucoma analysis did not yield any associations with $\log_{10}$ BF greater than 5, the joint analyses identified missense and PTVs in both \textit{ANGPTL7} and \textit{WNT10A} as associated with protection against glaucoma.
-These results illustrate the ability of applying model comparison to multiple traits simultaneously to identify protective modifiers of disease.
-
-% \begin{figure}
-% \centering
-% \includegraphics[width=\textwidth]{glaucoma_combined.pdf}
-%   \caption{$\log_{10}$ Bayes Factors from applying MRP to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
-(E) shows the results of a joint analysis focused on finding rare variants that protect against glaucoma.
-The genes outside of chromosome 6 with with Bayes Factor greater than three are indicated by arrows.
-Only $\log_{10}$ Bayes Factors greater than zero are plotted.
-(F) $-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{ANGPTL7} for all three phenotypes.}
-%   \label{glaucoma_combined}
-% \end{figure}
-
+%% FIGURE: ASTHMA MANHATTAN AND FOREST
 \begin{figure}[!h]
-% \includegraphics[width=\textwidth]{FIG1FINAL.pdf}
-\caption{{\bf Results for glaucoma application.}
-$\log_{10}$ Bayes Factors from applying MRP to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
-(E) shows the results of a joint analysis focused on finding rare variants that protect against glaucoma.
-The genes outside of chromosome 6 with with Bayes Factor greater than three are indicated by arrows.
-Only $\log_{10}$ Bayes Factors greater than zero are plotted.
-F: $-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{ANGPTL7} for all three phenotypes.}
+\caption{{\bf Results for asthma application.}
+$\log_{10}$ Bayes Factors from applying MRP to summary statistics for missense and protein-truncating variants from (A) asthma (HC382), (B) eosinophil counts (INI30150), (C) forced vital capacity (FVC, INI3062), (D) forced expiratory volume in 1-second (FEV1, INI3063), (E) all four traits jointly, and (F) all four traits jointly with focus on protective effects.
+The four genes outside of chromosome 6 with the largest Bayes Factors are labeled in each plot.
+Only $\log_{10}$ Bayes Factors greater than -5 are plotted.
+(F) $-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{IL33} for each phenotype}
 \label{fig3}
 \end{figure}
-
-% %% TABLE: GLAUCOMA GENES OF INTEREST
-\begin{table}[!ht]
-\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
-\centering
-\caption{
-{\bf $\log_{10}$ Bayes Factors for genes highlighted in Figure \ref{fig3}.}}
-\begin{tabular}{|l+p{3cm}|p{1cm}|p{1.5cm}|p{3.8cm}|p{3.5cm}|}
-\hline
-{\bf Gene} & {\bf Joint, protective} & {\bf Joint} & {\bf Glaucoma} & {\bf Intra-ocular pressure, Goldmann-correlated} & {\bf Corneal resistance factor} \\
-\thickhline
-\textit{ANGPTL7} &              10.1 &   9.4 &      1.7 &                                       10.1 &                       6.2 \\
-\hline
-\textit{KLHL22}  &               3.1 &   3.6 &     -0.2 &                                        3.9 &                       2.2 \\
-\hline
-\textit{GLI3}    &               2.9 &   2.9 &     -0.3 &                                       -0.7 &                       3.2 \\
-\hline
-\textit{WNT10A}  &               2.2 &   3.0 &     -0.3 &                                       -0.5 &                       3.5 \\
-\hline
-\textit{RNASE2}  &              -1.3 &  -0.3 &      0.4 &                                       -1.0 &                      -1.1 \\
-\hline
-\textit{FUK}     &              -6.5 &  -2.3 &      2.5 &                                       -2.1 &                      -2.1 \\
-\hline
-\end{tabular}
-\label{table1}
-\end{adjustwidth}
-\end{table}
 
@@
 \caption{
-{\bf $\log_{10}$ Bayes Factors for genes highlighted in Figure \ref{fig4}.}}
+{\bf $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{fig3}.}}
 \begin{tabular}{|l+l|l|l|l|l|l|}
@@
 \textit{CHI3L1}       &              -4.4 &  -2.0 &             -2.0 &  -1.9 &  -2.0 &    3.0 \\
+\hline
+\end{tabular}
+\label{table1}
+\end{adjustwidth}
+\end{table}
+
+\subsubsection*{Model comparison applied to glaucoma and related phenotypes}
+We also applied MRP to missense variants and PTVs for glaucoma, intra-ocular pressure, and corneal resistance factor as well as performing a joint analysis and a joint analysis focused on protective associations.
+While the individual glaucoma analysis did not yield any associations with $\log_{10}$ BF greater than 5, the joint analyses identified missense and PTVs in both \textit{ANGPTL7} and \textit{WNT10A} as associated with protection against glaucoma.
+These results illustrate the ability of applying model comparison to multiple traits simultaneously to identify protective modifiers of disease.
+
+\begin{figure}[!h]
+\caption{{\bf Results for glaucoma application.}
+$\log_{10}$ Bayes Factors from applying MRP to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
+(E) shows the results of a joint analysis focused on finding rare variants that protect against glaucoma.
+The genes outside of chromosome 6 with with Bayes Factor greater than three are indicated by arrows.
+Only $\log_{10}$ Bayes Factors greater than zero are plotted.
+F: $-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{ANGPTL7} for all three phenotypes.}
+\label{fig4}
+\end{figure}
+
+% %% TABLE: GLAUCOMA GENES OF INTEREST
+\begin{table}[!ht]
+\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+\centering
+\caption{
+{\bf $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{fig4}.}}
+\begin{tabular}{|l+p{3cm}|p{1cm}|p{1.5cm}|p{3.8cm}|p{3.5cm}|}
+\hline
+{\bf Gene} & {\bf Joint, protective} & {\bf Joint} & {\bf Glaucoma} & {\bf Intra-ocular pressure, Goldmann-correlated} & {\bf Corneal resistance factor} \\
+\thickhline
+\textit{ANGPTL7} &              10.1 &   9.4 &      1.7 &                                       10.1 &                       6.2 \\
+\hline
+\textit{KLHL22}  &               3.1 &   3.6 &     -0.2 &                                        3.9 &                       2.2 \\
+\hline
+\textit{GLI3}    &               2.9 &   2.9 &     -0.3 &                                       -0.7 &                       3.2 \\
+\hline
+\textit{WNT10A}  &               2.2 &   3.0 &     -0.3 &                                       -0.5 &                       3.5 \\
+\hline
+\textit{RNASE2}  &              -1.3 &  -0.3 &      0.4 &                                       -1.0 &                      -1.1 \\
+\hline
+\textit{FUK}     &              -6.5 &  -2.3 &      2.5 &                                       -2.1 &                      -2.1 \\
 \hline
@@
 We also applied the MRP model comparison framework to summary statistic data from four traits from the UK Biobank: asthma, eosinophil counts, forced expiratory volume in 1-second (FEV1), and forced vital capacity (FVC).
-We identified strong evidence for the previously described association between \textit{IL33} and asthma\cite{DeBoever179762, 10.1371/journal.pgen.1006659} as well as evidence of association for several additional genes that have not been previously reported (Table~\ref{asthma_table}).
+We identified strong evidence for the previously described association between \textit{IL33} and asthma\cite{DeBoever179762, 10.1371/journal.pgen.1006659} as well as evidence of association for several additional genes that have not been previously reported (Table~\ref{table2}).
 

### 2018-01-26 tex/manuscript.tex
Text updates

@@
 In this paper, we present a Bayesian multiple rare variants and phenotypes (MRP) model comparison approach for identifying rare variant associations as an alternative to current widely-used statistical tests.
-The MRP framework exploits correlation, scale, or location (direction) of genetic effects in a broad range of rare variant association study designs including: case-control; multiple diseases and shared controls; single continuous phenotype; multiple continuous phenotypes; or a mixture of case-control and multiple continuous phenotypes (Fig~\ref{fig1}).
+The MRP framework exploits correlation, scale, or location (direction) of genetic effects in a broad range of rare variant association study designs including: case-control; multiple diseases and shared controls; single continuous phenotype; multiple continuous phenotypes; or a mixture of case-control and multiple continuous phenotypes (Fig~\ref{overview}).
 MRP makes use of Bayesian model comparison, whereby we compute a Bayes Factor (BF) defined as the ratio of the marginal likelihoods of the observed data under two models: 1) a pre-specified null where all genetic effects are zero; and 2) an alternative model where factors like correlation, scale, or location of genetic effects are considered.
@@
 B: Diagram of factors considered in rare variant association analysis including the correlation matrices: $\mathbf{R_{\textrm{study}}}$  (expected correlation of genetic effects among a group of studies), $\mathbf{R_{\textrm{var}}}$ (expected correlation of genetic effects among a group of variants), and $\mathbf{R_{\textrm{phen}}}$ (expected correlation of genetic effects among a group of phenotypes); the scale parameter for genetic variant annotation; and the location of genetic effects, which may be used to prioritize or identify protective modifiers of disease risk.}
-\label{fig1}
+\label{overview}
 \end{figure}
@@
 We also apply MRP to summary statistics for two groups of related phenotypes from the UK Biobank.
-First, we apply MRP to asthma (HC382: the corresponding phenotype label in Global Biobank Engine [\url{https://biobankengine.stanford.edu}]), eosinophil count (INI30150), Forced Expiratory Volume in 1-second (FEV1, INI3063), and Forced Vital Capacity (FVC, INI3062).
+First, we apply MRP to asthma (HC382: the corresponding phenotype label in Global Biobank Engine [\url{https://biobankengine.stanford.edu}]), eosinophil count (INI30150), Forced Expiratory Volume in 1-second (FEV$_1$, INI3063), and Forced Vital Capacity (FVC, INI3062).
 We find that XXX.
@@
 \subsection*{HDF5 Tables}
-Although summary statistics are quicker to read and process than raw data, the number of studies meta-analyzed in this work is expected to be sufficiently large to require optimizations in data representation and processing (Fig~\ref{S1_Fig}).
+Although summary statistics are quicker to read and process than raw data, the number of studies meta-analyzed in this work is expected to be sufficiently large to require optimizations in data representation and processing (\nameref{hdf5_fig}).
 Our solution was the use of the HDF5 (Hierarchical Data Format 5) data representation to enable rapid processing of effect size, uncertainty, and cross-trait estimate data.
@@
 Asthma cases were defined using both Hospital Episode Statistics and verbal questionnaire responses.
-We used the provided values from the UK Biobank for eosinophil counts, forced vital capacity (FVC), and forced expiratory volume in 1-second (FEV1).
-The phenotype codes used throughout (asthma=HC382, eosinophil count=INI30150, Forced Expiratory Volume in 1-second (FEV1)=INI3063, and Forced Vital Capacity (FVC)=INI3062) correspond to the phenotype codes used in on the Global Biobank Engine [\url{https://biobankengine.stanford.edu}].
+We used the provided values from the UK Biobank for eosinophil counts, forced vital capacity (FVC), and forced expiratory volume in 1-second (FEV$_1$).
+The phenotype codes used throughout (asthma=HC382, eosinophil count=INI30150, Forced Expiratory Volume in 1-second (FEV$_1$)=INI3063, and Forced Vital Capacity (FVC)=INI3062) correspond to the phenotype codes used in on the Global Biobank Engine [\url{https://biobankengine.stanford.edu}].
 
 \subsubsection*{Genetic Correlations}
-We calculated the genetic between asthma, eosinophil counts, forced vital capacity (FVC), and forced expiratory volume in 1-second (FEV1) using the MultiVariate Polygenic Mixture Model (MVPMM)~\cite{DeBoever2017}.
+We calculated the genetic between asthma, eosinophil counts, forced vital capacity (FVC), and forced expiratory volume in 1-second (FEV$_1$) using the MultiVariate Polygenic Mixture Model (MVPMM)~\cite{DeBoever2017}.
 Briefly, MVPMM estimates genetic correlation given GWAS summary statistics (effect size and standard error of effect size estimate) by modeling GWAS summary statistics as generated from one of two mixture components.
@@
 
-Although we see advantages in adopting a Bayesian perspective, our approach could be used in a frequentist context by calculating a Bayes factor and using it as a test statistic to compute p-values (Methods, Fig~\ref{fig2}).
+Although we see advantages in adopting a Bayesian perspective, our approach could be used in a frequentist context by calculating a Bayes factor and using it as a test statistic to compute p-values (Methods, Fig~\ref{simresults}).
 
@@
 D: ROC curves for detecting gene association when incorporating prior mean of genetic effects (orange) to identify protective alleles.}
-\label{fig2}
+\label{simresults}
 \end{figure}
@@
 To compare the Bayesian models we compute p-values by using the Bayes Factor as the test statistic and approximating it using distribution properties of quadratic forms (Materials and Methods).
-As expected, an independent effects model has high correlation with the gene-based test SKAT ($r^2 = 0.99$), whereas the similar effects model has high correlation with the burden test ($r^2 = 0.93$, Fig~\ref{fig2}A).
+As expected, an independent effects model has high correlation with the gene-based test SKAT ($r^2 = 0.99$), whereas the similar effects model has high correlation with the burden test ($r^2 = 0.93$, Fig~\ref{simresults}A).
 
@@
 We conducted $1000$ simulation experiments where we let $K$ (the number of phenotypes) $=3$, $M$ (the number of variants) $=10$, $S$ (the number of studies) $=2$, $N_0$ (number of individuals in study with access to all the data) $= 10000$, $N_1$ (meta-analysis study 1) $= 5000$, $N_2$ (meta-analysis study 2) $= 5000$.
-We find that, under the scenario where similar effects are assumed across studies, the Bayes Factors obtained using summary statistics alone are strongly correlated ($r^2 = 1$) to Bayes Factors obtained by the full genotype and phenotype data (Fig~\ref{fig2}B).
+We find that, under the scenario where similar effects are assumed across studies, the Bayes Factors obtained using summary statistics alone are strongly correlated ($r^2 = 1$) to Bayes Factors obtained by the full genotype and phenotype data (Fig~\ref{simresults}B).
 
@@
 We simulated three studies and applied the model comparison unit jointly to summary statistic data obtained for each study (Supplementary Note).
-Overall, we observed that considering the joint effects across multiple studies in a group of variants and phenotypes may improve ability to detect gene-based signals (Fig~\ref{fig2}C), and that considering prior mean of genetic effects should aid in efforts to identify protective modifiers of disease risk (Fig~\ref{fig2}D).
+Overall, we observed that considering the joint effects across multiple studies in a group of variants and phenotypes may improve ability to detect gene-based signals (Fig~\ref{simresults}C), and that considering prior mean of genetic effects should aid in efforts to identify protective modifiers of disease risk (Fig~\ref{simresults}D).
 
 \subsection*{Applications}
-We also applied the MRP model comparison approach to summary statistic data generated from single variant logistic regression and linear regression analysis for coding variants on the UK Biobank array (Methods).
-We applied the MRP model comparison approach to asthma and three related traits and glaucoma and two related traits.
+We applied the MRP model comparison approach to summary statistic data generated from single variant logistic regression and linear regression analysis for coding variants on the UK Biobank array (Methods).
+We applied MRP to asthma and three related traits and glaucoma and two related traits.
 
@@
 FEV$_1$ and FVC are measures of pulmonary function that are used to diagnosis and classify pulmonary disease \cite{Swanney1046}.
-To demonstrate the advantage of considering the phenotypes jointly, we applied the model to missense variants and PTVs for each phenotype separately (Fig~\ref{asthma_combined}A-D) as well as to all phenotypes jointly (Fig~\ref{fig3}E) and obtained $\log_{10}$ BF for each gene.
-In agreement with previous studies, we observed evidence that missense variants and PTVs in \textit{IL33} affect eosinophil counts and offer protection from asthma from the single-phenotype analyses, though the evidence of association was strongest for the joint analysis (Table~\ref{asthma_table}) \cite{DeBoever179762, 10.1371/journal.pgen.1006659}.
-We also performed an analysis focused on identifying protective variants which also identified the \textit{IL33} association (Fig~\ref{fig3}).
+To demonstrate the advantage of considering the phenotypes jointly, we applied MRP to rare missense variants and PTVs (MAF $< 1\%$) for each phenotype separately (Fig~\ref{asthma_bma}A-D) as well as to all phenotypes jointly (Fig~\ref{asthma_bma}E,F) and obtained $\log_{10}$ BF for each gene.
+We applied both independent and similar effects models and used Bayesian model averaging to compute a single BF per gene \cite{10.2307/2676803}.
+In agreement with previous studies, we observed evidence that rare missense variants and/or PTVs in \textit{IL33} affect eosinophil counts and offer protection from asthma from the single-phenotype analyses, though the evidence of association was strongest for the joint analysis (Table~\ref{asthma_table}) \cite{DeBoever179762, 10.1371/journal.pgen.1006659}.
+We also performed an analysis focused on identifying protective variants which also identified the \textit{IL33} association (Fig~\ref{asthma_bma}).
+The results were similar using only either the independent effects (\nameref{asthma_independent}) or similar effects models (\nameref{asthma_similar}).
+We can inspect the effect sizes from the marginal GWAS regressions for the rare variants included in the analysis and see that the association identified by MRP is likely driven by the PTV rs146597587 (Fig~\ref{asthma_bma}G).
 
-We also found evidence for associations between missense variants and PTVs in several other genes and the four phenotypes in the joint model that were not identified when considering asthma alone.
-For instance, \textit{GPR20}, \textit{CSF2RB}, \textit{PACS2}, and \textit{CCR3} all had $\log_{10}$ BF greater than 9 in the eosinophil count-only analysis but $\log_{10}$ BF less than 0 for the asthma-only analysis.
-In the joint model, the $\log_{10}$ BFs for these genes were all greater than six indicating possible associations with these traits.
-Besides \textit{IL33}, none of the genes with $\log_{10}$ BF greater than six in the joint model were reported in a large GWAS for allergic disease including asthma\cite{Ferreira:2017ba} though \textit{CCR3} is near a locus associated with atopy in a previous meta-analysis\cite{Ober:2011jk}.
+We also found moderate evidence for associations between rare coding variants in \textit{CCR3} and asthma.
+The $\log_{10}$ BFs for \textit{CCR3} were 3.3 and 3.1 in the joint and joint protective analyses respectively compared to only -0.5 in the asthma-only analysis  (Fig~\ref{asthma_bma}, Table~\ref{asthma_table}).
+\textit{CCR3} was not reported in a large GWAS for allergic disease including asthma\cite{Ferreira:2017ba} though \textit{CCR3} is near a locus associated with atopy in a previous meta-analysis\cite{Ober:2011jk}.
+Other genes like \textit{IL18RAP}, \textit{ATP2A3}, and \textit{FLG} had $\log_{10}$ BFs greater than 4 in the asthma-only analysis but much smaller BFs in the joint analyses indicating that rare variants in these genes are less likely to affect this group of traits.
+Similarly, there were other genes like \textit{RP11-39K24.9} and \textit{IL17RA} that had relatively larger BFs in the eosinophil count-only analysis but small BFs for the joint analyses demonstrating MRP's ability to integrate information across all phenotypes considered.
 
-%% FIGURE: ASTHMA MANHATTAN AND FOREST
+%% FIGURE: ASTHMA MANHATTAN AND FOREST FOR BMA
 \begin{figure}[!h]
 \caption{{\bf Results for asthma application.}
-$\log_{10}$ Bayes Factors from applying MRP to summary statistics for missense and protein-truncating variants from (A) asthma (HC382), (B) eosinophil counts (INI30150), (C) forced vital capacity (FVC, INI3062), (D) forced expiratory volume in 1-second (FEV1, INI3063), (E) all four traits jointly, and (F) all four traits jointly with focus on protective effects.
-The four genes outside of chromosome 6 with the largest Bayes Factors are labeled in each plot.
+$\log_{10}$ Bayes Factors from applying MRP and Bayesian model averaging to summary statistics for missense and protein-truncating variants from (A) asthma (HC382), (B) eosinophil counts (INI30150), (C) forced vital capacity (FVC, INI3062), (D) forced expiratory volume in 1-second (FEV$_1$, INI3063), (E) all four traits jointly, and (F) all four traits jointly with focus on protective effects.
+The four genes outside of chromosome 6 with the largest Bayes Factors greater than three are labeled in each plot.
 Only $\log_{10}$ Bayes Factors greater than -5 are plotted.
 (F) $-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{IL33} for each phenotype}
-\label{fig3}
+\label{asthma_bma}
 \end{figure}
@@
 \caption{
-{\bf $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{fig3}.}}
+{\bf $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{asthma_bma}.}}
 \begin{tabular}{|l+l|l|l|l|l|l|}
 \hline
-{\bf Gene} & {\bf Joint, protective} & {\bf Joint} & {\bf Eosinophil count} & {\bf FVC} & {\bf FEV1} & {\bf Asthma} \\
+{\bf Gene} & {\bf Joint, protective} & {\bf Joint} & {\bf Eosinophil count} & {\bf FVC} & {\bf FEV$_1$} & {\bf Asthma} \\
 \thickhline
-\textit{IL33}         &              30.1 &  29.8 &             30.9 &  -3.1 &  -2.7 &    9.5 \\
+\textit{IL33}         &              29.4 &   29.3 &             30.6 &  -2.3 &  -2.2 &    8.1 \\
 \hline
-\textit{CCR3}         &               5.8 &   6.3 &              9.1 &  -1.7 &  -2.3 &   -0.5 \\
+\textit{CCR3}         &               3.1 &    3.3 &              7.4 &  -1.4 &  -1.6 &   -0.5 \\
 \hline
-\textit{IL17RA}       &               4.6 &   6.1 &             14.1 &  -4.1 &  -3.9 &   -1.4 \\
+\textit{RP11-39K24.9} &               0.8 &    1.8 &              4.9 &  -0.1 &  -0.4 &    0.3 \\
 \hline
-\textit{AL161450.1}   &               3.2 &   3.8 &              4.9 &  -0.1 &  -0.4 &    0.3 \\
+\textit{SCMH1}        &               0.5 &    4.7 &             -1.5 &   7.7 &   3.8 &   -0.7 \\
 \hline
-\textit{RP11-39K24.9} &               3.2 &   3.8 &              4.9 &  -0.1 &  -0.4 &    0.3 \\
+\textit{MUSTN1}       &               0.4 &    1.1 &             -1.2 &   2.9 &   2.9 &   -0.6 \\
 \hline
-\textit{ELN}          &               2.9 &   3.2 &             -1.0 &   5.6 &   2.9 &   -0.6 \\
+\textit{ZFAT}         &               0.3 &    1.3 &             -2.0 &   4.7 &   3.1 &   -0.4 \\
 \hline
-\textit{SCMH1}        &               2.7 &   3.9 &             -2.1 &   7.5 &   3.3 &   -0.7 \\
+\textit{ELN}          &               0.2 &    2.5 &             -1.0 &   5.6 &   2.9 &   -0.6 \\
 \hline
-\textit{TGFB2}        &               2.6 &   4.1 &             -2.8 &   1.7 &  -2.4 &    0.1 \\
+\textit{C14orf39}     &              -0.7 &   -0.0 &             -1.1 &   3.5 &   2.5 &    0.0 \\
 \hline
-\textit{ZFAT}         &               1.3 &   1.9 &             -2.8 &   6.7 &   4.6 &   -0.4 \\
+\textit{TMEM110}      &              -0.9 &    1.1 &             -1.0 &   3.3 &   3.1 &   -0.6 \\
 \hline
-\textit{STC2}         &               0.2 &   1.3 &             -0.8 &   3.9 &   2.3 &   -0.7 \\
+\textit{IL17RA}       &              -4.4 &    3.1 &              7.9 &  -2.7 &  -2.5 &   -1.1 \\
 \hline
-\textit{P4HTM}        &              -1.3 &  -0.7 &             -1.6 &   2.9 &   4.3 &   -1.0 \\
+\textit{IL18RAP}      &              -9.6 &   -0.9 &             -1.0 &  -1.6 &  -1.7 &    3.5 \\
 \hline
-\textit{IL18RAP}      &              -2.2 &  -1.0 &             -0.8 &  -2.0 &  -2.2 &    3.6 \\
+\textit{ATP2A3}       &             -11.9 &   -0.8 &             -1.2 &  -2.1 &  -2.2 &    5.8 \\
 \hline
-\textit{WDR6}         &              -4.1 &  -2.9 &             -2.3 &   1.9 &   3.4 &   -0.9 \\
-\hline
-\textit{ATP2A3}       &              -4.2 &  -1.5 &             -1.8 &  -3.1 &  -3.1 &    5.9 \\
-\hline
-\textit{CHI3L1}       &              -4.4 &  -2.0 &             -2.0 &  -1.9 &  -2.0 &    3.0 \\
+\textit{FLG}          &             -20.1 &  -17.2 &             -6.0 &  -7.4 &  -8.5 &    4.0 \\
 \hline
 \end{tabular}
-\label{table1}
+\label{asthma_table}
 \end{adjustwidth}
@@
 \subsubsection*{Model comparison applied to glaucoma and related phenotypes}
-We also applied MRP to missense variants and PTVs for glaucoma, intra-ocular pressure, and corneal resistance factor as well as performing a joint analysis and a joint analysis focused on protective associations.
-While the individual glaucoma analysis did not yield any associations with $\log_{10}$ BF greater than 5, the joint analyses identified missense and PTVs in both \textit{ANGPTL7} and \textit{WNT10A} as associated with protection against glaucoma.
+We also applied MRP to missense variants and PTVs for glaucoma, intra-ocular pressure, and corneal resistance factor as well as performing joint analyses.
+Elevated intra-ocular pressure is a measure of the fluid pressure in the eye.
+Elevated intra-ocular pressure is associated with glaucoma risk and has been linked to genetic variants associated with glaucoma \cite{doi:10.1001/jama.2014.3192}.
+Corneal resistance factor is a measure of the cornea's ability to resist mechanical stress and has been associated with glaucoma presence and severity \cite{CXO:CXO414, MANSOURI2012419, GriseDulac:2012bo}.
+While the individual glaucoma analysis did not yield any associations with $\log_{10}$ BF greater than three, the joint analysis identified rare coding variants in both \textit{ANGPTL7}, \textit{WNT10A}, and \textit{KLHL22} as associated with glaucoma (Fig~\ref{glaucoma_bma}A-E).
+Applying the MRP model to identify protective effects also identified \textit{ANGPTL7} and \textit{WNT10A} but not \textit{KLHL22}
 These results illustrate the ability of applying model comparison to multiple traits simultaneously to identify protective modifiers of disease.
@@
 \caption{{\bf Results for glaucoma application.}
-$\log_{10}$ Bayes Factors from applying MRP to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
+$\log_{10}$ Bayes Factors from applying MRP and Bayesian model averaging to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
 (E) shows the results of a joint analysis focused on finding rare variants that protect against glaucoma.
@@
 F: $-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{ANGPTL7} for all three phenotypes.}
-\label{fig4}
+\label{glaucoma_bma}
 \end{figure}
@@
 \caption{
-{\bf $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{fig4}.}}
-\begin{tabular}{|l+p{3cm}|p{1cm}|p{1.5cm}|p{3.8cm}|p{3.5cm}|}
+{\bf $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{glaucoma_bma}.}}
+\begin{tabular}{|l+p{3cm}|p{1cm}|p{1.6cm}|p{3.8cm}|p{3.4cm}|}
 \hline
@@
 We used simulations to compare MRP to the widely used burden and SKAT tests for identifying rare variant associations and found that jointly considering both variants and phenotypes can improve the ability to detect associations.
-We also applied the MRP model comparison framework to summary statistic data from four traits from the UK Biobank: asthma, eosinophil counts, forced expiratory volume in 1-second (FEV1), and forced vital capacity (FVC).
+We also applied the MRP model comparison framework to summary statistic data from four traits from the UK Biobank: asthma, eosinophil counts, forced expiratory volume in 1-second (FEV$_1$), and forced vital capacity (FVC).
 We identified strong evidence for the previously described association between \textit{IL33} and asthma\cite{DeBoever179762, 10.1371/journal.pgen.1006659} as well as evidence of association for several additional genes that have not been previously reported (Table~\ref{table2}).
@@
 \paragraph*{S1 Fig.}
-\label{S1_Fig}
+\label{hdf5_fig}
 {\bf HDF5 Implementation.} Our HDF5 implementation contained the following components: first, a group with one table per annotation file.
@@
 The index table encodes row/column position of each dataset.
+
+%% FIGURE: ASTHMA MANHATTAN AND FOREST FOR INDEPENDENT EFFECTS
+\paragraph*{S2 Fig.}
+\label{asthma_independent}
+{\bf Results for independent effects model applied to asthma, eosinophil counts, FEV$_1$, and FVC.} $\log_{10}$ Bayes Factors from applying MRP independent effects model to summary statistics for missense and protein-truncating variants from (A) asthma (HC382), (B) eosinophil counts (INI30150), (C) forced vital capacity (FVC, INI3062), (D) forced expiratory volume in 1-second (FEV$_1$, INI3063), (E) all four traits jointly, and (F) all four traits jointly with focus on protective effects.
+The four genes outside of chromosome 6 with the largest Bayes Factors greater than three are labeled in each plot.
+Only $\log_{10}$ Bayes Factors greater than -5 are plotted.
+
+%% FIGURE: ASTHMA MANHATTAN AND FOREST FOR INDEPENDENT EFFECTS
+\paragraph*{S3 Fig.}
+\label{asthma_similar}
+{\bf Results for independent effects model applied to asthma, eosinophil counts, FEV$_1$, and FVC.} $\log_{10}$ Bayes Factors from applying MRP similar effects model to summary statistics for missense and protein-truncating variants from (A) asthma (HC382), (B) eosinophil counts (INI30150), (C) forced vital capacity (FVC, INI3062), (D) forced expiratory volume in 1-second (FEV$_1$, INI3063), (E) all four traits jointly, and (F) all four traits jointly with focus on protective effects.
+The four genes outside of chromosome 6 with the largest Bayes Factors greater than three are labeled in each plot.
+Only $\log_{10}$ Bayes Factors greater than -5 are plotted.
 

### 2018-01-26 tex/manuscript.tex

@@
 We also apply MRP to summary statistics for two groups of related phenotypes from the UK Biobank.
-First, we apply MRP to asthma (HC382: the corresponding phenotype label in Global Biobank Engine [\url{https://biobankengine.stanford.edu}]), eosinophil count (INI30150), Forced Expiratory Volume in 1-second (FEV$_1$, INI3063), and Forced Vital Capacity (FVC, INI3062).
+First, we apply MRP to asthma (HC382: the corresponding phenotype label in Global Biobank Engine [\url{https://biobankengine.stanford.edu}]), eosinophil count (INI30150), Forced Expiratory Volume in 1-second (FEV1, INI3063), and Forced Vital Capacity (FVC, INI3062).
 We find that XXX.
@@
 Asthma cases were defined using both Hospital Episode Statistics and verbal questionnaire responses.
-We used the provided values from the UK Biobank for eosinophil counts, forced vital capacity (FVC), and forced expiratory volume in 1-second (FEV$_1$).
-The phenotype codes used throughout (asthma=HC382, eosinophil count=INI30150, Forced Expiratory Volume in 1-second (FEV$_1$)=INI3063, and Forced Vital Capacity (FVC)=INI3062) correspond to the phenotype codes used in on the Global Biobank Engine [\url{https://biobankengine.stanford.edu}].
+We used the provided values from the UK Biobank for eosinophil counts, forced vital capacity (FVC), and forced expiratory volume in 1-second (FEV1).
+The phenotype codes used throughout (asthma=HC382, eosinophil count=INI30150, Forced Expiratory Volume in 1-second (FEV1)=INI3063, and Forced Vital Capacity (FVC)=INI3062) correspond to the phenotype codes used in on the Global Biobank Engine [\url{https://biobankengine.stanford.edu}].
 
 \subsubsection*{Genetic Correlations}
-We calculated the genetic between asthma, eosinophil counts, forced vital capacity (FVC), and forced expiratory volume in 1-second (FEV$_1$) using the MultiVariate Polygenic Mixture Model (MVPMM)~\cite{DeBoever2017}.
+We calculated the genetic between asthma, eosinophil counts, forced vital capacity (FVC), and forced expiratory volume in 1-second (FEV1) using the MultiVariate Polygenic Mixture Model (MVPMM)~\cite{DeBoever2017}.
 Briefly, MVPMM estimates genetic correlation given GWAS summary statistics (effect size and standard error of effect size estimate) by modeling GWAS summary statistics as generated from one of two mixture components.
@@
 We applied the MRP model comparison approach to summary statistic data generated from single variant logistic regression and linear regression analysis for coding variants on the UK Biobank array (Methods).
-We applied MRP to asthma and three related traits and glaucoma and two related traits.
+We applied the MRP model comparison approach to asthma and three related traits and glaucoma and two related traits.
 
@@
 FEV$_1$ and FVC are measures of pulmonary function that are used to diagnosis and classify pulmonary disease \cite{Swanney1046}.
-To demonstrate the advantage of considering the phenotypes jointly, we applied MRP to rare missense variants and PTVs (MAF $< 1\%$) for each phenotype separately (Fig~\ref{asthma_bma}A-D) as well as to all phenotypes jointly (Fig~\ref{asthma_bma}E,F) and obtained $\log_{10}$ BF for each gene.
-We applied both independent and similar effects models and used Bayesian model averaging to compute a single BF per gene \cite{10.2307/2676803}.
-In agreement with previous studies, we observed evidence that rare missense variants and/or PTVs in \textit{IL33} affect eosinophil counts and offer protection from asthma from the single-phenotype analyses, though the evidence of association was strongest for the joint analysis (Table~\ref{asthma_table}) \cite{DeBoever179762, 10.1371/journal.pgen.1006659}.
+To demonstrate the advantage of considering the phenotypes jointly, we applied MRP to missense variants and PTVs for each phenotype separately (Fig~\ref{asthma_bma}A-D) as well as to all phenotypes jointly (Fig~\ref{asthma_bma}E,F) and obtained $\log_{10}$ BF for each gene.
+We applied both independent and similar effects models and used Bayesian model averaging to take both models into account.
+In agreement with previous studies, we observed evidence that missense variants and PTVs in \textit{IL33} affect eosinophil counts and offer protection from asthma from the single-phenotype analyses, though the evidence of association was strongest for the joint analysis (Table~\ref{asthma_table}) \cite{DeBoever179762, 10.1371/journal.pgen.1006659}.
 We also performed an analysis focused on identifying protective variants which also identified the \textit{IL33} association (Fig~\ref{asthma_bma}).
 The results were similar using only either the independent effects (\nameref{asthma_independent}) or similar effects models (\nameref{asthma_similar}).
-We can inspect the effect sizes from the marginal GWAS regressions for the rare variants included in the analysis and see that the association identified by MRP is likely driven by the PTV rs146597587 (Fig~\ref{asthma_bma}G).
 
-We also found moderate evidence for associations between rare coding variants in \textit{CCR3} and asthma.
+We also found moderate evidence for associations between rare missense variants and PTVs in \textit{CCR3} and asthma.
 The $\log_{10}$ BFs for \textit{CCR3} were 3.3 and 3.1 in the joint and joint protective analyses respectively compared to only -0.5 in the asthma-only analysis  (Fig~\ref{asthma_bma}, Table~\ref{asthma_table}).
@@
 \caption{{\bf Results for asthma application.}
-$\log_{10}$ Bayes Factors from applying MRP and Bayesian model averaging to summary statistics for missense and protein-truncating variants from (A) asthma (HC382), (B) eosinophil counts (INI30150), (C) forced vital capacity (FVC, INI3062), (D) forced expiratory volume in 1-second (FEV$_1$, INI3063), (E) all four traits jointly, and (F) all four traits jointly with focus on protective effects.
+$\log_{10}$ Bayes Factors from applying MRP and Bayesian model averaging to summary statistics for missense and protein-truncating variants from (A) asthma (HC382), (B) eosinophil counts (INI30150), (C) forced vital capacity (FVC, INI3062), (D) forced expiratory volume in 1-second (FEV1, INI3063), (E) all four traits jointly, and (F) all four traits jointly with focus on protective effects.
 The four genes outside of chromosome 6 with the largest Bayes Factors greater than three are labeled in each plot.
@@
 \hline
-{\bf Gene} & {\bf Joint, protective} & {\bf Joint} & {\bf Eosinophil count} & {\bf FVC} & {\bf FEV$_1$} & {\bf Asthma} \\
+{\bf Gene} & {\bf Joint, protective} & {\bf Joint} & {\bf Eosinophil count} & {\bf FVC} & {\bf FEV1} & {\bf Asthma} \\
 \thickhline
@@
 \subsubsection*{Model comparison applied to glaucoma and related phenotypes}
-We also applied MRP to missense variants and PTVs for glaucoma, intra-ocular pressure, and corneal resistance factor as well as performing joint analyses.
-Elevated intra-ocular pressure is a measure of the fluid pressure in the eye.
-Elevated intra-ocular pressure is associated with glaucoma risk and has been linked to genetic variants associated with glaucoma \cite{doi:10.1001/jama.2014.3192}.
-Corneal resistance factor is a measure of the cornea's ability to resist mechanical stress and has been associated with glaucoma presence and severity \cite{CXO:CXO414, MANSOURI2012419, GriseDulac:2012bo}.
-While the individual glaucoma analysis did not yield any associations with $\log_{10}$ BF greater than three, the joint analysis identified rare coding variants in both \textit{ANGPTL7}, \textit{WNT10A}, and \textit{KLHL22} as associated with glaucoma (Fig~\ref{glaucoma_bma}A-E).
-Applying the MRP model to identify protective effects also identified \textit{ANGPTL7} and \textit{WNT10A} but not \textit{KLHL22}
+We also applied MRP to missense variants and PTVs for glaucoma, intra-ocular pressure, and corneal resistance factor as well as performing a joint analysis and a joint analysis focused on protective associations.
+While the individual glaucoma analysis did not yield any associations with $\log_{10}$ BF greater than 5, the joint analyses identified missense and PTVs in both \textit{ANGPTL7} and \textit{WNT10A} as associated with protection against glaucoma.
 These results illustrate the ability of applying model comparison to multiple traits simultaneously to identify protective modifiers of disease.
@@
 We used simulations to compare MRP to the widely used burden and SKAT tests for identifying rare variant associations and found that jointly considering both variants and phenotypes can improve the ability to detect associations.
-We also applied the MRP model comparison framework to summary statistic data from four traits from the UK Biobank: asthma, eosinophil counts, forced expiratory volume in 1-second (FEV$_1$), and forced vital capacity (FVC).
+We also applied the MRP model comparison framework to summary statistic data from four traits from the UK Biobank: asthma, eosinophil counts, forced expiratory volume in 1-second (FEV1), and forced vital capacity (FVC).
 We identified strong evidence for the previously described association between \textit{IL33} and asthma\cite{DeBoever179762, 10.1371/journal.pgen.1006659} as well as evidence of association for several additional genes that have not been previously reported (Table~\ref{table2}).
@@
 \label{asthma_independent}
-{\bf Results for independent effects model applied to asthma, eosinophil counts, FEV$_1$, and FVC.} $\log_{10}$ Bayes Factors from applying MRP independent effects model to summary statistics for missense and protein-truncating variants from (A) asthma (HC382), (B) eosinophil counts (INI30150), (C) forced vital capacity (FVC, INI3062), (D) forced expiratory volume in 1-second (FEV$_1$, INI3063), (E) all four traits jointly, and (F) all four traits jointly with focus on protective effects.
+{\bf Results for independent effects model applied to asthma, eosinophil counts, FEV$_1$, and FVC.} $\log_{10}$ Bayes Factors from applying MRP independent effects model to summary statistics for missense and protein-truncating variants from (A) asthma (HC382), (B) eosinophil counts (INI30150), (C) forced vital capacity (FVC, INI3062), (D) forced expiratory volume in 1-second (FEV1, INI3063), (E) all four traits jointly, and (F) all four traits jointly with focus on protective effects.
 The four genes outside of chromosome 6 with the largest Bayes Factors greater than three are labeled in each plot.
@@
 \label{asthma_similar}
-{\bf Results for independent effects model applied to asthma, eosinophil counts, FEV$_1$, and FVC.} $\log_{10}$ Bayes Factors from applying MRP similar effects model to summary statistics for missense and protein-truncating variants from (A) asthma (HC382), (B) eosinophil counts (INI30150), (C) forced vital capacity (FVC, INI3062), (D) forced expiratory volume in 1-second (FEV$_1$, INI3063), (E) all four traits jointly, and (F) all four traits jointly with focus on protective effects.
+{\bf Results for independent effects model applied to asthma, eosinophil counts, FEV$_1$, and FVC.} $\log_{10}$ Bayes Factors from applying MRP similar effects model to summary statistics for missense and protein-truncating variants from (A) asthma (HC382), (B) eosinophil counts (INI30150), (C) forced vital capacity (FVC, INI3062), (D) forced expiratory volume in 1-second (FEV1, INI3063), (E) all four traits jointly, and (F) all four traits jointly with focus on protective effects.
 The four genes outside of chromosome 6 with the largest Bayes Factors greater than three are labeled in each plot.

### 2018-01-27 tex/manuscript.tex
Text updates

--- 2018-01-26 tex/manuscript.tex
@@
 Here we introduce a model comparison approach we refer to as MRP for rare variant association studies that considers correlation, scale, and location of genetic effects across a group of genetic variants, phenotypes, and studies.
-In so doing we consider the use of summary statistic data to apply 1) standard univariate and multivariate gene-based meta-analysis models, 2) models for assessing heterogeneity of genetic effects, which may be used in practice for downstream quality control, and 3) models for identifying protective protein-truncating variants, which can expedite drug discovery.
+We consider the use of summary statistic data to apply 1) standard univariate and multivariate gene-based meta-analysis models, 2) models for assessing heterogeneity of genetic effects, and 3) models for identifying protective protein-truncating variants that can expedite drug discovery.
 Through simulation studies, we demonstrate that the proposed model comparison approach may improve ability to detect rare variant association signals.
-Finally, we demonstrate its use with rare variant data combined with asthma diagnosis and haematological and spirometry measures for individuals in the UK Biobank.
-We show that we are able to retain useful features from widely-used meta-analysis approaches and prioritize protective modifiers of disease risk.
+Finally, we demonstrate the model on two groups of phenotypes from the UK Biobank: 1) asthma diagnosis, eosinophil counts, forced expiratory volume, and forced vital capacity; and 2) glaucoma diagnosis, intra-ocular pressure, and corneal resistance factor.
+We demonstrate that the model leverages information from multiple phenotypes to identify known and novel rare variant associations for asthma and glaucoma.
+Overall the model comparison approach is able to retain useful features from widely-used meta-analysis approaches and prioritize protective modifiers of disease risk.
 
@@
 This biological knowledge can be reflected in the choices of priors for PTVs in MRP.
-Since PTVs typically abolish gene function, there is particular interest in identifying protective PTV modifiers of human disease risk that may serve as targets for therapeutics\cite{pcsk9,cohen2006sequence,sullivan2012effect}.
+Since PTVs typically abolish or severely alter gene function, there is particular interest in identifying protective PTV modifiers of human disease risk that may serve as targets for therapeutics\cite{pcsk9,cohen2006sequence,sullivan2012effect}.
 We therefore demonstrate how the MRP model comparison approach can improve discovery of such protective signals by modeling the location (direction) of genetic effects which prioritizes variants or genes that are consistent with protecting against disease.
@@
 %%Apply methods to the UK Biobank
-We also apply MRP to summary statistics for two groups of related phenotypes from the UK Biobank.
-First, we apply MRP to asthma (HC382: the corresponding phenotype label in Global Biobank Engine [\url{https://biobankengine.stanford.edu}]), eosinophil count (INI30150), Forced Expiratory Volume in 1-second (FEV$_1$, INI3063), and Forced Vital Capacity (FVC, INI3062).
-We find that XXX.
-We next apply MRP to glaucoma (HC276), intra-ocular pressure (INI5263), and corneal resistance factor (INI5265) and find strong evidence that rare PTVs in \textit{ANGPTL7} protect against glaucoma.
+We applied MRP to summary statistics for two groups of related phenotypes from the UK Biobank.
+First, we apply MRP to asthma (HC382: the corresponding phenotype label in Global Biobank Engine [\url{https://biobankengine.stanford.edu}]), eosinophil count (INI30150), forced expiratory volume in 1-second (FEV$_1$, INI3063), and forced vital capacity (FVC, INI3062).
+We recover the reported association between a rare PTV in \textit{IL33} and asthma as well as a new association between \textit{CCR3} and asthma \cite{DeBoever179762, 10.1371/journal.pgen.1006659}.
+We also applied MRP to glaucoma (HC276), intra-ocular pressure (INI5263), and corneal resistance factor (INI5265) and find evidence that rare coding variants in \textit{ANGPTL7} and \textit{KLHL22} may protect against glaucoma.
 These analyses show that MRP recovers results from typical single variant-single phenotype analyses while identifying new rare variant associations that include protective modifiers of disease risk.
@@
 \subsubsection*{GWAS Summary Statistics}
-We performed genome-wide association analysis using PLINK v2.00a(17 July 2017) as previously described~\cite{DeBoever179762}.
-For asthma, we used the Firth-fallback in PLINK, a hybrid algorithm which normally uses the logistic regression code described in (Hill 2017), but switches to a port of logistf() (https://cran.r-project.org/web/packages/logistf/index.html) in two cases: (1) one of the cells in the 2x2 allele count by case/control status contingency table is empty (2) logistic regression was attempted since all the contingency table cells were nonzero, but it failed to converge within the usual number of steps.
+We performed genome-wide association analysis using PLINK v2.00a(17 July 2017) as previously described \cite{DeBoever179762}.
+For asthma, we used the Firth fallback in PLINK, a hybrid algorithm which normally uses the logistic regression code described in (Hill 2017), but switches to a port of logistf() (https://cran.r-project.org/web/packages/logistf/index.html) in two cases: (1) one of the cells in the 2x2 allele count by case/control status contingency table is empty (2) logistic regression was attempted since all the contingency table cells were nonzero, but it failed to converge within the usual number of steps.
 We used the following covariates in our analysis: age, sex, array type, and the first four principal components, where array type is a binary variable that represents whether an individual was genotyped with UK Biobank Axiom Array or UK BiLEVE Axiom Array.
@@
 While the individual glaucoma analysis did not yield any associations with $\log_{10}$ BF greater than three, the joint analysis identified rare coding variants in both \textit{ANGPTL7}, \textit{WNT10A}, and \textit{KLHL22} as associated with glaucoma (Fig~\ref{glaucoma_bma}A-E).
-Applying the MRP model to identify protective effects also identified \textit{ANGPTL7} and \textit{WNT10A} but not \textit{KLHL22}
-These results illustrate the ability of applying model comparison to multiple traits simultaneously to identify protective modifiers of disease.
+Applying the MRP model to identify protective effects also identified \textit{ANGPTL7} and \textit{WNT10A} but not \textit{KLHL22}.
+Expression of \textit{ANGPTL7} is upregulated in glaucome and has been proposed to regulate intra-ocular pressure and glaucoma risk \cite{Comes:2011ex, Kuchtey:2008ek}.
+\textit{KLHL22} has not been previously associated with glaucoma though a suggestive association was reported for retinopathy in individuals without diabetes \cite{10.1371/journal.pone.0054232}.
+These results illustrate that applying model comparison to multiple traits simultaneously can prioritize the discovery of protective modifiers of disease.
 
@@
 We used simulations to compare MRP to the widely used burden and SKAT tests for identifying rare variant associations and found that jointly considering both variants and phenotypes can improve the ability to detect associations.
-We also applied the MRP model comparison framework to summary statistic data from four traits from the UK Biobank: asthma, eosinophil counts, forced expiratory volume in 1-second (FEV$_1$), and forced vital capacity (FVC).
-We identified strong evidence for the previously described association between \textit{IL33} and asthma\cite{DeBoever179762, 10.1371/journal.pgen.1006659} as well as evidence of association for several additional genes that have not been previously reported (Table~\ref{table2}).
+We also applied the MRP model comparison framework to summary statistic data from two groups of traits from the UK Biobank: asthma diagnosis, eosinophil counts, FEV$_1$, and FVC; and glaucoma diagnosis, intra-ocular pressure, and corneal resistance factor.
+We identified strong evidence for the previously described association between the PTV rs146597587 in \textit{IL33} and asthma\cite{DeBoever179762,10.1371/journal.pgen.1006659} as well as evidence for a potentially novel association for \textit{CCR3} (Table~\ref{asthma_table}).
+We also identified genetic evidence for a link between rare variants in \textit{ANGPTL7} and glaucoma, consistent with previous experiments that suggested a role for \textit{ANGPTL7} in glaucoma \cite{Comes:2011ex, Kuchtey:2008ek}.
+These results demonstrate the ability of the MRP model comparison approach to leverage information across multiple phenotypes and variants to discover rare variant associations.
 
@@
 \label{asthma_similar}
-{\bf Results for independent effects model applied to asthma, eosinophil counts, FEV$_1$, and FVC.} $\log_{10}$ Bayes Factors from applying MRP similar effects model to summary statistics for missense and protein-truncating variants from (A) asthma (HC382), (B) eosinophil counts (INI30150), (C) forced vital capacity (FVC, INI3062), (D) forced expiratory volume in 1-second (FEV$_1$, INI3063), (E) all four traits jointly, and (F) all four traits jointly with focus on protective effects.
+{\bf Results for similar effects model applied to asthma, eosinophil counts, FEV$_1$, and FVC.} $\log_{10}$ Bayes Factors from applying MRP similar effects model to summary statistics for missense and protein-truncating variants from (A) asthma (HC382), (B) eosinophil counts (INI30150), (C) forced vital capacity (FVC, INI3062), (D) forced expiratory volume in 1-second (FEV$_1$, INI3063), (E) all four traits jointly, and (F) all four traits jointly with focus on protective effects.
 The four genes outside of chromosome 6 with the largest Bayes Factors greater than three are labeled in each plot.
 Only $\log_{10}$ Bayes Factors greater than -5 are plotted.
+
+\paragraph*{S4 Fig.}
+\label{glaucoma_similar}
+{\bf Results for similar effects model applied to glaucoma intra-ocular pressure, and corneal resistance factor.} $\log_{10}$ Bayes Factors from applying MRP similar effects model to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
+(E) shows the results of a joint analysis focused on finding rare variants that protect against glaucoma.
+The genes outside of chromosome 6 with with Bayes Factor greater than three are indicated by arrows.
+Only $\log_{10}$ Bayes Factors greater than zero are plotted.
+
+\paragraph*{S5 Fig.}
+\label{glaucoma_independent}
+{\bf Results for independent effects model applied to glaucoma intra-ocular pressure, and corneal resistance factor.} $\log_{10}$ Bayes Factors from applying MRP independent effects model to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
+(E) shows the results of a joint analysis focused on finding rare variants that protect against glaucoma.
+The genes outside of chromosome 6 with with Bayes Factor greater than three are indicated by arrows.
+Only $\log_{10}$ Bayes Factors greater than zero are plotted.
 
@@
 \section*{Author Contributions}
-M.A.R.
-and M.P.
-designed the method and derived all analytical calculations.
-M.A.R., M.P., and C.D.
-wrote the manuscript.
-Y.T., M.A., and C.D.
-provided analysis and designed figures.
-T.P.
-designed hDF5 tables and implementation of loaders.
-M.J.D.
-provided critical feedback on methodology.
-TODO: Add C.
-Spencer contributions.
+MAR and MP designed the method and derived all analytical calculations.
+MAR, MP, and CD wrote the manuscript.
+CCAS, YT, MA, and CD provided analysis and designed figures.
+TP designed HDF5 tables and implementation of loaders.
+MJD provided critical feedback on methodology.
 

### 2018-01-27 tex/manuscript.tex
Added author summary

@@
 \section*{Author summary}
-TODO
+Due to the continually decreasing cost of acquiring genetic data, we are now beginning to see large collections of individuals for which we have both genetic information and trait data such as disease status, physical measurements, biomarker levels, and more.
+These datasets offer new opportunities to find relationships between inherited genetic variation and disease.
+While it is known that there are relationships between different traits, typical genetic analyses only focus on analyzing one genetic variant and one phenotype at a time.
+Additionally, it is difficult to identify rare genetic variants that are associated with disease due to their scarcity, even among large sample sizes.
+In this work, we present a method for identifying associations between genetic variation and disease that considers multiple rare variants and phenotypes at the same time.
+By sharing information across rare variant and phenotypes, we improve our ability to identify rare variants associated with disease compared to considering a single rare variant and a single phenotype.
+The method can be used to identify candidate disease genes as well as genes that might represent attractive drug targets.
 

### 2018-01-27 tex/manuscript.tex
Glaucoma results update

@@
 Corneal resistance factor is a measure of the cornea's ability to resist mechanical stress and has been associated with glaucoma presence and severity \cite{CXO:CXO414, MANSOURI2012419, GriseDulac:2012bo}.
-While the individual glaucoma analysis did not yield any associations with $\log_{10}$ BF greater than three, the joint analysis identified rare coding variants in both \textit{ANGPTL7}, \textit{WNT10A}, and \textit{KLHL22} as associated with glaucoma (Fig~\ref{glaucoma_bma}A-E).
-Applying the MRP model to identify protective effects also identified \textit{ANGPTL7} and \textit{WNT10A} but not \textit{KLHL22}.
+While the individual glaucoma analysis did not yield any associations with $\log_{10}$ BF greater than three, the joint analysis identified rare coding variants in both \textit{ANGPTL7}, \textit{WNT10A}, and \textit{KLHL22} as associated with glaucoma (Fig~\ref{glaucoma_bma}A-D).
+Applying the MRP model to identify protective effects also identified \textit{ANGPTL7} and \textit{WNT10A} but not \textit{KLHL22} (Fig~\ref{glaucoma_bma}E).
 Expression of \textit{ANGPTL7} is upregulated in glaucome and has been proposed to regulate intra-ocular pressure and glaucoma risk \cite{Comes:2011ex, Kuchtey:2008ek}.
 \textit{KLHL22} has not been previously associated with glaucoma though a suggestive association was reported for retinopathy in individuals without diabetes \cite{10.1371/journal.pone.0054232}.
+The GWAS summary statistics for the rare variants in \textit{ANGPTL7} suggest that two missense variants, rs28991002 and rs28991009, appear to drive the association with glaucoma (Fig~\ref{glaucoma_bma}F).
 These results illustrate that applying model comparison to multiple traits simultaneously can prioritize the discovery of protective modifiers of disease.

### 2018-01-29 tex/appendix.tex
Updated methods and other small updates
I moved the "methods overview" into the methods section. I moved the
detailed methods into an appendix.

--- 2018-01-27 tex/manuscript.tex
@@
-% Template for PLoS
-% Version 3.4 January 2017
-%
-% % % % % % % % % % % % % % % % % % % % % %
-%
-% -- IMPORTANT NOTE
-%
-% This template contains comments intended 
-% to minimize problems and delays during our production 
-% process. Please follow the template instructions
-% whenever possible.
-%
-% % % % % % % % % % % % % % % % % % % % % % % 
-%
-% Once your paper is accepted for publication, 
-% PLEASE REMOVE ALL TRACKED CHANGES in this file 
-% and leave only the final text of your manuscript. 
-% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
-% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
-%
-%
-% There are no restrictions on package use within the LaTeX files except that 
-% no packages listed in the template may be deleted.
-%
-% Please do not include colors or graphics in the text.
-%
-% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
-%
-% % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- FIGURES AND TABLES
-%
-% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
-%
-% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
-% - Figures should be uploaded separately from your manuscript file. 
-% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
-% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
-% For figure citations, please use "Fig" instead of "Figure".
-% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
-%
-% Tables should be cell-based and may not contain:
-% - spacing/line breaks within cells to alter layout or alignment
-% - do not nest tabular environments (no tabular environments within tabular environments)
-% - no graphics or colored text (cell background color/shading OK)
-% See http://journals.plos.org/plosone/s/tables for table guidelines.
-%
-% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
-%
-% IMPORTANT
-% Below are a few tips to help format your equations and other special characters according to our specifications.
-For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
-%
-% For inline equations, please be sure to include all portions of an equation in the math environment.
-For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
-%
-% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
-%
-% Please add line breaks to long display equations when possible in order to fit size of the column. 
-%
-% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
-%
-% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
-%
-% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
-%
-% % % % % % % % % % % % % % % % % % % % % % % % 
-%
-% Please contact latex@plos.org with any questions.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-
-% % % % % % % % % % % % % % % % % % % % % % % %
-% Notes from Chris
-%
-% Example of a PLOS methods paper: http://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1005896
-%
-% Here are some examples I removed from the body of the document:
-%
-% Equation example
-% \begin{eqnarray}
-% \label{eq:schemeP}
-% 	\mathrm{P_Y} = \underbrace{H(Y_n) - H(Y_n|\mathbf{V}^{Y}_{n})}_{S_Y} + \underbrace{H(Y_n|\mathbf{V}^{Y}_{n})- H(Y_n|\mathbf{V}^{X,Y}_{n})}_{T_{X\rightarrow Y}},
-% \end{eqnarray}
-
-% For figure citations, please use "Fig" instead of "Figure".
-
-% % Place figure captions after the first paragraph in which they are cited.
-% \begin{figure}[!h]
-% \caption{{\bf Bold the figure title.}
-% Figure caption text here, please use this space for the figure panel descriptions instead of using subfigure commands. A: Lorem ipsum dolor sit amet. B: Consectetur adipiscing elit.}
-% \label{fig1}
-% \end{figure}
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-
 \documentclass[10pt,letterpaper]{article}
-\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+%\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+\setcounter{secnumdepth}{3}
 
@@
 
-% create "+" rule type for thick vertical lines
-\newcolumntype{+}{!{\vrule width 2pt}}
-
-% create \thickcline for thick horizontal lines of variable length
-\newlength\savedwidth
-\newcommand\thickcline[1]{%
-  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-  \cline{#1}%
-  \noalign{\vskip\arrayrulewidth}%
-  \noalign{\global\arrayrulewidth\savedwidth}%
-}
-
-% \thickhline command for thick horizontal lines that span the table
-\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-\hline
-\noalign{\global\arrayrulewidth\savedwidth}}
-
-% Remove comment for double spacing
-%\usepackage{setspace} 
-%\doublespacing
-
 % Text layout
@@
 
-% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
-% Captions will be left justified
-\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
-\renewcommand{\figurename}{Fig}
+% Leave date blank
+\date{}
 
@@
 
-% Remove brackets from numbering in List of References
-\makeatletter
-\renewcommand{\@biblabel}[1]{\quad#1.}
-\makeatother
+%% Include all macros below
 
-% Leave date blank
-\date{}
-
-% Header and Footer with logo
-\usepackage{lastpage,fancyhdr,graphicx}
-\usepackage{epstopdf}
-\pagestyle{myheadings}
-\pagestyle{fancy}
-\fancyhf{}
-\setlength{\headheight}{27.023pt}
-\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
-\rfoot{\thepage/\pageref{LastPage}}
-\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
-\fancyheadoffset[L]{2.25in}
-\fancyfootoffset[L]{2.25in}
-\lfoot{\sf PLOS}
-
-%% Include all macros below
+% This package only necessary if including Supporting Information figures in file.
+\usepackage{float}
 
@@
 
-% Title must be 250 characters or less.
-\begin{flushleft}
-{\Large
-\textbf\newline{Bayesian model comparison for rare variant association studies of multiple phenotypes} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+%\section*{Appendix S1}
+{\Huge
+\textbf\newline{Appendix S1} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
 }
 \newline
-% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
-\\
-\author{Christopher DeBoever$^{1}$, Matthew Aguirre$^{1}$, Yosuke Tanigawa$^{1}$, Chris C.
-A.
-Spencer$^{2}$, Timothy Poterba$^{3}$, Mark J.
-Daly$^{3,4}$,  Matti Pirinen$^{5}$\thanks{matti.pirinen@helsinki.fi}, and Manuel A.
-Rivas$^{1}$\thanks{mrivas@stanford.edu}}
-Christopher DeBoever\textsuperscript{1},
-Matthew Aguirre\textsuperscript{1},
-Yosuke Tanigawa\textsuperscript{1,2},
-Chris C. A. Spencer\textsuperscript{3},
-Timothy Poterba\textsuperscript{4},
-Mark J. Daly\textsuperscript{4,5},
-Matti Pirinen\textsuperscript{6},
-Manuel A. Rivas\textsuperscript{1*},
-\\
-\bigskip
-\textbf{1} Department of Biomedical Data Science, Stanford University, Stanford, CA, USA
-\\
-\textbf{2} Biomedical Informatics Training Program, Stanford University, Stanford, CA, USA
-\\
-\textbf{3} Genomics plc, Oxford, UK
-\\
-\textbf{4} Broad Institute of MIT and Harvard, Cambridge, MA, USA
-\\
-\textbf{5} Wellcome Trust Centre for Human Genetics, University of Oxford, Oxford, UK
-\\
-\textbf{6} Analytic and Translational Genetics Unit, Massachusetts General Hospital, Boston, MA, USA
-\\
-\textbf{7} Oxford Centre for Diabetes, Endocrinology and Metabolism, Oxford, UK
-\\
-\textbf{8} Institute for Molecular Medicine Finland, University of Helsinki, Finland
-\\
-% \textbf{1} Affiliation Dept/Program/Center, Institution Name, City, State, Country
-% \\
-\bigskip
 
-% Use the asterisk to denote corresponding authorship and provide email address in note below.
-* mrivas@stanford.edu
-
-\end{flushleft}
-% Please keep the abstract below 300 words
-\section*{Abstract}
-Whole genome sequencing studies applied to large populations or biobanks with extensive phenotyping raise new analytic challenges.
-The need to consider many variants at a locus or group of genes simultaneously and the potential to study many correlated phenotypes with shared genetic architecture provide opportunities for discovery and inference that are not addressed by the traditional one variant-one phenotype association study.
-Here we introduce a model comparison approach we refer to as MRP for rare variant association studies that considers correlation, scale, and location of genetic effects across a group of genetic variants, phenotypes, and studies.
-We consider the use of summary statistic data to apply 1) standard univariate and multivariate gene-based meta-analysis models, 2) models for assessing heterogeneity of genetic effects, and 3) models for identifying protective protein-truncating variants that can expedite drug discovery.
-Through simulation studies, we demonstrate that the proposed model comparison approach may improve ability to detect rare variant association signals.
-Finally, we demonstrate the model on two groups of phenotypes from the UK Biobank: 1) asthma diagnosis, eosinophil counts, forced expiratory volume, and forced vital capacity; and 2) glaucoma diagnosis, intra-ocular pressure, and corneal resistance factor.
-We demonstrate that the model leverages information from multiple phenotypes to identify known and novel rare variant associations for asthma and glaucoma.
-Overall the model comparison approach is able to retain useful features from widely-used meta-analysis approaches and prioritize protective modifiers of disease risk.
-
-% Please keep the Author Summary between 150 and 200 words
-% Use first person. PLOS ONE authors please skip this step. 
-% Author Summary not valid for PLOS ONE submissions.   
-\section*{Author summary}
-TODO
-
-\linenumbers
-
-% Use "Eq" instead of "Equation" for equation citations.
-\section*{Introduction}
-%% Sequencing technologies
-Sequencing technologies are quickly transforming human genetic studies of complex traits: it is increasingly possible to obtain whole genome sequence data on thousands of samples at manageable costs.
-As a result, the genome-wide study of rare variants (minor allele frequency [MAF] $< 1\%$) and their contribution to disease susceptibility and phenotype variation is now feasible\cite{ifih1,altshuler2010map,rivas2011deep,10002012integrated}.
-
-%% Rare variant methods
-In genetic studies of diseases or continuous phenotypes, rare variants are hard to assess individually due to the limited number of copies of each rare variant.
-Hence, to boost the ability to detect a signal, evidence is usually `aggregated' across variants.
-When designing an `aggregation' method, there are three questions that are usually considered.
-First, across which biological units should variants be combined; second, which variants mapping within those units should be included\cite{majithia2014rare}; and third, which statistical model should be used\cite{lee2014rare}?
-Given the widespread observations of shared genetic risk factors across distinct diseases, there is also considerable motivation to use gene discovery approaches that leverage the information from multiple phenotypes jointly.
-In other words, rather than only aggregating variants that may have effects on a single phenotype, we can also bring together sets of phenotypes for which a single variant or sets of variants might have effects.
-
-%% The model comparison method
-In this paper, we present a Bayesian multiple rare variants and phenotypes (MRP) model comparison approach for identifying rare variant associations as an alternative to current widely-used statistical tests.
-The MRP framework exploits correlation, scale, or location (direction) of genetic effects in a broad range of rare variant association study designs including: case-control; multiple diseases and shared controls; single continuous phenotype; multiple continuous phenotypes; or a mixture of case-control and multiple continuous phenotypes (Fig~\ref{overview}).
-MRP makes use of Bayesian model comparison, whereby we compute a Bayes Factor (BF) defined as the ratio of the marginal likelihoods of the observed data under two models: 1) a pre-specified null where all genetic effects are zero; and 2) an alternative model where factors like correlation, scale, or location of genetic effects are considered.
-The BF is an alternative to $p$-values from traditional hypothesis testing.
-For MRP, the BF represents the statistical evidence for a non-zero effect for a particular group of rare variants on the phenotype(s) of interest.
-
-% Place figure captions after the first paragraph in which they are cited.
-\begin{figure}[!h]
-\caption{{\bf Schematic overview of MRP.}
-A: MRP is suitable for a broad range of rare variant association study designs including (from left to right): i) case-control, ii) multiple diseases with shared controls, iii) single quantitative phenotype, and iv) mixture of case-control and quantitative phenotypes.
-B: Diagram of factors considered in rare variant association analysis including the correlation matrices: $\mathbf{R_{\textrm{study}}}$  (expected correlation of genetic effects among a group of studies), $\mathbf{R_{\textrm{var}}}$ (expected correlation of genetic effects among a group of variants), and $\mathbf{R_{\textrm{phen}}}$ (expected correlation of genetic effects among a group of phenotypes); the scale parameter for genetic variant annotation; and the location of genetic effects, which may be used to prioritize or identify protective modifiers of disease risk.}
-\label{overview}
-\end{figure}
-
-%% Use of summary statistic data
-While many large genetic consortia generally collect both raw genotype and phenotype data, in practice, sharing of individual genotype and phenotype data across groups is difficult to achieve.
-To address this, MRP can take summary statistics, such as estimates of effect size and the corresponding standard error from typical single variant-single phenotype linear or logistic regressions, as input data.
-Furthermore, we use insights from Liu et al.\cite{liu2014meta} and Cichonska et al.\cite{cichonska2016metacca} who suggest the use of additional summary statistics, like covariance estimates across variants and studies, respectively, that would enable lossless ability to detect gene-based association signals using summary statistics alone.
-
-%% Genome variant annotation and protein truncating variants and  protective modifiers of disease risk
-Aggregation techniques rely on variant annotations to assign variants to groups for analysis.
-MRP allows for the inclusion of priors on the spread of effect sizes that can be adjusted depending on what type of variants are included in the analysis.
-For instance, protein truncating variants (PTVs)\cite{rivas2013assessing,rivas2015effect} are an important class of variants that are more likely to be functional because they often disrupt the normal function of a gene.
-This biological knowledge can be reflected in the choices of priors for PTVs in MRP.
-Since PTVs typically abolish or severely alter gene function, there is particular interest in identifying protective PTV modifiers of human disease risk that may serve as targets for therapeutics\cite{pcsk9,cohen2006sequence,sullivan2012effect}.
-We therefore demonstrate how the MRP model comparison approach can improve discovery of such protective signals by modeling the location (direction) of genetic effects which prioritizes variants or genes that are consistent with protecting against disease.
-
-%%Comparing to other methods 
-To evaluate the performance of MRP and to study its behavior we use simulations and compare it to other commonly used approaches.
-Some simple alternatives to MRP include univariate approaches for rare variant association studies including the sequence kernel association test (SKAT)\cite{skat}, and the burden test, which we show are special cases of the MRP model comparison when we assign the prior correlation of genetic effects across different variants to be zero or one.
-Furthermore, we show that MRP is able to detect the presence of heterogeneity of effects, which in some circumstances, such as those where effects are observed to be heterogeneous across studies, may indicate the presence of technical sources of error.
-
-%%Apply methods to the UK Biobank
-We applied MRP to summary statistics for two groups of related phenotypes from the UK Biobank.
-First, we apply MRP to asthma (HC382: the corresponding phenotype label in Global Biobank Engine [\url{https://biobankengine.stanford.edu}]), eosinophil count (INI30150), forced expiratory volume in 1-second (FEV$_1$, INI3063), and forced vital capacity (FVC, INI3062).
-We recover the reported association between a rare PTV in \textit{IL33} and asthma as well as a new association between \textit{CCR3} and asthma \cite{DeBoever179762, 10.1371/journal.pgen.1006659}.
-We also applied MRP to glaucoma (HC276), intra-ocular pressure (INI5263), and corneal resistance factor (INI5265) and find evidence that rare coding variants in \textit{ANGPTL7} and \textit{KLHL22} may protect against glaucoma.
-These analyses show that MRP recovers results from typical single variant-single phenotype analyses while identifying new rare variant associations that include protective modifiers of disease risk.
-
-\section*{Materials and methods}
-\subsection*{MRP model comparison for association testing}
+\section{MRP model comparison for association testing}
 We consider the multivariate linear regression model 
@@
 The prior distribution for the null model, $p\left({\bs {\beta}}|\textrm{NULL}\right)$, is simply the point mass at $\bs \beta = 0$.
-In subsection~\ref{approxlikelihood} we show how we approximate the likelihood function for $\bs \beta$, $p\left(\mathbf{\textrm{Data}} | {\bs {\beta}}\right)$,
-in subsection~\ref{prioralt} we define the prior distribution $p\left({\bs {\beta}}|\textrm{ALT}\right)$ for the alternative model, 
-and finally, in subsection~\ref{bfmodel}, we compute the BF.
+In section~\ref{approxlikelihood} we show how we approximate the likelihood function for $\bs \beta$, $p\left(\mathbf{\textrm{Data}} | {\bs {\beta}}\right)$,
+in section~\ref{prioralt} we define the prior distribution $p\left({\bs {\beta}}|\textrm{ALT}\right)$ for the alternative model, 
+and finally, in section~\ref{bfmodel}, we compute the BF.
 
-\subsubsection*{Likelihood function} 
-\label{approxlikelihood}
+\section{Likelihood function} \label{approxlikelihood}
 A maximum likelihood estimator of $\mathbf{\textrm{B}}$ is given by the ordinary least-squares method
@@
 
-\subsubsection*{Prior of $\bs{\beta}$ in the alternative model} 
+\section{Prior of $\bs{\beta}$ in the alternative model} 
 \label{prioralt}
@@
 
-\subsubsection*{$\textrm{BF}_{\textrm{MRP}}$} 
+\section{$\textrm{BF}_{\textrm{MRP}}$} 
 \label{bfmodel}
@@
 
-The Bayes Factor (derived is below) is given by
+The Bayes Factor (derived below in section~\ref{app1}) is given by
 $$\textrm{BF}_{\textrm{MRP}} = \frac{\det\left(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U} \right)^{-\frac{1}{2}}
@@
 
-\subsubsection*{MRP Bayes Factor derivation} \label{app1}
+\subsection{MRP Bayes Factor derivation} \label{app1}
 To compute the Bayes Factor 
@@
 
-\subsection*{HDF5 Tables}
-Although summary statistics are quicker to read and process than raw data, the number of studies meta-analyzed in this work is expected to be sufficiently large to require optimizations in data representation and processing (\nameref{hdf5_fig}).
-Our solution was the use of the HDF5 (Hierarchical Data Format 5) data representation to enable rapid processing of effect size, uncertainty, and cross-trait estimate data.
-HDF5 is a fast and lightweight file format designed for scientific data.
-It has bindings for R, Python, C/C++, Java, and nearly every other population programming language.
-Reading data from a table within a HDF5 file can be an order of magnitude faster than reading text files from a Unix file, and it makes it easier to organize data within an internal structure.
-
-\subsection*{UK Biobank Data}
-\subsubsection*{GWAS Summary Statistics}
-We performed genome-wide association analysis using PLINK v2.00a(17 July 2017) as previously described \cite{DeBoever179762}.
-For asthma, we used the Firth fallback in PLINK, a hybrid algorithm which normally uses the logistic regression code described in (Hill 2017), but switches to a port of logistf() (https://cran.r-project.org/web/packages/logistf/index.html) in two cases: (1) one of the cells in the 2x2 allele count by case/control status contingency table is empty (2) logistic regression was attempted since all the contingency table cells were nonzero, but it failed to converge within the usual number of steps.
-We used the following covariates in our analysis: age, sex, array type, and the first four principal components, where array type is a binary variable that represents whether an individual was genotyped with UK Biobank Axiom Array or UK BiLEVE Axiom Array.
-For variants that were specific to one array, we did not use array as a covariate.
-Asthma cases were defined using both Hospital Episode Statistics and verbal questionnaire responses.
-We used the provided values from the UK Biobank for eosinophil counts, forced vital capacity (FVC), and forced expiratory volume in 1-second (FEV$_1$).
-The phenotype codes used throughout (asthma=HC382, eosinophil count=INI30150, Forced Expiratory Volume in 1-second (FEV$_1$)=INI3063, and Forced Vital Capacity (FVC)=INI3062) correspond to the phenotype codes used in on the Global Biobank Engine [\url{https://biobankengine.stanford.edu}].
-
-\subsubsection*{Genetic Correlations}
-We calculated the genetic between asthma, eosinophil counts, forced vital capacity (FVC), and forced expiratory volume in 1-second (FEV$_1$) using the MultiVariate Polygenic Mixture Model (MVPMM)~\cite{DeBoever2017}.
-Briefly, MVPMM estimates genetic correlation given GWAS summary statistics (effect size and standard error of effect size estimate) by modeling GWAS summary statistics as generated from one of two mixture components.
-Summary statistics from variants in the null component are modeled as being drawn from a multivariate normal distribution with zero mean and covariance matrix that captures correlation in the summary statistics due to the use of shared subjects or other sources of correlation.
-Summary statistics from variants in the non-null component are modeled as being drawn from a multivariate normal distribution with zero mean, but the covariance matrix for the non-null component combines the covariance matrix from the null component with another covariance matrix that captures the genetic correlation between the phenotypes being considered.
-
-\subsubsection*{UK Biobank Asthma Application}
-For the Manhattan plots and Table~\ref{asthma_table}, we removed any genes with non-unique gene symbols.
-We also removed the sense overlapping gene \textit{CTD-3064M3.3} and and the antisense gene \textit{ZFAT-AS1} since these genes overlap \textit{GPR20} and \textit{ZFAT} respectively and therefore have the same $\log_{10}$ Bayes Factors as those genes.
-
-% Results and Discussion can be combined.
-\section*{Results}
-
-\subsection*{Method overview}
-A complete description of MRP, along with an analytical derivation, is given in the Materials and Methods. Briefly, we describe the approach.
-
-In Bayesian model comparison, we compare the null and alternative models via the Bayes Factor (BF) defined as the ratio of the marginal likelihoods of the observed data under two models.
-Here, the null model is that the effect sizes obtained across all studies for a group of variants and a group of phenotypes is zero.
-To complete the specification of the alternative model we consider the correlation structure, the scale, and the location of the effect sizes (which we find to be informative for alternative models where we seek to identify variants with protective effects).
-
-Let $N$ be the number of individuals and $K$ the number of quantitative phenotype measurements on each individual, for example, here in the applications we focus on asthma, eosinophil count, Forced Expiratory Volume in 1-second, and Forced Vital Capacity, in which case $K=4$.
-Let $M$ be the number of variants in a testing unit ${\bs G}$, where ${\bs G}$ can be, for example, a gene, pathway, or a network.
-Let $S$ be the number of studies where data is obtained from - this data may be in the form of raw genotypes and phenotypes or summary statistics including linkage-disequilibrium, effect sizes (or odds ratio), and standard error of the effect size.
-
-In the multivariate setting we define the prior correlation structure of the effect sizes, denoted by a SMK-by-SMK matrix $\mathbf{U}$.
-In practice, we define $\mathbf{U}$ as a Kronecker product, an operation of matrices of arbitrary size, of three sub-matrices: an S-by-S matrix $\mathbf{R_{\textrm{study}}}$ being the correlations of genetic effects among studies where different values can be used to compare different models of association, such as for identifying heterogeneity of effect sizes between populations~\cite{band2013imputation}; an M-by-M matrix $\mathbf{R_{\textrm{var}}}$ being the correlations of genetic effects among genetic variants, which may reflect the assumption that all the PTVs in a gene may have the same biological consequence\cite{macarthur,rivas2013assessing,rivas2015effect} or prior information obtained through integration of additional data sources, such as functional assay data\cite{majithia2014rare,findlay2014saturation}, otherwise zero correlation of genetic effects may be assumed, which is used in dispersion tests like C-alpha\cite{calpha,clarke2013flexible} and SKAT\cite{skat}; and a K-by-K matrix $\mathbf{R_{\textrm{phen}}}$ being the correlations of genetic effects among phenotypes, which may be obtained from common variant data\cite{cotsapas2011pervasive,solovieff2013pleiotropy,gencorr2015}.
-The variance-covariance matrix of the effect sizes may be obtained from readily available summary statistic data such as in-study LD matrices, effect size estimate (or log odds ratio), and the standard error of the effect size estimate (Online Methods).
-
-For gene discovery, the scale may be used to denote the prior spread of effect sizes.
-For instance, emerging empirical genetic studies have shown that within a gene protein truncating variants may have stronger effects than missense variants\cite{do2015exome}.
-
-Thus far we have assumed that the prior mean, or location, of genetic effects is zero.
-This makes it feasible to analyze a large number of phenotypes without enumerating the prior mean across all phenotypes.
-However, in practice, we may want to proactively identify genetic variants that have effects that are consistent with a protective profile for a disease, which may exploit information from mendelian randomization studies of common variants, such as recent rare variant findings in {\it PCSK9} where truncating loss-of-function variants are found to decrease LDL and triglyceride levels, and decrease CAD risk\cite{cohen2005low,pcsk9,do2013common,cohorts2014loss}.
-This is straightforward to consider by including a non-zero vector as a prior mean of genetic effect (Online Methods).
-
-Although we see advantages in adopting a Bayesian perspective, our approach could be used in a frequentist context by calculating a Bayes factor and using it as a test statistic to compute p-values (Methods, Fig~\ref{simresults}).
-
-\begin{figure}[!h]
-\caption{{\bf Simulation studies.}
-A: Comparison of $-\log_{10}$(p-values) from frequentist $\textrm{BF}_{\textrm{MRP}}$ approximation for an independent effects and a similar effects model to commonly used gene-based statistical tests (skat and burden).
-B: Comparison of log10(Bayes Factors) obtained when raw genotype and phenotype data is available to a scenario where summary statistics only was available and similar effects across studies is assumed.
-C: From single variant and single phenotype to multiple variants and multiple phenotypes gene discovery: ROC curves for detecting gene association to any of the phenotypes using single variant/single phenotype association (blue) to multiple variants and multiple phenotypes association (purple).
-D: ROC curves for detecting gene association when incorporating prior mean of genetic effects (orange) to identify protective alleles.}
-\label{simresults}
-\end{figure}
-
-\subsection*{Simulation studies}
-We first verified the analytical derivations and examined the properties of the approach under a simulation framework.
-
-%% Comparison to other widely used methods
-\subsubsection*{Comparison to frequentist gene tests}
-For the analysis of multiple rare variants and a single phenotype we compared it to the burden test and the SKAT test, commonly used statistical tests in rare variant association studies of a single phenotype.
-We observe concordance between the frequentist methods and the Bayesian models.
-To compare the Bayesian models we compute p-values by using the Bayes Factor as the test statistic and approximating it using distribution properties of quadratic forms (Materials and Methods).
-As expected, an independent effects model has high correlation with the gene-based test SKAT ($r^2 = 0.99$), whereas the similar effects model has high correlation with the burden test ($r^2 = 0.93$, Fig~\ref{simresults}A).
-
-\subsubsection*{Summary statistic data}
-To study the behavior of MRP using summary statistics we simulate two scenarios: first, the scenario where analysts have access to all the raw genotype and phenotype data; and second, the scenario where analysts only have access to summary statistics data\cite{liu2014meta}.
-We conducted $1000$ simulation experiments where we let $K$ (the number of phenotypes) $=3$, $M$ (the number of variants) $=10$, $S$ (the number of studies) $=2$, $N_0$ (number of individuals in study with access to all the data) $= 10000$, $N_1$ (meta-analysis study 1) $= 5000$, $N_2$ (meta-analysis study 2) $= 5000$.
-We find that, under the scenario where similar effects are assumed across studies, the Bayes Factors obtained using summary statistics alone are strongly correlated ($r^2 = 1$) to Bayes Factors obtained by the full genotype and phenotype data (Fig~\ref{simresults}B).
-
-\subsubsection*{From single variant and single phenotype analysis to multiple variants and multiple phenotypes}
-To validate the flexibility of the approach we conducted a simulation experiment where we assumed an allelic architecture consistent to that discovered for {\it APOC3} in relation to coronary artery disease (CAD), triglycerides (TG), low-density lipoprotein cholesterol (LDL-C), and high-density lipoprotein cholesterol (HDL-C)\cite{apoc3,apoc32,jorgensen2014loss,cohorts2014loss}.
-We simulated three studies and applied the model comparison unit jointly to summary statistic data obtained for each study (Supplementary Note).
-Overall, we observed that considering the joint effects across multiple studies in a group of variants and phenotypes may improve ability to detect gene-based signals (Fig~\ref{simresults}C), and that considering prior mean of genetic effects should aid in efforts to identify protective modifiers of disease risk (Fig~\ref{simresults}D).
-
-\subsection*{Applications}
-We applied the MRP model comparison approach to summary statistic data generated from single variant logistic regression and linear regression analysis for coding variants on the UK Biobank array (Methods).
-We applied MRP to asthma and three related traits and glaucoma and two related traits.
-
-\subsubsection*{Model comparison applied to asthma and related phenotypes}
-We first applied MRP to GWAS summary statistics for asthma, eosinophil count, forced expiratory volume in 1-second (FEV$_1$), and forced vital capacity (FVC) phenotypes.
-Recent work has identified associations between the PTV rs146597587 in \textit{IL33} and asthma and eosinophil counts \cite{DeBoever179762, 10.1371/journal.pgen.1006659}.
-FEV$_1$ and FVC are measures of pulmonary function that are used to diagnosis and classify pulmonary disease \cite{Swanney1046}.
-To demonstrate the advantage of considering the phenotypes jointly, we applied MRP to rare missense variants and PTVs (MAF $< 1\%$) for each phenotype separately (Fig~\ref{asthma_bma}A-D) as well as to all phenotypes jointly (Fig~\ref{asthma_bma}E,F) and obtained $\log_{10}$ BF for each gene.
-We applied both independent and similar effects models and used Bayesian model averaging to compute a single BF per gene \cite{10.2307/2676803}.
-In agreement with previous studies, we observed evidence that rare missense variants and/or PTVs in \textit{IL33} affect eosinophil counts and offer protection from asthma from the single-phenotype analyses, though the evidence of association was strongest for the joint analysis (Table~\ref{asthma_table}) \cite{DeBoever179762, 10.1371/journal.pgen.1006659}.
-We also performed an analysis focused on identifying protective variants which also identified the \textit{IL33} association (Fig~\ref{asthma_bma}).
-The results were similar using only either the independent effects (\nameref{asthma_independent}) or similar effects models (\nameref{asthma_similar}).
-We can inspect the effect sizes from the marginal GWAS regressions for the rare variants included in the analysis and see that the association identified by MRP is likely driven by the PTV rs146597587 (Fig~\ref{asthma_bma}G).
-
-We also found moderate evidence for associations between rare coding variants in \textit{CCR3} and asthma.
-The $\log_{10}$ BFs for \textit{CCR3} were 3.3 and 3.1 in the joint and joint protective analyses respectively compared to only -0.5 in the asthma-only analysis  (Fig~\ref{asthma_bma}, Table~\ref{asthma_table}).
-\textit{CCR3} was not reported in a large GWAS for allergic disease including asthma\cite{Ferreira:2017ba} though \textit{CCR3} is near a locus associated with atopy in a previous meta-analysis\cite{Ober:2011jk}.
-Other genes like \textit{IL18RAP}, \textit{ATP2A3}, and \textit{FLG} had $\log_{10}$ BFs greater than 4 in the asthma-only analysis but much smaller BFs in the joint analyses indicating that rare variants in these genes are less likely to affect this group of traits.
-Similarly, there were other genes like \textit{RP11-39K24.9} and \textit{IL17RA} that had relatively larger BFs in the eosinophil count-only analysis but small BFs for the joint analyses demonstrating MRP's ability to integrate information across all phenotypes considered.
-
-%% FIGURE: ASTHMA MANHATTAN AND FOREST FOR BMA
-\begin{figure}[!h]
-\caption{{\bf Results for asthma application.}
-$\log_{10}$ Bayes Factors from applying MRP and Bayesian model averaging to summary statistics for missense and protein-truncating variants from (A) asthma (HC382), (B) eosinophil counts (INI30150), (C) forced vital capacity (FVC, INI3062), (D) forced expiratory volume in 1-second (FEV$_1$, INI3063), (E) all four traits jointly, and (F) all four traits jointly with focus on protective effects.
-The four genes outside of chromosome 6 with the largest Bayes Factors greater than three are labeled in each plot.
-Only $\log_{10}$ Bayes Factors greater than -5 are plotted.
-(F) $-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{IL33} for each phenotype}
-\label{asthma_bma}
-\end{figure}
-
-\begin{table}[!ht]
-\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
-\centering
-\caption{
-{\bf $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{asthma_bma}.}}
-\begin{tabular}{|l+l|l|l|l|l|l|}
-\hline
-{\bf Gene} & {\bf Joint, protective} & {\bf Joint} & {\bf Eosinophil count} & {\bf FVC} & {\bf FEV$_1$} & {\bf Asthma} \\
-\thickhline
-\textit{IL33}         &              29.4 &   29.3 &             30.6 &  -2.3 &  -2.2 &    8.1 \\
-\hline
-\textit{CCR3}         &               3.1 &    3.3 &              7.4 &  -1.4 &  -1.6 &   -0.5 \\
-\hline
-\textit{RP11-39K24.9} &               0.8 &    1.8 &              4.9 &  -0.1 &  -0.4 &    0.3 \\
-\hline
-\textit{SCMH1}        &               0.5 &    4.7 &             -1.5 &   7.7 &   3.8 &   -0.7 \\
-\hline
-\textit{MUSTN1}       &               0.4 &    1.1 &             -1.2 &   2.9 &   2.9 &   -0.6 \\
-\hline
-\textit{ZFAT}         &               0.3 &    1.3 &             -2.0 &   4.7 &   3.1 &   -0.4 \\
-\hline
-\textit{ELN}          &               0.2 &    2.5 &             -1.0 &   5.6 &   2.9 &   -0.6 \\
-\hline
-\textit{C14orf39}     &              -0.7 &   -0.0 &             -1.1 &   3.5 &   2.5 &    0.0 \\
-\hline
-\textit{TMEM110}      &              -0.9 &    1.1 &             -1.0 &   3.3 &   3.1 &   -0.6 \\
-\hline
-\textit{IL17RA}       &              -4.4 &    3.1 &              7.9 &  -2.7 &  -2.5 &   -1.1 \\
-\hline
-\textit{IL18RAP}      &              -9.6 &   -0.9 &             -1.0 &  -1.6 &  -1.7 &    3.5 \\
-\hline
-\textit{ATP2A3}       &             -11.9 &   -0.8 &             -1.2 &  -2.1 &  -2.2 &    5.8 \\
-\hline
-\textit{FLG}          &             -20.1 &  -17.2 &             -6.0 &  -7.4 &  -8.5 &    4.0 \\
-\hline
-\end{tabular}
-\label{asthma_table}
-\end{adjustwidth}
-\end{table}
-
-\subsubsection*{Model comparison applied to glaucoma and related phenotypes}
-We also applied MRP to missense variants and PTVs for glaucoma, intra-ocular pressure, and corneal resistance factor as well as performing joint analyses.
-Elevated intra-ocular pressure is a measure of the fluid pressure in the eye.
-Elevated intra-ocular pressure is associated with glaucoma risk and has been linked to genetic variants associated with glaucoma \cite{doi:10.1001/jama.2014.3192}.
-Corneal resistance factor is a measure of the cornea's ability to resist mechanical stress and has been associated with glaucoma presence and severity \cite{CXO:CXO414, MANSOURI2012419, GriseDulac:2012bo}.
-While the individual glaucoma analysis did not yield any associations with $\log_{10}$ BF greater than three, the joint analysis identified rare coding variants in both \textit{ANGPTL7}, \textit{WNT10A}, and \textit{KLHL22} as associated with glaucoma (Fig~\ref{glaucoma_bma}A-E).
-Applying the MRP model to identify protective effects also identified \textit{ANGPTL7} and \textit{WNT10A} but not \textit{KLHL22}.
-Expression of \textit{ANGPTL7} is upregulated in glaucome and has been proposed to regulate intra-ocular pressure and glaucoma risk \cite{Comes:2011ex, Kuchtey:2008ek}.
-\textit{KLHL22} has not been previously associated with glaucoma though a suggestive association was reported for retinopathy in individuals without diabetes \cite{10.1371/journal.pone.0054232}.
-These results illustrate that applying model comparison to multiple traits simultaneously can prioritize the discovery of protective modifiers of disease.
-
-\begin{figure}[!h]
-\caption{{\bf Results for glaucoma application.}
-$\log_{10}$ Bayes Factors from applying MRP and Bayesian model averaging to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
-(E) shows the results of a joint analysis focused on finding rare variants that protect against glaucoma.
-The genes outside of chromosome 6 with with Bayes Factor greater than three are indicated by arrows.
-Only $\log_{10}$ Bayes Factors greater than zero are plotted.
-F: $-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{ANGPTL7} for all three phenotypes.}
-\label{glaucoma_bma}
-\end{figure}
-
-% %% TABLE: GLAUCOMA GENES OF INTEREST
-\begin{table}[!ht]
-\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
-\centering
-\caption{
-{\bf $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{glaucoma_bma}.}}
-\begin{tabular}{|l+p{3cm}|p{1cm}|p{1.6cm}|p{3.8cm}|p{3.4cm}|}
-\hline
-{\bf Gene} & {\bf Joint, protective} & {\bf Joint} & {\bf Glaucoma} & {\bf Intra-ocular pressure, Goldmann-correlated} & {\bf Corneal resistance factor} \\
-\thickhline
-\textit{ANGPTL7} &              10.1 &   9.4 &      1.7 &                                       10.1 &                       6.2 \\
-\hline
-\textit{KLHL22}  &               3.1 &   3.6 &     -0.2 &                                        3.9 &                       2.2 \\
-\hline
-\textit{GLI3}    &               2.9 &   2.9 &     -0.3 &                                       -0.7 &                       3.2 \\
-\hline
-\textit{WNT10A}  &               2.2 &   3.0 &     -0.3 &                                       -0.5 &                       3.5 \\
-\hline
-\textit{RNASE2}  &              -1.3 &  -0.3 &      0.4 &                                       -1.0 &                      -1.1 \\
-\hline
-\textit{FUK}     &              -6.5 &  -2.3 &      2.5 &                                       -2.1 &                      -2.1 \\
-\hline
-\end{tabular}
-\label{table2}
-\end{adjustwidth}
-\end{table}
-
-% % Place tables after the first paragraph in which they are cited.
-% \begin{table}[!ht]
-% \begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
-% \centering
-% \caption{
-% {\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
-% \begin{tabular}{|l+l|l|l|l|l|l|l|}
-% \hline
-% \multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \thickhline
-% $cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
-% $cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
-% $cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
-% \end{tabular}
-% \begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl.
-Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
-% \end{flushleft}
-% \label{table1}
-% \end{adjustwidth}
-% \end{table}
-
-%PLOS does not support heading levels beyond the 3rd (no 4th level headings).
-% \subsection*{\lorem\ and \ipsum\ nunc blandit a tortor}
-% \subsubsection*{3rd level heading}
-
-% \begin{enumerate}
-% 	\item{react}
-% 	\item{diffuse free particles}
-% 	\item{increment time by dt and go to 1}
-% \end{enumerate}
-
-% \begin{itemize}
-% 	\item First bulleted item.
-% 	\item Second bulleted item.
-% 	\item Third bulleted item.
-% \end{itemize}
-
-\section*{Discussion}
-In this study, we developed a Bayesian model comparison approach MRP that shares information across both variants and phenotypes to identify rare variant associations.
-We used simulations to compare MRP to the widely used burden and SKAT tests for identifying rare variant associations and found that jointly considering both variants and phenotypes can improve the ability to detect associations.
-We also applied the MRP model comparison framework to summary statistic data from two groups of traits from the UK Biobank: asthma diagnosis, eosinophil counts, FEV$_1$, and FVC; and glaucoma diagnosis, intra-ocular pressure, and corneal resistance factor.
-We identified strong evidence for the previously described association between the PTV rs146597587 in \textit{IL33} and asthma\cite{DeBoever179762,10.1371/journal.pgen.1006659} as well as evidence for a potentially novel association for \textit{CCR3} (Table~\ref{asthma_table}).
-We also identified genetic evidence for a link between rare variants in \textit{ANGPTL7} and glaucoma, consistent with previous experiments that suggested a role for \textit{ANGPTL7} in glaucoma \cite{Comes:2011ex, Kuchtey:2008ek}.
-These results demonstrate the ability of the MRP model comparison approach to leverage information across multiple phenotypes and variants to discover rare variant associations.
-
-As genetic data linked to high-dimensional phenotype data continues to be made available through biobanks, health systems, and research programs, there is a large need for statistical approaches that can leverage information across different genetic variants, phenotypes, and studies to make strong inferences about disease-associated genes.
-The approach presented here relies only on summary statistics from marginal association analyses which can be shared with less privacy concerns compared to raw genotype and phenotype data.
-Combining joint analysis of variants and phenotypes with meta-analysis across studies offers new opportunities to identify gene-disease associations.
-
-% Conclusion is optional
-% \section*{Conclusion}
-
-% CO\textsubscript{2}
-
-\section*{Supporting information}
-
-% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
-\paragraph*{S1 Fig.}
-\label{hdf5_fig}
-{\bf HDF5 Implementation.} Our HDF5 implementation contained the following components: first, a group with one table per annotation file.
-All effect size (beta) values and study-specific annotations were contained here, and the number of tables is limited by $S\ \textrm{(the number of studies)} \times K\ \textrm{(the number of traits)}$.
-Second, a group with site-site covariance data.
-While these covariance matrices may have dimension $M\ \textrm{(the number of variants)} \times M$, we store the data as tables, each row specifiying the covariance between two variants.
-The number of tables should be the same as the previous set, capped by $S\ \textrm{(the number of studies)} \times K\ \textrm{(the number of traits)}$.
-Third, we store one table with sigma values for each study/phenotype combination.
-In the event that the traits were rank-normal transformation was performed these sigma values are equal to 1.
-These are used to compute correlation between two datasets.
-Finally, we store a matrix/table pair for Vy null and its index.
-The Vy null matrix has dimensions $ (S \times K) \times (S \times K)$ each entry specifying the estimated correlation of effect sizes between two datasets.
-The index table encodes row/column position of each dataset.
-
-%% FIGURE: ASTHMA MANHATTAN AND FOREST FOR INDEPENDENT EFFECTS
-\paragraph*{S2 Fig.}
-\label{asthma_independent}
-{\bf Results for independent effects model applied to asthma, eosinophil counts, FEV$_1$, and FVC.} $\log_{10}$ Bayes Factors from applying MRP independent effects model to summary statistics for missense and protein-truncating variants from (A) asthma (HC382), (B) eosinophil counts (INI30150), (C) forced vital capacity (FVC, INI3062), (D) forced expiratory volume in 1-second (FEV$_1$, INI3063), (E) all four traits jointly, and (F) all four traits jointly with focus on protective effects.
-The four genes outside of chromosome 6 with the largest Bayes Factors greater than three are labeled in each plot.
-Only $\log_{10}$ Bayes Factors greater than -5 are plotted.
-
-%% FIGURE: ASTHMA MANHATTAN AND FOREST FOR INDEPENDENT EFFECTS
-\paragraph*{S3 Fig.}
-\label{asthma_similar}
-{\bf Results for similar effects model applied to asthma, eosinophil counts, FEV$_1$, and FVC.} $\log_{10}$ Bayes Factors from applying MRP similar effects model to summary statistics for missense and protein-truncating variants from (A) asthma (HC382), (B) eosinophil counts (INI30150), (C) forced vital capacity (FVC, INI3062), (D) forced expiratory volume in 1-second (FEV$_1$, INI3063), (E) all four traits jointly, and (F) all four traits jointly with focus on protective effects.
-The four genes outside of chromosome 6 with the largest Bayes Factors greater than three are labeled in each plot.
-Only $\log_{10}$ Bayes Factors greater than -5 are plotted.
-
-\paragraph*{S4 Fig.}
-\label{glaucoma_similar}
-{\bf Results for similar effects model applied to glaucoma intra-ocular pressure, and corneal resistance factor.} $\log_{10}$ Bayes Factors from applying MRP similar effects model to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
-(E) shows the results of a joint analysis focused on finding rare variants that protect against glaucoma.
-The genes outside of chromosome 6 with with Bayes Factor greater than three are indicated by arrows.
-Only $\log_{10}$ Bayes Factors greater than zero are plotted.
-
-\paragraph*{S5 Fig.}
-\label{glaucoma_independent}
-{\bf Results for independent effects model applied to glaucoma intra-ocular pressure, and corneal resistance factor.} $\log_{10}$ Bayes Factors from applying MRP independent effects model to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
-(E) shows the results of a joint analysis focused on finding rare variants that protect against glaucoma.
-The genes outside of chromosome 6 with with Bayes Factor greater than three are indicated by arrows.
-Only $\log_{10}$ Bayes Factors greater than zero are plotted.
-
-% \paragraph*{S2 Fig.}
-% \label{S2_Fig}
-% {\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-% 
-% \paragraph*{S1 File.}
-% \label{S1_File}
-% {\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-% 
-% \paragraph*{S1 Video.}
-% \label{S1_Video}
-% {\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-% 
-% \paragraph*{S1 Appendix.}
-% \label{S1_Appendix}
-% {\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-% 
-% \paragraph*{S1 Table.}
-% \label{S1_Table}
-% {\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
-% I added in the Author Contributions section
-\section*{Author Contributions}
-MAR and MP designed the method and derived all analytical calculations.
-MAR, MP, and CD wrote the manuscript.
-CCAS, YT, MA, and CD provided analysis and designed figures.
-TP designed HDF5 tables and implementation of loaders.
-MJD provided critical feedback on methodology.
-
-\section*{Acknowledgments}
-This research has been conducted using the UK Biobank Resource under Application Number 24983.
-We thank all the participants in the UK Biobank study.
-M.A.R.
-and C.D.
-are supported by Stanford University and a National Institute of Health center for Multi- and Trans-ethnic Mapping of Mendelian and Complex Diseases grant (5U01 HG009080).
-C.D.
-is supported by a postdoctoral fellowship from the Stanford Center for Computational, Evolutionary, and Human Genomics and the Stanford ChEM-H Institute.
-Y .T.
-is supported by Funai Overseas Scholarship from Funai Foundation for Information Technology and the Stanford University Biomedical Informatics Training Program.
-The primary and processed data used to generate the analyses presented here are available in the UK Biobank access management system (https://amsportal.ukbiobank.ac.uk/) for application 24983, ``Generating effective therapeutic hypotheses from genomic and hospital linkage data'' (http://www.ukbiobank.ac.uk/wp-content/uploads/2017/06/24983-Dr-Manuel-Rivas.pdf), and the results are displayed in the Global Biobank Engine (https://biobankengine.stanford.edu).
-We would like to thank the Customer Solutions Team from Paradigm4 who helped us implement efficient databases for queries and application of inference methods to the data, and also implemented optimized versions of truncated singular value decomposition.
-M.A.R.
-is a paid consultant in Genomics PLC.
-
-\nolinenumbers
-
-% From the PLOS website (http://journals.plos.org/plosone/s/latex#loc-plos-template):
-% As stated in the PLOS template, your reference information should be included in your .tex file (not submitted separately as .bib or .bbl).
-Here is a step-by-step way to include your reference list directly within your .tex file:
-% 
-% Compose your LaTeX manuscript as normal, using \cite{bibkey} for reference citations.
-% Compile your manuscript using your .bib file (\bibliography{references}) and the plos2015.bst style file.  This process should output a .bbl file into the same folder as your manuscript.
-% Open the .bbl file and copy/paste its contents into the appropriate position within the manuscript .tex.  Comment out or delete the \bibliography command.
-% Compile your manuscript again. The PDF output should be the same as before.
-
-% Either type in your references using
-% \begin{thebibliography}{}
-% \bibitem{}
-% Text
-% \end{thebibliography}
-%
-% or
-%
-% Compile your BiBTeX database using our plos2015.bst
-% style file and paste the contents of your .bbl file
-% here. See http://journals.plos.org/plosone/s/latex for 
-% step-by-step instructions.
-%
-
-% \begin{thebibliography}{10}
-% 
-% \bibitem{bib1}
-% Conant GC, Wolfe KH.
-% \newblock {{T}urning a hobby into a job: how duplicated genes find new
-%   functions}.
-% \newblock Nat Rev Genet. 2008 Dec;9(12):938--950.
-% 
-% \bibitem{bib2}
-% Ohno S.
-% \newblock Evolution by gene duplication.
-% \newblock London: George Alien \& Unwin Ltd. Berlin, Heidelberg and New York:
-%   Springer-Verlag.; 1970.
-% 
-% \bibitem{bib3}
-% Magwire MM, Bayer F, Webster CL, Cao C, Jiggins FM.
-% \newblock {{S}uccessive increases in the resistance of {D}rosophila to viral
-%   infection through a transposon insertion followed by a {D}uplication}.
-% \newblock PLoS Genet. 2011 Oct;7(10):e1002337.
-% 
-% \end{thebibliography}
-
-\bibliography{references}
+\bibliography{appendix_references}
 

### 2018-01-29 tex/manuscript.tex
Added figures to manuscript
I've included the graphics in the manuscript file. I've renamed the
"final" figures.
Updated methods and other small updates
I moved the "methods overview" into the methods section. I moved the
detailed methods into an appendix.

@@
 
+% This package only necessary if including Supporting Information figures in file.
+\usepackage{float}
+
 \newcommand{\lorem}{{\bf LOREM}}
@@
 \begin{figure}[!h]
+\includegraphics[width=\textwidth]{../figures/Figure_01.pdf}
 \caption{{\bf Schematic overview of MRP.}
@@
 
-\section*{Materials and methods}
-\subsection*{MRP model comparison for association testing}
-We consider the multivariate linear regression model 
-$$\underset{\left(N\times K\right)}{\mathbf{\textrm{Y}}} 
-= \underset{\left(N\times K\right)}{\mathbf{\Psi}}  + \underset{\left(N\times M\right)}{\mathbf{\textrm{X}}}\underset{\left(M \times K\right)}{\mathbf{\textrm{B}}} 
-+ \underset{\left(N\times K\right)}{\mathbf{\textrm{E}}},$$
-where the matrices $\mathbf{\textrm{Y}} = \left[{y}_{i k}\right]$, $\mathbf{\textrm{X}} = \left[ x_{i m}\right]$, 
-$\mathbf{\textrm{B}} = \left[ \beta_{m k} \right]$ and $\mathbf{\textrm{E}} = \left[e_{i k}\right]$  describe the phenotype values ($y_{i k}$),
-copies of minor allele ($x_{i m}$), variant-phenotype effects ($\beta_{m k}$), and residual errors ($e_{i k}$), 
-for individual $i$, phenotype $k$, and variant $m$. 
-We assume that each phenotype has been transformed to a standard normal
-distribution and that the columns of ${\mathbf{{\textrm{X}}}}$ 
-have been centered, which means that the estimate for the intercept term ${\mathbf{\Psi}}$ is 0 and independent of the estimate of ${\mathbf{\textrm{B}}}$.
-We use vectorized notation where the rows of ${\mathbf{\textrm{B}}}$ form vector 
-$\bs {{ \beta}} = \left(\bs {{ \beta}}_{1},\dots,\bs {{\beta}}_{M}\right)^{\intercal}$
-of length $MK$.
+\section*{Methods}
 
-We define the MRP model comparison as a Bayes factor (BF) between the alternative model, where at least one variant
-affects at least one phenotype, and the null model, where all variant-phenotype effects are zero.  
-BF is the ratio of the marginal likelihoods for these two models: 
-$$ \textrm{BF} = \frac{\int_{{\bs {\beta}}}p\left(\mathbf{\textrm{Data}}|{\bs {\beta}}\right)p\left({\bs {\beta}}|\textrm{ALT}\right)d{\bs {\beta}}}{\int_{{\bs {\beta}}}p\left(\mathbf{\textrm{Data}}|{\bs {\beta}}\right)p\left({\bs {\beta}}|\textrm{NULL}\right)d{\bs {\beta}}},$$
-where Data can correspond either to the effect size estimates $\bs {\widehat{ \beta}}$ and the estimated variance-covariance matrix of  $\bs {\widehat{ \beta}}$, $\widehat{\mathbf{\textrm{V}}}_{\bs{\beta}}$, or to the original phenotypes and genotypes,
-$\underset{\left(N\times K\right)}{\mathbf{\textrm{\textrm{Y}}}}$ and $ \underset{\left(N\times M\right)}{\mathbf{\textrm{\textrm{X}}}}$, and any other covariates that we
-want to regress out from the phenotypes.
+\subsection*{Description of MRP}
+In this section, we provide an overview of the MRP model comparison approach.
+Refer to \nameref{S1_Appendix} for detailed derivations.
+MRP models GWAS summary statistics as being distributed according to one of two models.
+The null model is that the regression effect sizes obtained across all studies for a group of variants and a group of phenotypes is zero.
+The alternative model is that summary statistics are distributed according to a multivariate normal distribution with mean zero and covariance matrix described below.
+MRP compares the evidence for the null and alternative model using a Bayes Factor (BF) that quantifies the amount of evidence for each model as the ratio of the marginal likelihoods of the observed data under two models.
 
-The prior distribution for the null model, $p\left({\bs {\beta}}|\textrm{NULL}\right)$, is simply the point mass at $\bs \beta = 0$.
-In subsection~\ref{approxlikelihood} we show how we approximate the likelihood function for $\bs \beta$, $p\left(\mathbf{\textrm{Data}} | {\bs {\beta}}\right)$,
-in subsection~\ref{prioralt} we define the prior distribution $p\left({\bs {\beta}}|\textrm{ALT}\right)$ for the alternative model, 
-and finally, in subsection~\ref{bfmodel}, we compute the BF.
+ To define the alternative model, we must specify the prior correlation structure, scale, and location (direction) of the effect sizes.
+Let $N$ be the number of individuals and $K$ the number of phenotype measurements on each individual.
+Let $M$ be the number of variants in a testing unit ${\bs G}$, where ${\bs G}$ can be, for example, a gene, pathway, or a network.
+Let $S$ be the number of studies where data is obtained from - this data may be in the form of raw genotypes and phenotypes or summary statistics including linkage-disequilibrium, effect sizes (or odds ratio), and standard error of the effect size.
+When considering multiple phenotypes ($K > 1$), we define the prior correlation structure of the effect sizes as an $SMK\times SMK$ matrix $\mathbf{U}$.
+In practice, we define $\mathbf{U}$ as a Kronecker product, an operation of matrices of arbitrary size, of three sub-matrices
+\begin{itemize}
+\item an $S\times S$ matrix $\mathbf{R_{\textrm{study}}}$ containing the correlations of genetic effects among studies where different values can be used to compare different models of association, such as for identifying heterogeneity of effect sizes between populations~\cite{band2013imputation}
+\item an $M\times M$ matrix $\mathbf{R_{\textrm{var}}}$ containing the correlations of genetic effects among genetic variants, which may reflect the assumption that all the PTVs in a gene may have the same biological consequence\cite{macarthur,rivas2013assessing,rivas2015effect} or prior information obtained through integration of additional data sources, such as functional assay data\cite{majithia2014rare,findlay2014saturation}, otherwise zero correlation of genetic effects may be assumed, which is used in dispersion tests like C-alpha\cite{calpha,clarke2013flexible} and SKAT\cite{skat}
+\item a $K\times K$ matrix $\mathbf{R_{\textrm{phen}}}$ containing the correlations of genetic effects among phenotypes, which may be obtained from common variant data\cite{cotsapas2011pervasive,solovieff2013pleiotropy,gencorr2015}
+\end{itemize}
+The variance-covariance matrix of the effect sizes may be obtained from readily available summary statistic data such as in-study LD matrices, effect size estimates (or log odds ratios), and the standard errors of the effect size estimates (Online Methods).
 
-\subsubsection*{Likelihood function} 
-\label{approxlikelihood}
-A maximum likelihood estimator of $\mathbf{\textrm{B}}$ is given by the ordinary least-squares method
-$$\widehat{\mathbf{\textrm{B}}} = \left(\mathbf{\textrm{X}}^\intercal\mathbf{\textrm{X}}\right)^{-1}\mathbf{\textrm{X}}^{\intercal}\mathbf{\textrm{Y}},$$
-that in vectorized form is denoted by   
-$\bs {\widehat{ \beta}} = \left(\bs {\widehat{ \beta}}_{1},\dots,\bs {\widehat{\beta}}_{M}\right)^{\intercal}.$
-An estimator of the variance-covariance of $\widehat{\bs \beta}$  is given by 
-$$\widehat{\textrm{V}}_{\mathbf{\bs \beta}} =  \left(\mathbf{\textrm{X}}^\intercal\mathbf{\textrm{X}}\right)^{-1}\otimes \widehat{\textrm{V}}_{\mathbf{\textrm{Y}}},$$
-where $\widehat{\textrm{V}}_{\mathbf{\textrm{Y}}}$ is the estimated residual variance-covariance matrix of $\mathbf{\textrm{Y}}$ given $\mathbf{\textrm{X}}$.
+MRP allows users to specify priors that reflect knowledge of the variants and phenotypes under study.
+The prior on the scale of effect sizes can be used to denote which variants may have larger effect sizes.
+For instance, emerging empirical genetic studies have shown that within a gene, PTVs may have stronger effects than missense variants\cite{do2015exome}.
+This can be reflected by providing larger priors for PTVs.
 
-%Let $\bs {\widehat{ \beta}} = \left(\bs {\widehat{ \beta}}^{\intercal}_{1},\dots,\bs {\widehat{\beta}}^{\intercal}_{m}\right)^{\intercal}$ be a vector of point estimates of all $MK$ effects. 
-Following Band et al.\cite{band2013imputation}, we approximate the likelihood function of $\bs {\beta}$ by a multivariate normal distribution with mean $\bs {\widehat{\beta}}$ and variance-covariance matrix $\widehat{\textrm{V}}_{\bs {\beta}}$.
-Note that by approximating $\Vy$ by the trait correlation matrix, this likelihood approximation does not require access to the individual level data $\mathbf{\textrm{X}}$ and $\mathbf{\textrm{Y}}$ but only to the
-summary data of effect sizes $\bs {\widehat{\beta}}$, LD-matrix $\mathbf{\textrm{X}}^\intercal\mathbf{\textrm{X}}$ and a trait correlation estimate.
+Similarly, we can utilize a prior on the location (direction) of effects to specify alternative models where we seek to identify variants with protective effects against disease.
+Thus far we have assumed that the prior mean, or location, of genetic effects is zero which makes it feasible to analyze a large number of phenotypes without enumerating the prior mean across all phenotypes.
+To proactively identify genetic variants that have effects that are consistent with a protective profile for a disease, we can include a non-zero vector as a prior mean of genetic effect \nameref{S1_Appendix}.
+We can exploit information from Mendelian randomization studies of common variants, such as recent findings where rare truncating loss-of-function variants in {\it PCSK9} were found to decrease LDL and triglyceride levels and decrease CAD risk\cite{cohen2005low,pcsk9,do2013common,cohorts2014loss} to identify situations where such a prior is warranted.
 
-\subsubsection*{Prior of $\bs{\beta}$ in the alternative model} 
-\label{prioralt}
-We construct the prior distribution $p\left({\bs {\beta}} | \textrm{ALT} \right)$ for the alternative model in three steps allowing 
-user to specify correlations between effects of different variants on different traits across different studies.
+Applying MRP to variants from a testing unit ${\bs G}$ yields a BF for that testing unit that describes the evidence that rare variants in that testing unit have a nonzero effect on the traits used in the model.
+For instance, consider genes as testing units.
+By running MRP, we obtain a BF for each gene that represents the evidence that rare variants in that gene affect the traits of interest.
+These BF can be used to identify specific genes that may be linked to disease.
+Although we see advantages in adopting a Bayesian perspective for MRP, our approach could be used in a frequentist context by calculating a BF and using it as a test statistic to compute p-values (\nameref{S1_Appendix}, Fig~\ref{simresults}).
 
-In a single study, the prior density for ${\bs {\beta}}$ incorporates the expected correlation of genetic effects among 
-a group of variants ($\mathbf{R}_{\textrm{var}}$) and among a group of phenotypes ($\mathbf{R}_{\textrm{phen}}$). 
-In addition, we  incorporate an expected spread of the effect size of each variant by scaling $\mathbf{R}_{\textrm{var}}$ as 
-$$\mathbf{S}_{\textrm{var}} = \Delta\left(\sigma_m\right) \mathbf{R}_{\textrm{var}} \Delta\left(\sigma_m\right),$$
-where $\Delta\left(\sigma_m\right)$ is a diagonal matrix with entries $\sigma_m$ determining the spread of the effect size distribution for each variant $m \leq M$.
-Thus, we can model settings where, e.g., protein-truncating variants have larger effect sizes ($\sigma = 1$) than missense variants ($\sigma = 0.1$).
-Note that when $\sigma_m = 1$ for all $m$ then $\mathbf{S}_{\textrm{var}} = \mathbf{R}_{\textrm{var}}$.
-
-All in all, our prior density for ${\bs \beta}$ under alternative model is 
-$${\bs {\beta}}|\textrm{ALT} \sim \mathcal{N}\left(\mathbf{0}, \mathbf{U}\right), \textrm{ where } \mathbf{U} = \mathbf{S}_{\textrm{var}} \otimes \mathbf{R}_{\textrm{phen}}.$$
-
-When we have data from multiple studies 
-we allow for possible differences in genetic effects across ethnicities or populations 
-extending the Approximate Bayes Factors of Band et al.~\cite{band2013imputation}
-and the summary statistics approach of RAREMETAL~\cite{liu2014meta} from univariate to multivariate phenotypes.
-Let $\bs {\widehat{ \beta}} = \left(\widehat{ \beta}_{s,m,k}\right)=\left(\widehat{\beta}_{1,1,1}, \widehat{\beta}_{1,1,2}, \dots, \widehat{\beta}_{1,1,K}, \widehat{\beta}_{1,2,1}, \dots, \widehat{\beta}_{1,2,K}, \dots, \widehat{\beta}_{1,M,K}, \widehat{\beta}_{2,1,1}, \dots, \widehat{\beta}_{S,M,K}\right),$ where $S$ is the number of studies, $M$ is the number of variants, and $K$ is the number of phenotypes.
-As with a single study, we incorporate the expected correlation of genetic effects between a pair of variants and a single phenotype using the matrix $\mathbf{S_{\textrm{var}}}$, between a variant and a pair of phenotypes using the matrix $\mathbf{R_{\textrm{phen}}}$, and we introduce the matrix $\mathbf{R_{\textrm{study}}}$ to specify prior on the similarity in effect sizes across the studies.
-Thus, the prior is
-$${\bs {\beta}} \sim \mathcal{N}\left(\mathbf{0}, \mathbf{U}\right), \textrm{ where } 
-\mathbf{U} = \mathbf{R}_{\textrm{study}}\otimes\left(\mathbf{S}_{\textrm{var}} \otimes \mathbf{R}_{\textrm{phen}}\right).$$
-
-It is straightforward to include a non-zero vector ${\bs\mu}$ as a prior mean of genetic effects, in which case the prior is
-$${\bs {\beta}} \sim \mathcal{N}\left({\bs \mu},\mathbf{U}\right).$$
-We use this, for example, when screening for protective rare variants that have a pre-specified beneficial profile on a set of risk factors.
-
-\subsubsection*{$\textrm{BF}_{\textrm{MRP}}$} 
-\label{bfmodel}
-The Bayes Factor is the ratio of the marginal likelihoods between the alternative and the null model. 
-The marginal likelihood for the alternative model is 
-$$ \int_{{\bs {\beta}}} p\left(\mathbf{\textrm{Data}}|{\bs {\beta}}\right) p\left({\bs {\beta}}|\textrm{ALT}\right)d{\bs {\beta}} = c \times \mathcal{N}\left({ {\widehat{\bs \beta}}}; {\bs \mu},  \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U}\right)$$
-and the marginal likelihood for the null model is 
-$$\int_{{\bs {\beta}}} p\left(\mathbf{\textrm{Data}}|{\bs {\beta}}\right) p\left({\bs {\beta}}|\textrm{NULL}\right)d{\bs {\beta}} = c \times \mathcal{N}\left({ {\widehat{\bs \beta}}}; 0,  \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}\right).$$
-
-The Bayes Factor (derived is below) is given by
-$$\textrm{BF}_{\textrm{MRP}} = \frac{\det\left(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U} \right)^{-\frac{1}{2}}
-\exp\left[-\frac{1}{2}\left({\bs {\widehat{\beta}} - {\bs \mu}}\right)^{\intercal}\left(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U}\right)^{-1}\left({\bs {\widehat{\beta}}} - {\bs \mu}\right)\right]}
-{\det \left( \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}\right)^{-\frac{1}{2}}\exp \left[-\frac{1}{2}{\bs {\widehat{\beta}}}^{\intercal}\, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} 
-\,{\bs {\widehat{\beta}}}\right]}.$$
-
-When $\bs \mu=0$, $\textrm{BF}_{\textrm{MRP}}$ is an increasing function of the following quadratic form
-\begin{equation}\label{quadratic_form}
-Q(\bs{\widehat{\beta}};\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}},\mathbf{U})=
-\bs{\widehat{\beta}^\intercal} \left(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}^{-1} - (\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}+\mathbf{U})^{-1}\right) \bs{\widehat{\beta}}.
-\end{equation}
-Furthermore, this quadratic form is the only part of the $\textrm{BF}_{\textrm{MRP}}$ that depends on $\bs {\widehat{\beta}}$.
-Thus, by deriving a distribution of $Q(\bs{\widehat{\beta}};\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}},\mathbf{U})$
-under the null model we can compute a p-value when $\textrm{BF}_{\textrm{MRP}}$ is used as a test statistic.
-According to basic properties of quadratic forms of Gaussian variables,
-$Q(\bs{\widehat{\beta}};\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}},\mathbf{U}) \sim \sum_{i=1}^n d_i \chi^2_i$, where $\chi_i^2$
-are an independent sample from $\chi_1^2$ distribution (chi-square with one degree of freedom),
-and $d_i$ are the eigenvalues of matrix
-$I-(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}+\mathbf{U})^{-1}\, \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}$.
-The distribution function for a mixture of chi-squares can be numerically evaluated by
-the R-package `CompQuadForm'~\cite{Duchesne2010}.
-
-\subsubsection*{MRP Bayes Factor derivation} \label{app1}
-To compute the Bayes Factor 
-$$\textrm{BF}_{\textrm{MRP}} = \frac{\det\left(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U} \right)^{-\frac{1}{2}}
-\exp\left[-\frac{1}{2}\left({\bs {\widehat{\beta}} - {\bs \mu}}\right)^{\intercal}\left(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U}\right)^{-1}\left({\bs {\widehat{\beta}}} - {\bs \mu}\right)\right]}
-{\det \left( \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}\right)^{-\frac{1}{2}}\exp \left[-\frac{1}{2}{\bs {\widehat{\beta}}}^{\intercal}\, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} 
-\,{\bs {\widehat{\beta}}}\right]},$$
-we first consider the term inside the exponential function:
-$$ \mathcal{E} \left(\widehat{\bs \beta}, \bs \mu, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}, \mathbf{U}\right) = 
-\frac{1}{2}\, {\bs {\widehat{\beta}}}^{\intercal}\, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} \,{\bs {\widehat{\beta}}}
--\frac{1}{2}\left({\bs {\widehat{\beta}} - {\bs \mu}}\right)^{\intercal}\left(\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U}\right)^{-1}\left({\bs {\widehat{\beta}}} - {\bs \mu}\right).
-$$
-Since $\widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}$ and $\mathbf{U}$ are typically defined through Kronecker products of smaller matrices, their inverses
-are easier to compute than the inverse of their sum. Hence we use Woodbury matrix identity to write
-$$ 
-\mathcal{E} \left(\widehat{\bs \beta}, \bs \mu, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}, \mathbf{U}\right) = 
-\frac{1}{2}\, {\bs {\widehat{\beta}}}^{\intercal}\, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} \,{\bs {\widehat{\beta}}}
--\frac{1}{2}\left({\bs {\widehat{\beta}} - {\bs \mu}}\right)^{\intercal} \left( \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} - \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} 
-\left(\mathbf{U}^{-1} + \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} \right)^{-1} \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}^{-1} \right ) 
-\left({\bs {\widehat{\beta}} - {\bs \mu}}\right).
-$$
-
-To simplify the determinant calculation we write 
-$$\det\left( \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}} + \mathbf{U} \right) = \det\left( \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}\right) \det\left( \mathbf{I} + \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}^{-1} \mathbf{U}\right).$$
-The logarithm of the Bayes Factor is then
-$$\log \left( \textrm{BF}_{\textrm{MRP}} \right) = 
--\frac{1}{2} \log \left( \det\left( \mathbf{I} + \widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}^{-1} \mathbf{U} \right) \right)  + \mathcal{E} \left(\widehat{\bs \beta}, \bs \mu, \widehat{\mathbf{\textrm{V}}}_{\bs {\beta}}, \mathbf{U}\right).$$
-
-If studies do not share individuals, $\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}$ is a block-diagonal matrix 
-$$\widehat{\mathbf{\textrm{V}}}_{{\bs {\beta}}}=
- \left[\begin{array}{c|c|c|c}
- \widehat{\mathbf{\textrm{V}}}^{1}_{{\bs \beta}} & 0 & \cdots &0\\ \hline
-0 & \widehat{\mathbf{\textrm{V}}}^{2}_{{\bs \beta}}&\cdots &0 \\ \hline
-\vdots & &\ddots &\vdots \\ \hline
-0 & 0 &\cdots & \widehat{\mathbf{\textrm{V}}}^{S}_{{\bs {\beta}}}\\
-\end{array}\right].$$
-If studies share individuals, e.g., controls, 
-we take the approach of Cichonska et al.~\cite{cichonska2016metacca} to use summary level data
-to estimate the correlation structure of the non-diagonal blocks caused by overlapping individuals.
+\begin{figure}[!h]
+\includegraphics[width=\textwidth]{../figures/Figure_02.pdf}
+\caption{{\bf Simulation studies.}
+A: Comparison of $-\log_{10}$(p-values) from frequentist $\textrm{BF}_{\textrm{MRP}}$ approximation for an independent effects and a similar effects model to commonly used gene-based statistical tests (skat and burden).
+B: Comparison of log10(Bayes Factors) obtained when raw genotype and phenotype data is available to a scenario where summary statistics only was available and similar effects across studies is assumed.
+C: From single variant and single phenotype to multiple variants and multiple phenotypes gene discovery: ROC curves for detecting gene association to any of the phenotypes using single variant/single phenotype association (blue) to multiple variants and multiple phenotypes association (purple).
+D: ROC curves for detecting gene association when incorporating prior mean of genetic effects (orange) to identify protective alleles.}
+\label{simresults}
+\end{figure}
 
@@
 For variants that were specific to one array, we did not use array as a covariate.
-Asthma cases were defined using both Hospital Episode Statistics and verbal questionnaire responses.
-We used the provided values from the UK Biobank for eosinophil counts, forced vital capacity (FVC), and forced expiratory volume in 1-second (FEV$_1$).
-The phenotype codes used throughout (asthma=HC382, eosinophil count=INI30150, Forced Expiratory Volume in 1-second (FEV$_1$)=INI3063, and Forced Vital Capacity (FVC)=INI3062) correspond to the phenotype codes used in on the Global Biobank Engine [\url{https://biobankengine.stanford.edu}].
+
+Asthma and glaucoma cases were defined using both Hospital Episode Statistics and verbal questionnaire responses.
+We used the provided values from the UK Biobank for eosinophil counts, forced vital capacity (FVC), forced expiratory volume in 1-second (FEV$_1$), intra-ocular pressure, and corneal resistance factor.
+The phenotype codes used throughout (asthma=HC382, eosinophil count=INI30150, FEV$_1$=INI3063, FVC=INI3062, glaucoma=HC276, intra-ocular pressure=INI5263, and corneal resistance factor=INI5265) correspond to the phenotype codes used in on the Global Biobank Engine [\url{https://biobankengine.stanford.edu}].
 
 \subsubsection*{Genetic Correlations}
-We calculated the genetic between asthma, eosinophil counts, forced vital capacity (FVC), and forced expiratory volume in 1-second (FEV$_1$) using the MultiVariate Polygenic Mixture Model (MVPMM)~\cite{DeBoever2017}.
+We calculated the genetic between the two groups of traits (asthma, eosinophil counts, FVC, FEV$_1$ and glaucoma, intra-ocular pressure, corneal resistance factor) using the MultiVariate Polygenic Mixture Model (MVPMM)~\cite{DeBoever2017}.
 Briefly, MVPMM estimates genetic correlation given GWAS summary statistics (effect size and standard error of effect size estimate) by modeling GWAS summary statistics as generated from one of two mixture components.
@@
 
-\subsubsection*{UK Biobank Asthma Application}
-For the Manhattan plots and Table~\ref{asthma_table}, we removed any genes with non-unique gene symbols.
-We also removed the sense overlapping gene \textit{CTD-3064M3.3} and and the antisense gene \textit{ZFAT-AS1} since these genes overlap \textit{GPR20} and \textit{ZFAT} respectively and therefore have the same $\log_{10}$ Bayes Factors as those genes.
+\subsubsection*{UK Biobank Asthma and Glaucoma Applications}
+We applied MRP for both sets of traits assuming both an independent effects model and a similar effects model.
+We applied Bayesian model averaging to these results by averaging the BF from each model for each gene; these results are reported in the main text.
+The results for each individual model are included in the Supporting Information.
+For the Manhattan plots and tables, we removed any genes with non-unique gene symbols.
+In cases where genes overlapped such that they shared rare variants and therefore the same BF, we removed one gene.
 
@@
 \section*{Results}
-
-\subsection*{Method overview}
-A complete description of MRP, along with an analytical derivation, is given in the Materials and Methods. Briefly, we describe the approach.
-
-In Bayesian model comparison, we compare the null and alternative models via the Bayes Factor (BF) defined as the ratio of the marginal likelihoods of the observed data under two models.
-Here, the null model is that the effect sizes obtained across all studies for a group of variants and a group of phenotypes is zero.
-To complete the specification of the alternative model we consider the correlation structure, the scale, and the location of the effect sizes (which we find to be informative for alternative models where we seek to identify variants with protective effects).
-
-Let $N$ be the number of individuals and $K$ the number of quantitative phenotype measurements on each individual, for example, here in the applications we focus on asthma, eosinophil count, Forced Expiratory Volume in 1-second, and Forced Vital Capacity, in which case $K=4$.
-Let $M$ be the number of variants in a testing unit ${\bs G}$, where ${\bs G}$ can be, for example, a gene, pathway, or a network.
-Let $S$ be the number of studies where data is obtained from - this data may be in the form of raw genotypes and phenotypes or summary statistics including linkage-disequilibrium, effect sizes (or odds ratio), and standard error of the effect size.
-
-In the multivariate setting we define the prior correlation structure of the effect sizes, denoted by a SMK-by-SMK matrix $\mathbf{U}$.
-In practice, we define $\mathbf{U}$ as a Kronecker product, an operation of matrices of arbitrary size, of three sub-matrices: an S-by-S matrix $\mathbf{R_{\textrm{study}}}$ being the correlations of genetic effects among studies where different values can be used to compare different models of association, such as for identifying heterogeneity of effect sizes between populations~\cite{band2013imputation}; an M-by-M matrix $\mathbf{R_{\textrm{var}}}$ being the correlations of genetic effects among genetic variants, which may reflect the assumption that all the PTVs in a gene may have the same biological consequence\cite{macarthur,rivas2013assessing,rivas2015effect} or prior information obtained through integration of additional data sources, such as functional assay data\cite{majithia2014rare,findlay2014saturation}, otherwise zero correlation of genetic effects may be assumed, which is used in dispersion tests like C-alpha\cite{calpha,clarke2013flexible} and SKAT\cite{skat}; and a K-by-K matrix $\mathbf{R_{\textrm{phen}}}$ being the correlations of genetic effects among phenotypes, which may be obtained from common variant data\cite{cotsapas2011pervasive,solovieff2013pleiotropy,gencorr2015}.
-The variance-covariance matrix of the effect sizes may be obtained from readily available summary statistic data such as in-study LD matrices, effect size estimate (or log odds ratio), and the standard error of the effect size estimate (Online Methods).
-
-For gene discovery, the scale may be used to denote the prior spread of effect sizes.
-For instance, emerging empirical genetic studies have shown that within a gene protein truncating variants may have stronger effects than missense variants\cite{do2015exome}.
-
-Thus far we have assumed that the prior mean, or location, of genetic effects is zero.
-This makes it feasible to analyze a large number of phenotypes without enumerating the prior mean across all phenotypes.
-However, in practice, we may want to proactively identify genetic variants that have effects that are consistent with a protective profile for a disease, which may exploit information from mendelian randomization studies of common variants, such as recent rare variant findings in {\it PCSK9} where truncating loss-of-function variants are found to decrease LDL and triglyceride levels, and decrease CAD risk\cite{cohen2005low,pcsk9,do2013common,cohorts2014loss}.
-This is straightforward to consider by including a non-zero vector as a prior mean of genetic effect (Online Methods).
-
-Although we see advantages in adopting a Bayesian perspective, our approach could be used in a frequentist context by calculating a Bayes factor and using it as a test statistic to compute p-values (Methods, Fig~\ref{simresults}).
-
-\begin{figure}[!h]
-\caption{{\bf Simulation studies.}
-A: Comparison of $-\log_{10}$(p-values) from frequentist $\textrm{BF}_{\textrm{MRP}}$ approximation for an independent effects and a similar effects model to commonly used gene-based statistical tests (skat and burden).
-B: Comparison of log10(Bayes Factors) obtained when raw genotype and phenotype data is available to a scenario where summary statistics only was available and similar effects across studies is assumed.
-C: From single variant and single phenotype to multiple variants and multiple phenotypes gene discovery: ROC curves for detecting gene association to any of the phenotypes using single variant/single phenotype association (blue) to multiple variants and multiple phenotypes association (purple).
-D: ROC curves for detecting gene association when incorporating prior mean of genetic effects (orange) to identify protective alleles.}
-\label{simresults}
-\end{figure}
 
@@
 We observe concordance between the frequentist methods and the Bayesian models.
-To compare the Bayesian models we compute p-values by using the Bayes Factor as the test statistic and approximating it using distribution properties of quadratic forms (Materials and Methods).
+To compare the Bayesian models we compute p-values by using the BF as the test statistic and approximating it using distribution properties of quadratic forms (\nameref{S1_Appendix}).
 As expected, an independent effects model has high correlation with the gene-based test SKAT ($r^2 = 0.99$), whereas the similar effects model has high correlation with the burden test ($r^2 = 0.93$, Fig~\ref{simresults}A).
@@
 
-\subsubsection*{Model comparison applied to asthma and related phenotypes}
+\subsubsection*{Asthma, eosinophil counts, forced expiratory volume, and forced vital capacity}
 We first applied MRP to GWAS summary statistics for asthma, eosinophil count, forced expiratory volume in 1-second (FEV$_1$), and forced vital capacity (FVC) phenotypes.
@@
 \begin{figure}[!h]
+\includegraphics[width=\textwidth]{../figures/Figure_03.pdf}
 \caption{{\bf Results for asthma application.}
@@
 
-\subsubsection*{Model comparison applied to glaucoma and related phenotypes}
+\subsubsection*{Glaucoma, intra-ocular pressure, and corneal resistance factor}
 We also applied MRP to missense variants and PTVs for glaucoma, intra-ocular pressure, and corneal resistance factor as well as performing joint analyses.
@@
 \begin{figure}[!h]
+\includegraphics[width=\textwidth]{../figures/Figure_04.pdf}
 \caption{{\bf Results for glaucoma application.}
@@
 % Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
+
+\paragraph*{S1 Appendix.}
+\label{S1_Appendix}
+{\bf MRP model details.} Specification of the MRP model including the likelihood function, priors, and Bayes factor calculation.
+
+\begin{figure}[H]
+\includegraphics[width=\textwidth]{../figures/Figure_S01.pdf}
+\end{figure}
 \paragraph*{S1 Fig.}
@@
 %% FIGURE: ASTHMA MANHATTAN AND FOREST FOR INDEPENDENT EFFECTS
+\begin{figure}[H]
+\includegraphics[width=\textwidth]{../figures/Figure_S02.pdf}
+\end{figure}
 \paragraph*{S2 Fig.}
@@
 
-%% FIGURE: ASTHMA MANHATTAN AND FOREST FOR INDEPENDENT EFFECTS
+%% FIGURE: ASTHMA MANHATTAN AND FOREST FOR SIMILAR EFFECTS
+\begin{figure}[H]
+\includegraphics[width=\textwidth]{../figures/Figure_S03.pdf}
+\end{figure}
 \paragraph*{S3 Fig.}
@@
 
+\begin{figure}[H]
+\includegraphics[width=\textwidth]{../figures/Figure_S04.pdf}
+\end{figure}
 \paragraph*{S4 Fig.}
@@
 
+\begin{figure}[H]
+\includegraphics[width=\textwidth]{../figures/Figure_S05.pdf}
+\end{figure}
 \paragraph*{S5 Fig.}

### 2018-01-29 tex/manuscriptbiorxiv.tex
changes to main text ; added a biorxiv version and edited the plos submission version; edited appendix

--- 2018-01-29 tex/manuscript.tex
@@
 
-\documentclass[10pt,letterpaper]{article}
-\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+\documentclass[10pt]{article}
+%\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
 
@@
 \setlength{\parindent}{0.5cm}
-\textwidth 5.25in 
-\textheight 8.75in
+%\textwidth 5.25in 
+%\textheight 8.75in
 
@@
 \pagestyle{myheadings}
-\pagestyle{fancy}
+%\pagestyle{fancy}
 \fancyhf{}
-\setlength{\headheight}{27.023pt}
-\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
+%\setlength{\headheight}{27.023pt}
+%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
 \rfoot{\thepage/\pageref{LastPage}}
 \renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
-\fancyheadoffset[L]{2.25in}
-\fancyfootoffset[L]{2.25in}
-\lfoot{\sf PLOS}
+%\fancyheadoffset[L]{2.25in}
+%\fancyfootoffset[L]{2.25in}
+%\lfoot{\sf PLOS}
 
@@
 \begin{document}
-\vspace*{0.2in}
+%\vspace*{0.2in}
 
@@
 Spencer$^{2}$, Timothy Poterba$^{3}$, Mark J.
-Daly$^{3,4}$,  Matti Pirinen$^{5}$\thanks{matti.pirinen@helsinki.fi}, and Manuel A.
+Daly$^{4,5}$,  Matti Pirinen$^{5}$\thanks{matti.pirinen@helsinki.fi}, and Manuel A.
 Rivas$^{1}$\thanks{mrivas@stanford.edu}}
@@
 Matthew Aguirre\textsuperscript{1},
-Yosuke Tanigawa\textsuperscript{1,2},
-Chris C. A. Spencer\textsuperscript{3},
-Timothy Poterba\textsuperscript{4},
-Mark J. Daly\textsuperscript{4,5},
-Matti Pirinen\textsuperscript{6},
+Yosuke Tanigawa\textsuperscript{1},
+Chris C. A. Spencer\textsuperscript{2},
+Timothy Poterba\textsuperscript{3},
+Mark J. Daly\textsuperscript{3,4},
+Matti Pirinen\textsuperscript{5},
 Manuel A. Rivas\textsuperscript{1*},
@@
 \\
-\textbf{2} Biomedical Informatics Training Program, Stanford University, Stanford, CA, USA
+\textbf{2} Genomics plc, Oxford, UK
 \\
-\textbf{3} Genomics plc, Oxford, UK
+\textbf{3} Broad Institute of MIT and Harvard, Cambridge, MA, USA
 \\
-\textbf{4} Broad Institute of MIT and Harvard, Cambridge, MA, USA
+\textbf{4} Analytic and Translational Genetics Unit, Massachusetts General Hospital, Boston, MA, USA
 \\
-\textbf{5} Wellcome Trust Centre for Human Genetics, University of Oxford, Oxford, UK
-\\
-\textbf{6} Analytic and Translational Genetics Unit, Massachusetts General Hospital, Boston, MA, USA
-\\
-\textbf{7} Oxford Centre for Diabetes, Endocrinology and Metabolism, Oxford, UK
-\\
-\textbf{8} Institute for Molecular Medicine Finland, University of Helsinki, Finland
+\textbf{5} Institute for Molecular Medicine Finland, University of Helsinki, Finland
 \\
@@
 % Use the asterisk to denote corresponding authorship and provide email address in note below.
+* matti.pirinen@helsinki.fi
+\newline
 * mrivas@stanford.edu
@@
 Some simple alternatives to MRP include univariate approaches for rare variant association studies including the sequence kernel association test (SKAT)\cite{skat}, and the burden test, which we show are special cases of the MRP model comparison when we assign the prior correlation of genetic effects across different variants to be zero or one.
-Furthermore, we show that MRP is able to detect the presence of heterogeneity of effects, which in some circumstances, such as those where effects are observed to be heterogeneous across studies, may indicate the presence of technical sources of error.
+Furthermore, MRP can be extended to detect the presence of heterogeneity of effects, which in some circumstances, such as those where effects are observed to be heterogeneous across studies, may indicate the presence of technical sources of error.
 
@@
 \label{glaucoma_similar}
-{\bf Results for similar effects model applied to glaucoma intra-ocular pressure, and corneal resistance factor.} $\log_{10}$ Bayes Factors from applying MRP similar effects model to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
+{\bf Results for independent effects model applied to glaucoma intra-ocular pressure, and corneal resistance factor.} $\log_{10}$ Bayes Factors from applying MRP similar effects model to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
 (E) shows the results of a joint analysis focused on finding rare variants that protect against glaucoma.
@@
 \label{glaucoma_independent}
-{\bf Results for independent effects model applied to glaucoma intra-ocular pressure, and corneal resistance factor.} $\log_{10}$ Bayes Factors from applying MRP independent effects model to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
+{\bf Results for similar effects model applied to glaucoma intra-ocular pressure, and corneal resistance factor.} $\log_{10}$ Bayes Factors from applying MRP independent effects model to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
 (E) shows the results of a joint analysis focused on finding rare variants that protect against glaucoma.
@@
 The primary and processed data used to generate the analyses presented here are available in the UK Biobank access management system (https://amsportal.ukbiobank.ac.uk/) for application 24983, ``Generating effective therapeutic hypotheses from genomic and hospital linkage data'' (http://www.ukbiobank.ac.uk/wp-content/uploads/2017/06/24983-Dr-Manuel-Rivas.pdf), and the results are displayed in the Global Biobank Engine (https://biobankengine.stanford.edu).
-We would like to thank the Customer Solutions Team from Paradigm4 who helped us implement efficient databases for queries and application of inference methods to the data, and also implemented optimized versions of truncated singular value decomposition.
+We would like to thank the Customer Solutions Team from Paradigm4 who helped us implement efficient databases for queries and application of inference methods to the data.
 M.A.R.

### 2018-01-29 tex/manuscript.tex
Updated manuscript and final figures
I added information about the variant filtering and fixed the problems
with the supporting figures.

@@
 \begin{figure}[!h]
-\includegraphics[width=\textwidth]{../figures/Figure_01.pdf}
+\includegraphics[width=\textwidth]{../figures/final/Figure_01.pdf}
 \caption{{\bf Schematic overview of MRP.}
@@
 \begin{figure}[!h]
-\includegraphics[width=\textwidth]{../figures/Figure_02.pdf}
+\includegraphics[width=\textwidth]{../figures/final/Figure_02.pdf}
 \caption{{\bf Simulation studies.}
@@
 
-\subsubsection*{UK Biobank Asthma and Glaucoma Applications}
+\subsection*{UK Biobank Asthma and Glaucoma Applications}
 We applied MRP for both sets of traits assuming both an independent effects model and a similar effects model.
@@
 In cases where genes overlapped such that they shared rare variants and therefore the same BF, we removed one gene.
+
+\subsubsection*{Variant Filtering}
+We used the $\texttt{variant\_filter\_table.tsv}$ file available at https://github.com/rivas-lab/public-resources (6f9f726) to filter variants on the UK Biobank array for use with MRP.
+We first chose variants with minor allele frequency less than $1\%$.
+We then filtered out all variants with $\texttt{all\_filters}$ less than one.
+This removes variants with missingness greater than $1\%$ (calculated on an array-specific basis for array-specific variants) or Hardy-Weinberg equilibrium $p < 10^{-7}$.
+This also removes some PTVs for which manual inspection revealed irregular cluster plots \cite{DeBoever179762}.
+We LD pruned the variants by only using variants with $\texttt{ld}$ equal to one.
+We included missense variants and PTVs indicated by the following annotations: $\texttt{missense\_variant}$,
+$\texttt{stop\_gained}$, $\texttt{frameshift\_variant}$, $\texttt{splice\_acceptor\_variant}$, $\texttt{splice\_donor\_variant}$, $\texttt{splice\_region\_variant}$, $\texttt{start\_lost}$, $\texttt{stop\_lost}$.
 
@@
 \begin{figure}[!h]
-\includegraphics[width=\textwidth]{../figures/Figure_03.pdf}
+\includegraphics[width=\textwidth]{../figures/final/Figure_03.pdf}
 \caption{{\bf Results for asthma application.}
@@
 We also applied MRP to missense variants and PTVs for glaucoma, intra-ocular pressure, and corneal resistance factor as well as performing joint analyses.
-Elevated intra-ocular pressure is a measure of the fluid pressure in the eye.
-Elevated intra-ocular pressure is associated with glaucoma risk and has been linked to genetic variants associated with glaucoma \cite{doi:10.1001/jama.2014.3192}.
+Intra-ocular pressure is a measure of the fluid pressure in the eye, is associated with glaucoma risk, and has been linked to genetic variants associated with glaucoma \cite{doi:10.1001/jama.2014.3192}.
 Corneal resistance factor is a measure of the cornea's ability to resist mechanical stress and has been associated with glaucoma presence and severity \cite{CXO:CXO414, MANSOURI2012419, GriseDulac:2012bo}.
@@
 Applying the MRP model to identify protective effects also identified \textit{ANGPTL7} and \textit{WNT10A} but not \textit{KLHL22} (Fig~\ref{glaucoma_bma}E).
-Expression of \textit{ANGPTL7} is upregulated in glaucome and has been proposed to regulate intra-ocular pressure and glaucoma risk \cite{Comes:2011ex, Kuchtey:2008ek}.
+We obtained similar results using the independent effects (\nameref{glaucoma_independent}) or similar effects models (\nameref{glaucoma_similar}).
+
+Expression of \textit{ANGPTL7} is upregulated in glaucoma and has been proposed to regulate intra-ocular pressure and glaucoma risk \cite{Comes:2011ex, Kuchtey:2008ek}.
 \textit{KLHL22} has not been previously associated with glaucoma though a suggestive association was reported for retinopathy in individuals without diabetes \cite{10.1371/journal.pone.0054232}.
@@
 \begin{figure}[!h]
-\includegraphics[width=\textwidth]{../figures/Figure_04.pdf}
+\includegraphics[width=\textwidth]{../figures/final/Figure_04.pdf}
 \caption{{\bf Results for glaucoma application.}
@@
 \begin{figure}[H]
-\includegraphics[width=\textwidth]{../figures/Figure_S01.pdf}
+\includegraphics[width=\textwidth]{../figures/final/Figure_S01.pdf}
 \end{figure}
@@
 \begin{figure}[H]
-\includegraphics[width=\textwidth]{../figures/Figure_S02.pdf}
+\includegraphics[width=\textwidth]{../figures/final/Figure_S02.pdf}
 \end{figure}
@@
 \begin{figure}[H]
-\includegraphics[width=\textwidth]{../figures/Figure_S03.pdf}
+\includegraphics[width=\textwidth]{../figures/final/Figure_S03.pdf}
 \end{figure}
@@
 \begin{figure}[H]
-\includegraphics[width=\textwidth]{../figures/Figure_S04.pdf}
+\includegraphics[width=\textwidth]{../figures/final/Figure_S04.pdf}
 \end{figure}
 \paragraph*{S4 Fig.}
-\label{glaucoma_similar}
-{\bf Results for similar effects model applied to glaucoma intra-ocular pressure, and corneal resistance factor.} $\log_{10}$ Bayes Factors from applying MRP similar effects model to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
+\label{glaucoma_independent}
+{\bf Results for independent effects model applied to glaucoma intra-ocular pressure, and corneal resistance factor.} $\log_{10}$ Bayes Factors from applying MRP independent effects model to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
 (E) shows the results of a joint analysis focused on finding rare variants that protect against glaucoma.
@@
 \begin{figure}[H]
-\includegraphics[width=\textwidth]{../figures/Figure_S05.pdf}
+\includegraphics[width=\textwidth]{../figures/final/Figure_S05.pdf}
 \end{figure}
 \paragraph*{S5 Fig.}
-\label{glaucoma_independent}
-{\bf Results for independent effects model applied to glaucoma intra-ocular pressure, and corneal resistance factor.} $\log_{10}$ Bayes Factors from applying MRP independent effects model to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
+\label{glaucoma_similar}
+{\bf Results for similar effects model applied to glaucoma intra-ocular pressure, and corneal resistance factor.} $\log_{10}$ Bayes Factors from applying MRP similar effects model to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
 (E) shows the results of a joint analysis focused on finding rare variants that protect against glaucoma.

### 2018-01-29 tex/manuscript.tex
Updates for comments from MP

@@
 Mark J. Daly\textsuperscript{3,4},
-Matti Pirinen\textsuperscript{5},
+Matti Pirinen\textsuperscript{5,6,7*},
 Manuel A. Rivas\textsuperscript{1*},
@@
 \\
-\textbf{5} Institute for Molecular Medicine Finland, University of Helsinki, Finland
+\textbf{5} Institute for Molecular Medicine Finland (FIMM), University of Helsinki, Helsinki, Finland
 \\
+\textbf{6} Department of Public Health, University of Helsinki, Helsinki, Finland
+\\
+\textbf{7} Helsinki Institute for Information Technology HIIT and Department of Mathematics and Statistics, University of Helsinki, Helsinki, Finland
 % \textbf{1} Affiliation Dept/Program/Center, Institution Name, City, State, Country
@@
 B: Comparison of log10(Bayes Factors) obtained when raw genotype and phenotype data is available to a scenario where summary statistics only was available and similar effects across studies is assumed.
-C: From single variant and single phenotype to multiple variants and multiple phenotypes gene discovery: ROC curves for detecting gene association to any of the phenotypes using single variant/single phenotype association (blue) to multiple variants and multiple phenotypes association (purple).
+C: From single variant and single phenotype to multiple variants and multiple phenotypes gene discovery: ROC curves for detecting gene association to any of the phenotypes using single variant/single phenotype association (green) to multiple variants and multiple phenotypes association (purple).
 D: ROC curves for detecting gene association when incorporating prior mean of genetic effects (orange) to identify protective alleles.}
@@
 \subsubsection*{Genetic Correlations}
-We calculated the genetic between the two groups of traits (asthma, eosinophil counts, FVC, FEV$_1$ and glaucoma, intra-ocular pressure, corneal resistance factor) using the MultiVariate Polygenic Mixture Model (MVPMM)~\cite{DeBoever2017}.
+We calculated the genetic correlation between the two groups of traits (asthma, eosinophil counts, FVC, FEV$_1$ and glaucoma, intra-ocular pressure, corneal resistance factor) using the MultiVariate Polygenic Mixture Model (MVPMM)~\cite{DeBoever2017}.
 Briefly, MVPMM estimates genetic correlation given GWAS summary statistics (effect size and standard error of effect size estimate) by modeling GWAS summary statistics as generated from one of two mixture components.
@@
 The GWAS summary statistics for the rare variants in \textit{ANGPTL7} suggest that two missense variants, rs28991002 and rs28991009, appear to drive the association with glaucoma (Fig~\ref{glaucoma_bma}F).
-These results illustrate that applying model comparison to multiple traits simultaneously can prioritize the discovery of protective modifiers of disease.
 
@@
 We thank all the participants in the UK Biobank study.
+M.P.
+is financially supported by the Academy of Finland [288509 and 294050].
 M.A.R.
@@
 M.A.R.
-is a paid consultant in Genomics PLC.
+and M.P.
+are paid consultants in Genomics PLC.
 

### 2018-01-29 tex/manuscriptbiorxiv.tex
Updates for comments from MP

@@
 
-\documentclass[10pt]{article}
+\documentclass[10pt,letterpaper]{article}
 %\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
@@
 %\setlength{\headheight}{27.023pt}
-%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
+\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
 \rfoot{\thepage/\pageref{LastPage}}
@@
 \begin{table}[!ht]
-\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
 \centering
@@
 \label{asthma_table}
-\end{adjustwidth}
+%\end{adjustwidth}
 \end{table}
@@
 Applying the MRP model to identify protective effects also identified \textit{ANGPTL7} and \textit{WNT10A} but not \textit{KLHL22} (Fig~\ref{glaucoma_bma}E).
-Expression of \textit{ANGPTL7} is upregulated in glaucome and has been proposed to regulate intra-ocular pressure and glaucoma risk \cite{Comes:2011ex, Kuchtey:2008ek}.
+We obtained similar results using the independent effects (\nameref{glaucoma_independent}) or similar effects models (\nameref{glaucoma_similar}).
+
+Expression of \textit{ANGPTL7} is upregulated in glaucoma and has been proposed to regulate intra-ocular pressure and glaucoma risk \cite{Comes:2011ex, Kuchtey:2008ek}.
 \textit{KLHL22} has not been previously associated with glaucoma though a suggestive association was reported for retinopathy in individuals without diabetes \cite{10.1371/journal.pone.0054232}.
 The GWAS summary statistics for the rare variants in \textit{ANGPTL7} suggest that two missense variants, rs28991002 and rs28991009, appear to drive the association with glaucoma (Fig~\ref{glaucoma_bma}F).
-These results illustrate that applying model comparison to multiple traits simultaneously can prioritize the discovery of protective modifiers of disease.
 
 \begin{figure}[!h]
-\includegraphics[width=\textwidth]{../figures/Figure_04.pdf}
+\includegraphics[width=\textwidth]{../figures/final/Figure_04.pdf}
 \caption{{\bf Results for glaucoma application.}
@@
 \begin{table}[!ht]
-\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+\begin{adjustwidth}{-1.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
 \centering
@@
 \begin{figure}[H]
-\includegraphics[width=\textwidth]{../figures/Figure_S04.pdf}
+\includegraphics[width=\textwidth]{../figures/final/Figure_S04.pdf}
 \end{figure}
 \paragraph*{S4 Fig.}
-\label{glaucoma_similar}
-{\bf Results for independent effects model applied to glaucoma intra-ocular pressure, and corneal resistance factor.} $\log_{10}$ Bayes Factors from applying MRP similar effects model to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
+\label{glaucoma_independent}
+{\bf Results for independent effects model applied to glaucoma intra-ocular pressure, and corneal resistance factor.} $\log_{10}$ Bayes Factors from applying MRP independent effects model to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
 (E) shows the results of a joint analysis focused on finding rare variants that protect against glaucoma.
@@
 \begin{figure}[H]
-\includegraphics[width=\textwidth]{../figures/Figure_S05.pdf}
+\includegraphics[width=\textwidth]{../figures/final/Figure_S05.pdf}
 \end{figure}
 \paragraph*{S5 Fig.}
-\label{glaucoma_independent}
-{\bf Results for similar effects model applied to glaucoma intra-ocular pressure, and corneal resistance factor.} $\log_{10}$ Bayes Factors from applying MRP independent effects model to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
+\label{glaucoma_similar}
+{\bf Results for similar effects model applied to glaucoma intra-ocular pressure, and corneal resistance factor.} $\log_{10}$ Bayes Factors from applying MRP similar effects model to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
 (E) shows the results of a joint analysis focused on finding rare variants that protect against glaucoma.

### 2018-01-30 tex/manuscript.tex
Manuscript updates

@@
 Here we introduce a model comparison approach we refer to as MRP for rare variant association studies that considers correlation, scale, and location of genetic effects across a group of genetic variants, phenotypes, and studies.
-We consider the use of summary statistic data to apply 1) standard univariate and multivariate gene-based meta-analysis models, 2) models for assessing heterogeneity of genetic effects, and 3) models for identifying protective protein-truncating variants that can expedite drug discovery.
-Through simulation studies, we demonstrate that the proposed model comparison approach may improve ability to detect rare variant association signals.
-Finally, we demonstrate the model on two groups of phenotypes from the UK Biobank: 1) asthma diagnosis, eosinophil counts, forced expiratory volume, and forced vital capacity; and 2) glaucoma diagnosis, intra-ocular pressure, and corneal resistance factor.
-We demonstrate that the model leverages information from multiple phenotypes to identify known and novel rare variant associations for asthma and glaucoma.
-Overall the model comparison approach is able to retain useful features from widely-used meta-analysis approaches and prioritize protective modifiers of disease risk.
+We consider the use of summary statistic data to apply univariate and multivariate gene-based meta-analysis models for identifying rare variant associations with an emphasis on protective protein-truncating variants that can expedite drug discovery.
+Through simulation studies, we demonstrate that the proposed model comparison approach can improve ability to detect rare variant association signals.
+Finally, we apply the model to two groups of phenotypes from the UK Biobank: 1) asthma diagnosis, eosinophil counts, forced expiratory volume, and forced vital capacity; and 2) glaucoma diagnosis, intra-ocular pressure, and corneal resistance factor.
+We are able to recover known associations such as the protective association between rs146597587 in \textit{IL33} and asthma.
+We also find evidence for an association between rare variants in the drug target \textit{CCR3} and asthma.
+We identified novel protective associations between rare variants in \textit{ANGPTL7} and glaucoma as well.
+Overall, we show that the model comparison approach is able to retain useful features from widely-used meta-analysis approaches and prioritize protective modifiers of disease risk.
 
@@
 Some simple alternatives to MRP include univariate approaches for rare variant association studies including the sequence kernel association test (SKAT)\cite{skat}, and the burden test, which we show are special cases of the MRP model comparison when we assign the prior correlation of genetic effects across different variants to be zero or one.
-Furthermore, MRP can be extended to detect the presence of heterogeneity of effects, which in some circumstances, such as those where effects are observed to be heterogeneous across studies, may indicate the presence of technical sources of error.
 
@@
 
-We also found moderate evidence for associations between rare coding variants in \textit{CCR3} and asthma.
+We also found moderate evidence for a protective association between rare coding variants in \textit{CCR3} and asthma.
 The $\log_{10}$ BFs for \textit{CCR3} were 3.3 and 3.1 in the joint and joint protective analyses respectively compared to only -0.5 in the asthma-only analysis  (Fig~\ref{asthma_bma}, Table~\ref{asthma_table}).
+\textit{CCR3} is a chemokine receptor that is highly expressed on eosinophils and has been a therapeutic focus for asthma \cite{Neighbour:2014jw, Pease:2014em}.
 \textit{CCR3} was not reported in a large GWAS for allergic disease including asthma\cite{Ferreira:2017ba} though \textit{CCR3} is near a locus associated with atopy in a previous meta-analysis\cite{Ober:2011jk}.
-Other genes like \textit{IL18RAP}, \textit{ATP2A3}, and \textit{FLG} had $\log_{10}$ BFs greater than 4 in the asthma-only analysis but much smaller BFs in the joint analyses indicating that rare variants in these genes are less likely to affect this group of traits.
+These results demonstrate that MRP can identify biologically meaningful therapeutic targets that may be missed by standard GWAS approaches.
+
+Considering multiple phenotypes jointly allows for the efficient prioritization of disease genes.
+For instance, some genes like \textit{IL18RAP}, \textit{ATP2A3}, and \textit{FLG} had $\log_{10}$ BFs greater than 4 in the asthma-only analysis but much smaller BFs in the joint analyses indicating that rare variants in these genes are less likely to affect this group of traits.
 Similarly, there were other genes like \textit{RP11-39K24.9} and \textit{IL17RA} that had relatively larger BFs in the eosinophil count-only analysis but small BFs for the joint analyses demonstrating MRP's ability to integrate information across all phenotypes considered.
@@
 \end{figure}
-
-\begin{table}[!ht]
-\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
-\centering
-\caption{
-{\bf $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{asthma_bma}.}}
-\begin{tabular}{|l+l|l|l|l|l|l|}
-\hline
-{\bf Gene} & {\bf Joint, protective} & {\bf Joint} & {\bf Eosinophil count} & {\bf FVC} & {\bf FEV$_1$} & {\bf Asthma} \\
-\thickhline
-\textit{IL33}         &              29.4 &   29.3 &             30.6 &  -2.3 &  -2.2 &    8.1 \\
-\hline
-\textit{CCR3}         &               3.1 &    3.3 &              7.4 &  -1.4 &  -1.6 &   -0.5 \\
-\hline
-\textit{RP11-39K24.9} &               0.8 &    1.8 &              4.9 &  -0.1 &  -0.4 &    0.3 \\
-\hline
-\textit{SCMH1}        &               0.5 &    4.7 &             -1.5 &   7.7 &   3.8 &   -0.7 \\
-\hline
-\textit{MUSTN1}       &               0.4 &    1.1 &             -1.2 &   2.9 &   2.9 &   -0.6 \\
-\hline
-\textit{ZFAT}         &               0.3 &    1.3 &             -2.0 &   4.7 &   3.1 &   -0.4 \\
-\hline
-\textit{ELN}          &               0.2 &    2.5 &             -1.0 &   5.6 &   2.9 &   -0.6 \\
-\hline
-\textit{C14orf39}     &              -0.7 &   -0.0 &             -1.1 &   3.5 &   2.5 &    0.0 \\
-\hline
-\textit{TMEM110}      &              -0.9 &    1.1 &             -1.0 &   3.3 &   3.1 &   -0.6 \\
-\hline
-\textit{IL17RA}       &              -4.4 &    3.1 &              7.9 &  -2.7 &  -2.5 &   -1.1 \\
-\hline
-\textit{IL18RAP}      &              -9.6 &   -0.9 &             -1.0 &  -1.6 &  -1.7 &    3.5 \\
-\hline
-\textit{ATP2A3}       &             -11.9 &   -0.8 &             -1.2 &  -2.1 &  -2.2 &    5.8 \\
-\hline
-\textit{FLG}          &             -20.1 &  -17.2 &             -6.0 &  -7.4 &  -8.5 &    4.0 \\
-\hline
-\end{tabular}
-\label{asthma_table}
-\end{adjustwidth}
-\end{table}
 
@@
 \end{figure}
-
-% %% TABLE: GLAUCOMA GENES OF INTEREST
-\begin{table}[!ht]
-\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
-\centering
-\caption{
-{\bf $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{glaucoma_bma}.}}
-\begin{tabular}{|l+p{3cm}|p{1cm}|p{1.6cm}|p{3.8cm}|p{3.4cm}|}
-\hline
-{\bf Gene} & {\bf Joint, protective} & {\bf Joint} & {\bf Glaucoma} & {\bf Intra-ocular pressure, Goldmann-correlated} & {\bf Corneal resistance factor} \\
-\thickhline
-\textit{ANGPTL7} &              10.1 &   9.4 &      1.7 &                                       10.1 &                       6.2 \\
-\hline
-\textit{KLHL22}  &               3.1 &   3.6 &     -0.2 &                                        3.9 &                       2.2 \\
-\hline
-\textit{GLI3}    &               2.9 &   2.9 &     -0.3 &                                       -0.7 &                       3.2 \\
-\hline
-\textit{WNT10A}  &               2.2 &   3.0 &     -0.3 &                                       -0.5 &                       3.5 \\
-\hline
-\textit{RNASE2}  &              -1.3 &  -0.3 &      0.4 &                                       -1.0 &                      -1.1 \\
-\hline
-\textit{FUK}     &              -6.5 &  -2.3 &      2.5 &                                       -2.1 &                      -2.1 \\
-\hline
-\end{tabular}
-\label{table2}
-\end{adjustwidth}
-\end{table}
 
@@
 
+% %% TABLE: GLAUCOMA GENES OF INTEREST
+\begin{table}[!ht]
+\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+\centering
+\caption{
+{\bf $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{glaucoma_bma}.}}
+\begin{tabular}{|l+p{3cm}|p{1cm}|p{1.6cm}|p{3.8cm}|p{3.4cm}|}
+\hline
+{\bf Gene} & {\bf Joint, protective} & {\bf Joint} & {\bf Glaucoma} & {\bf Intra-ocular pressure, Goldmann-correlated} & {\bf Corneal resistance factor} \\
+\thickhline
+\textit{ANGPTL7} &              10.1 &   9.4 &      1.7 &                                       10.1 &                       6.2 \\
+\hline
+\textit{KLHL22}  &               3.1 &   3.6 &     -0.2 &                                        3.9 &                       2.2 \\
+\hline
+\textit{GLI3}    &               2.9 &   2.9 &     -0.3 &                                       -0.7 &                       3.2 \\
+\hline
+\textit{WNT10A}  &               2.2 &   3.0 &     -0.3 &                                       -0.5 &                       3.5 \\
+\hline
+\textit{RNASE2}  &              -1.3 &  -0.3 &      0.4 &                                       -1.0 &                      -1.1 \\
+\hline
+\textit{FUK}     &              -6.5 &  -2.3 &      2.5 &                                       -2.1 &                      -2.1 \\
+\hline
+\end{tabular}
+\label{table2}
+\end{adjustwidth}
+\end{table}
+
 %% FIGURE: ASTHMA MANHATTAN AND FOREST FOR INDEPENDENT EFFECTS
@@
 Only $\log_{10}$ Bayes Factors greater than -5 are plotted.
+
+\begin{table}[!ht]
+\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+\centering
+\caption{
+{\bf $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{asthma_bma}.}}
+\begin{tabular}{|l+l|l|l|l|l|l|}
+\hline
+{\bf Gene} & {\bf Joint, protective} & {\bf Joint} & {\bf Eosinophil count} & {\bf FVC} & {\bf FEV$_1$} & {\bf Asthma} \\
+\thickhline
+\textit{IL33}         &              29.4 &   29.3 &             30.6 &  -2.3 &  -2.2 &    8.1 \\
+\hline
+\textit{CCR3}         &               3.1 &    3.3 &              7.4 &  -1.4 &  -1.6 &   -0.5 \\
+\hline
+\textit{RP11-39K24.9} &               0.8 &    1.8 &              4.9 &  -0.1 &  -0.4 &    0.3 \\
+\hline
+\textit{SCMH1}        &               0.5 &    4.7 &             -1.5 &   7.7 &   3.8 &   -0.7 \\
+\hline
+\textit{MUSTN1}       &               0.4 &    1.1 &             -1.2 &   2.9 &   2.9 &   -0.6 \\
+\hline
+\textit{ZFAT}         &               0.3 &    1.3 &             -2.0 &   4.7 &   3.1 &   -0.4 \\
+\hline
+\textit{ELN}          &               0.2 &    2.5 &             -1.0 &   5.6 &   2.9 &   -0.6 \\
+\hline
+\textit{C14orf39}     &              -0.7 &   -0.0 &             -1.1 &   3.5 &   2.5 &    0.0 \\
+\hline
+\textit{TMEM110}      &              -0.9 &    1.1 &             -1.0 &   3.3 &   3.1 &   -0.6 \\
+\hline
+\textit{IL17RA}       &              -4.4 &    3.1 &              7.9 &  -2.7 &  -2.5 &   -1.1 \\
+\hline
+\textit{IL18RAP}      &              -9.6 &   -0.9 &             -1.0 &  -1.6 &  -1.7 &    3.5 \\
+\hline
+\textit{ATP2A3}       &             -11.9 &   -0.8 &             -1.2 &  -2.1 &  -2.2 &    5.8 \\
+\hline
+\textit{FLG}          &             -20.1 &  -17.2 &             -6.0 &  -7.4 &  -8.5 &    4.0 \\
+\hline
+\end{tabular}
+\label{asthma_table}
+\end{adjustwidth}
+\end{table}
 

### 2018-01-30 tex/manuscript.tex
Updates

@@
 Through simulation studies, we demonstrate that the proposed model comparison approach can improve ability to detect rare variant association signals.
-Finally, we apply the model to two groups of phenotypes from the UK Biobank: 1) asthma diagnosis, eosinophil counts, forced expiratory volume, and forced vital capacity; and 2) glaucoma diagnosis, intra-ocular pressure, and corneal resistance factor.
+We also apply the model to two groups of phenotypes from the UK Biobank: 1) asthma diagnosis, eosinophil counts, forced expiratory volume, and forced vital capacity; and 2) glaucoma diagnosis, intra-ocular pressure, and corneal resistance factor.
 We are able to recover known associations such as the protective association between rs146597587 in \textit{IL33} and asthma.
-We also find evidence for an association between rare variants in the drug target \textit{CCR3} and asthma.
-We identified novel protective associations between rare variants in \textit{ANGPTL7} and glaucoma as well.
-Overall, we show that the model comparison approach is able to retain useful features from widely-used meta-analysis approaches and prioritize protective modifiers of disease risk.
+We also find evidence for an association between rare variants in the drug target \textit{CCR3} and asthma as well as novel protective associations between rare variants in \textit{ANGPTL7} and \textit{WNT10A} and glaucoma.
+Overall, we show that the MRP model comparison approach is able to retain and improve upon useful features from widely-used meta-analysis approaches for rare variant association analyses and prioritize protective modifiers of disease risk.
 
@@
 We recover the reported association between a rare PTV in \textit{IL33} and asthma as well as a new association between \textit{CCR3} and asthma \cite{DeBoever179762, 10.1371/journal.pgen.1006659}.
-We also applied MRP to glaucoma (HC276), intra-ocular pressure (INI5263), and corneal resistance factor (INI5265) and find evidence that rare coding variants in \textit{ANGPTL7} and \textit{KLHL22} may protect against glaucoma.
+We also applied MRP to glaucoma (HC276), intra-ocular pressure (INI5263), and corneal resistance factor (INI5265) and find evidence that rare coding variants in \textit{ANGPTL7} and \textit{WNT10A} may protect against glaucoma.
 These analyses show that MRP recovers results from typical single variant-single phenotype analyses while identifying new rare variant associations that include protective modifiers of disease risk.
@@
 Summary statistics from variants in the non-null component are modeled as being drawn from a multivariate normal distribution with zero mean, but the covariance matrix for the non-null component combines the covariance matrix from the null component with another covariance matrix that captures the genetic correlation between the phenotypes being considered.
+We observed similar genetic correlations using LD score regression %(\nameref{gcorr_fig}) \cite{gencorr2015}.
 
@@
 In cases where genes overlapped such that they shared rare variants and therefore the same BF, we removed one gene.
+\textit{ANGPTL7} protein expression was assessed using the HIPED protein expression database accessed through genecards.org on 2017/1/29 \cite{Fishilevich:2016hx}.
 
@@
 We applied the MRP model comparison approach to summary statistic data generated from single variant logistic regression and linear regression analysis for coding variants on the UK Biobank array (Methods).
-We applied MRP to asthma and three related traits and glaucoma and two related traits.
+We applied MRP separately to asthma and three related traits as well as glaucoma and two related traits.
 
@@
 We applied both independent and similar effects models and used Bayesian model averaging to compute a single BF per gene \cite{10.2307/2676803}.
-In agreement with previous studies, we observed evidence that rare missense variants and/or PTVs in \textit{IL33} affect eosinophil counts and offer protection from asthma from the single-phenotype analyses, though the evidence of association was strongest for the joint analysis (Table~\ref{asthma_table}) \cite{DeBoever179762, 10.1371/journal.pgen.1006659}.
-We also performed an analysis focused on identifying protective variants which also identified the \textit{IL33} association (Fig~\ref{asthma_bma}).
+In agreement with previous studies, we observed evidence that rare missense variants and/or PTVs in \textit{IL33} affect eosinophil counts and offer protection from asthma from the single-phenotype analyses, though the evidence of association was strongest for the joint analysis (\nameref{asthma_table}) \cite{DeBoever179762, 10.1371/journal.pgen.1006659}.
+We performed an analysis focused on identifying protective variants which also identified the \textit{IL33} association (Fig~\ref{asthma_bma}F).
 The results were similar using only either the independent effects (\nameref{asthma_independent}) or similar effects models (\nameref{asthma_similar}).
-We can inspect the effect sizes from the marginal GWAS regressions for the rare variants included in the analysis and see that the association identified by MRP is likely driven by the PTV rs146597587 (Fig~\ref{asthma_bma}G).
+We inspected the effect sizes from the marginal GWAS regressions for the rare variants included in the analysis and found that the association identified by MRP is likely driven by the PTV rs146597587 (Fig~\ref{asthma_bma}G).
 
 We also found moderate evidence for a protective association between rare coding variants in \textit{CCR3} and asthma.
-The $\log_{10}$ BFs for \textit{CCR3} were 3.3 and 3.1 in the joint and joint protective analyses respectively compared to only -0.5 in the asthma-only analysis  (Fig~\ref{asthma_bma}, Table~\ref{asthma_table}).
+The $\log_{10}$ BFs for \textit{CCR3} were 3.3 and 3.1 in the joint and joint protective analyses respectively compared to only -0.5 in the asthma-only analysis  (Fig~\ref{asthma_bma}, \nameref{asthma_table}).
 \textit{CCR3} is a chemokine receptor that is highly expressed on eosinophils and has been a therapeutic focus for asthma \cite{Neighbour:2014jw, Pease:2014em}.
@@
 Corneal resistance factor is a measure of the cornea's ability to resist mechanical stress and has been associated with glaucoma presence and severity \cite{CXO:CXO414, MANSOURI2012419, GriseDulac:2012bo}.
-While the individual glaucoma analysis did not yield any associations with $\log_{10}$ BF greater than three, the joint analysis identified rare coding variants in both \textit{ANGPTL7}, \textit{WNT10A}, and \textit{KLHL22} as associated with glaucoma (Fig~\ref{glaucoma_bma}A-D).
-Applying the MRP model to identify protective effects also identified \textit{ANGPTL7} and \textit{WNT10A} but not \textit{KLHL22} (Fig~\ref{glaucoma_bma}E).
+While the individual glaucoma analysis did not yield any associations with $\log_{10}$ BF greater than three, the joint analysis identified rare coding variants in both \textit{ANGPTL7} and \textit{WNT10A} as associated with glaucoma (Fig~\ref{glaucoma_bma}A-D, \nameref{glaucoma_table}).
+Applying the MRP model to identify protective effects also identified \textit{ANGPTL7} and \textit{WNT10A} (Fig~\ref{glaucoma_bma}E).
 We obtained similar results using the independent effects (\nameref{glaucoma_independent}) or similar effects models (\nameref{glaucoma_similar}).
@@
 Expression of \textit{ANGPTL7} is upregulated in glaucoma and has been proposed to regulate intra-ocular pressure and glaucoma risk \cite{Comes:2011ex, Kuchtey:2008ek}.
-\textit{KLHL22} has not been previously associated with glaucoma though a suggestive association was reported for retinopathy in individuals without diabetes \cite{10.1371/journal.pone.0054232}.
 The GWAS summary statistics for the rare variants in \textit{ANGPTL7} suggest that two missense variants, rs28991002 and rs28991009, appear to drive the association with glaucoma (Fig~\ref{glaucoma_bma}F).
+According to the HIPED protein expression database, \textit{ANGPTL7} protein is expressed at $\sim0.7$ parts per million in vitreous humor, the material between the lens and retina of the eyeball; in contrast, the expression of \textit{ANGPTL7} protein is less than $0.01$ parts per million in 68 other normal tissues \cite{Fishilevich:2016hx}.
+Such tissue-specific activity may make \textit{ANGPTL7} a useful therapeutic target.
+\textit{WNT10A} has not been previously associated with glaucoma though an exonic variant rs121908120 in \textit{WNT10A} is associated with central cornea thickness and increased risk of keratoconus, a disease of the cornea, indicating that this gene may play a role in ocular diseases.
+
+%\textit{KLHL22} has not been previously associated with glaucoma though a suggestive association was reported for retinopathy in individuals without diabetes \cite{10.1371/journal.pone.0054232}.
 
@@
 We also applied the MRP model comparison framework to summary statistic data from two groups of traits from the UK Biobank: asthma diagnosis, eosinophil counts, FEV$_1$, and FVC; and glaucoma diagnosis, intra-ocular pressure, and corneal resistance factor.
-We identified strong evidence for the previously described association between the PTV rs146597587 in \textit{IL33} and asthma\cite{DeBoever179762,10.1371/journal.pgen.1006659} as well as evidence for a potentially novel association for \textit{CCR3} (Table~\ref{asthma_table}).
+We identified strong evidence for the previously described association between the PTV rs146597587 in \textit{IL33} and asthma\cite{DeBoever179762,10.1371/journal.pgen.1006659} as well as evidence for a potentially novel genetic association for the drug target \textit{CCR3} (\nameref{asthma_table}).
 We also identified genetic evidence for a link between rare variants in \textit{ANGPTL7} and glaucoma, consistent with previous experiments that suggested a role for \textit{ANGPTL7} in glaucoma \cite{Comes:2011ex, Kuchtey:2008ek}.
@@
 
-% %% TABLE: GLAUCOMA GENES OF INTEREST
-\begin{table}[!ht]
-\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
-\centering
-\caption{
-{\bf $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{glaucoma_bma}.}}
-\begin{tabular}{|l+p{3cm}|p{1cm}|p{1.6cm}|p{3.8cm}|p{3.4cm}|}
-\hline
-{\bf Gene} & {\bf Joint, protective} & {\bf Joint} & {\bf Glaucoma} & {\bf Intra-ocular pressure, Goldmann-correlated} & {\bf Corneal resistance factor} \\
-\thickhline
-\textit{ANGPTL7} &              10.1 &   9.4 &      1.7 &                                       10.1 &                       6.2 \\
-\hline
-\textit{KLHL22}  &               3.1 &   3.6 &     -0.2 &                                        3.9 &                       2.2 \\
-\hline
-\textit{GLI3}    &               2.9 &   2.9 &     -0.3 &                                       -0.7 &                       3.2 \\
-\hline
-\textit{WNT10A}  &               2.2 &   3.0 &     -0.3 &                                       -0.5 &                       3.5 \\
-\hline
-\textit{RNASE2}  &              -1.3 &  -0.3 &      0.4 &                                       -1.0 &                      -1.1 \\
-\hline
-\textit{FUK}     &              -6.5 &  -2.3 &      2.5 &                                       -2.1 &                      -2.1 \\
-\hline
-\end{tabular}
-\label{table2}
-\end{adjustwidth}
-\end{table}
-
-%% FIGURE: ASTHMA MANHATTAN AND FOREST FOR INDEPENDENT EFFECTS
 \begin{figure}[H]
@@
 \paragraph*{S2 Fig.}
+\label{gcorr_fig}
+{\bf Genetic correlations.} Genetic correlations for (A) asthma and related phenotypes and (B) glaucoma and related phenotypes estimated using MVPMM.
+Genetic correlations for (C) asthma and related phenotypes and (D) glaucoma and related phenotypes estimated using LD score regression.
+
+%% FIGURE: ASTHMA MANHATTAN AND FOREST FOR INDEPENDENT EFFECTS
+\begin{figure}[H]
+\includegraphics[width=\textwidth]{../figures/final/Figure_S03.pdf}
+\end{figure}
+\paragraph*{S3 Fig.}
 \label{asthma_independent}
@@
 \begin{figure}[H]
-\includegraphics[width=\textwidth]{../figures/final/Figure_S03.pdf}
+\includegraphics[width=\textwidth]{../figures/final/Figure_S04.pdf}
 \end{figure}
-\paragraph*{S3 Fig.}
+\paragraph*{S4 Fig.}
 \label{asthma_similar}
@@
 
+\begin{figure}[H]
+\includegraphics[width=\textwidth]{../figures/final/Figure_S05.pdf}
+\end{figure}
+\paragraph*{S5 Fig.}
+\label{glaucoma_independent}
+{\bf Results for independent effects model applied to glaucoma intra-ocular pressure, and corneal resistance factor.} $\log_{10}$ Bayes Factors from applying MRP independent effects model to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
+(E) shows the results of a joint analysis focused on finding rare variants that protect against glaucoma.
+The genes outside of chromosome 6 with with Bayes Factor greater than three are indicated by arrows.
+Only $\log_{10}$ Bayes Factors greater than zero are plotted.
+
+\begin{figure}[H]
+\includegraphics[width=\textwidth]{../figures/final/Figure_S06.pdf}
+\end{figure}
+\paragraph*{S6 Fig.}
+\label{glaucoma_similar}
+{\bf Results for similar effects model applied to glaucoma intra-ocular pressure, and corneal resistance factor.} $\log_{10}$ Bayes Factors from applying MRP similar effects model to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
+(E) shows the results of a joint analysis focused on finding rare variants that protect against glaucoma.
+The genes outside of chromosome 6 with with Bayes Factor greater than three are indicated by arrows.
+Only $\log_{10}$ Bayes Factors greater than zero are plotted.
+
 \begin{table}[!ht]
@@
 \centering
-\caption{
-{\bf $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{asthma_bma}.}}
+%\caption{
+%{\bf $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{asthma_bma}.}}
 \begin{tabular}{|l+l|l|l|l|l|l|}
@@
 \end{table}
+\paragraph*{S1 Table.}
+\label{asthma_table}
+{\bf Highlighted genes from asthma analysis.} $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{asthma_bma}.
 
-\begin{figure}[H]
-\includegraphics[width=\textwidth]{../figures/final/Figure_S04.pdf}
-\end{figure}
-\paragraph*{S4 Fig.}
-\label{glaucoma_independent}
-{\bf Results for independent effects model applied to glaucoma intra-ocular pressure, and corneal resistance factor.} $\log_{10}$ Bayes Factors from applying MRP independent effects model to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
-(E) shows the results of a joint analysis focused on finding rare variants that protect against glaucoma.
-The genes outside of chromosome 6 with with Bayes Factor greater than three are indicated by arrows.
-Only $\log_{10}$ Bayes Factors greater than zero are plotted.
-
-\begin{figure}[H]
-\includegraphics[width=\textwidth]{../figures/final/Figure_S05.pdf}
-\end{figure}
-\paragraph*{S5 Fig.}
-\label{glaucoma_similar}
-{\bf Results for similar effects model applied to glaucoma intra-ocular pressure, and corneal resistance factor.} $\log_{10}$ Bayes Factors from applying MRP similar effects model to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
-(E) shows the results of a joint analysis focused on finding rare variants that protect against glaucoma.
-The genes outside of chromosome 6 with with Bayes Factor greater than three are indicated by arrows.
-Only $\log_{10}$ Bayes Factors greater than zero are plotted.
+% %% TABLE: GLAUCOMA GENES OF INTEREST
+\begin{table}[!ht]
+\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+\centering
+%\caption{
+%{\bf $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{glaucoma_bma}.}}
+\begin{tabular}{|l+p{3cm}|p{1cm}|p{1.6cm}|p{3.8cm}|p{3.4cm}|}
+\hline
+{\bf Gene} & {\bf Joint, protective} & {\bf Joint} & {\bf Glaucoma} & {\bf Intra-ocular pressure, Goldmann-correlated} & {\bf Corneal resistance factor} \\
+\thickhline
+\textit{ANGPTL7} &              11.9 &  11.9 &      1.7 &                                       10.1 &                       6.2 \\
+\hline
+\textit{KLHL22}  &               3.5 &   3.5 &     -0.2 &                                        3.9 &                       2.2 \\
+\hline
+\textit{WNT10A}  &               2.5 &   2.5 &     -0.3 &                                       -0.5 &                       3.5 \\
+\hline
+\textit{GLI3}    &               1.5 &   1.5 &     -0.3 &                                       -0.7 &                       3.2 \\
+\hline
+\textit{RNASE2}  &              -0.5 &  -0.5 &      0.4 &                                       -1.0 &                      -1.1 \\
+\hline
+\textit{FUK}     &              -1.8 &  -1.8 &      2.5 &                                       -2.1 &                      -2.1 \\
+\hline
+\end{tabular}
+\end{adjustwidth}
+\end{table}
+\paragraph*{S2 Table.}
+\label{glaucoma_table}
+{\bf Highlighted genes from glaucoma analysis.} $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{glaucoma_bma}.
 

### 2018-01-30 tex/manuscriptbiorxiv.tex
Updates

@@
 \pagestyle{myheadings}
-%\pagestyle{fancy}
+\pagestyle{fancy}
 \fancyhf{}
 %\setlength{\headheight}{27.023pt}
-\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
+%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
 \rfoot{\thepage/\pageref{LastPage}}
@@
 Here we introduce a model comparison approach we refer to as MRP for rare variant association studies that considers correlation, scale, and location of genetic effects across a group of genetic variants, phenotypes, and studies.
-We consider the use of summary statistic data to apply 1) standard univariate and multivariate gene-based meta-analysis models, 2) models for assessing heterogeneity of genetic effects, and 3) models for identifying protective protein-truncating variants that can expedite drug discovery.
-Through simulation studies, we demonstrate that the proposed model comparison approach may improve ability to detect rare variant association signals.
-Finally, we demonstrate the model on two groups of phenotypes from the UK Biobank: 1) asthma diagnosis, eosinophil counts, forced expiratory volume, and forced vital capacity; and 2) glaucoma diagnosis, intra-ocular pressure, and corneal resistance factor.
-We demonstrate that the model leverages information from multiple phenotypes to identify known and novel rare variant associations for asthma and glaucoma.
-Overall the model comparison approach is able to retain useful features from widely-used meta-analysis approaches and prioritize protective modifiers of disease risk.
+We consider the use of summary statistic data to apply univariate and multivariate gene-based meta-analysis models for identifying rare variant associations with an emphasis on protective protein-truncating variants that can expedite drug discovery.
+Through simulation studies, we demonstrate that the proposed model comparison approach can improve ability to detect rare variant association signals.
+We also apply the model to two groups of phenotypes from the UK Biobank: 1) asthma diagnosis, eosinophil counts, forced expiratory volume, and forced vital capacity; and 2) glaucoma diagnosis, intra-ocular pressure, and corneal resistance factor.
+We are able to recover known associations such as the protective association between rs146597587 in \textit{IL33} and asthma.
+We also find evidence for an association between rare variants in the drug target \textit{CCR3} and asthma as well as novel protective associations between rare variants in \textit{ANGPTL7} and \textit{WNT10A} and glaucoma.
+Overall, we show that the MRP model comparison approach is able to retain and improve upon useful features from widely-used meta-analysis approaches for rare variant association analyses and prioritize protective modifiers of disease risk.
 
@@
 We applied both independent and similar effects models and used Bayesian model averaging to compute a single BF per gene \cite{10.2307/2676803}.
-In agreement with previous studies, we observed evidence that rare missense variants and/or PTVs in \textit{IL33} affect eosinophil counts and offer protection from asthma from the single-phenotype analyses, though the evidence of association was strongest for the joint analysis (Table~\ref{asthma_table}) \cite{DeBoever179762, 10.1371/journal.pgen.1006659}.
-We also performed an analysis focused on identifying protective variants which also identified the \textit{IL33} association (Fig~\ref{asthma_bma}).
+In agreement with previous studies, we observed evidence that rare missense variants and/or PTVs in \textit{IL33} affect eosinophil counts and offer protection from asthma from the single-phenotype analyses, though the evidence of association was strongest for the joint analysis (\nameref{asthma_table}) \cite{DeBoever179762, 10.1371/journal.pgen.1006659}.
+We performed an analysis focused on identifying protective variants which also identified the \textit{IL33} association (Fig~\ref{asthma_bma}F).
 The results were similar using only either the independent effects (\nameref{asthma_independent}) or similar effects models (\nameref{asthma_similar}).
-We can inspect the effect sizes from the marginal GWAS regressions for the rare variants included in the analysis and see that the association identified by MRP is likely driven by the PTV rs146597587 (Fig~\ref{asthma_bma}G).
+We inspected the effect sizes from the marginal GWAS regressions for the rare variants included in the analysis and found that the association identified by MRP is likely driven by the PTV rs146597587 (Fig~\ref{asthma_bma}G).
 
-We also found moderate evidence for associations between rare coding variants in \textit{CCR3} and asthma.
-The $\log_{10}$ BFs for \textit{CCR3} were 3.3 and 3.1 in the joint and joint protective analyses respectively compared to only -0.5 in the asthma-only analysis  (Fig~\ref{asthma_bma}, Table~\ref{asthma_table}).
+We also found moderate evidence for a protective association between rare coding variants in \textit{CCR3} and asthma.
+The $\log_{10}$ BFs for \textit{CCR3} were 3.3 and 3.1 in the joint and joint protective analyses respectively compared to only -0.5 in the asthma-only analysis  (Fig~\ref{asthma_bma}, \nameref{asthma_table}).
+\textit{CCR3} is a chemokine receptor that is highly expressed on eosinophils and has been a therapeutic focus for asthma \cite{Neighbour:2014jw, Pease:2014em}.
 \textit{CCR3} was not reported in a large GWAS for allergic disease including asthma\cite{Ferreira:2017ba} though \textit{CCR3} is near a locus associated with atopy in a previous meta-analysis\cite{Ober:2011jk}.
-Other genes like \textit{IL18RAP}, \textit{ATP2A3}, and \textit{FLG} had $\log_{10}$ BFs greater than 4 in the asthma-only analysis but much smaller BFs in the joint analyses indicating that rare variants in these genes are less likely to affect this group of traits.
+These results demonstrate that MRP can identify biologically meaningful therapeutic targets that may be missed by standard GWAS approaches.
+
+Considering multiple phenotypes jointly allows for the efficient prioritization of disease genes.
+For instance, some genes like \textit{IL18RAP}, \textit{ATP2A3}, and \textit{FLG} had $\log_{10}$ BFs greater than 4 in the asthma-only analysis but much smaller BFs in the joint analyses indicating that rare variants in these genes are less likely to affect this group of traits.
 Similarly, there were other genes like \textit{RP11-39K24.9} and \textit{IL17RA} that had relatively larger BFs in the eosinophil count-only analysis but small BFs for the joint analyses demonstrating MRP's ability to integrate information across all phenotypes considered.
@@
 
-\begin{table}[!ht]
-%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
-\centering
-\caption{
-{\bf $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{asthma_bma}.}}
-\begin{tabular}{|l+l|l|l|l|l|l|}
-\hline
-{\bf Gene} & {\bf Joint, protective} & {\bf Joint} & {\bf Eosinophil count} & {\bf FVC} & {\bf FEV$_1$} & {\bf Asthma} \\
-\thickhline
-\textit{IL33}         &              29.4 &   29.3 &             30.6 &  -2.3 &  -2.2 &    8.1 \\
-\hline
-\textit{CCR3}         &               3.1 &    3.3 &              7.4 &  -1.4 &  -1.6 &   -0.5 \\
-\hline
-\textit{RP11-39K24.9} &               0.8 &    1.8 &              4.9 &  -0.1 &  -0.4 &    0.3 \\
-\hline
-\textit{SCMH1}        &               0.5 &    4.7 &             -1.5 &   7.7 &   3.8 &   -0.7 \\
-\hline
-\textit{MUSTN1}       &               0.4 &    1.1 &             -1.2 &   2.9 &   2.9 &   -0.6 \\
-\hline
-\textit{ZFAT}         &               0.3 &    1.3 &             -2.0 &   4.7 &   3.1 &   -0.4 \\
-\hline
-\textit{ELN}          &               0.2 &    2.5 &             -1.0 &   5.6 &   2.9 &   -0.6 \\
-\hline
-\textit{C14orf39}     &              -0.7 &   -0.0 &             -1.1 &   3.5 &   2.5 &    0.0 \\
-\hline
-\textit{TMEM110}      &              -0.9 &    1.1 &             -1.0 &   3.3 &   3.1 &   -0.6 \\
-\hline
-\textit{IL17RA}       &              -4.4 &    3.1 &              7.9 &  -2.7 &  -2.5 &   -1.1 \\
-\hline
-\textit{IL18RAP}      &              -9.6 &   -0.9 &             -1.0 &  -1.6 &  -1.7 &    3.5 \\
-\hline
-\textit{ATP2A3}       &             -11.9 &   -0.8 &             -1.2 &  -2.1 &  -2.2 &    5.8 \\
-\hline
-\textit{FLG}          &             -20.1 &  -17.2 &             -6.0 &  -7.4 &  -8.5 &    4.0 \\
-\hline
-\end{tabular}
-\label{asthma_table}
-%\end{adjustwidth}
-\end{table}
-
 \subsubsection*{Glaucoma, intra-ocular pressure, and corneal resistance factor}
@@
 \end{figure}
-
-% %% TABLE: GLAUCOMA GENES OF INTEREST
-\begin{table}[!ht]
-\begin{adjustwidth}{-1.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
-\centering
-\caption{
-{\bf $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{glaucoma_bma}.}}
-\begin{tabular}{|l+p{3cm}|p{1cm}|p{1.6cm}|p{3.8cm}|p{3.4cm}|}
-\hline
-{\bf Gene} & {\bf Joint, protective} & {\bf Joint} & {\bf Glaucoma} & {\bf Intra-ocular pressure, Goldmann-correlated} & {\bf Corneal resistance factor} \\
-\thickhline
-\textit{ANGPTL7} &              10.1 &   9.4 &      1.7 &                                       10.1 &                       6.2 \\
-\hline
-\textit{KLHL22}  &               3.1 &   3.6 &     -0.2 &                                        3.9 &                       2.2 \\
-\hline
-\textit{GLI3}    &               2.9 &   2.9 &     -0.3 &                                       -0.7 &                       3.2 \\
-\hline
-\textit{WNT10A}  &               2.2 &   3.0 &     -0.3 &                                       -0.5 &                       3.5 \\
-\hline
-\textit{RNASE2}  &              -1.3 &  -0.3 &      0.4 &                                       -1.0 &                      -1.1 \\
-\hline
-\textit{FUK}     &              -6.5 &  -2.3 &      2.5 &                                       -2.1 &                      -2.1 \\
-\hline
-\end{tabular}
-\label{table2}
-\end{adjustwidth}
-\end{table}
 
@@
 
-%% FIGURE: ASTHMA MANHATTAN AND FOREST FOR INDEPENDENT EFFECTS
 \begin{figure}[H]
@@
 \begin{figure}[H]
-\includegraphics[width=\textwidth]{../figures/final/Figure_S04.pdf}
+\includegraphics[width=\textwidth]{../figures/final/Figure_S05.pdf}
 \end{figure}
-\paragraph*{S4 Fig.}
+\paragraph*{S5 Fig.}
 \label{glaucoma_independent}
@@
 \begin{figure}[H]
-\includegraphics[width=\textwidth]{../figures/final/Figure_S05.pdf}
+\includegraphics[width=\textwidth]{../figures/final/Figure_S06.pdf}
 \end{figure}
-\paragraph*{S5 Fig.}
+\paragraph*{S6 Fig.}
 \label{glaucoma_similar}
@@
 Only $\log_{10}$ Bayes Factors greater than zero are plotted.
+
+\begin{table}[!ht]
+\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+\centering
+%\caption{
+%{\bf $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{asthma_bma}.}}
+\begin{tabular}{|l+l|l|l|l|l|l|}
+\hline
+{\bf Gene} & {\bf Joint, protective} & {\bf Joint} & {\bf Eosinophil count} & {\bf FVC} & {\bf FEV$_1$} & {\bf Asthma} \\
+\thickhline
+\textit{IL33}         &              29.4 &   29.3 &             30.6 &  -2.3 &  -2.2 &    8.1 \\
+\hline
+\textit{CCR3}         &               3.1 &    3.3 &              7.4 &  -1.4 &  -1.6 &   -0.5 \\
+\hline
+\textit{RP11-39K24.9} &               0.8 &    1.8 &              4.9 &  -0.1 &  -0.4 &    0.3 \\
+\hline
+\textit{SCMH1}        &               0.5 &    4.7 &             -1.5 &   7.7 &   3.8 &   -0.7 \\
+\hline
+\textit{MUSTN1}       &               0.4 &    1.1 &             -1.2 &   2.9 &   2.9 &   -0.6 \\
+\hline
+\textit{ZFAT}         &               0.3 &    1.3 &             -2.0 &   4.7 &   3.1 &   -0.4 \\
+\hline
+\textit{ELN}          &               0.2 &    2.5 &             -1.0 &   5.6 &   2.9 &   -0.6 \\
+\hline
+\textit{C14orf39}     &              -0.7 &   -0.0 &             -1.1 &   3.5 &   2.5 &    0.0 \\
+\hline
+\textit{TMEM110}      &              -0.9 &    1.1 &             -1.0 &   3.3 &   3.1 &   -0.6 \\
+\hline
+\textit{IL17RA}       &              -4.4 &    3.1 &              7.9 &  -2.7 &  -2.5 &   -1.1 \\
+\hline
+\textit{IL18RAP}      &              -9.6 &   -0.9 &             -1.0 &  -1.6 &  -1.7 &    3.5 \\
+\hline
+\textit{ATP2A3}       &             -11.9 &   -0.8 &             -1.2 &  -2.1 &  -2.2 &    5.8 \\
+\hline
+\textit{FLG}          &             -20.1 &  -17.2 &             -6.0 &  -7.4 &  -8.5 &    4.0 \\
+\hline
+\end{tabular}
+\label{asthma_table}
+\end{adjustwidth}
+\end{table}
+\paragraph*{S1 Table.}
+\label{asthma_table}
+{\bf Highlighted genes from asthma analysis.} $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{asthma_bma}.
+
+% %% TABLE: GLAUCOMA GENES OF INTEREST
+\begin{table}[!ht]
+\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+\centering
+%\caption{
+%{\bf $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{glaucoma_bma}.}}
+\begin{tabular}{|l+p{3cm}|p{1cm}|p{1.6cm}|p{3.8cm}|p{3.4cm}|}
+\hline
+{\bf Gene} & {\bf Joint, protective} & {\bf Joint} & {\bf Glaucoma} & {\bf Intra-ocular pressure, Goldmann-correlated} & {\bf Corneal resistance factor} \\
+\thickhline
+\textit{ANGPTL7} &              11.9 &  11.9 &      1.7 &                                       10.1 &                       6.2 \\
+\hline
+\textit{KLHL22}  &               3.5 &   3.5 &     -0.2 &                                        3.9 &                       2.2 \\
+\hline
+\textit{WNT10A}  &               2.5 &   2.5 &     -0.3 &                                       -0.5 &                       3.5 \\
+\hline
+\textit{GLI3}    &               1.5 &   1.5 &     -0.3 &                                       -0.7 &                       3.2 \\
+\hline
+\textit{RNASE2}  &              -0.5 &  -0.5 &      0.4 &                                       -1.0 &                      -1.1 \\
+\hline
+\textit{FUK}     &              -1.8 &  -1.8 &      2.5 &                                       -2.1 &                      -2.1 \\
+\hline
+\end{tabular}
+\end{adjustwidth}
+\end{table}
+\paragraph*{S2 Table.}
+\label{glaucoma_table}
+{\bf Highlighted genes from glaucoma analysis.} $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{glaucoma_bma}.
 

### 2018-01-30 tex/manuscript.tex

@@
 We are able to recover known associations such as the protective association between rs146597587 in \textit{IL33} and asthma.
-We also find evidence for an association between rare variants in the drug target \textit{CCR3} and asthma as well as novel protective associations between rare variants in \textit{ANGPTL7} and \textit{WNT10A} and glaucoma.
+We also find evidence for an association between rare variants in the drug target \textit{CCR3} and asthma as well as novel protective associations between rare variants in \textit{ANGPTL7} and glaucoma.
 Overall, we show that the MRP model comparison approach is able to retain and improve upon useful features from widely-used meta-analysis approaches for rare variant association analyses and prioritize protective modifiers of disease risk.
@@
 %% Use of summary statistic data
-While many large genetic consortia generally collect both raw genotype and phenotype data, in practice, sharing of individual genotype and phenotype data across groups is difficult to achieve.
+While many large genetic consortia collect both raw genotype and phenotype data, in practice, sharing of individual genotype and phenotype data across groups is difficult to achieve.
 To address this, MRP can take summary statistics, such as estimates of effect size and the corresponding standard error from typical single variant-single phenotype linear or logistic regressions, as input data.
@@
 Summary statistics from variants in the non-null component are modeled as being drawn from a multivariate normal distribution with zero mean, but the covariance matrix for the non-null component combines the covariance matrix from the null component with another covariance matrix that captures the genetic correlation between the phenotypes being considered.
-We observed similar genetic correlations using LD score regression %(\nameref{gcorr_fig}) \cite{gencorr2015}.
+We observed similar genetic correlations using LD score regression (\nameref{gcorr_fig}) \cite{gencorr2015}.
 
@@
 
-We also found moderate evidence for a protective association between rare coding variants in \textit{CCR3} and asthma.
-The $\log_{10}$ BFs for \textit{CCR3} were 3.3 and 3.1 in the joint and joint protective analyses respectively compared to only -0.5 in the asthma-only analysis  (Fig~\ref{asthma_bma}, \nameref{asthma_table}).
+We also found moderate evidence for association between rare coding variants in \textit{CCR3} and asthma.
+The $\log_{10}$ BFs for \textit{CCR3} was 3.3 in the joint model compared to only -0.5 in the asthma-only analysis  (Fig~\ref{asthma_bma}, \nameref{asthma_table}).
 \textit{CCR3} is a chemokine receptor that is highly expressed on eosinophils and has been a therapeutic focus for asthma \cite{Neighbour:2014jw, Pease:2014em}.
@@
 For instance, some genes like \textit{IL18RAP}, \textit{ATP2A3}, and \textit{FLG} had $\log_{10}$ BFs greater than 4 in the asthma-only analysis but much smaller BFs in the joint analyses indicating that rare variants in these genes are less likely to affect this group of traits.
-Similarly, there were other genes like \textit{RP11-39K24.9} and \textit{IL17RA} that had relatively larger BFs in the eosinophil count-only analysis but small BFs for the joint analyses demonstrating MRP's ability to integrate information across all phenotypes considered.
+Similarly, there were other genes like \textit{RP11-39K24.9} and \textit{IL17RA} that had larger BFs in the eosinophil count-only analysis but small BFs for the joint analyses demonstrating MRP's ability to integrate information across all phenotypes considered.
 
@@
 We identified strong evidence for the previously described association between the PTV rs146597587 in \textit{IL33} and asthma\cite{DeBoever179762,10.1371/journal.pgen.1006659} as well as evidence for a potentially novel genetic association for the drug target \textit{CCR3} (\nameref{asthma_table}).
-We also identified genetic evidence for a link between rare variants in \textit{ANGPTL7} and glaucoma, consistent with previous experiments that suggested a role for \textit{ANGPTL7} in glaucoma \cite{Comes:2011ex, Kuchtey:2008ek}.
+We also identified genetic evidence for a link between rare variants in \textit{ANGPTL7} and glaucoma with effects consistent with protection to disease.
 These results demonstrate the ability of the MRP model comparison approach to leverage information across multiple phenotypes and variants to discover rare variant associations.
@@
 MAR, MP, and CD wrote the manuscript.
-CCAS, YT, MA, and CD provided analysis and designed figures.
+MAR, MP, CCAS, YT, MA, and CD provided analysis and designed figures.
 TP designed HDF5 tables and implementation of loaders.

### 2018-01-30 tex/manuscriptbiorxiv.tex
Manuscript update

@@
 \setlength{\parindent}{0.5cm}
-%\textwidth 5.25in 
-%\textheight 8.75in
+\textwidth 5.25in 
+\textheight 8.75in
 
@@
 We are able to recover known associations such as the protective association between rs146597587 in \textit{IL33} and asthma.
-We also find evidence for an association between rare variants in the drug target \textit{CCR3} and asthma as well as novel protective associations between rare variants in \textit{ANGPTL7} and \textit{WNT10A} and glaucoma.
+We also find evidence for novel protective associations between rare variants in \textit{ANGPTL7} and \textit{WNT10A} and glaucoma.
 Overall, we show that the MRP model comparison approach is able to retain and improve upon useful features from widely-used meta-analysis approaches for rare variant association analyses and prioritize protective modifiers of disease risk.
@@
 We applied MRP to summary statistics for two groups of related phenotypes from the UK Biobank.
-First, we apply MRP to asthma (HC382: the corresponding phenotype label in Global Biobank Engine [\url{https://biobankengine.stanford.edu}]), eosinophil count (INI30150), forced expiratory volume in 1-second (FEV$_1$, INI3063), and forced vital capacity (FVC, INI3062).
-We recover the reported association between a rare PTV in \textit{IL33} and asthma as well as a new association between \textit{CCR3} and asthma \cite{DeBoever179762, 10.1371/journal.pgen.1006659}.
+First, we applied MRP to asthma (HC382: the corresponding phenotype label in Global Biobank Engine [\url{https://biobankengine.stanford.edu}]), eosinophil count (INI30150), forced expiratory volume in 1-second (FEV$_1$, INI3063), and forced vital capacity (FVC, INI3062) and recovered the reported association between a rare PTV in \textit{IL33} and asthma \cite{DeBoever179762, 10.1371/journal.pgen.1006659}.
 We also applied MRP to glaucoma (HC276), intra-ocular pressure (INI5263), and corneal resistance factor (INI5265) and find evidence that rare coding variants in \textit{ANGPTL7} and \textit{WNT10A} may protect against glaucoma.
@@
 We performed genome-wide association analysis using PLINK v2.00a(17 July 2017) as previously described \cite{DeBoever179762}.
-For asthma, we used the Firth fallback in PLINK, a hybrid algorithm which normally uses the logistic regression code described in (Hill 2017), but switches to a port of logistf() (https://cran.r-project.org/web/packages/logistf/index.html) in two cases: (1) one of the cells in the 2x2 allele count by case/control status contingency table is empty (2) logistic regression was attempted since all the contingency table cells were nonzero, but it failed to converge within the usual number of steps.
+For asthma, we used the Firth fallback in PLINK, a hybrid algorithm which normally uses the logistic regression code described in (Hill 2017), but switches to a port of $\texttt{logistf()}$ (\url{https://cran.r-project.org/web/packages/logistf/index.html}) in two cases: (1) one of the cells in the 2x2 allele count by case/control status contingency table is empty (2) logistic regression was attempted since all the contingency table cells were nonzero, but it failed to converge within the usual number of steps.
 We used the following covariates in our analysis: age, sex, array type, and the first four principal components, where array type is a binary variable that represents whether an individual was genotyped with UK Biobank Axiom Array or UK BiLEVE Axiom Array.
@@
 \textit{ANGPTL7} protein expression was assessed using the HIPED protein expression database accessed through genecards.org on 2017/1/29 \cite{Fishilevich:2016hx}.
+We identified the protein 1JC9\_A as homologous to the \textit{ANGPTL7} protein using the ``3D structure mapping'' link from dbSNP (\url{https://www.ncbi.nlm.nih.gov/projects/SNP/snp_ref.cgi?rs=28991009}).
+We retrieved the 3D structure image from the iCn3D Structure Viewer (\url{https://www.ncbi.nlm.nih.gov/Structure/icn3d/icn3d.html}).
+
+structure for protein XXX which was homologous to the \textit{ANGPTL7} protein
 
 \subsubsection*{Variant Filtering}
-We used the $\texttt{variant\_filter\_table.tsv}$ file available at https://github.com/rivas-lab/public-resources (6f9f726) to filter variants on the UK Biobank array for use with MRP.
+We used the $\texttt{variant\_filter\_table.tsv}$ file available at \url{https://github.com/rivas-lab/public-resources} (6f9f726) to filter variants on the UK Biobank array for use with MRP.
 We first chose variants with minor allele frequency less than $1\%$.
@@
 Expression of \textit{ANGPTL7} is upregulated in glaucoma and has been proposed to regulate intra-ocular pressure and glaucoma risk \cite{Comes:2011ex, Kuchtey:2008ek}.
-The GWAS summary statistics for the rare variants in \textit{ANGPTL7} suggest that two missense variants, rs28991002 and rs28991009, appear to drive the association with glaucoma (Fig~\ref{glaucoma_bma}F).
+The GWAS summary statistics for the rare variants in \textit{ANGPTL7} suggest that the association with glaucoma is driven by the missense variant rs28991009 that changes residue 175 from glutamine to histidine (Fig~\ref{glaucoma_bma}F, G).
 According to the HIPED protein expression database, \textit{ANGPTL7} protein is expressed at $\sim0.7$ parts per million in vitreous humor, the material between the lens and retina of the eyeball; in contrast, the expression of \textit{ANGPTL7} protein is less than $0.01$ parts per million in 68 other normal tissues \cite{Fishilevich:2016hx}.
@@
 Only $\log_{10}$ Bayes Factors greater than zero are plotted.
-F: $-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{ANGPTL7} for all three phenotypes.}
+F: $-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{ANGPTL7} for all three phenotypes.
+G: Location of rs28991009 variant (green) for the protein 1JC9\_A homologous to \textit{ANGPTL7}.}
 \label{glaucoma_bma}
@@
 We also applied the MRP model comparison framework to summary statistic data from two groups of traits from the UK Biobank: asthma diagnosis, eosinophil counts, FEV$_1$, and FVC; and glaucoma diagnosis, intra-ocular pressure, and corneal resistance factor.
-We identified strong evidence for the previously described association between the PTV rs146597587 in \textit{IL33} and asthma\cite{DeBoever179762,10.1371/journal.pgen.1006659} as well as evidence for a potentially novel genetic association for the drug target \textit{CCR3} (\nameref{asthma_table}).
-We also identified genetic evidence for a link between rare variants in \textit{ANGPTL7} and glaucoma, consistent with previous experiments that suggested a role for \textit{ANGPTL7} in glaucoma \cite{Comes:2011ex, Kuchtey:2008ek}.
+We identified strong evidence for the previously described association between the PTV rs146597587 in \textit{IL33} and asthma\cite{DeBoever179762,10.1371/journal.pgen.1006659}.
+We also found evidence for a link between rare variants in \textit{ANGPTL7} and glaucoma, consistent with previous experiments that suggested a role for \textit{ANGPTL7} in glaucoma \cite{Comes:2011ex, Kuchtey:2008ek}.
 These results demonstrate the ability of the MRP model comparison approach to leverage information across multiple phenotypes and variants to discover rare variant associations.
@@
 is supported by Funai Overseas Scholarship from Funai Foundation for Information Technology and the Stanford University Biomedical Informatics Training Program.
-The primary and processed data used to generate the analyses presented here are available in the UK Biobank access management system (https://amsportal.ukbiobank.ac.uk/) for application 24983, ``Generating effective therapeutic hypotheses from genomic and hospital linkage data'' (http://www.ukbiobank.ac.uk/wp-content/uploads/2017/06/24983-Dr-Manuel-Rivas.pdf), and the results are displayed in the Global Biobank Engine (https://biobankengine.stanford.edu).
+The primary and processed data used to generate the analyses presented here are available in the UK Biobank access management system (\url{https://amsportal.ukbiobank.ac.uk/}) for application 24983, ``Generating effective therapeutic hypotheses from genomic and hospital linkage data'' (\url{http://www.ukbiobank.ac.uk/wp-content/uploads/2017/06/24983-Dr-Manuel-Rivas.pdf}), and the results are displayed in the Global Biobank Engine (\url{https://biobankengine.stanford.edu}).
 We would like to thank the Customer Solutions Team from Paradigm4 who helped us implement efficient databases for queries and application of inference methods to the data.

### 2018-01-30 tex/manuscript.tex
small updates to manuscript

@@
 We are able to recover known associations such as the protective association between rs146597587 in \textit{IL33} and asthma.
-We also find evidence for an association between rare variants in the drug target \textit{CCR3} and asthma as well as novel protective associations between rare variants in \textit{ANGPTL7} and glaucoma.
+We also find evidence for novel protective associations between rare variants in \textit{ANGPTL7} and glaucoma.
 Overall, we show that the MRP model comparison approach is able to retain and improve upon useful features from widely-used meta-analysis approaches for rare variant association analyses and prioritize protective modifiers of disease risk.
@@
 We applied MRP to summary statistics for two groups of related phenotypes from the UK Biobank.
-First, we apply MRP to asthma (HC382: the corresponding phenotype label in Global Biobank Engine [\url{https://biobankengine.stanford.edu}]), eosinophil count (INI30150), forced expiratory volume in 1-second (FEV$_1$, INI3063), and forced vital capacity (FVC, INI3062).
-We recover the reported association between a rare PTV in \textit{IL33} and asthma as well as a new association between \textit{CCR3} and asthma \cite{DeBoever179762, 10.1371/journal.pgen.1006659}.
-We also applied MRP to glaucoma (HC276), intra-ocular pressure (INI5263), and corneal resistance factor (INI5265) and find evidence that rare coding variants in \textit{ANGPTL7} and \textit{WNT10A} may protect against glaucoma.
+First, we applied MRP to asthma (HC382: the corresponding phenotype label in Global Biobank Engine [\url{https://biobankengine.stanford.edu}]), eosinophil count (INI30150), forced expiratory volume in 1-second (FEV$_1$, INI3063), and forced vital capacity (FVC, INI3062) and recovered the reported association between a rare PTV in \textit{IL33} and asthma \cite{DeBoever179762, 10.1371/journal.pgen.1006659}.
+We also applied MRP to glaucoma (HC276), intra-ocular pressure (INI5263), and corneal resistance factor (INI5265) and find evidence that rare coding variants in \textit{ANGPTL7} protect against glaucoma.
 These analyses show that MRP recovers results from typical single variant-single phenotype analyses while identifying new rare variant associations that include protective modifiers of disease risk.
@@
 While the individual glaucoma analysis did not yield any associations with $\log_{10}$ BF greater than three, the joint analysis identified rare coding variants in both \textit{ANGPTL7} and \textit{WNT10A} as associated with glaucoma (Fig~\ref{glaucoma_bma}A-D, \nameref{glaucoma_table}).
-Applying the MRP model to identify protective effects also identified \textit{ANGPTL7} and \textit{WNT10A} (Fig~\ref{glaucoma_bma}E).
+Applying the MRP model to identify protective effects also identified \textit{ANGPTL7} and additional association to \textit{WNT10A} (Fig~\ref{glaucoma_bma}E).
 We obtained similar results using the independent effects (\nameref{glaucoma_independent}) or similar effects models (\nameref{glaucoma_similar}).
@@
 We also applied the MRP model comparison framework to summary statistic data from two groups of traits from the UK Biobank: asthma diagnosis, eosinophil counts, FEV$_1$, and FVC; and glaucoma diagnosis, intra-ocular pressure, and corneal resistance factor.
-We identified strong evidence for the previously described association between the PTV rs146597587 in \textit{IL33} and asthma\cite{DeBoever179762,10.1371/journal.pgen.1006659} as well as evidence for a potentially novel genetic association for the drug target \textit{CCR3} (\nameref{asthma_table}).
-We also identified genetic evidence for a link between rare variants in \textit{ANGPTL7} and glaucoma with effects consistent with protection to disease.
+We identified strong evidence for the previously described association between the PTV rs146597587 in \textit{IL33} and asthma\cite{DeBoever179762,10.1371/journal.pgen.1006659}.
+We also found evidence for a link between rare variants in \textit{ANGPTL7} and glaucoma, consistent with previous experiments that suggested a role for \textit{ANGPTL7} in glaucoma \cite{Comes:2011ex, Kuchtey:2008ek}.
 These results demonstrate the ability of the MRP model comparison approach to leverage information across multiple phenotypes and variants to discover rare variant associations.

### 2018-01-30 tex/manuscript.tex
small updates to manuscript

@@
 While the individual glaucoma analysis did not yield any associations with $\log_{10}$ BF greater than three, the joint analysis identified rare coding variants in both \textit{ANGPTL7} and \textit{WNT10A} as associated with glaucoma (Fig~\ref{glaucoma_bma}A-D, \nameref{glaucoma_table}).
-Applying the MRP model to identify protective effects also identified \textit{ANGPTL7} and additional association to \textit{WNT10A} (Fig~\ref{glaucoma_bma}E).
+Applying the MRP model identified \textit{ANGPTL7} with effects consistent with protection to glaucoma, and additional association to \textit{WNT10A} (Fig~\ref{glaucoma_bma}E).
 We obtained similar results using the independent effects (\nameref{glaucoma_independent}) or similar effects models (\nameref{glaucoma_similar}).

### 2018-01-30 tex/manuscript.tex
Updates

@@
 In this section, we provide an overview of the MRP model comparison approach.
-Refer to \nameref{S1_Appendix} for detailed derivations.
+Refer to \nameref{S1_Appendix} for a detailed description.
 MRP models GWAS summary statistics as being distributed according to one of two models.
@@
 MRP allows users to specify priors that reflect knowledge of the variants and phenotypes under study.
-The prior on the scale of effect sizes can be used to denote which variants may have larger effect sizes.
+For instance, we can define an independent effects model where each variant in the model may have different effect sizes.
+In this case, $\mathbf{R_{\textrm{var}}}$ is the identity matrix which reflects the assumption that the effect sizes of the variants are not correlated.
+This could be true for missense variants where some variants may impact important functional domains but others do not.
+We can also define a similar effects model by setting every value of $\mathbf{R_{\textrm{var}}}$ to $\sim 1$.
+This model assumes that all variants under consideration have the same effect sizes.
+This model may be appropriate for PTVs where each variant completely disrupts the function of the gene, leading to a gene knockout.
+The prior on the scale of effect sizes can also be used to denote which variants may have larger effect sizes.
 For instance, emerging empirical genetic studies have shown that within a gene, PTVs may have stronger effects than missense variants\cite{do2015exome}.
@@
 We retrieved the 3D structure image from the iCn3D Structure Viewer (\url{https://www.ncbi.nlm.nih.gov/Structure/icn3d/icn3d.html}).
-
-structure for protein XXX which was homologous to the \textit{ANGPTL7} protein
 

### 2018-01-30 tex/manuscript.tex
Updates

@@
 In this case, $\mathbf{R_{\textrm{var}}}$ is the identity matrix which reflects the assumption that the effect sizes of the variants are not correlated.
-This could be true for missense variants where some variants may impact important functional domains but others do not.
 We can also define a similar effects model by setting every value of $\mathbf{R_{\textrm{var}}}$ to $\sim 1$.
-This model assumes that all variants under consideration have the same effect sizes.
+This model assumes that all variants under consideration have similar effect sizes (with possibly differences in scale).
 This model may be appropriate for PTVs where each variant completely disrupts the function of the gene, leading to a gene knockout.
@@
 For instance, emerging empirical genetic studies have shown that within a gene, PTVs may have stronger effects than missense variants\cite{do2015exome}.
-This can be reflected by providing larger priors for PTVs.
+This can be reflected by adjusting the prior spread of effect sizes ($\sigma$) for PTVs (\nameref{S1_Appendix}).
 

### 2018-01-30 tex/manuscript.tex
final changes

@@
 Let $S$ be the number of studies where data is obtained from - this data may be in the form of raw genotypes and phenotypes or summary statistics including linkage-disequilibrium, effect sizes (or odds ratio), and standard error of the effect size.
-When considering multiple phenotypes ($K > 1$), we define the prior correlation structure of the effect sizes as an $SMK\times SMK$ matrix $\mathbf{U}$.
-In practice, we define $\mathbf{U}$ as a Kronecker product, an operation of matrices of arbitrary size, of three sub-matrices
+When considering multiple studies ($S > 1$), multiple rare variants ($M > 1$), and multiple phenotypes ($K > 1$), we define the prior correlation structure of the effect sizes as an $SMK\times SMK$ matrix $\mathbf{U}$.
+In practice, we define $\mathbf{U}$ as a Kronecker product, an operation of matrices of arbitrary size, of three sub-matrices:
 \begin{itemize}
-\item an $S\times S$ matrix $\mathbf{R_{\textrm{study}}}$ containing the correlations of genetic effects among studies where different values can be used to compare different models of association, such as for identifying heterogeneity of effect sizes between populations~\cite{band2013imputation}
-\item an $M\times M$ matrix $\mathbf{R_{\textrm{var}}}$ containing the correlations of genetic effects among genetic variants, which may reflect the assumption that all the PTVs in a gene may have the same biological consequence\cite{macarthur,rivas2013assessing,rivas2015effect} or prior information obtained through integration of additional data sources, such as functional assay data\cite{majithia2014rare,findlay2014saturation}, otherwise zero correlation of genetic effects may be assumed, which is used in dispersion tests like C-alpha\cite{calpha,clarke2013flexible} and SKAT\cite{skat}
-\item a $K\times K$ matrix $\mathbf{R_{\textrm{phen}}}$ containing the correlations of genetic effects among phenotypes, which may be obtained from common variant data\cite{cotsapas2011pervasive,solovieff2013pleiotropy,gencorr2015}
+\item an $S\times S$ matrix $\mathbf{R_{\textrm{study}}}$ containing the correlations of genetic effects among studies where different values can be used to compare different models of association, such as for identifying heterogeneity of effect sizes between populations~\cite{band2013imputation};
+\item an $M\times M$ matrix $\mathbf{R_{\textrm{var}}}$ containing the correlations of genetic effects among genetic variants, which may reflect the assumption that all the PTVs in a gene may have the same biological consequence\cite{macarthur,rivas2013assessing,rivas2015effect} or prior information obtained through integration of additional data sources, such as functional assay data\cite{majithia2014rare,findlay2014saturation}, otherwise zero correlation of genetic effects may be assumed, which is used in dispersion tests like C-alpha\cite{calpha,clarke2013flexible} and SKAT\cite{skat}; and
+\item a $K\times K$ matrix $\mathbf{R_{\textrm{phen}}}$ containing the correlations of genetic effects among phenotypes, which may be obtained from common variant data\cite{cotsapas2011pervasive,solovieff2013pleiotropy,gencorr2015}.
 \end{itemize}
-The variance-covariance matrix of the effect sizes may be obtained from readily available summary statistic data such as in-study LD matrices, effect size estimates (or log odds ratios), and the standard errors of the effect size estimates (Online Methods).
+The variance-covariance matrix of the effect sizes may be obtained from readily available summary statistic data such as in-study LD matrices, effect size estimates (or log odds ratios), and the standard errors of the effect size estimates (\nameref{S1_Appendix}).
 
@@
 Thus far we have assumed that the prior mean, or location, of genetic effects is zero which makes it feasible to analyze a large number of phenotypes without enumerating the prior mean across all phenotypes.
-To proactively identify genetic variants that have effects that are consistent with a protective profile for a disease, we can include a non-zero vector as a prior mean of genetic effect \nameref{S1_Appendix}.
+To proactively identify genetic variants that have effects that are consistent with a protective profile for a disease, we can include a non-zero vector as a prior mean of genetic effect (\nameref{S1_Appendix}).
 We can exploit information from Mendelian randomization studies of common variants, such as recent findings where rare truncating loss-of-function variants in {\it PCSK9} were found to decrease LDL and triglyceride levels and decrease CAD risk\cite{cohen2005low,pcsk9,do2013common,cohorts2014loss} to identify situations where such a prior is warranted.
@@
 We performed genome-wide association analysis using PLINK v2.00a(17 July 2017) as previously described \cite{DeBoever179762}.
-For asthma, we used the Firth fallback in PLINK, a hybrid algorithm which normally uses the logistic regression code described in (Hill 2017), but switches to a port of $\texttt{logistf()}$ (\url{https://cran.r-project.org/web/packages/logistf/index.html}) in two cases: (1) one of the cells in the 2x2 allele count by case/control status contingency table is empty (2) logistic regression was attempted since all the contingency table cells were nonzero, but it failed to converge within the usual number of steps.
+For asthma, we used the Firth fallback in PLINK, a hybrid algorithm which normally uses the logistic regression code described in \cite{hill2017stepwise}, but switches to a port of $\texttt{logistf()}$ (\url{https://cran.r-project.org/web/packages/logistf/index.html}) in two cases: (1) one of the cells in the 2x2 allele count by case/control status contingency table is empty (2) logistic regression was attempted since all the contingency table cells were nonzero, but it failed to converge within the usual number of steps.
 We used the following covariates in our analysis: age, sex, array type, and the first four principal components, where array type is a binary variable that represents whether an individual was genotyped with UK Biobank Axiom Array or UK BiLEVE Axiom Array.

### 2018-01-30 tex/manuscriptbiorxiv.tex
Updates

@@
 \begin{document}
-%\vspace*{0.2in}
+\vspace*{0.2in}
 
@@
 Spencer$^{2}$, Timothy Poterba$^{3}$, Mark J.
-Daly$^{4,5}$,  Matti Pirinen$^{5}$\thanks{matti.pirinen@helsinki.fi}, and Manuel A.
+Daly$^{4,5}$, Matti Pirinen$^{5,6,7}$\thanks{matti.pirinen@helsinki.fi}, and Manuel A.
 Rivas$^{1}$\thanks{mrivas@stanford.edu}}
@@
 We are able to recover known associations such as the protective association between rs146597587 in \textit{IL33} and asthma.
-We also find evidence for novel protective associations between rare variants in \textit{ANGPTL7} and \textit{WNT10A} and glaucoma.
+We also find evidence for novel protective associations between rare variants in \textit{ANGPTL7} and glaucoma.
 Overall, we show that the MRP model comparison approach is able to retain and improve upon useful features from widely-used meta-analysis approaches for rare variant association analyses and prioritize protective modifiers of disease risk.
@@
 First, we applied MRP to asthma (HC382: the corresponding phenotype label in Global Biobank Engine [\url{https://biobankengine.stanford.edu}]), eosinophil count (INI30150), forced expiratory volume in 1-second (FEV$_1$, INI3063), and forced vital capacity (FVC, INI3062) and recovered the reported association between a rare PTV in \textit{IL33} and asthma \cite{DeBoever179762, 10.1371/journal.pgen.1006659}.
-We also applied MRP to glaucoma (HC276), intra-ocular pressure (INI5263), and corneal resistance factor (INI5265) and find evidence that rare coding variants in \textit{ANGPTL7} and \textit{WNT10A} may protect against glaucoma.
+We also applied MRP to glaucoma (HC276), intra-ocular pressure (INI5263), and corneal resistance factor (INI5265) and find evidence that rare coding variants in \textit{ANGPTL7} protect against glaucoma.
 These analyses show that MRP recovers results from typical single variant-single phenotype analyses while identifying new rare variant associations that include protective modifiers of disease risk.
@@
 Let $S$ be the number of studies where data is obtained from - this data may be in the form of raw genotypes and phenotypes or summary statistics including linkage-disequilibrium, effect sizes (or odds ratio), and standard error of the effect size.
-When considering multiple phenotypes ($K > 1$), we define the prior correlation structure of the effect sizes as an $SMK\times SMK$ matrix $\mathbf{U}$.
-In practice, we define $\mathbf{U}$ as a Kronecker product, an operation of matrices of arbitrary size, of three sub-matrices
+When considering multiple studies ($S > 1$), multiple rare variants ($M > 1$), and multiple phenotypes ($K > 1$), we define the prior correlation structure of the effect sizes as an $SMK\times SMK$ matrix $\mathbf{U}$.
+In practice, we define $\mathbf{U}$ as a Kronecker product, an operation of matrices of arbitrary size, of three sub-matrices:
 \begin{itemize}
-\item an $S\times S$ matrix $\mathbf{R_{\textrm{study}}}$ containing the correlations of genetic effects among studies where different values can be used to compare different models of association, such as for identifying heterogeneity of effect sizes between populations~\cite{band2013imputation}
-\item an $M\times M$ matrix $\mathbf{R_{\textrm{var}}}$ containing the correlations of genetic effects among genetic variants, which may reflect the assumption that all the PTVs in a gene may have the same biological consequence\cite{macarthur,rivas2013assessing,rivas2015effect} or prior information obtained through integration of additional data sources, such as functional assay data\cite{majithia2014rare,findlay2014saturation}, otherwise zero correlation of genetic effects may be assumed, which is used in dispersion tests like C-alpha\cite{calpha,clarke2013flexible} and SKAT\cite{skat}
-\item a $K\times K$ matrix $\mathbf{R_{\textrm{phen}}}$ containing the correlations of genetic effects among phenotypes, which may be obtained from common variant data\cite{cotsapas2011pervasive,solovieff2013pleiotropy,gencorr2015}
+\item an $S\times S$ matrix $\mathbf{R_{\textrm{study}}}$ containing the correlations of genetic effects among studies where different values can be used to compare different models of association, such as for identifying heterogeneity of effect sizes between populations~\cite{band2013imputation};
+\item an $M\times M$ matrix $\mathbf{R_{\textrm{var}}}$ containing the correlations of genetic effects among genetic variants, which may reflect the assumption that all the PTVs in a gene may have the same biological consequence\cite{macarthur,rivas2013assessing,rivas2015effect} or prior information obtained through integration of additional data sources, such as functional assay data\cite{majithia2014rare,findlay2014saturation}, otherwise zero correlation of genetic effects may be assumed, which is used in dispersion tests like C-alpha\cite{calpha,clarke2013flexible} and SKAT\cite{skat}; and
+\item a $K\times K$ matrix $\mathbf{R_{\textrm{phen}}}$ containing the correlations of genetic effects among phenotypes, which may be obtained from common variant data\cite{cotsapas2011pervasive,solovieff2013pleiotropy,gencorr2015}.
 \end{itemize}
-The variance-covariance matrix of the effect sizes may be obtained from readily available summary statistic data such as in-study LD matrices, effect size estimates (or log odds ratios), and the standard errors of the effect size estimates (Online Methods).
+The variance-covariance matrix of the effect sizes may be obtained from readily available summary statistic data such as in-study LD matrices, effect size estimates (or log odds ratios), and the standard errors of the effect size estimates (\nameref{S1_Appendix}).
 
 MRP allows users to specify priors that reflect knowledge of the variants and phenotypes under study.
-The prior on the scale of effect sizes can be used to denote which variants may have larger effect sizes.
+For instance, we can define an independent effects model where each variant in the model may have different effect sizes.
+In this case, $\mathbf{R_{\textrm{var}}}$ is the identity matrix which reflects the assumption that the effect sizes of the variants are not correlated.
+We can also define a similar effects model by setting every value of $\mathbf{R_{\textrm{var}}}$ to $\sim 1$.
+This model assumes that all variants under consideration have similar effect sizes (with possibly differences in scale).
+This model may be appropriate for PTVs where each variant completely disrupts the function of the gene, leading to a gene knockout.
+The prior on the scale of effect sizes can also be used to denote which variants may have larger effect sizes.
 For instance, emerging empirical genetic studies have shown that within a gene, PTVs may have stronger effects than missense variants\cite{do2015exome}.
-This can be reflected by providing larger priors for PTVs.
+This can be reflected by adjusting the prior spread of effect sizes ($\sigma$) for PTVs (\nameref{S1_Appendix}).
 
@@
 While the individual glaucoma analysis did not yield any associations with $\log_{10}$ BF greater than three, the joint analysis identified rare coding variants in both \textit{ANGPTL7} and \textit{WNT10A} as associated with glaucoma (Fig~\ref{glaucoma_bma}A-D, \nameref{glaucoma_table}).
-Applying the MRP model to identify protective effects also identified \textit{ANGPTL7} and \textit{WNT10A} (Fig~\ref{glaucoma_bma}E).
+Applying the MRP model identified \textit{ANGPTL7} with effects consistent with protection to glaucoma, and additional association to \textit{WNT10A} (Fig~\ref{glaucoma_bma}E).
 We obtained similar results using the independent effects (\nameref{glaucoma_independent}) or similar effects models (\nameref{glaucoma_similar}).

### 2018-03-12 tex/manuscript.tex
Updates
Minor figure and text updates
Updated text based on feedback and comments. Fixed a few small things in
figures.

@@
 Timothy Poterba\textsuperscript{3},
-Mark J. Daly\textsuperscript{3,4},
-Matti Pirinen\textsuperscript{5,6,7*},
+Carlos D. Bustamante \textsuperscript{1,4},
+Mark J. Daly\textsuperscript{3,5},
+Matti Pirinen\textsuperscript{6,7,8*},
 Manuel A. Rivas\textsuperscript{1*},
@@
 \\
-\textbf{4} Analytic and Translational Genetics Unit, Massachusetts General Hospital, Boston, MA, USA
+\textbf{4} Department of Genetics, Stanford University, Stanford, CA, USA
 \\
-\textbf{5} Institute for Molecular Medicine Finland (FIMM), University of Helsinki, Helsinki, Finland
+\textbf{5} Analytic and Translational Genetics Unit, Massachusetts General Hospital, Boston, MA, USA
 \\
-\textbf{6} Department of Public Health, University of Helsinki, Helsinki, Finland
+\textbf{6} Institute for Molecular Medicine Finland (FIMM), University of Helsinki, Helsinki, Finland
 \\
-\textbf{7} Helsinki Institute for Information Technology HIIT and Department of Mathematics and Statistics, University of Helsinki, Helsinki, Finland
+\textbf{7} Department of Public Health, University of Helsinki, Helsinki, Finland
+\\
+\textbf{8} Helsinki Institute for Information Technology HIIT and Department of Mathematics and Statistics, University of Helsinki, Helsinki, Finland
 % \textbf{1} Affiliation Dept/Program/Center, Institution Name, City, State, Country
@@
 We used the provided values from the UK Biobank for eosinophil counts, forced vital capacity (FVC), forced expiratory volume in 1-second (FEV$_1$), intra-ocular pressure, and corneal resistance factor.
-The phenotype codes used throughout (asthma=HC382, eosinophil count=INI30150, FEV$_1$=INI3063, FVC=INI3062, glaucoma=HC276, intra-ocular pressure=INI5263, and corneal resistance factor=INI5265) correspond to the phenotype codes used in on the Global Biobank Engine [\url{https://biobankengine.stanford.edu}].
+The phenotype codes used throughout (asthma=HC382, eosinophil count=INI30150, FEV$_1$=INI3063, FVC=INI3062, glaucoma=HC276, intra-ocular pressure=INI5263, and corneal resistance factor=INI5265) correspond to the phenotype codes used on the Global Biobank Engine [\url{https://biobankengine.stanford.edu}].
 
@@
 F: $-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{ANGPTL7} for all three phenotypes.
-G: Location of rs28991009 variant (green) for the protein 1JC9\_A homologous to \textit{ANGPTL7}.}
+G: Location of rs28991009 variant (green, p.Q175H, NM\_021146) for the protein 1JC9\_A homologous to \textit{ANGPTL7}.}
 \label{glaucoma_bma}
@@
 M.A.R.
+is supported by the GSP Coordinating Center (U24 HG008956).
+M.A.R.
 and C.D.
@@
 are paid consultants in Genomics PLC.
+CDB is a member of the scientific advisory boards for Liberty Biosecurity, Personalis, 23andMe Roots into the Future, Ancestry.com, IdentifyGenomics, and Etalon and is a founder of C.D.B.
+Consulting.
 

### 2018-03-13 tex/manuscript.tex
Update to acknowledgments

@@
 is financially supported by the Academy of Finland [288509 and 294050].
-M.A.R.
-is supported by the GSP Coordinating Center (U24 HG008956).
+C.D.B.
+and M.A.R.
+are supported by the GSP Coordinating Center (U24 HG008956).
 M.A.R.
 and C.D.
-are supported by Stanford University and a National Institute of Health center for Multi- and Trans-ethnic Mapping of Mendelian and Complex Diseases grant (5U01 HG009080).
+are supported by Stanford University and a National Institute of Health center for Multi- and Trans-ethnic Mapping of Mendelian and Complex Diseases grant (U01 HG009080).
+M.A.R.
+is a Faculty Fellow at the Stanford Center for Population Health Sciences.
 C.D.
@@
 Y .T.
-is supported by Funai Overseas Scholarship from Funai Foundation for Information Technology and the Stanford University Biomedical Informatics Training Program.
+is supported by Funai Overseas Scholarship from Funai Foundation for Information Technology and the Stanford University Biomedical Informatics Training Program (T32 LM012409).
 The primary and processed data used to generate the analyses presented here are available in the UK Biobank access management system (\url{https://amsportal.ukbiobank.ac.uk/}) for application 24983, ``Generating effective therapeutic hypotheses from genomic and hospital linkage data'' (\url{http://www.ukbiobank.ac.uk/wp-content/uploads/2017/06/24983-Dr-Manuel-Rivas.pdf}), and the results are displayed in the Global Biobank Engine (\url{https://biobankengine.stanford.edu}).

### 2018-03-16 tex/manuscriptbiorxiv.tex
Manuscript and cover letter updates
I updated the biorxiv version and pushed it to biorxiv. I also removed
the figures from manuscript.tex in preparation for submission. Minor
updates to cover letter.

@@
 \documentclass[10pt,letterpaper]{article}
-%\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
 
@@
 \begin{table}[!ht]
-%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
 \centering
@@
 \label{asthma_table}
-%\end{adjustwidth}
+\end{adjustwidth}
 \end{table}
@@
 \begin{table}[!ht]
-\begin{adjustwidth}{-1.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
 \centering
@@
 is financially supported by the Academy of Finland [288509 and 294050].
+C.D.B.
+and M.A.R.
+are supported by the GSP Coordinating Center (U24 HG008956).
 M.A.R.
 and C.D.
-are supported by Stanford University and a National Institute of Health center for Multi- and Trans-ethnic Mapping of Mendelian and Complex Diseases grant (5U01 HG009080).
+are supported by Stanford University and a National Institute of Health center for Multi- and Trans-ethnic Mapping of Mendelian and Complex Diseases grant (U01 HG009080).
+M.A.R.
+is a Faculty Fellow at the Stanford Center for Population Health Sciences.
 C.D.

### 2018-03-19 tex/manuscript.tex
Manuscript and cover letter updates
I updated the biorxiv version and pushed it to biorxiv. I also removed
the figures from manuscript.tex in preparation for submission. Minor
updates to cover letter.
Small update to author contributions
Updates

@@
 \begin{figure}[!h]
-\includegraphics[width=\textwidth]{../figures/final/Figure_01.pdf}
+%\includegraphics[width=\textwidth]{../figures/final/Figure_01.pdf}
 \caption{{\bf Schematic overview of MRP.}
@@
 \begin{figure}[!h]
-\includegraphics[width=\textwidth]{../figures/final/Figure_02.pdf}
+%\includegraphics[width=\textwidth]{../figures/final/Figure_02.pdf}
 \caption{{\bf Simulation studies.}
@@
 \begin{figure}[!h]
-\includegraphics[width=\textwidth]{../figures/final/Figure_03.pdf}
+%\includegraphics[width=\textwidth]{../figures/final/Figure_03.pdf}
 \caption{{\bf Results for asthma application.}
@@
 \begin{figure}[!h]
-\includegraphics[width=\textwidth]{../figures/final/Figure_04.pdf}
+%%\in\includegraphics[width=\textwidth]{../figures/final/Figure_04.pdf}
 \caption{{\bf Results for glaucoma application.}
@@
 
-\begin{figure}[H]
-\includegraphics[width=\textwidth]{../figures/final/Figure_S01.pdf}
-\end{figure}
+%\begin{figure}[H]
+%\includegraphics[width=\textwidth]{../figures/final/Figure_S01.pdf}
+%\end{figure}
 \paragraph*{S1 Fig.}
@@
 
-\begin{figure}[H]
-\includegraphics[width=\textwidth]{../figures/final/Figure_S02.pdf}
-\end{figure}
+%\begin{figure}[H]
+%\includegraphics[width=\textwidth]{../figures/final/Figure_S02.pdf}
+%\end{figure}
 \paragraph*{S2 Fig.}
@@
 %% FIGURE: ASTHMA MANHATTAN AND FOREST FOR INDEPENDENT EFFECTS
-\begin{figure}[H]
-\includegraphics[width=\textwidth]{../figures/final/Figure_S03.pdf}
-\end{figure}
+%\begin{{figure}[H]
+%\includegraphics[width=\textwidth]{../figures/final/Figure_S03.pdf}
+%\end{figure}
 \paragraph*{S3 Fig.}
@@
 %% FIGURE: ASTHMA MANHATTAN AND FOREST FOR SIMILAR EFFECTS
-\begin{figure}[H]
-\includegraphics[width=\textwidth]{../figures/final/Figure_S04.pdf}
-\end{figure}
+%\begin{figure}[H]
+%\includegraphics[width=\textwidth]{../figures/final/Figure_S04.pdf}
+%\end{figure}
 \paragraph*{S4 Fig.}
@@
 
-\begin{figure}[H]
-\includegraphics[width=\textwidth]{../figures/final/Figure_S05.pdf}
-\end{figure}
+%\begin{figure}[H]
+%\includegraphics[width=\textwidth]{../figures/final/Figure_S05.pdf}
+%\end{figure}
 \paragraph*{S5 Fig.}
@@
 
-\begin{figure}[H]
-\includegraphics[width=\textwidth]{../figures/final/Figure_S06.pdf}
-\end{figure}
+%\begin{figure}[H]
+%\includegraphics[width=\textwidth]{../figures/final/Figure_S06.pdf}
+%\end{figure}
 \paragraph*{S6 Fig.}
@@
 MAR, MP, and CD wrote the manuscript.
-MAR, MP, CCAS, YT, MA, and CD provided analysis and designed figures.
+MAR, MP, CCAS, YT, MA and CD provided analysis and designed figures.
 TP designed HDF5 tables and implementation of loaders.
-MJD provided critical feedback on methodology.
+MJD and CDB provided critical feedback on methodology.
 
@@
 are supported by the GSP Coordinating Center (U24 HG008956).
-M.A.R.
-and C.D.
+M.A.R., C.D., and C.D.B.
 are supported by Stanford University and a National Institute of Health center for Multi- and Trans-ethnic Mapping of Mendelian and Complex Diseases grant (U01 HG009080).
@@
 Consulting.
+The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.
 

### 2018-03-20 tex/manuscript.tex
Formatting update
I've changed the formatting according to the requests from the plos
genetics editor.

@@
 % Place figure captions after the first paragraph in which they are cited.
-\begin{figure}[!h]
-%\includegraphics[width=\textwidth]{../figures/final/Figure_01.pdf}
-\caption{{\bf Schematic overview of MRP.}
-A: MRP is suitable for a broad range of rare variant association study designs including (from left to right): i) case-control, ii) multiple diseases with shared controls, iii) single quantitative phenotype, and iv) mixture of case-control and quantitative phenotypes.
-B: Diagram of factors considered in rare variant association analysis including the correlation matrices: $\mathbf{R_{\textrm{study}}}$  (expected correlation of genetic effects among a group of studies), $\mathbf{R_{\textrm{var}}}$ (expected correlation of genetic effects among a group of variants), and $\mathbf{R_{\textrm{phen}}}$ (expected correlation of genetic effects among a group of phenotypes); the scale parameter for genetic variant annotation; and the location of genetic effects, which may be used to prioritize or identify protective modifiers of disease risk.}
-\label{overview}
-\end{figure}
 
@@
 
-\section*{Methods}
+\section*{Materials and Methods}
 
@@
 Although we see advantages in adopting a Bayesian perspective for MRP, our approach could be used in a frequentist context by calculating a BF and using it as a test statistic to compute p-values (\nameref{S1_Appendix}, Fig~\ref{simresults}).
-
-\begin{figure}[!h]
-%\includegraphics[width=\textwidth]{../figures/final/Figure_02.pdf}
-\caption{{\bf Simulation studies.}
-A: Comparison of $-\log_{10}$(p-values) from frequentist $\textrm{BF}_{\textrm{MRP}}$ approximation for an independent effects and a similar effects model to commonly used gene-based statistical tests (skat and burden).
-B: Comparison of log10(Bayes Factors) obtained when raw genotype and phenotype data is available to a scenario where summary statistics only was available and similar effects across studies is assumed.
-C: From single variant and single phenotype to multiple variants and multiple phenotypes gene discovery: ROC curves for detecting gene association to any of the phenotypes using single variant/single phenotype association (green) to multiple variants and multiple phenotypes association (purple).
-D: ROC curves for detecting gene association when incorporating prior mean of genetic effects (orange) to identify protective alleles.}
-\label{simresults}
-\end{figure}
 
@@
 
-%% FIGURE: ASTHMA MANHATTAN AND FOREST FOR BMA
-\begin{figure}[!h]
-%\includegraphics[width=\textwidth]{../figures/final/Figure_03.pdf}
-\caption{{\bf Results for asthma application.}
-$\log_{10}$ Bayes Factors from applying MRP and Bayesian model averaging to summary statistics for missense and protein-truncating variants from (A) asthma (HC382), (B) eosinophil counts (INI30150), (C) forced vital capacity (FVC, INI3062), (D) forced expiratory volume in 1-second (FEV$_1$, INI3063), (E) all four traits jointly, and (F) all four traits jointly with focus on protective effects.
-The four genes outside of chromosome 6 with the largest Bayes Factors greater than three are labeled in each plot.
-Only $\log_{10}$ Bayes Factors greater than -5 are plotted.
-(F) $-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{IL33} for each phenotype}
-\label{asthma_bma}
-\end{figure}
-
 \subsubsection*{Glaucoma, intra-ocular pressure, and corneal resistance factor}
@@
 %\textit{KLHL22} has not been previously associated with glaucoma though a suggestive association was reported for retinopathy in individuals without diabetes \cite{10.1371/journal.pone.0054232}.
-
-\begin{figure}[!h]
-%%\in\includegraphics[width=\textwidth]{../figures/final/Figure_04.pdf}
-\caption{{\bf Results for glaucoma application.}
-$\log_{10}$ Bayes Factors from applying MRP and Bayesian model averaging to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
-(E) shows the results of a joint analysis focused on finding rare variants that protect against glaucoma.
-The genes outside of chromosome 6 with with Bayes Factor greater than three are indicated by arrows.
-Only $\log_{10}$ Bayes Factors greater than zero are plotted.
-F: $-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{ANGPTL7} for all three phenotypes.
-G: Location of rs28991009 variant (green, p.Q175H, NM\_021146) for the protein 1JC9\_A homologous to \textit{ANGPTL7}.}
-\label{glaucoma_bma}
-\end{figure}
 
@@
 
-\section*{Supporting information}
+\section*{Acknowledgments}
+This research has been conducted using the UK Biobank Resource under Application Number 24983.
+We thank all the participants in the UK Biobank study.
+M.P.
+is financially supported by the Academy of Finland [288509 and 294050].
+C.D.B.
+and M.A.R.
+are supported by the GSP Coordinating Center (U24 HG008956).
+M.A.R., C.D., and C.D.B.
+are supported by Stanford University and a National Institute of Health center for Multi- and Trans-ethnic Mapping of Mendelian and Complex Diseases grant (U01 HG009080).
+M.A.R.
+is a Faculty Fellow at the Stanford Center for Population Health Sciences.
+C.D.
+is supported by a postdoctoral fellowship from the Stanford Center for Computational, Evolutionary, and Human Genomics and the Stanford ChEM-H Institute.
+Y .T.
+is supported by Funai Overseas Scholarship from Funai Foundation for Information Technology and the Stanford University Biomedical Informatics Training Program (T32 LM012409).
+The primary and processed data used to generate the analyses presented here are available in the UK Biobank access management system (\url{https://amsportal.ukbiobank.ac.uk/}) for application 24983, ``Generating effective therapeutic hypotheses from genomic and hospital linkage data'' (\url{http://www.ukbiobank.ac.uk/wp-content/uploads/2017/06/24983-Dr-Manuel-Rivas.pdf}), and the results are displayed in the Global Biobank Engine (\url{https://biobankengine.stanford.edu}).
+We would like to thank the Customer Solutions Team from Paradigm4 who helped us implement efficient databases for queries and application of inference methods to the data.
+M.A.R.
+and M.P.
+are paid consultants in Genomics PLC.
+CDB is a member of the scientific advisory boards for Liberty Biosecurity, Personalis, 23andMe Roots into the Future, Ancestry.com, IdentifyGenomics, and Etalon and is a founder of C.D.B.
+Consulting.
+The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.
+
+% \paragraph*{S2 Fig.}
+% \label{S2_Fig}
+% {\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+% 
+% \paragraph*{S1 File.}
+% \label{S1_File}
+% {\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+% 
+% \paragraph*{S1 Video.}
+% \label{S1_Video}
+% {\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+% 
+% \paragraph*{S1 Appendix.}
+% \label{S1_Appendix}
+% {\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+% 
+% \paragraph*{S1 Table.}
+% \label{S1_Table}
+% {\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+
+% I added in the Author Contributions section
+\section*{Author Contributions}
+MAR and MP designed the method and derived all analytical calculations.
+MAR, MP, and CD wrote the manuscript.
+MAR, MP, CCAS, YT, MA and CD provided analysis and designed figures.
+TP designed HDF5 tables and implementation of loaders.
+MJD and CDB provided critical feedback on methodology.
+
+\nolinenumbers
+
+% From the PLOS website (http://journals.plos.org/plosone/s/latex#loc-plos-template):
+% As stated in the PLOS template, your reference information should be included in your .tex file (not submitted separately as .bib or .bbl).
+Here is a step-by-step way to include your reference list directly within your .tex file:
+% 
+% Compose your LaTeX manuscript as normal, using \cite{bibkey} for reference citations.
+% Compile your manuscript using your .bib file (\bibliography{references}) and the plos2015.bst style file.  This process should output a .bbl file into the same folder as your manuscript.
+% Open the .bbl file and copy/paste its contents into the appropriate position within the manuscript .tex.  Comment out or delete the \bibliography command.
+% Compile your manuscript again. The PDF output should be the same as before.
+
+% Either type in your references using
+% \begin{thebibliography}{}
+% \bibitem{}
+% Text
+% \end{thebibliography}
+%
+% or
+%
+% Compile your BiBTeX database using our plos2015.bst
+% style file and paste the contents of your .bbl file
+% here. See http://journals.plos.org/plosone/s/latex for 
+% step-by-step instructions.
+%
+
+% \begin{thebibliography}{10}
+% 
+% \bibitem{bib1}
+% Conant GC, Wolfe KH.
+% \newblock {{T}urning a hobby into a job: how duplicated genes find new
+%   functions}.
+% \newblock Nat Rev Genet. 2008 Dec;9(12):938--950.
+% 
+% \bibitem{bib2}
+% Ohno S.
+% \newblock Evolution by gene duplication.
+% \newblock London: George Alien \& Unwin Ltd. Berlin, Heidelberg and New York:
+%   Springer-Verlag.; 1970.
+% 
+% \bibitem{bib3}
+% Magwire MM, Bayer F, Webster CL, Cao C, Jiggins FM.
+% \newblock {{S}uccessive increases in the resistance of {D}rosophila to viral
+%   infection through a transposon insertion followed by a {D}uplication}.
+% \newblock PLoS Genet. 2011 Oct;7(10):e1002337.
+% 
+% \end{thebibliography}
+
+\bibliography{references}
+
+\section*{Figure Legends}
+
+%\begin{figure}[!h]
+%\includegraphics[width=\textwidth]{../figures/final/Figure_01.pdf}
+%\caption{{\bf Schematic overview of MRP.}
+\paragraph*{Fig 1. Schematic overview of MRP.}
+\label{overview}
+A: MRP is suitable for a broad range of rare variant association study designs including (from left to right): i) case-control, ii) multiple diseases with shared controls, iii) single quantitative phenotype, and iv) mixture of case-control and quantitative phenotypes.
+B: Diagram of factors considered in rare variant association analysis including the correlation matrices: $\mathbf{R_{\textrm{study}}}$  (expected correlation of genetic effects among a group of studies), $\mathbf{R_{\textrm{var}}}$ (expected correlation of genetic effects among a group of variants), and $\mathbf{R_{\textrm{phen}}}$ (expected correlation of genetic effects among a group of phenotypes); the scale parameter for genetic variant annotation; and the location of genetic effects, which may be used to prioritize or identify protective modifiers of disease risk.
+%\end{figure}
+
+%\begin{figure}[!h]
+%\includegraphics[width=\textwidth]{../figures/final/Figure_02.pdf}
+%\caption{{\bf Simulation studies.}
+\paragraph*{Fig 2. Simulation studies.}
+\label{simresults}
+A: Comparison of $-\log_{10}$(p-values) from frequentist $\textrm{BF}_{\textrm{MRP}}$ approximation for an independent effects and a similar effects model to commonly used gene-based statistical tests (skat and burden).
+B: Comparison of log10(Bayes Factors) obtained when raw genotype and phenotype data is available to a scenario where summary statistics only was available and similar effects across studies is assumed.
+C: From single variant and single phenotype to multiple variants and multiple phenotypes gene discovery: ROC curves for detecting gene association to any of the phenotypes using single variant/single phenotype association (green) to multiple variants and multiple phenotypes association (purple).
+D: ROC curves for detecting gene association when incorporating prior mean of genetic effects (orange) to identify protective alleles.
+%\end{figure}
+
+%% FIGURE: ASTHMA MANHATTAN AND FOREST FOR BMA
+%\begin{figure}[!h]
+%\includegraphics[width=\textwidth]{../figures/final/Figure_03.pdf}
+%\caption{{\bf Results for asthma application.}
+\paragraph*{Fig 3. Results for asthma application.}
+\label{asthma_bma}
+$\log_{10}$ Bayes Factors from applying MRP and Bayesian model averaging to summary statistics for missense and protein-truncating variants from (A) asthma (HC382), (B) eosinophil counts (INI30150), (C) forced vital capacity (FVC, INI3062), (D) forced expiratory volume in 1-second (FEV$_1$, INI3063), (E) all four traits jointly, and (F) all four traits jointly with focus on protective effects.
+The four genes outside of chromosome 6 with the largest Bayes Factors greater than three are labeled in each plot.
+Only $\log_{10}$ Bayes Factors greater than -5 are plotted.
+(F) $-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{IL33} for each phenotype
+%\end{figure}
+
+%\begin{figure}[!h]
+%%\in\includegraphics[width=\textwidth]{../figures/final/Figure_04.pdf}
+%\caption{{\bf Results for glaucoma application.}
+\paragraph*{Fig 4. Results for glaucoma application.}
+\label{glaucoma_bma}
+$\log_{10}$ Bayes Factors from applying MRP and Bayesian model averaging to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
+(E) shows the results of a joint analysis focused on finding rare variants that protect against glaucoma.
+The genes outside of chromosome 6 with with Bayes Factor greater than three are indicated by arrows.
+Only $\log_{10}$ Bayes Factors greater than zero are plotted.
+F: $-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{ANGPTL7} for all three phenotypes.
+G: Location of rs28991009 variant (green, p.Q175H, NM\_021146) for the protein 1JC9\_A homologous to \textit{ANGPTL7}.
+%\end{figure}
+
+\section*{Supporting Information Legends}
 
@@
 Only $\log_{10}$ Bayes Factors greater than zero are plotted.
+
+\section*{Tables}
 
@@
 
-% \paragraph*{S2 Fig.}
-% \label{S2_Fig}
-% {\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-% 
-% \paragraph*{S1 File.}
-% \label{S1_File}
-% {\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-% 
-% \paragraph*{S1 Video.}
-% \label{S1_Video}
-% {\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-% 
-% \paragraph*{S1 Appendix.}
-% \label{S1_Appendix}
-% {\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-% 
-% \paragraph*{S1 Table.}
-% \label{S1_Table}
-% {\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
-% I added in the Author Contributions section
-\section*{Author Contributions}
-MAR and MP designed the method and derived all analytical calculations.
-MAR, MP, and CD wrote the manuscript.
-MAR, MP, CCAS, YT, MA and CD provided analysis and designed figures.
-TP designed HDF5 tables and implementation of loaders.
-MJD and CDB provided critical feedback on methodology.
-
-\section*{Acknowledgments}
-This research has been conducted using the UK Biobank Resource under Application Number 24983.
-We thank all the participants in the UK Biobank study.
-M.P.
-is financially supported by the Academy of Finland [288509 and 294050].
-C.D.B.
-and M.A.R.
-are supported by the GSP Coordinating Center (U24 HG008956).
-M.A.R., C.D., and C.D.B.
-are supported by Stanford University and a National Institute of Health center for Multi- and Trans-ethnic Mapping of Mendelian and Complex Diseases grant (U01 HG009080).
-M.A.R.
-is a Faculty Fellow at the Stanford Center for Population Health Sciences.
-C.D.
-is supported by a postdoctoral fellowship from the Stanford Center for Computational, Evolutionary, and Human Genomics and the Stanford ChEM-H Institute.
-Y .T.
-is supported by Funai Overseas Scholarship from Funai Foundation for Information Technology and the Stanford University Biomedical Informatics Training Program (T32 LM012409).
-The primary and processed data used to generate the analyses presented here are available in the UK Biobank access management system (\url{https://amsportal.ukbiobank.ac.uk/}) for application 24983, ``Generating effective therapeutic hypotheses from genomic and hospital linkage data'' (\url{http://www.ukbiobank.ac.uk/wp-content/uploads/2017/06/24983-Dr-Manuel-Rivas.pdf}), and the results are displayed in the Global Biobank Engine (\url{https://biobankengine.stanford.edu}).
-We would like to thank the Customer Solutions Team from Paradigm4 who helped us implement efficient databases for queries and application of inference methods to the data.
-M.A.R.
-and M.P.
-are paid consultants in Genomics PLC.
-CDB is a member of the scientific advisory boards for Liberty Biosecurity, Personalis, 23andMe Roots into the Future, Ancestry.com, IdentifyGenomics, and Etalon and is a founder of C.D.B.
-Consulting.
-The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.
-
-\nolinenumbers
-
-% From the PLOS website (http://journals.plos.org/plosone/s/latex#loc-plos-template):
-% As stated in the PLOS template, your reference information should be included in your .tex file (not submitted separately as .bib or .bbl).
-Here is a step-by-step way to include your reference list directly within your .tex file:
-% 
-% Compose your LaTeX manuscript as normal, using \cite{bibkey} for reference citations.
-% Compile your manuscript using your .bib file (\bibliography{references}) and the plos2015.bst style file.  This process should output a .bbl file into the same folder as your manuscript.
-% Open the .bbl file and copy/paste its contents into the appropriate position within the manuscript .tex.  Comment out or delete the \bibliography command.
-% Compile your manuscript again. The PDF output should be the same as before.
-
-% Either type in your references using
-% \begin{thebibliography}{}
-% \bibitem{}
-% Text
-% \end{thebibliography}
-%
-% or
-%
-% Compile your BiBTeX database using our plos2015.bst
-% style file and paste the contents of your .bbl file
-% here. See http://journals.plos.org/plosone/s/latex for 
-% step-by-step instructions.
-%
-
-% \begin{thebibliography}{10}
-% 
-% \bibitem{bib1}
-% Conant GC, Wolfe KH.
-% \newblock {{T}urning a hobby into a job: how duplicated genes find new
-%   functions}.
-% \newblock Nat Rev Genet. 2008 Dec;9(12):938--950.
-% 
-% \bibitem{bib2}
-% Ohno S.
-% \newblock Evolution by gene duplication.
-% \newblock London: George Alien \& Unwin Ltd. Berlin, Heidelberg and New York:
-%   Springer-Verlag.; 1970.
-% 
-% \bibitem{bib3}
-% Magwire MM, Bayer F, Webster CL, Cao C, Jiggins FM.
-% \newblock {{S}uccessive increases in the resistance of {D}rosophila to viral
-%   infection through a transposon insertion followed by a {D}uplication}.
-% \newblock PLoS Genet. 2011 Oct;7(10):e1002337.
-% 
-% \end{thebibliography}
-
-\bibliography{references}
-
 \end{document}

### 2018-03-20 tex/manuscript.tex
Minor updates

@@
 Through simulation studies, we demonstrate that the proposed model comparison approach can improve ability to detect rare variant association signals.
-We also apply the model to two groups of phenotypes from the UK Biobank: 1) asthma diagnosis, eosinophil counts, forced expiratory volume, and forced vital capacity; and 2) glaucoma diagnosis, intra-ocular pressure, and corneal resistance factor.
-We are able to recover known associations such as the protective association between rs146597587 in \textit{IL33} and asthma.
-We also find evidence for novel protective associations between rare variants in \textit{ANGPTL7} and glaucoma.
+We also apply the model to two groups of phenotypes from the UK Biobank: 1) asthma diagnosis (43,626 cases), eosinophil counts, forced expiratory volume, and forced vital capacity; and 2) glaucoma diagnosis (5,863 cases), intra-ocular pressure, and corneal resistance factor.
+We are able to recover known associations such as the protective association between rs146597587 in \textit{IL33} and asthma ($\log_{10} \left( \textrm{Bayes Factor}\right) = 29.3$).
+We also find evidence for novel protective associations between rare variants in \textit{ANGPTL7} and glaucoma ($\log_{10} \left( \textrm{Bayes Factor}\right) = 11.9$).
 Overall, we show that the MRP model comparison approach is able to retain and improve upon useful features from widely-used meta-analysis approaches for rare variant association analyses and prioritize protective modifiers of disease risk.

### 2018-04-14 tex/appendix.tex

@@
 }
-\newline
+
+{\LARGE
+\textbf\newline{Manuel A.
+Rivas and Matti Pirinen} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+}
+{\large
+\textbf\newline{January 28, 2018} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+}
+
+%\newline
 
@@
 where $\Delta\left(\sigma_m\right)$ is a diagonal matrix with entries $\sigma_m$ determining the spread of the effect size distribution for each variant $m \leq M$.
-Thus, we can model settings where, e.g., protein-truncating variants have larger effect sizes ($\sigma = 1$) than missense variants ($\sigma = 0.1$).
+Thus, we can model settings where, e.g., protein-truncating variants have larger effect sizes ($\sigma = 0.5$) than missense variants ($\sigma = 0.2$).
 Note that when $\sigma_m = 1$ for all $m$ then $\mathbf{S}_{\textrm{var}} = \mathbf{R}_{\textrm{var}}$.

### 2018-04-14 tex/manuscript.tex

@@
 We also apply the model to two groups of phenotypes from the UK Biobank: 1) asthma diagnosis (43,626 cases), eosinophil counts, forced expiratory volume, and forced vital capacity; and 2) glaucoma diagnosis (5,863 cases), intra-ocular pressure, and corneal resistance factor.
-We are able to recover known associations such as the protective association between rs146597587 in \textit{IL33} and asthma ($\log_{10} \left( \textrm{Bayes Factor}\right) = 29.3$).
+We are able to recover known associations such as the protective association between rs146597587 in \textit{IL33} and asthma ($\log_{10} \left( \textrm{Bayes Factor}\right) = 29.4$).
 We also find evidence for novel protective associations between rare variants in \textit{ANGPTL7} and glaucoma ($\log_{10} \left( \textrm{Bayes Factor}\right) = 11.9$).
@@
 We applied both independent and similar effects models and used Bayesian model averaging to compute a single BF per gene \cite{10.2307/2676803}.
-In agreement with previous studies, we observed evidence that rare missense variants and/or PTVs in \textit{IL33} affect eosinophil counts and offer protection from asthma from the single-phenotype analyses, though the evidence of association was strongest for the joint analysis (\nameref{asthma_table}) \cite{DeBoever179762, 10.1371/journal.pgen.1006659}.
-We performed an analysis focused on identifying protective variants which also identified the \textit{IL33} association (Fig~\ref{asthma_bma}F).
+In agreement with previous studies, we observed evidence that rare missense variants and/or PTVs in \textit{IL33} affect eosinophil counts and offer protection from asthma from the single-phenotype analyses, though the evidence of association was strongest for the joint analysis ($\log_{10} \textrm{BF} = 29.3$, \nameref{asthma_table}) \cite{DeBoever179762, 10.1371/journal.pgen.1006659}.
+We performed an analysis focused on identifying protective variants which also identified the \textit{IL33} association ($\log_{10} \textrm{BF} = 29.4$, Fig~\ref{asthma_bma}F).
 The results were similar using only either the independent effects (\nameref{asthma_independent}) or similar effects models (\nameref{asthma_similar}).
@@
 Corneal resistance factor is a measure of the cornea's ability to resist mechanical stress and has been associated with glaucoma presence and severity \cite{CXO:CXO414, MANSOURI2012419, GriseDulac:2012bo}.
-While the individual glaucoma analysis did not yield any associations with $\log_{10}$ BF greater than three, the joint analysis identified rare coding variants in both \textit{ANGPTL7} and \textit{WNT10A} as associated with glaucoma (Fig~\ref{glaucoma_bma}A-D, \nameref{glaucoma_table}).
-Applying the MRP model identified \textit{ANGPTL7} with effects consistent with protection to glaucoma, and additional association to \textit{WNT10A} (Fig~\ref{glaucoma_bma}E).
+While the individual glaucoma analysis did not yield any associations with $\log_{10}$ BF greater than three, the joint analysis identified rare coding variants in both \textit{ANGPTL7} ($\log_{10} \textrm{BF} = 11.9$) and \textit{WNT10A} ($\log_{10} \textrm{BF} = 2.5$) as associated with glaucoma (Fig~\ref{glaucoma_bma}A-D, \nameref{glaucoma_table}).
+Applying the protective MRP model also identified the protective association for \textit{ANGPTL7} against glaucoma and added support for an association for \textit{WNT10A} (Fig~\ref{glaucoma_bma}E).
 We obtained similar results using the independent effects (\nameref{glaucoma_independent}) or similar effects models (\nameref{glaucoma_similar}).
@@
 Such tissue-specific activity may make \textit{ANGPTL7} a useful therapeutic target.
-\textit{WNT10A} has not been previously associated with glaucoma though an exonic variant rs121908120 in \textit{WNT10A} is associated with central cornea thickness and increased risk of keratoconus, a disease of the cornea, indicating that this gene may play a role in ocular diseases.
+\textit{WNT10A} has not been previously associated with glaucoma though an exonic variant rs121908120 in \textit{WNT10A} is associated with central cornea thickness and increased risk of keratoconus, a disease of the cornea, indicating that this gene may play a role in ocular diseases \cite{CuellarPartida:2015hb}.
 
@@
 \thickhline
-\textit{ANGPTL7} &              11.9 &  11.9 &      1.7 &                                       10.1 &                       6.2 \\
+\textit{PGLYRP4} &              23.9 &   7.6 &     -0.1 &                                       -1.0 &                      -1.1 \\
 \hline
-\textit{KLHL22}  &               3.5 &   3.5 &     -0.2 &                                        3.9 &                       2.2 \\
+\textit{ANGPTL7} &              12.9 &  12.2 &      1.7 &                                       10.5 &                       6.4 \\
 \hline
-\textit{WNT10A}  &               2.5 &   2.5 &     -0.3 &                                       -0.5 &                       3.5 \\
+\textit{JPH4}    &               6.4 &   2.7 &      0.6 &                                       -1.1 &                      -1.2 \\
 \hline
-\textit{GLI3}    &               1.5 &   1.5 &     -0.3 &                                       -0.7 &                       3.2 \\
+\textit{WNT10A}  &               3.5 &   4.1 &     -0.2 &                                       -0.6 &                       3.1 \\
 \hline
-\textit{RNASE2}  &              -0.5 &  -0.5 &      0.4 &                                       -1.0 &                      -1.1 \\
+\textit{KLHL22}  &               3.1 &   3.7 &     -0.2 &                                        3.9 &                       2.2 \\
 \hline
-\textit{FUK}     &              -1.8 &  -1.8 &      2.5 &                                       -2.1 &                      -2.1 \\
+\textit{GLI3}    &               1.4 &   2.5 &     -0.4 &                                       -0.1 &                       4.3 \\
+\hline
+\textit{C4orf29} &              -0.0 &   6.4 &     -0.1 &                                       -0.7 &                      -1.1 \\
 \hline

### 2018-04-14 tex/manuscriptbiorxiv.tex
Small update to author contributions

@@
 Through simulation studies, we demonstrate that the proposed model comparison approach can improve ability to detect rare variant association signals.
-We also apply the model to two groups of phenotypes from the UK Biobank: 1) asthma diagnosis, eosinophil counts, forced expiratory volume, and forced vital capacity; and 2) glaucoma diagnosis, intra-ocular pressure, and corneal resistance factor.
-We are able to recover known associations such as the protective association between rs146597587 in \textit{IL33} and asthma.
-We also find evidence for novel protective associations between rare variants in \textit{ANGPTL7} and glaucoma.
+We also apply the model to two groups of phenotypes from the UK Biobank: 1) asthma diagnosis (43,626 cases), eosinophil counts, forced expiratory volume, and forced vital capacity; and 2) glaucoma diagnosis (5,863 cases), intra-ocular pressure, and corneal resistance factor.
+We are able to recover known associations such as the protective association between rs146597587 in \textit{IL33} and asthma ($\log_{10} \left( \textrm{Bayes Factor}\right) = 29.4$).
+We also find evidence for novel protective associations between rare variants in \textit{ANGPTL7} and glaucoma ($\log_{10} \left( \textrm{Bayes Factor}\right) = 11.9$).
 Overall, we show that the MRP model comparison approach is able to retain and improve upon useful features from widely-used meta-analysis approaches for rare variant association analyses and prioritize protective modifiers of disease risk.
@@
 % Place figure captions after the first paragraph in which they are cited.
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{../figures/final/Figure_01.pdf}
-\caption{{\bf Schematic overview of MRP.}
-A: MRP is suitable for a broad range of rare variant association study designs including (from left to right): i) case-control, ii) multiple diseases with shared controls, iii) single quantitative phenotype, and iv) mixture of case-control and quantitative phenotypes.
-B: Diagram of factors considered in rare variant association analysis including the correlation matrices: $\mathbf{R_{\textrm{study}}}$  (expected correlation of genetic effects among a group of studies), $\mathbf{R_{\textrm{var}}}$ (expected correlation of genetic effects among a group of variants), and $\mathbf{R_{\textrm{phen}}}$ (expected correlation of genetic effects among a group of phenotypes); the scale parameter for genetic variant annotation; and the location of genetic effects, which may be used to prioritize or identify protective modifiers of disease risk.}
-\label{overview}
-\end{figure}
 
@@
 Although we see advantages in adopting a Bayesian perspective for MRP, our approach could be used in a frequentist context by calculating a BF and using it as a test statistic to compute p-values (\nameref{S1_Appendix}, Fig~\ref{simresults}).
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{../figures/final/Figure_02.pdf}
-\caption{{\bf Simulation studies.}
-A: Comparison of $-\log_{10}$(p-values) from frequentist $\textrm{BF}_{\textrm{MRP}}$ approximation for an independent effects and a similar effects model to commonly used gene-based statistical tests (skat and burden).
-B: Comparison of log10(Bayes Factors) obtained when raw genotype and phenotype data is available to a scenario where summary statistics only was available and similar effects across studies is assumed.
-C: From single variant and single phenotype to multiple variants and multiple phenotypes gene discovery: ROC curves for detecting gene association to any of the phenotypes using single variant/single phenotype association (green) to multiple variants and multiple phenotypes association (purple).
-D: ROC curves for detecting gene association when incorporating prior mean of genetic effects (orange) to identify protective alleles.}
-\label{simresults}
-\end{figure}
 
@@
 
-%% FIGURE: ASTHMA MANHATTAN AND FOREST FOR BMA
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{../figures/final/Figure_03.pdf}
-\caption{{\bf Results for asthma application.}
-$\log_{10}$ Bayes Factors from applying MRP and Bayesian model averaging to summary statistics for missense and protein-truncating variants from (A) asthma (HC382), (B) eosinophil counts (INI30150), (C) forced vital capacity (FVC, INI3062), (D) forced expiratory volume in 1-second (FEV$_1$, INI3063), (E) all four traits jointly, and (F) all four traits jointly with focus on protective effects.
-The four genes outside of chromosome 6 with the largest Bayes Factors greater than three are labeled in each plot.
-Only $\log_{10}$ Bayes Factors greater than -5 are plotted.
-(F) $-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{IL33} for each phenotype}
-\label{asthma_bma}
-\end{figure}
-
 \subsubsection*{Glaucoma, intra-ocular pressure, and corneal resistance factor}
@@
 Such tissue-specific activity may make \textit{ANGPTL7} a useful therapeutic target.
-\textit{WNT10A} has not been previously associated with glaucoma though an exonic variant rs121908120 in \textit{WNT10A} is associated with central cornea thickness and increased risk of keratoconus, a disease of the cornea, indicating that this gene may play a role in ocular diseases.
+\textit{WNT10A} has not been previously associated with glaucoma though an exonic variant rs121908120 in \textit{WNT10A} is associated with central cornea thickness and increased risk of keratoconus, a disease of the cornea, indicating that this gene may play a role in ocular diseases \cite{CuellarPartida:2015hb}.
 
 %\textit{KLHL22} has not been previously associated with glaucoma though a suggestive association was reported for retinopathy in individuals without diabetes \cite{10.1371/journal.pone.0054232}.
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{../figures/final/Figure_04.pdf}
-\caption{{\bf Results for glaucoma application.}
-$\log_{10}$ Bayes Factors from applying MRP and Bayesian model averaging to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly.
-(E) shows the results of a joint analysis focused on finding rare variants that protect against glaucoma.
-The genes outside of chromosome 6 with with Bayes Factor greater than three are indicated by arrows.
-Only $\log_{10}$ Bayes Factors greater than zero are plotted.
-F: $-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{ANGPTL7} for all three phenotypes.
-G: Location of rs28991009 variant (green, p.Q175H, NM\_021146) for the protein 1JC9\_A homologous to \textit{ANGPTL7}.}
-\label{glaucoma_bma}
-\end{figure}
 
@@
 
-% \paragraph*{S2 Fig.}
-% \label{S2_Fig}
-% {\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-% 
-% \paragraph*{S1 File.}
-% \label{S1_File}
-% {\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-% 
-% \paragraph*{S1 Video.}
-% \label{S1_Video}
-% {\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-% 
-% \paragraph*{S1 Appendix.}
-% \label{S1_Appendix}
-% {\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-% 
-% \paragraph*{S1 Table.}
-% \label{S1_Table}
-% {\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
-% I added in the Author Contributions section
-\section*{Author Contributions}
-MAR and MP designed the method and derived all analytical calculations.
-MAR, MP, and CD wrote the manuscript.
-MAR, MP, CCAS, YT, MA, and CD provided analysis and designed figures.
-TP designed HDF5 tables and implementation of loaders.
-MJD provided critical feedback on methodology.
-
-\section*{Acknowledgments}
-This research has been conducted using the UK Biobank Resource under Application Number 24983.
-We thank all the participants in the UK Biobank study.
-M.P.
-is financially supported by the Academy of Finland [288509 and 294050].
-C.D.B.
-and M.A.R.
-are supported by the GSP Coordinating Center (U24 HG008956).
-M.A.R.
-and C.D.
-are supported by Stanford University and a National Institute of Health center for Multi- and Trans-ethnic Mapping of Mendelian and Complex Diseases grant (U01 HG009080).
-M.A.R.
-is a Faculty Fellow at the Stanford Center for Population Health Sciences.
-C.D.
-is supported by a postdoctoral fellowship from the Stanford Center for Computational, Evolutionary, and Human Genomics and the Stanford ChEM-H Institute.
-Y .T.
-is supported by Funai Overseas Scholarship from Funai Foundation for Information Technology and the Stanford University Biomedical Informatics Training Program (T32 LM012409).
-The primary and processed data used to generate the analyses presented here are available in the UK Biobank access management system (\url{https://amsportal.ukbiobank.ac.uk/}) for application 24983, ``Generating effective therapeutic hypotheses from genomic and hospital linkage data'' (\url{http://www.ukbiobank.ac.uk/wp-content/uploads/2017/06/24983-Dr-Manuel-Rivas.pdf}), and the results are displayed in the Global Biobank Engine (\url{https://biobankengine.stanford.edu}).
-We would like to thank the Customer Solutions Team from Paradigm4 who helped us implement efficient databases for queries and application of inference methods to the data.
-M.A.R.
-and M.P.
-are paid consultants in Genomics PLC.
-CDB is a member of the scientific advisory boards for Liberty Biosecurity, Personalis, 23andMe Roots into the Future, Ancestry.com, IdentifyGenomics, and Etalon and is a founder of C.D.B.
-Consulting.
-
-\nolinenumbers
-
-% From the PLOS website (http://journals.plos.org/plosone/s/latex#loc-plos-template):
-% As stated in the PLOS template, your reference information should be included in your .tex file (not submitted separately as .bib or .bbl).
-Here is a step-by-step way to include your reference list directly within your .tex file:
-% 
-% Compose your LaTeX manuscript as normal, using \cite{bibkey} for reference citations.
-% Compile your manuscript using your .bib file (\bibliography{references}) and the plos2015.bst style file.  This process should output a .bbl file into the same folder as your manuscript.
-% Open the .bbl file and copy/paste its contents into the appropriate position within the manuscript .tex.  Comment out or delete the \bibliography command.
-% Compile your manuscript again. The PDF output should be the same as before.
-
-% Either type in your references using
-% \begin{thebibliography}{}
-% \bibitem{}
-% Text
-% \end{thebibliography}
-%
-% or
-%
-% Compile your BiBTeX database using our plos2015.bst
-% style file and paste the contents of your .bbl file
-% here. See http://journals.plos.org/plosone/s/latex for 
-% step-by-step instructions.
-%
-
-% \begin{thebibliography}{10}
-% 
-% \bibitem{bib1}
-% Conant GC, Wolfe KH.
-% \newblock {{T}urning a hobby into a job: how duplicated genes find new
-%   functions}.
-% \newblock Nat Rev Genet. 2008 Dec;9(12):938--950.
-% 
-% \bibitem{bib2}
-% Ohno S.
-% \newblock Evolution by gene duplication.
-% \newblock London: George Alien \& Unwin Ltd. Berlin, Heidelberg and New York:
-%   Springer-Verlag.; 1970.
-% 
-% \bibitem{bib3}
-% Magwire MM, Bayer F, Webster CL, Cao C, Jiggins FM.
-% \newblock {{S}uccessive increases in the resistance of {D}rosophila to viral
-%   infection through a transposon insertion followed by a {D}uplication}.
-% \newblock PLoS Genet. 2011 Oct;7(10):e1002337.
-% 
-% \end{thebibliography}
-
-\bibliography{references}
-
 \end{document}

### 2019-10-20 tex/manuscript.tex
Updates, new files

@@
 We are able to recover known associations such as the protective association between rs146597587 in \textit{IL33} and asthma ($\log_{10} \left( \textrm{Bayes Factor}\right) = 29.4$).
-We also find evidence for novel protective associations between rare variants in \textit{ANGPTL7} and glaucoma ($\log_{10} \left( \textrm{Bayes Factor}\right) = 11.9$).
+We also find evidence for novel protective associations between rare variants in \textit{ANGPTL7} and glaucoma ($\log_{10} \left( \textrm{Bayes Factor}\right) = 13.1$).
 Overall, we show that the MRP model comparison approach is able to retain and improve upon useful features from widely-used meta-analysis approaches for rare variant association analyses and prioritize protective modifiers of disease risk.
@@
 \subsection*{UK Biobank Asthma and Glaucoma Applications}
-We applied MRP for both sets of traits assuming both an independent effects model and a similar effects model.
-We applied Bayesian model averaging to these results by averaging the BF from each model for each gene; these results are reported in the main text.
-The results for each individual model are included in the Supporting Information.
+For each group of traits (asthma, eosinophil counts, FVC, FEV$_1$ and glaucoma, intra-ocular pressure, corneal resistance factor), we applied MRP individually to each phenotype as well as performing a joint analysis using all traits.
+We also applied a model that prioritizes protective variants where we used non-zero priors for the variant effect size of $-0.5$ for PTVs and $-0.2$ for missense alleles.
+For each analysis, we applied MRP assuming an independent effects model and a similar effects model.
+We applied Bayesian model averaging to the results of the independent and similar effects models by summing the $\log_{10}$ BF for each gene from each model and dividing by two.
+The Bayesian model averaging results are reported in the main text while the results for each individual model are included in the Supporting Information.
+
 For the Manhattan plots and tables, we removed any genes with non-unique gene symbols.
@@
 $\texttt{stop\_gained}$, $\texttt{frameshift\_variant}$, $\texttt{splice\_acceptor\_variant}$, $\texttt{splice\_donor\_variant}$, $\texttt{splice\_region\_variant}$, $\texttt{start\_lost}$, $\texttt{stop\_lost}$.
+We removed variants whose regression effect size had standard error greater than 0.15.
 
@@
 Corneal resistance factor is a measure of the cornea's ability to resist mechanical stress and has been associated with glaucoma presence and severity \cite{CXO:CXO414, MANSOURI2012419, GriseDulac:2012bo}.
-While the individual glaucoma analysis did not yield any associations with $\log_{10}$ BF greater than three, the joint analysis identified rare coding variants in both \textit{ANGPTL7} ($\log_{10} \textrm{BF} = 11.9$) and \textit{WNT10A} ($\log_{10} \textrm{BF} = 2.5$) as associated with glaucoma (Fig~\ref{glaucoma_bma}A-D, \nameref{glaucoma_table}).
-Applying the protective MRP model also identified the protective association for \textit{ANGPTL7} against glaucoma and added support for an association for \textit{WNT10A} (Fig~\ref{glaucoma_bma}E).
+While the individual glaucoma analysis did not yield any associations with $\log_{10}$ BF greater than three, the joint analysis identified rare coding variants in \textit{ANGPTL7} ($\log_{10} \textrm{BF} = 12.2$), \textit{KLHL22} ($\log_{10} \textrm{BF} = 3.7$), and \textit{WNT10A} ($\log_{10} \textrm{BF} = 2.6$) as associated with glaucoma (Fig~\ref{glaucoma_bma}A-D, \nameref{glaucoma_table}).
+Applying the protective MRP model also identified the protective association for \textit{ANGPTL7} against glaucoma and added support for associations for \textit{KLHL22} and \textit{WNT10A} (Fig~\ref{glaucoma_bma}E).
 We obtained similar results using the independent effects (\nameref{glaucoma_independent}) or similar effects models (\nameref{glaucoma_similar}).
@@
 Such tissue-specific activity may make \textit{ANGPTL7} a useful therapeutic target.
-\textit{WNT10A} has not been previously associated with glaucoma though an exonic variant rs121908120 in \textit{WNT10A} is associated with central cornea thickness and increased risk of keratoconus, a disease of the cornea, indicating that this gene may play a role in ocular diseases \cite{CuellarPartida:2015hb}.
-
-%\textit{KLHL22} has not been previously associated with glaucoma though a suggestive association was reported for retinopathy in individuals without diabetes \cite{10.1371/journal.pone.0054232}.
+\textit{KLHL22} has not been previously associated with glaucoma though a suggestive association was reported for retinopathy in individuals without diabetes \cite{10.1371/journal.pone.0054232}.
+\textit{WNT10A} also has not been previously associated with glaucoma though an exonic variant rs121908120 in \textit{WNT10A} is associated with central cornea thickness and increased risk of keratoconus, a disease of the cornea, indicating that this gene may play a role in ocular diseases \cite{CuellarPartida:2015hb}.
 
@@
 
+\pagebreak
+
 \section*{Figure Legends}
 
-%\begin{figure}[!h]
+\begin{figure}[!h]
 %\includegraphics[width=\textwidth]{../figures/final/Figure_01.pdf}
-%\caption{{\bf Schematic overview of MRP.}
-\paragraph*{Fig 1. Schematic overview of MRP.}
+\caption{{\bf Schematic overview of MRP.}}
+%\paragraph*{Fig 1. Schematic overview of MRP.}
 \label{overview}
@@
 B: Diagram of factors considered in rare variant association analysis including the correlation matrices: $\mathbf{R_{\textrm{study}}}$  (expected correlation of genetic effects among a group of studies), $\mathbf{R_{\textrm{var}}}$ (expected correlation of genetic effects among a group of variants), and $\mathbf{R_{\textrm{phen}}}$ (expected correlation of genetic effects among a group of phenotypes); the scale parameter for genetic variant annotation; and the location of genetic effects, which may be used to prioritize or identify protective modifiers of disease risk.
-%\end{figure}
+\end{figure}
 
-%\begin{figure}[!h]
+\begin{figure}[!h]
 %\includegraphics[width=\textwidth]{../figures/final/Figure_02.pdf}
-%\caption{{\bf Simulation studies.}
-\paragraph*{Fig 2. Simulation studies.}
+\caption{{\bf Simulation studies.}}
 \label{simresults}
@@
 D: ROC curves for detecting gene association when incorporating prior mean of genetic effects (orange) to identify protective alleles.
-%\end{figure}
+\end{figure}
 
 %% FIGURE: ASTHMA MANHATTAN AND FOREST FOR BMA
-%\begin{figure}[!h]
+\begin{figure}[!h]
 %\includegraphics[width=\textwidth]{../figures/final/Figure_03.pdf}
-%\caption{{\bf Results for asthma application.}
-\paragraph*{Fig 3. Results for asthma application.}
+\caption{{\bf Results for asthma application.}}
 \label{asthma_bma}
@@
 (F) $-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{IL33} for each phenotype
-%\end{figure}
+\end{figure}
 
-%\begin{figure}[!h]
+\begin{figure}[!h]
 %%\in\includegraphics[width=\textwidth]{../figures/final/Figure_04.pdf}
-%\caption{{\bf Results for glaucoma application.}
-\paragraph*{Fig 4. Results for glaucoma application.}
+\caption{{\bf Results for glaucoma application.}}
 \label{glaucoma_bma}
@@
 G: Location of rs28991009 variant (green, p.Q175H, NM\_021146) for the protein 1JC9\_A homologous to \textit{ANGPTL7}.
-%\end{figure}
+\end{figure}
+
+\pagebreak
 
@@
 %% FIGURE: ASTHMA MANHATTAN AND FOREST FOR INDEPENDENT EFFECTS
-%\begin{{figure}[H]
+%\begin{figure}[H]
 %\includegraphics[width=\textwidth]{../figures/final/Figure_S03.pdf}
@@
 
+\pagebreak
+
 \section*{Tables}
@@
 \hline
-\textit{ATP2A3}       &             -11.9 &   -0.8 &             -1.2 &  -2.1 &  -2.2 &    5.8 \\
+\textit{ATP2A3}       &             \textendash11.9 &   -0.8 &             -1.2 &  -2.1 &  -2.2 &    5.8 \\
 \hline
@@
 \thickhline
-\textit{PGLYRP4} &              23.9 &   7.6 &     -0.1 &                                       -1.0 &                      -1.1 \\
-\hline
-\textit{ANGPTL7} &              12.9 &  12.2 &      1.7 &                                       10.5 &                       6.4 \\
-\hline
-\textit{JPH4}    &               6.4 &   2.7 &      0.6 &                                       -1.1 &                      -1.2 \\
-\hline
-\textit{WNT10A}  &               3.5 &   4.1 &     -0.2 &                                       -0.6 &                       3.1 \\
+\textit{ANGPTL7} &              13.1 &  12.2 &      1.7 &                                       10.5 &                       6.4 \\
 \hline
@@
 \hline
-\textit{GLI3}    &               1.4 &   2.5 &     -0.4 &                                       -0.1 &                       4.3 \\
+\textit{WNT10A}  &               2.6 &   2.6 &     -0.2 &                                       -0.6 &                       3.1 \\
 \hline
-\textit{C4orf29} &              -0.0 &   6.4 &     -0.1 &                                       -0.7 &                      -1.1 \\
+\textit{GLI3}    &               0.5 &   2.3 &     -0.4 &                                       -0.1 &                       4.3 \\
 \hline

### 2019-10-20 tex/manuscriptbiorxiv.tex
Updates, new files

@@
 
+\pagebreak
+
 \section*{Figure Legends}
 
-%\begin{figure}[!h]
-%\includegraphics[width=\textwidth]{../figures/final/Figure_01.pdf}
-%\caption{{\bf Schematic overview of MRP.}
-\paragraph*{Fig 1. Schematic overview of MRP.}
+\begin{figure}[!h]
+\includegraphics[width=\textwidth]{../figures/final/Figure_01.pdf}
+\caption{{\bf Schematic overview of MRP.}}
 \label{overview}
@@
 B: Diagram of factors considered in rare variant association analysis including the correlation matrices: $\mathbf{R_{\textrm{study}}}$  (expected correlation of genetic effects among a group of studies), $\mathbf{R_{\textrm{var}}}$ (expected correlation of genetic effects among a group of variants), and $\mathbf{R_{\textrm{phen}}}$ (expected correlation of genetic effects among a group of phenotypes); the scale parameter for genetic variant annotation; and the location of genetic effects, which may be used to prioritize or identify protective modifiers of disease risk.
-%\end{figure}
+\end{figure}
 
-%\begin{figure}[!h]
-%\includegraphics[width=\textwidth]{../figures/final/Figure_02.pdf}
-%\caption{{\bf Simulation studies.}
-\paragraph*{Fig 2. Simulation studies.}
+\pagebreak
+
+\begin{figure}[!h]
+\includegraphics[width=\textwidth]{../figures/final/Figure_02.pdf}
+\caption{{\bf Simulation studies.}}
 \label{simresults}
@@
 D: ROC curves for detecting gene association when incorporating prior mean of genetic effects (orange) to identify protective alleles.
-%\end{figure}
+\end{figure}
+
+\pagebreak
 
 %% FIGURE: ASTHMA MANHATTAN AND FOREST FOR BMA
-%\begin{figure}[!h]
-%\includegraphics[width=\textwidth]{../figures/final/Figure_03.pdf}
-%\caption{{\bf Results for asthma application.}
-\paragraph*{Fig 3. Results for asthma application.}
+\begin{figure}[!h]
+\includegraphics[width=\textwidth]{../figures/final/Figure_03.pdf}
+\caption{{\bf Results for asthma application.}}
 \label{asthma_bma}
@@
 (F) $-\log_{10}$ $p$-values (left panel) and estimated effect sizes with 95\% confidence intervals (right panel) for missense variants and PTVs in \textit{IL33} for each phenotype
-%\end{figure}
+\end{figure}
 
-%\begin{figure}[!h]
-%%\in\includegraphics[width=\textwidth]{../figures/final/Figure_04.pdf}
-%\caption{{\bf Results for glaucoma application.}
-\paragraph*{Fig 4. Results for glaucoma application.}
+\pagebreak
+
+\begin{figure}[!h]
+\includegraphics[width=\textwidth]{../figures/final/Figure_04.pdf}
+\caption{{\bf Results for glaucoma application.}}
 \label{glaucoma_bma}
@@
 
-%\begin{figure}[H]
-%\includegraphics[width=\textwidth]{../figures/final/Figure_S01.pdf}
-%\end{figure}
+\pagebreak
+
+\begin{figure}[H]
+\includegraphics[width=\textwidth]{../figures/final/Figure_S01.pdf}
+\end{figure}
 \paragraph*{S1 Fig.}
@@
 
-%\begin{figure}[H]
-%\includegraphics[width=\textwidth]{../figures/final/Figure_S02.pdf}
-%\end{figure}
+\pagebreak
+
+\begin{figure}[H]
+\includegraphics[width=\textwidth]{../figures/final/Figure_S02.pdf}
+\end{figure}
 \paragraph*{S2 Fig.}
@@
 
+\pagebreak
+
 %% FIGURE: ASTHMA MANHATTAN AND FOREST FOR INDEPENDENT EFFECTS
-%\begin{{figure}[H]
-%\includegraphics[width=\textwidth]{../figures/final/Figure_S03.pdf}
-%\end{figure}
+\begin{figure}[H]
+\includegraphics[width=\textwidth]{../figures/final/Figure_S03.pdf}
+\end{figure}
 \paragraph*{S3 Fig.}
@@
 
+\pagebreak
+
 %% FIGURE: ASTHMA MANHATTAN AND FOREST FOR SIMILAR EFFECTS
-%\begin{figure}[H]
-%\includegraphics[width=\textwidth]{../figures/final/Figure_S04.pdf}
-%\end{figure}
+\begin{figure}[H]
+\includegraphics[width=\textwidth]{../figures/final/Figure_S04.pdf}
+\end{figure}
 \paragraph*{S4 Fig.}
@@
 
-%\begin{figure}[H]
-%\includegraphics[width=\textwidth]{../figures/final/Figure_S05.pdf}
-%\end{figure}
+\pagebreak
+
+\begin{figure}[H]
+\includegraphics[width=\textwidth]{../figures/final/Figure_S05.pdf}
+\end{figure}
 \paragraph*{S5 Fig.}
@@
 
-%\begin{figure}[H]
-%\includegraphics[width=\textwidth]{../figures/final/Figure_S06.pdf}
-%\end{figure}
+\pagebreak
+
+\begin{figure}[H]
+\includegraphics[width=\textwidth]{../figures/final/Figure_S06.pdf}
+\end{figure}
 \paragraph*{S6 Fig.}
@@
 Only $\log_{10}$ Bayes Factors greater than zero are plotted.
+
+\pagebreak
 
